---
title: "COVID eBirding: Preliminary analyses"
date: "2022/03/23"
author: "Karthik Thrikkadeeri"
editor_options: 
  chunk_output_type: console
bibliography: covid-ebirding.bib
link-citations: yes
output: 
  html_document:
    toc: true
    toc_depth: 3  
    number_sections: false  
    theme: united  
    highlight: tango  # specifies the syntax highlighting style
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(patchwork)
library(boot)

theme_set(theme_bw())

covid_palette <- c("#1B9E77", "#E89005", "#EF4050", "#9678B6")
covid_palette3 <- c("#1B9E77", "#E89005", "#9678B6")

# covid_palette2 <- c("#1B9E77", "#D95F02", "#7570B3", "#555555")
knitr::opts_chunk$set(echo = FALSE, warnings = FALSE, 
                      messages = FALSE, cache.lazy = FALSE)

```

```{r data_filt, cache=TRUE}

datapath <- "data/ebd_IN_relFeb-2022.RData"
groupaccspath <- "data/ebd_users_GA_relDec-2021.csv"
covidclasspath <- "data/covid_classification.csv"

source("scripts/functions.R")
# filtering the data for current use
dataqual_filt(datapath, groupaccspath, covidclasspath, 
              maxvel = 20, minsut = 2)

# load("data/data0.RData")

```

```{r data_subset, eval=FALSE, cache=TRUE, dependson="data_filt"}

# random subset for analysis
set.seed(10)
data_sub <- data0 %>% 
  filter(GROUP.ID %in% sample(unique(data0$GROUP.ID), 100000))
save(data_sub, file = "data/data_sub.RData")

# load("data/data_sub.RData")

```


# Introduction

This study will focus on the impacts of the COVID-19 pandemic on both birds and the birding community in the Indian context. It will also try to understand the consequences of these changes on citizen science and the inferences that can be drawn from it. Lastly, it will propose guidelines for making the most of the available data while accounting for pandemic-driven biases and deviations from the pre-COVID normal. 

To ensure a balanced time frame, this study will focus on the years from 2018 to 2021. The time period of 2018–19 will be considered as “pre-COVID” while 2020–21 will be considered “COVID”. In order to account for within-year variations in bird and birder behaviour, the comparisons will be made on the month level. For instance, various metrics of birding behaviour in the month of June will be compared between 2018–19 and 2020–21, and will not be confounded with patterns seen in other months. 

***This notebook is the exploratory analysis notebook for the study. Here, I explore, mainly via visualisations, some of the interesting patterns/metrics discovered from 01_prelim, in order to narrow down on what to include and analyse statistically.***

## Questions

Are there any changes and gaps in the pandemic eBird data? These questions will be answered using numerous metrics calculated from the focal data point in eBird data, a checklist.

- What has happened to overall birding activity? 

- What has happened to group birding? Has it decreased, as would be expected?

- Has birding effort changed with the pandemic? 

- Are there biases in the pandemic birding activity in terms of spatial and temporal spread?


# Birder behaviour (focus level: checklists)

Changes in eBirder behaviour will be analysed using the focal data point, checklists. A number of metrics based on checklists will be calculated and analysed (**at the monthly scale**), such as:

1. Number of submitted checklists  
    1. Total lists  
    1. Shared lists (use no. of obs.)  
1. Number of submitted checklists per observer  
    1. Total lists  
    1. Shared lists (use no. of obs.)  
1. Hotspot birding  
1. Birding protocol  
1. Birding distance
1. Site fidelity per observer  
1. Habitats surveyed  
1. Spatial spread of birding activity. These spatial patterns will be explored at two scales too: nationwide and regional.  
    1. Spatial spread by absences  
    1. Spatial spread by densities  
1. List duration
1. List length  
1. Birding time per observer (monthly for prelim, later maybe daily)  
1. Temporal spread of birding activity  
    1. Temporal spread within day (time-of-day)  
    1. Temporal spread within week (day-of-week)  
1. Media  
1. Number of new birders

When running the actual models, observer will be used as a random effect because all of these metrics would vary differently from observer to observer.

Aside from comparing trends before and during the pandemic, at a finer scale we will also try to find patterns matching the various waves of spread and complete lockdown, as well as the vaccination progress.

Post-presentation: regarding averaging national level patterns by state for some metrics, this means that means are manually calculated for states and the points plotted in the graph are only these single points per state (instead of all the raw data points).


### Group birding 

Doesn't include duplicate lists as we are interested in Boolean group/solo, and otherwise proportion will be inflated just by having more birders in the group birding session.

```{r no_lists, cache=TRUE, message=FALSE}

# only considers unique lists

# since we need to look at each month separately and since this is not per observer, there is no source of variation in the counts (no. of lists).
# hence, stick to proportion CIs

temp1 <- data0 %>% 
  group_by(GROUP.ID) %>% 
  slice(1) %>% ungroup() %>% 
  group_by(COVID, YEAR, MONTH, STATE) %>% 
  summarise(NO.LISTS = n_distinct(GROUP.ID)) %>%
  ungroup() 

temp2 <- data0 %>% 
  group_by(GROUP.ID) %>% 
  slice(1) %>% ungroup() %>% 
  filter(NUMBER.OBSERVERS > 1) %>% 
  group_by(COVID, MONTH, STATE) %>% 
  summarise(NO.SLISTS = n_distinct(GROUP.ID)) %>%
  ungroup() %>% 
  complete(COVID, MONTH, STATE, fill = list(NO.SLISTS = 0))

# unequal values for lower and upper: bracketing the CI(?)
# https://en.wikipedia.org/wiki/Binomial_proportion_confidence_interval#Bracketing_the_confidence_interval
# cannot use this because here we need to propagate. hence, assume Gaussian.
#
# SE of proportions = sqrt((p)(1-p)/n)
# z = 1.96 for 95% CI


nl_sw <- temp1 %>% 
  left_join(temp2) %>% 
  mutate(PROP.SLISTS = NO.SLISTS/NO.LISTS,
         PROP.SE = sqrt((PROP.SLISTS)*(1 - PROP.SLISTS)/NO.LISTS),
         PROP.CI.L = PROP.SLISTS - 1.96*PROP.SE,
         PROP.CI.U = PROP.SLISTS + 1.96*PROP.SE)

rm(list = c("temp1","temp2"))

nl_nw <- nl_sw %>% ungroup() %>% 
  group_by(COVID, YEAR, MONTH) %>% 
  summarise(PROP.SLISTS = mean(PROP.SLISTS),
            PROP.SE = sqrt(sum((PROP.SE)^2))/n(),
            PROP.CI.L = PROP.SLISTS - 1.96*PROP.SE,
            PROP.CI.U = PROP.SLISTS + 1.96*PROP.SE)


(ggplot(nl_nw, aes(MONTH, PROP.SLISTS, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Change in monthly group birding with pandemic at national level",
       x = "Month", y = "")) /
((ggplot(filter(nl_sw, STATE == "Kerala"), 
       aes(MONTH, PROP.SLISTS, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Kerala",
       x = "Month", y = "Proportion of group birding lists")) |
(ggplot(filter(nl_sw, STATE == "Karnataka"), 
       aes(MONTH, PROP.SLISTS, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Karnataka",
       x = "Month", y = ""))) /
((ggplot(filter(nl_sw, STATE == "Gujarat"), 
       aes(MONTH, PROP.SLISTS, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Gujarat",
       x = "Month", y = "")) |
(ggplot(filter(nl_sw, STATE == "Maharashtra"), 
       aes(MONTH, PROP.SLISTS, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Maharashtra",
       x = "Month", y = ""))) +
plot_layout(guides = "collect") -> no_lists

no_lists # plot

ggsave("02_explor_figs/01_no_lists.png", plot = no_lists,
       width = 8, height = 9, units = "in")

rm(list = c("nl_sw","nl_nw","no_lists"))

```

* At the national level, proportion of group birdin decreased drastically with pandemic, especially in the months of April, May and June. It also had a lasting effect in the months of Aug--Nov. In 2021 there is a very slight suggestion that group birding is rising back up.   
* Kerala has also shown a decrease in group birding proportion, strongest during the national lockdown and the second wave, but month-to-month variation is very high.  
* In Karnataka the pandemic affected mostly only the months of April, May and June. Proportions of group birding otherwise were similar to pre-COVID levels.  
* Gujarat's story was similar to Karnataka's but slightly more exaggerated.  
* Maharashtra shows very similar patterns to Karnataka's.


### Group birding per observer

```{r no_lists_po, cache=TRUE, message=FALSE}

temp1 <- data0 %>% 
  group_by(GROUP.ID) %>% 
  slice(1) %>% ungroup() %>% 
  group_by(COVID, YEAR, MONTH, STATE, OBSERVER.ID) %>% 
  summarise(NO.LISTS = n_distinct(GROUP.ID)) %>%
  ungroup() 

temp2 <- data0 %>% 
  group_by(GROUP.ID) %>% 
  slice(1) %>% ungroup() %>% 
  filter(NUMBER.OBSERVERS > 1) %>% 
  group_by(COVID, MONTH, STATE, OBSERVER.ID) %>% 
  summarise(NO.SLISTS = n_distinct(GROUP.ID)) %>%
  ungroup() %>% 
  complete(COVID, MONTH, STATE, fill = list(NO.SLISTS = 0))


# this is for each observer, but we are interested in plotting mean value of all observers combined
# hence, averaging proportions and its SE by propagation of error

nl_po_sw <- temp1 %>% 
  left_join(temp2) %>% 
  mutate(NO.SLISTS = ifelse(is.na(NO.SLISTS), 0, NO.SLISTS),
         PROP.SLISTS = NO.SLISTS/NO.LISTS,
         PROP.SE = sqrt((PROP.SLISTS)*(1 - PROP.SLISTS)/NO.LISTS)) %>%
  ungroup() %>% 
  group_by(COVID, YEAR, MONTH, STATE) %>% 
  summarise(PROP.SLISTS = mean(PROP.SLISTS),
            PROP.SE = sqrt(sum((PROP.SE)^2))/n(),
            PROP.CI.L = PROP.SLISTS - 1.96*PROP.SE,
            PROP.CI.U = PROP.SLISTS + 1.96*PROP.SE)

rm(list = c("temp1","temp2"))

nl_po_nw <- nl_po_sw %>% ungroup() %>% 
  group_by(COVID, YEAR, MONTH) %>% 
  summarise(PROP.SLISTS = mean(PROP.SLISTS),
            PROP.SE = sqrt(sum((PROP.SE)^2))/n(),
            PROP.CI.L = PROP.SLISTS - 1.96*PROP.SE,
            PROP.CI.U = PROP.SLISTS + 1.96*PROP.SE)


(ggplot(nl_po_nw, aes(MONTH, PROP.SLISTS, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Change in monthly group birding per observer with pandemic at national level",
       x = "Month", y = "")) /
((ggplot(filter(nl_po_sw, STATE == "Kerala"), 
       aes(MONTH, PROP.SLISTS, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Kerala",
       x = "Month", y = "Proportion of group birding lists per observer")) |
(ggplot(filter(nl_po_sw, STATE == "Karnataka"), 
       aes(MONTH, PROP.SLISTS, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Karnataka",
       x = "Month", y = ""))) /
((ggplot(filter(nl_po_sw, STATE == "Gujarat"), 
       aes(MONTH, PROP.SLISTS, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Gujarat",
       x = "Month", y = "")) |
(ggplot(filter(nl_po_sw, STATE == "Maharashtra"), 
       aes(MONTH, PROP.SLISTS, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Maharashtra",
       x = "Month", y = ""))) +
plot_layout(guides = "collect") -> no_lists_po

no_lists_po # plot

ggsave("02_explor_figs/02_no_lists_po.png", plot = no_lists_po,
       width = 8, height = 9, units = "in")

rm(list = c("nl_po_sw","nl_po_nw","no_lists_po"))

```

* At the observer level, group birding patterns are very similar to those at the level of lists, and are just more exaggerated version with much smaller confidence intervals.  
* There is a suggestion that levels are returning to pre-COVID levels, but are not quite there yet.  
* Only Kerala, among the four state compared here, seems to have seen a major impact of the pandemic on group birding *outside* of the months coinciding with the national lockdown and the second wave. i.e., only Kerala showed declines in group birding in the months in the second half of the year.  


### Site fidelity per observer  

```{r fidelity, cache=TRUE, message=FALSE}

# 111 km = 1 deg
# 111 * n * 1 km = n deg
# 111 * n (5K) = n deg * 5
# 111/5 * n (5K) = n deg 
#
# here, doing 25km x 25km (SoIB and also more lenient and appropriate)

set.seed(213)

fidel_sw <- data0 %>% 
  group_by(SAMPLING.EVENT.IDENTIFIER) %>% 
  slice(1) %>% ungroup() %>% 
  # grouping in terms of 5kmx5km and 25kmx25km cells (not 1degx1deg)
  mutate(LAT.5K = LATITUDE*111/5, 
         LON.5K = LONGITUDE*111/5,
         LAT.25K = LATITUDE*111/25, 
         LON.25K = LONGITUDE*111/25) %>% 
  # pasting lat-lon as string to find truly unique sites
  mutate(COORD.5K = paste0(ceiling(LON.5K), ",", ceiling(LAT.5K)),
         COORD.25K = paste0(ceiling(LON.25K), ",", ceiling(LAT.25K))) %>% 
  group_by(COVID, YEAR, MONTH, STATE, OBSERVER.ID) %>% 
  summarise(NO.SITES = n_distinct(COORD.25K)) %>% 
  ungroup() %>% 
  mutate(NO.SITES = ifelse(is.na(NO.SITES), 0, NO.SITES)) %>% 
  group_by(COVID, YEAR, MONTH, STATE) %>% 
  summarise(NO.SE = boot_se(NO.SITES),
            NO.SITES = mean(NO.SITES),
            NO.CI.L = NO.SITES - 1.96*NO.SE,
            NO.CI.U = NO.SITES + 1.96*NO.SE)

fidel_nw <- fidel_sw %>% 
  ungroup() %>% 
  group_by(COVID, YEAR, MONTH) %>% 
  summarise(NO.SITES = mean(NO.SITES),
            NO.SE = sqrt(sum((NO.SE)^2))/n(),
            NO.CI.L = NO.SITES - 1.96*NO.SE,
            NO.CI.U = NO.SITES + 1.96*NO.SE)

  
(ggplot(fidel_nw, aes(MONTH, NO.SITES, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = NO.CI.L, ymax = NO.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "No. of sites (unique 25 km x 25 km cells) visited by observers at national level",
         x = "Month", y = "No. of sites")) /
((ggplot(filter(fidel_sw, STATE == "Kerala"), 
         aes(MONTH, NO.SITES, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = NO.CI.L, ymax = NO.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Kerala",
         x = "Month", y = "No. of sites")) |
(ggplot(filter(fidel_sw, STATE == "Karnataka"), 
         aes(MONTH, NO.SITES, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = NO.CI.L, ymax = NO.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Karnataka",
         x = "Month", y = "No. of sites"))) /
((ggplot(filter(fidel_sw, STATE == "Gujarat"), 
         aes(MONTH, NO.SITES, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = NO.CI.L, ymax = NO.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Gujarat",
         x = "Month", y = "No. of sites")) |
(ggplot(filter(fidel_sw, STATE == "Maharashtra"), 
         aes(MONTH, NO.SITES, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = NO.CI.L, ymax = NO.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Maharashtra",
         x = "Month", y = "No. of sites"))) +
plot_layout(guides = "collect") -> fidelity

fidelity # plot

ggsave("02_explor_figs/03_fidelity.png", plot = fidelity,
       width = 9, height = 9, units = "in")
  
rm(list = c("fidel_nw","fidel_sw","fidelity"))

```

* At the national level, site fidelity was generally higher during the pandemic than before but mostly only during lockdown/second wave. In other months there was no noticeable difference.    
* Among the four states compared, only Kerala and Karnataka showed this pattern, while in Gujarat and Maharashtra there seemed to be no difference with the pandemic at all.  

### Birding time per observer

```{r time, cache=TRUE, message=FALSE}

set.seed(423) 

time_sw <- data0 %>% 
  group_by(SAMPLING.EVENT.IDENTIFIER) %>% 
  slice(1) %>% ungroup() %>% 
  group_by(COVID, YEAR, MONTH, STATE, OBSERVER.ID) %>% 
  summarise(B.TIME = sum(DURATION.MINUTES)) %>% 
  ungroup() %>% 
  mutate(B.TIME = ifelse(is.na(B.TIME), 0, B.TIME)) %>% 
  group_by(COVID, YEAR, MONTH, STATE) %>% 
  summarise(B.TIME.SE = boot_se(B.TIME),
            B.TIME = mean(B.TIME),
            B.TIME.CI.L = B.TIME - 1.96*B.TIME.SE,
            B.TIME.CI.U = B.TIME + 1.96*B.TIME.SE)

time_nw <- time_sw %>% 
  ungroup() %>% 
  group_by(COVID, YEAR, MONTH) %>% 
  summarise(B.TIME = mean(B.TIME),
            B.TIME.SE = sqrt(sum((B.TIME.SE)^2))/n(),
            B.TIME.CI.L = B.TIME - 1.96*B.TIME.SE,
            B.TIME.CI.U = B.TIME + 1.96*B.TIME.SE)


(ggplot(time_nw, aes(MONTH, B.TIME, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = B.TIME.CI.L, ymax = B.TIME.CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Monthly birding times of observers at national level",
       x = "Month", y = "Time (mins)")) /
((ggplot(filter(time_sw, STATE == "Kerala"), 
         aes(MONTH, B.TIME, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = B.TIME.CI.L, ymax = B.TIME.CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Kerala",
       x = "Month", y = "Time (mins)")) |
(ggplot(filter(time_sw, STATE == "Karnataka"), 
         aes(MONTH, B.TIME, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = B.TIME.CI.L, ymax = B.TIME.CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Karnataka",
       x = "Month", y = "Time (mins)"))) /
((ggplot(filter(time_sw, STATE == "Gujarat"), 
         aes(MONTH, B.TIME, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = B.TIME.CI.L, ymax = B.TIME.CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Gujarat",
       x = "Month", y = "Time (mins)")) |
(ggplot(filter(time_sw, STATE == "Maharashtra"), 
         aes(MONTH, B.TIME, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = B.TIME.CI.L, ymax = B.TIME.CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Maharashtra",
       x = "Month", y = "Time (mins)"))) +
plot_layout(guides = "collect") -> time

time # plot

ggsave("02_explor_figs/04_time.png", plot = time,
       width = 9, height = 9, units = "in")

rm(time_nw, time_sw, time)

```

* At the national level, birding times were and are lowest during monsoon. There was no noticeable negative impact of the pandemic on monthly birding times, except in March 2020 when the pandemic was a novelty. In fact, in the monsoon months of July and August, birding times even increased slightly during the pandemic.  
* The only other major pattern visible at the state level is in Karnataka, where birding times have increased in the months of August--October.   

### Hotspot birding  

Not "per observer", as this metric from data POV. Includes duplicate lists, of course.

```{r hotspot, cache=TRUE, message=FALSE}

temp1 <- data0 %>% 
  group_by(SAMPLING.EVENT.IDENTIFIER) %>% 
  slice(1) %>% ungroup() %>% 
  group_by(COVID, YEAR, MONTH, STATE) %>% 
  summarise(NO.LISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>%
  ungroup() 

temp2 <- data0 %>% 
  group_by(SAMPLING.EVENT.IDENTIFIER) %>% 
  slice(1) %>% ungroup() %>% 
  filter(LOCALITY.TYPE == "H") %>% 
  group_by(COVID, YEAR, MONTH, STATE) %>% 
  summarise(NO.HLISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>%
  ungroup() %>% 
  complete(COVID, MONTH, STATE, fill = list(NO.SLISTS = 0))


hot_sw <- temp1 %>% 
  left_join(temp2) %>% 
  group_by(COVID, YEAR, MONTH, STATE) %>% 
  mutate(NO.HLISTS = ifelse(is.na(NO.HLISTS), 0, NO.HLISTS),
         PROP.HLISTS = NO.HLISTS/NO.LISTS,
         PROP.SE = sqrt((PROP.HLISTS)*(1 - PROP.HLISTS)/NO.LISTS),
         PROP.CI.L = PROP.HLISTS - 1.96*PROP.SE,
         PROP.CI.U = PROP.HLISTS + 1.96*PROP.SE) 

rm(list = c("temp1","temp2"))

hot_nw <- hot_sw %>% ungroup() %>% 
  filter(NO.LISTS != 0) %>% 
  group_by(COVID, YEAR, MONTH) %>% 
  summarise(PROP.HLISTS = mean(PROP.HLISTS),
            PROP.SE = sqrt(sum((PROP.SE)^2))/n(),
            PROP.CI.L = PROP.HLISTS - 1.96*PROP.SE,
            PROP.CI.U = PROP.HLISTS + 1.96*PROP.SE)

# Daman and Diu lists not appearing from 2020--2021


(ggplot(hot_nw, aes(MONTH, PROP.HLISTS, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Change in monthly hotspot birding with pandemic at national level",
       x = "Month", y = "")) /
((ggplot(filter(hot_sw, STATE == "Kerala"),
         aes(MONTH, PROP.HLISTS, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Kerala",
       x = "Month", y = "Proportion of hotspot lists")) |
(ggplot(filter(hot_sw, STATE == "Karnataka"),
         aes(MONTH, PROP.HLISTS, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Karnataka",
       x = "Month", y = ""))) /
((ggplot(filter(hot_sw, STATE == "Gujarat"),
         aes(MONTH, PROP.HLISTS, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Gujarat",
       x = "Month", y = "")) |
(ggplot(filter(hot_sw, STATE == "Maharashtra"),
         aes(MONTH, PROP.HLISTS, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Maharashtra",
       x = "Month", y = ""))) +
plot_layout(guides = "collect") -> hotspot

hotspot # plot

ggsave("02_explor_figs/05_hotspot.png", plot = hotspot,
       width = 9, height = 9, units = "in")

rm(list = c("hot_nw","hot_sw", "hotspot"))

```

* The national level saw lower hotspot birding during the lockdown/second wave months, with some spillover into the months of June and July.  
* Kerala mirrored the national level pattern with slightly higher month-to-month variability.  
* In Karnataka, this impact surprisingly lasted through all of 2020 but returned almost exactly to pre-COVID levels in 2021.  
* Gujarat and Maharashtra basically mirror the national pattern.   

### Birding protocol  

Not "per observer", as this metric from data POV. Includes duplicate lists, of course.

```{r protocol, cache=TRUE, message=FALSE}

temp1 <- data0 %>% 
  group_by(SAMPLING.EVENT.IDENTIFIER) %>% 
  slice(1) %>% ungroup() %>% 
  group_by(COVID, YEAR, MONTH, STATE) %>% 
  summarise(NO.LISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>%
  ungroup() 

temp2 <- data0 %>% 
  group_by(SAMPLING.EVENT.IDENTIFIER) %>% 
  slice(1) %>% ungroup() %>% 
  group_by(COVID, MONTH, STATE) %>% 
  filter(PROTOCOL.TYPE == "Traveling") %>% 
  summarise(NO.TLISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>%
  ungroup() %>% 
  complete(COVID, MONTH, STATE, fill = list(NO.TLISTS = 0))


prot_sw <- temp1 %>% 
  left_join(temp2) %>% 
  group_by(COVID, YEAR, MONTH, STATE) %>% 
  mutate(NO.TLISTS = ifelse(is.na(NO.TLISTS), 0, NO.TLISTS),
         PROP.TLISTS = NO.TLISTS/NO.LISTS,
         PROP.SE = sqrt((PROP.TLISTS)*(1 - PROP.TLISTS)/NO.LISTS),
         PROP.CI.L = PROP.TLISTS - 1.96*PROP.SE,
         PROP.CI.U = PROP.TLISTS + 1.96*PROP.SE) 

rm(list = c("temp1","temp2"))

prot_nw <- prot_sw %>% ungroup() %>% 
  filter(NO.LISTS != 0) %>% 
  group_by(COVID, YEAR, MONTH) %>% 
  summarise(PROP.TLISTS = mean(PROP.TLISTS),
            PROP.SE = sqrt(sum((PROP.SE)^2))/n(),
            PROP.CI.L = PROP.TLISTS - 1.96*PROP.SE,
            PROP.CI.U = PROP.TLISTS + 1.96*PROP.SE)

# Daman and Diu lists not appearing from 2020--2021

(ggplot(prot_nw, aes(MONTH, PROP.TLISTS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Change in monthly travelling birding with pandemic at national level",
         x = "Month", y = "Proportion of travelling lists")) /
((ggplot(filter(prot_sw, STATE == "Kerala"), 
         aes(MONTH, PROP.TLISTS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Kerala",
         x = "Month", y = "Proportion of travelling lists")) |
(ggplot(filter(prot_sw, STATE == "Karnataka"), 
         aes(MONTH, PROP.TLISTS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Karnataka",
         x = "Month", y = "Proportion of travelling lists"))) /
((ggplot(filter(prot_sw, STATE == "Gujarat"), 
         aes(MONTH, PROP.TLISTS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Gujarat",
         x = "Month", y = "Proportion of travelling lists")) |
(ggplot(filter(prot_sw, STATE == "Maharashtra"), 
         aes(MONTH, PROP.TLISTS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Maharashtra",
         x = "Month", y = "Proportion of travelling lists"))) +
plot_layout(guides = "collect") -> protocol

protocol 

ggsave("02_explor_figs/06_protocol.png", plot = protocol,
       width = 9, height = 9, units = "in")

rm(list = c("prot_nw","prot_sw","protocol"))

```

* At the national level, mobility of birders, in terms of proportion of travelling lists submitted, decreased sharply during April--May as expected, and had some spillover into June and July, but matched pre-COVID levels well in the other months of the year.  
* In Kerala the pandemic has impacted all months of the year, except January and February which continue to have high proportions of travelling lists, possibly due mass birding events such as GBBC and AWC. Otherwise, even "after" the pandemic, mobility is much lower than before the pandemic.  
* Karnataka was mostly only affected during the months of March--May, with spillover into June and July. Moreover, this impact was only in 2020 and not 2021.  
* Gujarat and Maharashtra show high month-to-month variability but suggest somewhat similar patterns as at the national level.  

### Birding distance

Not "per observer", as this metric from data POV. Includes duplicate lists, of course. Only considers travelling lists.

```{r distance, cache=TRUE, message=FALSE}

set.seed(192) 

dist_sw <- data0 %>% 
  group_by(SAMPLING.EVENT.IDENTIFIER) %>% 
  slice(1) %>% ungroup() %>% 
  filter(PROTOCOL.TYPE == "Traveling", !is.na(EFFORT.DISTANCE.KM)) %>% 
  group_by(COVID, YEAR, MONTH, STATE) %>% 
  summarise(DIST = mean(EFFORT.DISTANCE.KM),
            DIST.SE = boot_se(EFFORT.DISTANCE.KM),
            DIST.CI.L = DIST - 1.96*DIST.SE,
            DIST.CI.U = DIST + 1.96*DIST.SE)

dist_nw <- dist_sw %>% ungroup() %>% 
  group_by(COVID, YEAR, MONTH) %>% 
  summarise(DIST = mean(DIST),
            DIST.SE = sqrt(sum((DIST.SE)^2))/n(),
            DIST.CI.L = DIST - 1.96*DIST.SE,
            DIST.CI.U = DIST + 1.96*DIST.SE)


(ggplot(dist_nw, aes(MONTH, DIST, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = DIST.CI.L, ymax = DIST.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Average distance travelled per travelling checklist at national level",
         x = "Month", y = "Distance travelled (km)")) /
((ggplot(filter(dist_sw, STATE == "Kerala"), 
         aes(MONTH, DIST, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = DIST.CI.L, ymax = DIST.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Kerala",
         x = "Month", y = "Distance travelled (km)")) |
(ggplot(filter(dist_sw, STATE == "Karnataka"), 
         aes(MONTH, DIST, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = DIST.CI.L, ymax = DIST.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Karnataka",
         x = "Month", y = "Distance travelled (km)"))) /
((ggplot(filter(dist_sw, STATE == "Gujarat"), 
         aes(MONTH, DIST, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = DIST.CI.L, ymax = DIST.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Gujarat",
         x = "Month", y = "Distance travelled (km)")) |
(ggplot(filter(dist_sw, STATE == "Maharashtra"), 
         aes(MONTH, DIST, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = DIST.CI.L, ymax = DIST.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Maharashtra",
         x = "Month", y = "Distance travelled (km)"))) +
plot_layout(guides = "collect") -> distance

distance

ggsave("02_explor_figs/07_distance.png", plot = distance,
       width = 9, height = 9, units = "in")

rm(dist_sw, dist_nw, distance)

```

* At the national level, checklist distances were highest during summer and winter before the pandemic. During the pandemic, distances remained fairly similar for the winter months of October--December, but were drastically lower during the summer months of April and May, which coincided with the lockdown/second wave.  
* In Kerala, before the pandemic, birders travelled longest in checklists during late summer and early winter and shortest in late winter and peak monsoon. This contrasts with the pattern seen in proportion of stationary lists (more stationary in early winter and less in late winter). This means that in early winter the lists were either stationary or travelling larger distances, while in late winter lists converged to shorter distances. With the pandemic, effort distances reduced in late summer and early monsoon, coinciding with the lockdown/second wave.  
* Karnataka showed small decreases in the months of April--June.  
* In Gujarat, interestingly, there is no difference with the pandemic except in the month of April.  
* Maharashtra seemed to very marginally affected by the pandemic, in terms of birding mobility. 


### List duration

Not "per observer", as this metric from data POV. Includes duplicate lists, of course.

```{r duration, cache=TRUE, message=FALSE}

set.seed(834)

dur_sw <- data0 %>% 
  group_by(SAMPLING.EVENT.IDENTIFIER) %>% 
  slice(1) %>% 
  filter(!is.na(DURATION.MINUTES)) %>% 
  ungroup() %>% 
  group_by(COVID, YEAR, MONTH, STATE) %>% 
  summarise(DURATION = mean(DURATION.MINUTES),
            DURATION.SE = boot_se(DURATION.MINUTES),
            DURATION.CI.L = DURATION - 1.96*DURATION.SE,
            DURATION.CI.U = DURATION + 1.96*DURATION.SE)

dur_nw <- dur_sw %>% 
  ungroup() %>% 
  group_by(COVID, YEAR, MONTH) %>% 
  summarise(DURATION = mean(DURATION),
            DURATION.SE = sqrt(sum((DURATION.SE)^2))/n(),
            DURATION.CI.L = DURATION - 1.96*DURATION.SE,
            DURATION.CI.U = DURATION + 1.96*DURATION.SE) 


(ggplot(dur_nw, aes(MONTH, DURATION, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = DURATION.CI.L, ymax = DURATION.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Average checklist duration at national level",
         x = "Month", y = "Duration (mins)")) /
((ggplot(filter(dur_sw, STATE == "Kerala"), 
         aes(MONTH, DURATION, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = DURATION.CI.L, ymax = DURATION.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Kerala",
         x = "Month", y = "Duration (mins)")) |
(ggplot(filter(dur_sw, STATE == "Karnataka"), 
         aes(MONTH, DURATION, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = DURATION.CI.L, ymax = DURATION.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Karnataka",
         x = "Month", y = "Duration (mins)"))) /
((ggplot(filter(dur_sw, STATE == "Gujarat"), 
         aes(MONTH, DURATION, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = DURATION.CI.L, ymax = DURATION.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Gujarat",
         x = "Month", y = "Duration (mins)")) |
(ggplot(filter(dur_sw, STATE == "Maharashtra"), 
         aes(MONTH, DURATION, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = DURATION.CI.L, ymax = DURATION.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Maharashtra",
         x = "Month", y = "Duration (mins)"))) +
plot_layout(guides = "collect") -> duration

duration

ggsave("02_explor_figs/08_duration.png", plot = duration,
       width = 9, height = 9, units = "in")

rm(dur_sw, dur_nw, duration)

```

* In normal circumstances, checklists become longer in the winter, and shorter in the monsoon. The only notable change during the pandemic at the national level was low checklist duration in April--June, due to the national lockdown/second wave.  
* In Kerala, while most of the pattern remains similar before and during the pandemic, the strong dip in checklist duration during the national lockdown is still present. Moreover, the winter month of January during the pandemic saw longer checklists than before.   
* Karnataka doesn't show very clear patterns apart from a possibly lower participation in the February GBBC (leading to higher average duration) during the pandemic, and shorter lists in April--May.  
* Gujarat and Maharashtra were mostly affected only in the March--May period.  
* A point of interest is that the average checklist duration varied greatly between states. Among the 4 compared here, Kerala on average had shortest lists, followed by Karnataka and then Maharashtra.  

### List length

Not "per observer", as this metric from data POV. Includes duplicate lists, of course.

```{r length, cache=TRUE, message=FALSE}

# var(mean(x))/var(median(x)) = 4n/(pi*(2n + 1))

# used mean here because it was proving extremely difficult to work with errors in the case of median

set.seed(98)

length_sw <- data0 %>% 
  mutate(CATEGORY = case_when(CATEGORY == "domestic" & 
                                 COMMON.NAME == "Rock Pigeon" ~ "species",
                               TRUE ~ CATEGORY)) %>% 
  filter(CATEGORY %in% c("issf","species")) %>% 
  group_by(COVID, YEAR, MONTH, STATE, SAMPLING.EVENT.IDENTIFIER) %>% 
  summarise(LENGTH = n_distinct(COMMON.NAME)) %>% 
  ungroup() %>% 
  group_by(COVID, YEAR, MONTH, STATE) %>% 
  summarise(LENGTH.SE = boot_se(LENGTH),
            LENGTH = mean(LENGTH),
            LENGTH.CI.L = LENGTH - 1.96*LENGTH.SE,
            LENGTH.CI.U = LENGTH + 1.96*LENGTH.SE)

length_nw <- length_sw %>% 
  ungroup() %>% 
  group_by(COVID, YEAR, MONTH) %>% 
  summarise(LENGTH = mean(LENGTH),
            LENGTH.SE = sqrt(sum((LENGTH.SE)^2))/n(),
            LENGTH.CI.L = LENGTH - 1.96*LENGTH.SE,
            LENGTH.CI.U = LENGTH + 1.96*LENGTH.SE)


(ggplot(length_nw, aes(MONTH, LENGTH, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = LENGTH.CI.L, ymax = LENGTH.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Average checklist length at national level",
         x = "Month", y = "List length (no. of species)")) /
((ggplot(filter(length_sw, STATE == "Kerala"), 
         aes(MONTH, LENGTH, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = LENGTH.CI.L, ymax = LENGTH.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Kerala",
         x = "Month", y = "List length (no. of species)")) |
(ggplot(filter(length_sw, STATE == "Karnataka"), 
         aes(MONTH, LENGTH, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = LENGTH.CI.L, ymax = LENGTH.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Karnataka",
         x = "Month", y = "List length (no. of species)"))) /
((ggplot(filter(length_sw, STATE == "Gujarat"), 
         aes(MONTH, LENGTH, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = LENGTH.CI.L, ymax = LENGTH.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Gujarat",
         x = "Month", y = "List length (no. of species)")) |
(ggplot(filter(length_sw, STATE == "Maharashtra"), 
         aes(MONTH, LENGTH, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = LENGTH.CI.L, ymax = LENGTH.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Maharashtra",
         x = "Month", y = "List length (no. of species)"))) +
plot_layout(guides = "collect") -> length

length

ggsave("02_explor_figs/09_length.png", plot = length,
       width = 9, height = 9, units = "in")

rm(length_nw, length_sw, length)

```

* At the national level, list length decreased only in April--May. Over all years, the average checklist becomes bigger in the winter (due to winter migrants adding to the local species pool).  
* In Kerala, list length increased in the winter months of December--January during the pandemic, but otherwise mirrored national patterns.  
* Other states mirrored national patterns with more or less degree of variablity.  

### Habitats surveyed  

### Spatial spread and coverage 

```{r MODIS, cache=TRUE, message=FALSE}

###### Obtaining MODIS LULC maps for urban/non-urban classification ####


library(raster) # masks dplyr functions, so have used package::function()
require(terra)
library(gdalUtils)
library(geodata)
# install.packages("luna", repos = "https://rspatial.r-universe.dev") # requires Rtools 4.0
library(luna)


## Obtaining MODIS data from within R for reproducibility, then saving the cropped and masked raster of area of interest as .tif file.
## Walkthrough at https://rspatial.org/terra/modis/2-download.html

# The MODIS Land Cover Type Product (MCD12Q1) provides a suite of science data sets (SDSs) that map global land cover at 500 meter spatial resolution at annual time step for six different land cover legends. The maps were created from classifications of spectro-temporal features derived of data from the Moderate Resolution Imaging Spectroradiometer (MODIS).


# ## To see list of MODIS products. We already know which one we want, so ignore.
# MODISprod <- getProducts("^MOD|^MYD|^MCD")
# MODISprod


prod <- "MCD12Q1" # short name of product of interest
start <- "2019-12-31" # time period of mapping
end <- "2019-12-31" # time period of mapping

# shapefile of India as area of interest
aoi <- geodata::gadm("India", level = 1, path = "data")

# creating folder to store MODIS data that will be downloaded
MODISpath <- "data/in_LULC_MODIS/"
dir.create(MODISpath, showWarnings=FALSE)


# ## create RDS file containing login credentials for EOSDIS account
# cred <- data.frame(UN = "xxx",
#                    PW = "yyy")
# saveRDS(cred, "userpass.rds")

userpass <- readRDS("userpass.rds")


## downloading MODIS tiled data (16 tiles for our defined area of interest, India)
# resulting object is a list of file names
modis <- luna::getModis(prod, start, end,
                        aoi = aoi, download = T, path = MODISpath,
                        username = userpass$UN, password = userpass$PW)

# choose the band (SDS) that you want using sds[] option and write GTiff files.
# University of Maryland SDS (LC_Type2) is SDS1
for (i in (modis)) {
  sds <- get_subdatasets(i)
  modis2 <- gdal_translate(sds[1], dst_dataset = paste0(i, ".tif"))
  }


## listing all the different tiles present as .tif files, to later merge into one
rast_list <- list.files(path = MODISpath,
                        pattern = "MCD12Q1*.*tif",
                        all.files = TRUE, full.names = TRUE)

# merge into one raster
rast_full <- rast_list %>%
  lapply(raster::raster) %>% # rasterise the different files
  do.call(what = raster::merge) %>% # merge the files
  terra::rast()


## projecting from MODIS sinusoidal to WGS84 (flat 2D)
projto <- raster::crs(aoi)
# next step takes a few minutes
# "near" because LC values are categories despite being numerical
rast_proj <- terra::project(rast_full, projto, method = "near")


## cropping and masking the raster to our area of interest
rast_aoi <- terra::crop(rast_proj, aoi)
rast_aoi <- terra::mask(rast_aoi, aoi)


raster::writeRaster(rast_aoi, paste0(MODISpath, "/in_LULC_MODIS.tif"), overwrite = TRUE)

rm(prod, start, end, aoi, modispath, userpass, modis, sds, modis2, rast_list, rast_full, projto, rast_proj, rast_aoi)



##### Modifying the MODIS data ####


## read in cropped and masked .tif
rast <- raster::raster("data/in_LULC_MODIS/in_LULC_MODIS.tif")
# raster::plot(rast)


## reclassifying the raster as urban vs. non-urban 

# (help from: http://www.wvview.org/spatial_analytics/Raster_Analysis/_site/index.html ; http://zevross.com/blog/2015/03/30/map-and-analyze-raster-data-in-r/ )
# category 13 corresponds to "urban and built-up lands"
categs <- unique(values(rast))
# ocean also becomes "non-urban" but that's fine since we won't be looking at those locations
reclass <- matrix(c(categs, rep(0, 13), 1, rep(0, 4)), ncol = 2)

rast_UNU <- raster::reclassify(rast, rcl = reclass)


## changing resolution of data (MODIS data has 500m*500m resolution)
# https://stackoverflow.com/questions/37956422/resample-raster

# aggregating to 2km*2km
rast_UNU <- rast_UNU %>% 
  raster::aggregate(fact = 2, fun = rast_agg_fn) %>% 
  raster::aggregate(fact = 2, fun = rast_agg_fn)

save(rast_UNU, file = "data/rast_UNU.RData")

# aggregating to 24km*24km for SoIB inference
# (here, priority is getting grid cell information, not so much the UNU classification, as that is not directly going to be used; instead will use birding values from them which will be averaged to the SoIB scale)
rast_SoIB <- rast_UNU %>% 
  raster::aggregate(fact = 12, fun = rast_agg_fn)

save(rast_SoIB, file = "data/rast_SoIB.RData")


##### Leaflet overlay to verify urban/non- classification ####

# https://rstudio.github.io/leaflet/raster.html
# https://rpubs.com/mgei/drivingtimes

library(leaflet)

leafcols <- colorFactor(c(NA, "#0C2C84"), values(rast_UNU),
                         na.color = "#ffba00")

# leaflet uses different crs so must be projected; method should be made categorical

leaflet() %>% 
  addTiles() %>% 
  addRasterImage(rast_UNU, 
                 method = "ngb", # nearest neighbour
                 colors = leafcols,
                 opacity = 0.7,
                 maxBytes = 150 * 1024 * 1024) %>% 
  addLegend(pal = leafcols, values = values(rast_UNU))


##### Adding UNU classification to eBird data ####

# adding extracted land cover values at eBird data points with grid cell information at both scales
# Pakistan-administered Kashmir lists produce NA for URBAN
data_UNU <- data0 %>% 
  mutate(URBAN = raster::extract(rast_UNU, cbind(LONGITUDE, LATITUDE)),
         NONURBAN = if_else(URBAN == 1, 0, 1),
         # 2km*2km
         SUBCELL.ID = cellFromXY(rast_UNU, cbind(LONGITUDE, LATITUDE)), 
         # 24km*24km
         CELL.ID = cellFromXY(rast_SoIB, cbind(LONGITUDE, LATITUDE))) %>% 
  filter(!is.na(URBAN))

save(data_UNU, file = "data/data_UNU.RData")


```

```{r s_listsUNU, cache=TRUE, message=FALSE}

##### Checking for pandemic-induced change in ratio of checklists submitted in urban and non-urban areas

require(raster) # masks dplyr functions, so have used raster::function()
require(terra)
require(gdalUtils)
require(geodata)
# install.packages("luna", repos = "https://rspatial.r-universe.dev") # requires Rtools 4.0
require(luna)

# load("data/data_UNU.RData")


temp1 <- data_UNU %>% 
  group_by(SAMPLING.EVENT.IDENTIFIER) %>% 
  slice(1) %>% 
  group_by(COVID, YEAR, MONTH, STATE) %>% 
  dplyr::summarise(NO.LISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>%
  ungroup() 

temp2 <- data_UNU %>% 
  group_by(SAMPLING.EVENT.IDENTIFIER) %>% 
  slice(1) %>% 
  filter(URBAN == 1) %>% 
  group_by(COVID, YEAR, MONTH, STATE) %>% 
  dplyr::summarise(NO.ULISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>%
  ungroup() 

urban_sw <- temp1 %>% 
  left_join(temp2) %>% 
  mutate(NO.ULISTS = ifelse(is.na(NO.ULISTS), 0, NO.ULISTS),
         PROP.ULISTS = NO.ULISTS/NO.LISTS,
         PROP.SE = sqrt((PROP.ULISTS)*(1 - PROP.ULISTS)/NO.LISTS),
         PROP.CI.L = PROP.ULISTS - 1.96*PROP.SE,
         PROP.CI.U = PROP.ULISTS + 1.96*PROP.SE) 

rm(list = c("temp1","temp2"))

urban_nw <- urban_sw %>% 
  group_by(COVID, YEAR, MONTH) %>% 
  dplyr::summarise(PROP.ULISTS = mean(PROP.ULISTS),
                   PROP.SE = sqrt(sum((PROP.SE)^2))/n(),
                   PROP.CI.L = PROP.ULISTS - 1.96*PROP.SE,
                   PROP.CI.U = PROP.ULISTS + 1.96*PROP.SE)


(ggplot(urban_nw, aes(MONTH, PROP.ULISTS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "National level",
         x = "Month", y = "Proportion of urban lists")) /
((ggplot(filter(urban_sw, STATE == "Kerala"), 
         aes(MONTH, PROP.ULISTS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Kerala",
         x = "Month", y = "Proportion of urban lists")) |
(ggplot(filter(urban_sw, STATE == "Karnataka"), 
         aes(MONTH, PROP.ULISTS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Karnataka",
         x = "Month", y = "Proportion of urban lists"))) /
((ggplot(filter(urban_sw, STATE == "Gujarat"), 
         aes(MONTH, PROP.ULISTS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Gujarat",
         x = "Month", y = "Proportion of urban lists")) |
(ggplot(filter(urban_sw, STATE == "Maharashtra"), 
         aes(MONTH, PROP.ULISTS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Maharashtra",
         x = "Month", y = "Proportion of urban lists"))) +
plot_layout(guides = "collect") +
plot_annotation(title = "What proportion of checklists were from urban areas (2kmx2km cells)?") -> s_listsUNU


ggsave("02_explor_figs/12_s_listsUNU.png", 
       plot = s_listsUNU,
       width = 9, height = 9, units = "in")


rm(data_LULC, rast, rast_UNU, reclass, categs,
   urban_nw, urban_sw, s_spread_LULC_MODIS)


detach("package:gdalUtils", unload=T)
detach("package:geodata", unload=T)
detach("package:raster", unload=T, force = T)
detach("package:terra", unload=T, force = T)
detach("package:luna", unload=T)

```

```{r s_cover, cache=TRUE, message=FALSE}

##### Proportion of total grid cells covered ####

library(sp)
library(rgdal)

india <- readOGR(dsn = "data/in_2011", layer = "India_2011")
states <- readOGR(dsn = "data/in_states_2019", layer = "in_states_2019")

load("data/maps.RData")
load("data/clips.RData")

data_sp <- data0 %>% 
  group_by(SAMPLING.EVENT.IDENTIFIER) %>% 
  slice(1) %>% ungroup() 

temp <- data_sp %>% column_to_rownames("SAMPLING.EVENT.IDENTIFIER") 
coordinates(temp) <- ~LONGITUDE + LATITUDE 
proj4string(temp) <- "+proj=longlat +datum=WGS84"
temp <- temp %>% over(districtmap) %>%
  select(1:2) %>% rownames_to_column("SAMPLING.EVENT.IDENTIFIER") 
data_sp <- left_join(temp, data_sp)
  
# add columns with GRID ATTRIBUTES to main data
temp <- data_sp %>% group_by(SAMPLING.EVENT.IDENTIFIER) %>% slice(1)
temp <- temp %>% column_to_rownames("SAMPLING.EVENT.IDENTIFIER")
coordinates(temp) <- ~LONGITUDE + LATITUDE
temp <- temp %>% over(gridmapg1) %>% rownames_to_column("SAMPLING.EVENT.IDENTIFIER") 
data_sp <- left_join(temp, data_sp)
names(data_sp)[2] <- "GRIDG1" # 25km2

temp <- data_sp %>% group_by(SAMPLING.EVENT.IDENTIFIER) %>% slice(1)
temp <- temp %>% column_to_rownames("SAMPLING.EVENT.IDENTIFIER")
coordinates(temp) <- ~LONGITUDE + LATITUDE
temp <- temp %>% over(g2clip) %>% rownames_to_column("SAMPLING.EVENT.IDENTIFIER") 
data_sp <- left_join(temp, data_sp)
names(data_sp)[2] <- "G2CLIP"     # 50km2


data_cov <- data_sp %>% 
  ungroup() %>% 
  mutate(TOT.GRIDS = n_distinct(GRIDG1)) %>% 
  group_by(COVID, YEAR, MONTH, GRIDG1) %>% 
  dplyr::summarise(NO.LISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER),
            TOT.GRIDS = min(TOT.GRIDS)) %>% 
  group_by(COVID, YEAR, MONTH) %>% 
  dplyr::summarise(SD.LISTS = sd(NO.LISTS),
            TOT.GRIDS = min(TOT.GRIDS),
            NO.GRIDS = n_distinct(GRIDG1),
            PROP.GRIDS = NO.GRIDS/TOT.GRIDS,
            PROP.SE = sqrt((PROP.GRIDS)*(1 - PROP.GRIDS)/TOT.GRIDS),
            PROP.CI.L = PROP.GRIDS - 1.96*PROP.SE,
            PROP.CI.U = PROP.GRIDS + 1.96*PROP.SE)

(ggplot(data_cov, aes(MONTH, PROP.GRIDS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Change in spatial coverage of monthly birding with pandemic",
         x = "Month", y = "Proportion of grid cells with birding activity")) /
(ggplot(data_cov, aes(as.factor(MONTH), SD.LISTS, colour = COVID)) + 
  scale_color_manual(values = covid_palette) +
  geom_point(stat = "summary", fun.data = mean_cl_boot, 
             size = 4, position = position_dodge(0.5)) +
  geom_errorbar(stat = 'summary', fun.data = mean_cl_boot, 
                size = 1.25, width = 0.3, position = position_dodge(0.5)) +
  xlab("Month") + ylab("Standard deviation of lists across cells")) +
plot_layout(guides = "collect") -> s_cover

ggsave("02_explor_figs/10_s_cover.png", 
       plot = s_cover,
       width = 9, height = 9, units = "in")


##### Proportion of total urban vs. non-urban cells covered ####


##### Proportion of covered cells urban (vs. non-urban)  ####

## checking no. of cells with at least 1 complete list
cells_UNU_sw <- data_UNU %>% 
  group_by(SAMPLING.EVENT.IDENTIFIER) %>% 
  slice(1) %>% ungroup() %>% 
  # below removes rows of any cell with <1 complete list
  filter(ALL.SPECIES.REPORTED == 1) %>% 
  group_by(COVID, YEAR, MONTH, CELL.ID) %>% 
  slice(1) %>% ungroup() %>% 
  filter(!is.na(URBAN)) %>% 
  group_by(COVID, YEAR, MONTH, STATE) %>% 
  dplyr::summarise(NO.UCELLS = sum(URBAN),
                   TOT.CELLS = n_distinct(CELL.ID),
                   PROP.UCELLS = NO.UCELLS/TOT.CELLS,
                   PROP.SE = sqrt((PROP.UCELLS)*(1 - PROP.UCELLS)/TOT.CELLS),
                   PROP.CI.L = PROP.UCELLS - 1.96*PROP.SE,
                   PROP.CI.U = PROP.UCELLS + 1.96*PROP.SE)

cells_UNU_nw <- cells_UNU_sw %>% 
  group_by(COVID, YEAR, MONTH) %>% 
  dplyr::summarise(PROP.UCELLS = mean(PROP.UCELLS),
                   PROP.SE = sqrt(sum((PROP.SE)^2))/n(),
                   PROP.CI.L = PROP.UCELLS - 1.96*PROP.SE,
                   PROP.CI.U = PROP.UCELLS + 1.96*PROP.SE)



(ggplot(cells_UNU_nw, aes(MONTH, PROP.UCELLS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "National level",
         x = "Month", y = "Proportion of urban cells")) /
((ggplot(filter(cells_UNU_sw, STATE == "Kerala"), 
         aes(MONTH, PROP.UCELLS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Kerala",
         x = "Month", y = "Proportion of urban cells")) |
(ggplot(filter(cells_UNU_sw, STATE == "Karnataka"), 
         aes(MONTH, PROP.UCELLS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Karnataka",
         x = "Month", y = "Proportion of urban cells"))) /
((ggplot(filter(cells_UNU_sw, STATE == "Gujarat"), 
         aes(MONTH, PROP.UCELLS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Gujarat",
         x = "Month", y = "Proportion of urban cells")) |
(ggplot(filter(cells_UNU_sw, STATE == "Maharashtra"), 
         aes(MONTH, PROP.UCELLS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Maharashtra",
         x = "Month", y = "Proportion of urban cells"))) +
plot_layout(guides = "collect") +
plot_annotation(title = "Change in proportion of urban birding",
                subtitle = "Of grid cells (2kmx2km) with at least one complete list for a time period, how many are urban?",
  ) -> s_spread_UNU_cells


ggsave("02_explor_figs/12_s_spread_UNU_cells.png",
       plot = s_spread_UNU_cells,
       width = 9, height = 9, units = "in")



```

```{r s_spread, cache=TRUE, message=FALSE}

library(rasterVis)

# using 24kmx24km for better compatibility with the 2x2 UNU resolution (facilitating whole number aggregation/averaging)

##### Change in birding effort (no. of lists) per grid cell with COVID ####

# load("data/data_UNU.RData")
# load("data/rast_UNU.RData")
# load("data/rast_SoIB.RData")


# choosing only months with data from all 3 COVID categories
month_compar <- data0 %>% 
  group_by(MONTH) %>% 
  mutate(COVID = as.character(COVID)) %>% 
  mutate(COVID = factor(case_when(COVID %in% c("DUR_20","DUR_21") ~ "DUR",
                                  TRUE ~ COVID),
                        levels = c("BEF","DUR","AFT"))) %>% 
  dplyr::summarise(N = n_distinct(COVID)) %>% 
  filter(N == 3) # all 3 COVID categories


data_map0 <- data_UNU %>% 
  filter(MONTH %in% month_compar$MONTH) %>% 
  mutate(CELL.LONG = raster::xyFromCell(rast_SoIB, CELL.ID)[,"x"],
         CELL.LAT = raster::xyFromCell(rast_SoIB, CELL.ID)[,"y"],
         SUBCELL.LONG = raster::xyFromCell(rast_UNU, SUBCELL.ID)[,"x"],
         SUBCELL.LAT = raster::xyFromCell(rast_UNU, SUBCELL.ID)[,"y"]) %>%
  group_by(COVID, CELL.ID, CELL.LONG, CELL.LAT, MONTH) %>% 
  dplyr::summarise(NO.LISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>% 
  group_by(COVID, CELL.ID, CELL.LONG, CELL.LAT) %>% 
  dplyr::summarise(NO.LISTS = mean(NO.LISTS)) %>% 
  ungroup()


data_map1 <- data_map0 %>% 
  mutate(NO.LISTS = case_when(COVID == "BEF" ~ NO.LISTS,
                              TRUE ~ 0)) %>% 
  select(CELL.LONG, CELL.LAT, NO.LISTS) %>% 
  mutate(NO.LISTS = log10(NO.LISTS + 1))

rast_map1 <- raster::rasterFromXYZ(xyz = data_map1,
                                   crs = raster::crs(rast_SoIB))


data_map2 <- data_map0 %>% 
  mutate(NO.LISTS = case_when(COVID %in% c("DUR_20","DUR_21") ~ NO.LISTS,
                              TRUE ~ 0)) %>% 
  group_by(CELL.ID, CELL.LONG, CELL.LAT) %>% 
  dplyr::summarise(NO.LISTS = case_when(n() > 1 ~ mean(NO.LISTS), 
                                 TRUE ~ NO.LISTS),
            COVID = "DUR") %>% 
  ungroup() %>% 
  select(CELL.LONG, CELL.LAT, NO.LISTS) %>% 
  mutate(NO.LISTS = log10(NO.LISTS + 1))

rast_map2 <- raster::rasterFromXYZ(xyz = data_map2,
                                   crs = raster::crs(rast_SoIB))

data_map3 <- data_map0 %>% 
  mutate(NO.LISTS = case_when(COVID == "AFT" ~ NO.LISTS,
                              TRUE ~ 0)) %>% 
  select(CELL.LONG, CELL.LAT, NO.LISTS) %>% 
  mutate(NO.LISTS = log10(NO.LISTS + 1))

rast_map3 <- raster::rasterFromXYZ(xyz = data_map3,
                                   crs = raster::crs(rast_SoIB))



# rast_propchange <- function(x, y)  {
#   # x is first, y is second, so change from x to y
#   return((y-x)/x)
# }

rast_logpropchange <- function(x, y)  {
  # x is first, y is second, so change from x to y
  # when no. of lists has been log transformed
  # log(n2) - log(n1) = log(n2/n1)
  return(y-x)
}


aoi0 <- geodata::gadm("India", level = 0, path = "data")
in_bound <- as(aoi, "Spatial") # converting for sp::sp.polygons()


data_map <- raster::stack(
  raster::overlay(x = rast_map1, y = rast_map2, 
                               fun = rast_logpropchange),
  raster::overlay(x = rast_map2, y = rast_map3, 
                               fun = rast_logpropchange),
  raster::overlay(x = rast_map1, y = rast_map3, 
                               fun = rast_logpropchange))
names(data_map) <- c("COVID", "After", "BeforeAfter")


mytheme <- list(strip.border$col = 'grey80',
                strip.background$col = 'grey80',
                axis.line$col = 'grey80',
                axis.text$col = 'grey60',
                plot.symbol$pch = 20,
                plot.symbol$cex = .5,
                plot.symbol$col = '#7AC5CD',
                plot.symbol$alpha = .80)

levelplot(data_map, 
          par.settings = list(background = list(col = "black"), 
                              col.regions = viridis::cividis(n = 1000)), 
          # trial and error to decide breaks and limits
          margin = F, at = seq(-2.5, 3, 0.1),
          main = "Log prop. change in monthly (select months) no. of lists per 25kmx25km cell") +
  latticeExtra::layer(sp::sp.polygons(in_bound, lwd = 1.5))



rasterVis::gplot(data_map) + geom_tile(aes(fill = value))




s_spread

ggsave("02_explor_figs/11_s_spread.png", plot = s_spread,
       width = 14, height = 5, units = "in")

rm(data_cov, data_map, data_sp, s_cover, s_spread, districtmap,
   g2clip, g3clip, gridmapg1, gridmapg2, gridmapg3, gridmapg4, india,
   indiamap, statemap, states, temp, gridlevels)


```


* Both the maps and the graphs show that there were overall more cells covered during the pandemic, and even more after. This can be attributed to general growth in eBirding.  
* Proportion of grid cells covered was impacted in both 2020 and 2021 in May, while in 2020 only in April.  
* SD of number of lists per cell was higher during the lockdown/second wave months, but otherwise not very different between the three periods.


### Temporal spread

Not "per observer", as this metric from data POV. Includes duplicate lists, of course.

```{r t_spread_1, cache=TRUE, message=FALSE}

temp1 <- data0 %>% 
  group_by(SAMPLING.EVENT.IDENTIFIER) %>% 
  slice(1) %>% ungroup() %>% 
  group_by(COVID, YEAR, MONTH, STATE) %>% 
  summarise(NO.LISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>%
  ungroup()

temp2 <- data0 %>% 
  group_by(SAMPLING.EVENT.IDENTIFIER) %>% 
  slice(1) %>% ungroup() %>% 
  mutate(DAY.W = wday(OBSERVATION.DATE, 
                      week_start = getOption("lubridate.week.start", 1))) %>% 
  group_by(COVID, MONTH, STATE, DAY.W) %>% 
  summarise(DAY.LISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>%
  ungroup() %>% 
  complete(COVID, MONTH, STATE, DAY.W, fill = list(DAY.LISTS = 0))

temp3 <- data0 %>% 
  group_by(SAMPLING.EVENT.IDENTIFIER) %>% 
  slice(1) %>% ungroup() %>% 
  group_by(COVID, MONTH, STATE, TIME.D) %>% 
  summarise(TIME.LISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>%
  ungroup() %>% 
  complete(COVID, MONTH, STATE, TIME.D, fill = list(TIME.LISTS = 0))

# Daman and Diu issue

```

#### Day-of-week

```{r t_spread_2, cache=TRUE, message=FALSE}

# month is not important in this metric, so first only taking 9:12 which have "AFT" also, then dropping month identity and averaging across the months

t_dow_sw <- temp1 %>% 
  left_join(temp2) %>% 
  filter(MONTH %in% 9:12) %>% 
  mutate(DAY.LISTS = ifelse(is.na(DAY.LISTS), 0, DAY.LISTS)) %>% 
  group_by(COVID, YEAR, STATE, DAY.W) %>% 
  summarise(NO.LISTS = sum(NO.LISTS),
            DAY.LISTS = sum(DAY.LISTS),
            PROP.LISTS = DAY.LISTS/NO.LISTS,
            PROP.SE = sqrt((PROP.LISTS)*(1 - PROP.LISTS)/NO.LISTS),
            PROP.CI.L = PROP.LISTS - 1.96*PROP.SE,
            PROP.CI.U = PROP.LISTS + 1.96*PROP.SE)

t_dow_nw <- t_dow_sw %>% ungroup() %>% 
  group_by(COVID, YEAR, DAY.W) %>% 
  summarise(PROP.LISTS = mean(PROP.LISTS),
            PROP.SE = sqrt(sum((PROP.SE)^2))/n(),
            PROP.CI.L = PROP.LISTS - 1.96*PROP.SE,
            PROP.CI.U = PROP.LISTS + 1.96*PROP.SE)

(ggplot(t_dow_nw, 
       aes(as.factor(DAY.W), PROP.LISTS, colour = COVID)) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette3) +
    labs(title = "Temporal spread (DoW) of checklists at national level (Sep to Dec)",
         x = "Day", y = "Proportion of lists")) /
((ggplot(filter(t_dow_sw, STATE == "Kerala"), 
         aes(as.factor(DAY.W), PROP.LISTS, colour = COVID)) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette3) +
    labs(title = "Kerala",
         x = "Day", y = "Proportion of lists")) |
ggplot(filter(t_dow_sw, STATE == "Karnataka"), 
         aes(as.factor(DAY.W), PROP.LISTS, colour = COVID)) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette3) +
    labs(title = "Karnataka",
         x = "Day", y = "Proportion of lists")) /
((ggplot(filter(t_dow_sw, STATE == "Gujarat"), 
         aes(as.factor(DAY.W), PROP.LISTS, colour = COVID)) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette3) +
    labs(title = "Gujarat",
         x = "Day", y = "Proportion of lists")) |
ggplot(filter(t_dow_sw, STATE == "Maharashtra"), 
         aes(as.factor(DAY.W), PROP.LISTS, colour = COVID)) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette3) +
    labs(title = "Maharashtra",
         x = "Day", y = "Proportion of lists")) +
plot_layout(guides = "collect") -> t_spread2

t_spread2

ggsave("02_explor_figs/13_t_spread2.png", 
       plot = t_spread2,
       width = 11, height = 9, units = "in")


```

* At the national level, birding on Saturdays has decreased, and been balanced by more birding on Wednesdays, while Sunday birding seems to have also increased.  
* In Kerala, birding seems to have increased during Thursday--Saturday and decreased on Sunday.  
* In Karnataka there hasn't been much of a change with the pandemic, and there is still a heavy weekend bias in birding activity (Saturday and Sunday levels being similar).  
* In Gujarat and Maharashtra, Sundays see more birding than Saturdays, but there has been no major impact of the pandemic.  
* This might mean that the national level patterns might be driven by changes in states with generally low birding activity.  

#### Time-of-day

```{r t_spread_3, cache=TRUE, message=FALSE}

# month is not important in this metric, so first only taking 9:12 which have "AFT" also, then dropping month identity and averaging across the months

t_tod_sw <- temp1 %>% 
  left_join(temp3) %>% 
  filter(MONTH %in% 9:12) %>% 
  mutate(TIME.LISTS = ifelse(is.na(TIME.LISTS), 0, TIME.LISTS)) %>% 
  group_by(COVID, YEAR, STATE, TIME.D) %>% 
  summarise(NO.LISTS = sum(NO.LISTS),
            TIME.LISTS = sum(TIME.LISTS),
            PROP.LISTS = TIME.LISTS/NO.LISTS,
            PROP.SE = sqrt((PROP.LISTS)*(1 - PROP.LISTS)/NO.LISTS),
            PROP.CI.L = PROP.LISTS - 1.96*PROP.SE,
            PROP.CI.U = PROP.LISTS + 1.96*PROP.SE)

t_tod_nw <- t_tod_sw %>% ungroup() %>% 
  group_by(COVID, YEAR, TIME.D) %>% 
  summarise(PROP.LISTS = mean(PROP.LISTS),
            PROP.SE = sqrt(sum((PROP.SE)^2))/n(),
            PROP.CI.L = PROP.LISTS - 1.96*PROP.SE,
            PROP.CI.U = PROP.LISTS + 1.96*PROP.SE)


(ggplot(t_tod_nw, 
       aes(as.factor(TIME.D), PROP.LISTS, colour = COVID)) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette3) +
    labs(title = "Temporal spread (ToD) of checklists at national level (Sep to Dec)",
         x = "Time (hours)", y = "Proportion of lists")) /
((ggplot(filter(t_tod_sw, STATE == "Kerala"), 
         aes(as.factor(TIME.D), PROP.LISTS, colour = COVID)) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette3) +
    labs(title = "Kerala",
         x = "Time (hours)", y = "Proportion of lists")) |
ggplot(filter(t_tod_sw, STATE == "Karnataka"), 
         aes(as.factor(TIME.D), PROP.LISTS, colour = COVID)) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette3) +
    labs(title = "Karnataka",
         x = "Time (hours)", y = "Proportion of lists")) /
((ggplot(filter(t_tod_sw, STATE == "Gujarat"), 
         aes(as.factor(TIME.D), PROP.LISTS, colour = COVID)) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette3) +
    labs(title = "Gujarat",
         x = "Time (hours)", y = "Proportion of lists")) |
ggplot(filter(t_tod_sw, STATE == "Maharashtra"), 
         aes(as.factor(TIME.D), PROP.LISTS, colour = COVID)) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette3) +
    labs(title = "Maharashtra",
         x = "Time (hours)", y = "Proportion of lists")) +
plot_layout(guides = "collect") -> t_spread3

t_spread3

ggsave("02_explor_figs/14_t_spread3.png", 
       plot = t_spread3,
       width = 11, height = 9, units = "in")


rm(temp1, temp2, temp3, t_dow_nw, t_dow_sw, t_tod_nw, t_tod_sw,
   t_spread2, t_spread3)

```

* At the national level, there are two peaks in the day that see highest birding activity: early morning and evening.  
* In 2020, there was a sharp rise in birding activity at 17:00, and dip at 16:00.  
* In 2021, birding increased at 06:00 and 15:00, and decreased at 16:00.  
* The increase in evening birding was prominent in Karnataka, Gujarat and Maharashtra.

### Media

Not "per observer", as this metric from data POV. Includes duplicate lists, of course.

```{r media, cache=TRUE, message=FALSE}

temp1 <- data0 %>% 
  group_by(SAMPLING.EVENT.IDENTIFIER) %>% 
  slice(1) %>% ungroup() %>% 
  group_by(COVID, YEAR, MONTH, STATE) %>% 
  summarise(NO.LISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>%
  ungroup() 

temp2 <- data0 %>% 
  group_by(SAMPLING.EVENT.IDENTIFIER) %>% 
  slice(1) %>% ungroup() %>% 
  filter(HAS.MEDIA == 1) %>% 
  group_by(COVID, MONTH, STATE) %>% 
  summarise(NO.MLISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>%
  ungroup() %>% 
  complete(COVID, MONTH, STATE, fill = list(NO.MLISTS = 0))


media_sw <- temp1 %>% 
  left_join(temp2) %>% 
  mutate(NO.MLISTS = ifelse(is.na(NO.MLISTS), 0, NO.MLISTS),
         PROP.MLISTS = NO.MLISTS/NO.LISTS,
         PROP.SE = sqrt((PROP.MLISTS)*(1 - PROP.MLISTS)/NO.LISTS),
         PROP.CI.L = PROP.MLISTS - 1.96*PROP.SE,
         PROP.CI.U = PROP.MLISTS + 1.96*PROP.SE) 

rm(list = c("temp1","temp2"))

media_nw <- media_sw %>% ungroup() %>% 
  group_by(COVID, YEAR, MONTH) %>% 
  summarise(PROP.MLISTS = mean(PROP.MLISTS),
            PROP.SE = sqrt(sum((PROP.SE)^2))/n(),
            PROP.CI.L = PROP.MLISTS - 1.96*PROP.SE,
            PROP.CI.U = PROP.MLISTS + 1.96*PROP.SE)


# Daman and Diu lists not appearing from 2020--2021


(ggplot(media_nw, aes(MONTH, PROP.MLISTS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Change in proportion of media lists at national level",
         x = "Month", y = "Proportion of media lists")) /
((ggplot(filter(media_sw, STATE == "Kerala"), 
         aes(MONTH, PROP.MLISTS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Kerala",
         x = "Month", y = "Proportion of media lists")) |
(ggplot(filter(media_sw, STATE == "Karnataka"), 
         aes(MONTH, PROP.MLISTS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Karnataka",
         x = "Month", y = "Proportion of media lists"))) /
((ggplot(filter(media_sw, STATE == "Gujarat"), 
         aes(MONTH, PROP.MLISTS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Gujarat",
         x = "Month", y = "Proportion of media lists")) |
(ggplot(filter(media_sw, STATE == "Maharashtra"), 
         aes(MONTH, PROP.MLISTS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = PROP.CI.L, ymax = PROP.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Maharashtra",
         x = "Month", y = "Proportion of media lists"))) +
plot_layout(guides= "collect") -> media

media

ggsave("02_explor_figs/15_media.png", plot = media,
       width = 9, height = 9, units = "in")

rm(list = c("media_nw","media_sw","media"))

```

* There is no major difference in proportion of media lists before, during and after the pandemic.  

### New birders

```{r new, cache=TRUE, message=FALSE}

load("data/new_obsr_data.RData")

# for left joining YEAR with COVID
temp1 <- new_obsr_data %>% 
  distinct(COVID, LE.YEAR, LE.MONTH)
         
# not new to the state, but new overall, just grouped by state of first eBirding
temp2 <- new_obsr_data %>% 
  filter(COVID != "NA") %>% 
  group_by(COVID, LE.MONTH, STATE) %>% 
  summarise(NEW = n_distinct(OBSERVER.ID)) %>% 
  ungroup() %>% 
  complete(COVID, LE.MONTH, STATE, fill = list(NEW = 0))

new_obsr_sw <- temp1 %>% 
  left_join(temp2) %>% 
  mutate(NEW = ifelse(is.na(NEW), 0, NEW))

set.seed(765)

new_obsr_nw <- new_obsr_sw %>% 
  ungroup() %>% 
  group_by(COVID, LE.YEAR, LE.MONTH) %>% 
  summarise(NEW.SE = boot_se(NEW),
            NEW = mean(NEW),
            NEW.CI.L = NEW - 1.96*NEW.SE,
            NEW.CI.U = NEW + 1.96*NEW.SE)


(ggplot(new_obsr_nw, aes(LE.MONTH, NEW, colour = COVID)) + 
    scale_y_log10() +
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = NEW.CI.L, ymax = NEW.CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Monthly number of new eBirders at national level",
         x = "Month", y = "")) /
((ggplot(filter(new_obsr_sw, STATE == "Kerala"), 
         aes(LE.MONTH, NEW, colour = COVID)) + 
    scale_y_log10() +
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Kerala",
         x = "Month", y = "Number of new eBirders")) |
(ggplot(filter(new_obsr_sw, STATE == "Karnataka"), 
         aes(LE.MONTH, NEW, colour = COVID)) + 
    scale_y_log10() +
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Karnataka",
         x = "Month", y = ""))) /
((ggplot(filter(new_obsr_sw, STATE == "Gujarat"), 
         aes(LE.MONTH, NEW, colour = COVID)) + 
    scale_y_log10() +
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Gujarat",
         x = "Month", y = "")) |
(ggplot(filter(new_obsr_sw, STATE == "Maharashtra"), 
         aes(LE.MONTH, NEW, colour = COVID)) + 
    scale_y_log10() +
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Maharashtra",
         x = "Month", y = ""))) +
plot_layout(guides = "collect") -> new

new # plot

ggsave("02_explor_figs/16_new.png", plot = new,
       width = 9, height = 9, units = "in")


# Gujarat and ArP showing zero eBirders in certain cases. Probably due to random subsampling.


rm(list = c("new_obsr_nw","new_obsr_sw","new"))

```

* At the national level, the pandemic has seen a greater number of monthly new eBirders than before, presumably due to greater frequency and impact of outreach initiatives.  
* While Kerala and Karnataka both have seen increases in monthly new eBirders with the pandemic, the increase is more striking in Karnataka than in Kerala. 


# References

