---
title: "COVID eBirding: Analyses"
date: "2022/05/02"
author: "Karthik Thrikkadeeri"
editor_options: 
  chunk_output_type: console
bibliography: covid-ebirding.bib
link-citations: yes
output: 
  html_document:
    toc: true
    toc_depth: 3  
    number_sections: false  
    theme: united  
    highlight: tango  # specifies the syntax highlighting style
---

```{r setup, include=FALSE}

library(tidyverse)
library(lubridate)
library(patchwork)
library(boot)
library(lme4)
library(glue)
library(spdep)
library(sf)
library(parallel) # for bootstrapping error from GLMMs
library(colorspace) # for ability to set midpoint in ggplot scale colour
library(ggallin) # for inverse arcsine transformation of ggplot scale

theme_set(theme_classic())


covid_palette <- c("#1B9E77", "#EF4050", "#E89005", "#9678B6")
# covid_palette2 <- c("#1B9E77", "#D95F02", "#7570B3", "#555555")
covid_palette3 <- c("#1B9E77", "#E89005", "#9678B6")


anal_states <- data.frame(a = "Karnataka",
                          b = "Kerala",
                          c = "Maharashtra",
                          d = "Assam")


load("00_data/maps_sf.RData")


knitr::opts_chunk$set(echo = FALSE, warnings = FALSE, 
                      messages = FALSE, cache.lazy = FALSE)

```

```{r data, cache=TRUE}

source("00_scripts/functions.R")

# Only needed once in the beginning to obtain the data.

# Getting MODIS data ###
if (!file.exists("00_data/in_LULC_MODIS/in_LULC_MODIS.tif") & 
    !file.exists("00_data/rast_UNU.RData")) {

  getmodisdata()

} else {
  print("MODIS data is ready to use!")
}

# #  Getting EBD data + filtering and preparing for current use (61 mins on server) ###
# tictoc::tic("Preparing data for analyses")
# rawdatapath <- "00_data/ebd_IN_prv_relMay-2022.txt"
# senspath <- "00_data/ebd_sensitive_relMay-2022_IN.txt"
# datapath <- "00_data/ebd_IN_relMay-2022.RData"
# groupaccspath <- "00_data/ebd_users_GA_relMay-2022.csv"
# covidclasspath <- "00_data/covid_classification.csv"
# rast_UNU_path <- "00_data/rast_UNU.RData"
# 
# data_qualfilt_prep(rawdatapath, senspath, datapath, groupaccspath, covidclasspath,
#                    rast_UNU_path,
#                    maxvel = 20, minsut = 2)
# tictoc::toc()


load("00_data/data0_MY_d_slice.RData")
load("00_data/grids_st_sf.RData")


```

```{r setup2, include=FALSE}

# timeline metadata
timeline <- data0_MY_d_slice_G %>% distinct(COVID, YEAR, MONTH, M.YEAR)

# finding main state of birders (mainly for observer-level metrics)
obs_mainstates <- data0_MY_d_slice_S %>% 
  group_by(OBSERVER.ID, STATE) %>% 
  dplyr::summarise(LISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>% 
  arrange(desc(LISTS)) %>% 
  slice(1) %>% 
  ungroup() %>% 
  rename(MAIN.STATE = STATE) %>% 
  dplyr::select(-LISTS)

# choosing only months with data from all 3 COVID categories
month_compar <- data0_MY_d_slice_G %>% 
  group_by(MONTH) %>% 
  mutate(COVID = factor(case_when(as.character(COVID) %in% c("DUR_20","DUR_21") ~ "DUR",
                                  TRUE ~ as.character(COVID)),
                        levels = c("BEF","DUR","AFT"))) %>% 
  dplyr::summarise(N = n_distinct(COVID)) %>% 
  filter(N == 3) %>%  # all 3 COVID categories
  dplyr::select(MONTH)

# removing unnecessary objects from maps_sf.RData
rm(g2_in_sf, g3_in_sf, g4_in_sf,
   g2_st_sf, g3_st_sf, g4_st_sf)

```

# Introduction

This study will focus on the impacts of the COVID-19 pandemic on both birds and the birding community in the Indian context. It will also try to understand the consequences of these changes on citizen science and the inferences that can be drawn from it. Lastly, it will propose guidelines for making the most of the available data while accounting for pandemic-driven biases and deviations from the pre-COVID normal.

To ensure a balanced time frame, this study will focus on the years from 2018 to 2021. The time period of 2019--20 will be considered as "pre-COVID" while 2020--21 will be considered "COVID". In order to account for within-year variations in bird and birder behaviour, the comparisons will be made on the month level. For instance, various metrics of birding behaviour in the month of June will be compared between 2018--19 and 2020--21, and will not be confounded with patterns seen in other months.

***This notebook is the almost-final analysis notebook for the study. Here, following the interesting patterns/metrics discovered in 02_explor, I explore whether or not bird detectability has changed. I also build statistical models to prove that the patterns found in 02_explor are scientifically robust.***

## Questions

Are there any changes and gaps in the pandemic eBird data? These questions will be answered using numerous metrics calculated from the focal data point in eBird data, a checklist.

-   What has happened to overall birding activity?

    Overall birding activity has increased.

-   What has happened to group birding? Has it decreased, as would be expected?

    Group birding decreased as expected.

-   Has birding effort changed with the pandemic?

    Birding effort has not decreased with the pandemic.

-   Are there biases in the pandemic birding activity in terms of spatial and temporal spread?

    There do not seem to be many large spatial or temporal biases in pandemic birding activity.

-   Has there been a change in detectability of certain common urban birds during the pandemic?

-   Is this change due to the behavioural changes in birdwatchers (such as increased urban birding) in 2020, especially around the months of Mar--May, or did detectability of birds truly increase?

# Birder behaviour (focus level: checklists)

```{r child = "04_birder.Rmd"}

```

# Bird behaviour

```{r child = "05_bird.Rmd"}

```

# References
