---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Birder behaviour (focus level: checklists)

Changes in eBirder behaviour will be analysed using the focal data point, checklists. A number of metrics based on checklists will be calculated and analysed (**at the monthly scale**), such as:

1.  Number of submitted checklists
    1.  Total lists\
    2.  Shared lists (use no. of obs.)\
2.  Number of submitted checklists per observer
    1.  Total lists\
    2.  Shared lists (use no. of obs.)\
3.  Hotspot birding\
4.  Birding protocol\
5.  Birding distance
6.  Site fidelity per observer\
7.  Habitats surveyed\
8.  Spatial spread of birding activity. These spatial patterns will be explored at two scales too: nationwide and regional.
    1.  Spatial spread by absences\
    2.  Spatial spread by densities\
9.  List duration
10. List length\
11. Birding time per observer (monthly for prelim, later maybe daily)\
12. Temporal spread of birding activity
    1.  Temporal spread within day (time-of-day)\
    2.  Temporal spread within week (day-of-week)\
13. Media\
14. Number of new birders

When running the actual models, observer will be used as a random effect because all of these metrics would vary differently from observer to observer.

Post-presentation: regarding averaging national level patterns by state for some metrics, this means that means are manually calculated for states and the points plotted in the graph are only these single points per state (instead of all the raw data points).

GLMMs for birder behaviour run only for the following metrics:

-   Group birding per observer
-   Site fidelity per observer
-   Birding time per observer
-   Hotspot birding
-   Birding protocol
-   Birding distance
-   List duration
-   List length
-   Proportion of urban birding
-   Spatial coverage
-   Spatial spread (slightly different)
-   Temporal spread

These are usually driven by decision at the state level, so since the models are being run for the states individually, no need to consider district.

### Group birding (\>2 observer) per observer

```{r d1_group_po, cache=TRUE, message=FALSE}

##### Change in group birding per observer ####

temp1 <- data0_MY_slice_S %>% 
  group_by(COVID, M.YEAR, MONTH, STATE, OBSERVER.ID) %>% 
  dplyr::summarise(NO.LISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>%
  ungroup() 

temp2 <- data0_MY_slice_S %>% 
  filter(NUMBER.OBSERVERS > 2) %>% 
  group_by(COVID, M.YEAR, MONTH, STATE, OBSERVER.ID) %>% 
  dplyr::summarise(NO.SLISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>%
  ungroup()

# this is for each observer, but we are interested in plotting mean value of all observers combined
# hence, averaging proportions and its SE by propagation of error

nl_po_sw <- temp1 %>% 
  left_join(temp2) %>% 
  mutate(NO.SLISTS = replace_na(NO.SLISTS, 0)) %>% 
  mutate(PROP.SLISTS = NO.SLISTS/NO.LISTS,
         SE = sqrt((PROP.SLISTS)*(1 - PROP.SLISTS)/NO.LISTS)) %>% 
  group_by(COVID, M.YEAR, MONTH, STATE) %>% 
  dplyr::summarise(PROP.SLISTS = mean(PROP.SLISTS),
            SE = sqrt(sum((SE)^2))/n(),
            CI.L = PROP.SLISTS - 1.96*SE,
            CI.U = PROP.SLISTS + 1.96*SE)

nl_po_nw <- nl_po_sw %>% 
  group_by(COVID, M.YEAR, MONTH) %>% 
  dplyr::summarise(PROP.SLISTS = mean(PROP.SLISTS),
            SE = sqrt(sum((SE)^2))/n(),
            CI.L = PROP.SLISTS - 1.96*SE,
            CI.U = PROP.SLISTS + 1.96*SE)


(ggplot(nl_po_nw, aes(MONTH, PROP.SLISTS, colour = COVID)) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "National level",
       x = "Month", y = "Group birding proportion")) /
((ggplot(filter(nl_po_sw, STATE == "Kerala"), 
       aes(MONTH, PROP.SLISTS, colour = COVID)) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Kerala",
       x = "Month", y = "Group birding proportion")) |
(ggplot(filter(nl_po_sw, STATE == "Karnataka"), 
       aes(MONTH, PROP.SLISTS, colour = COVID)) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Karnataka",
       x = "Month", y = "Group birding proportion"))) /
((ggplot(filter(nl_po_sw, STATE == "Maharashtra"), 
       aes(MONTH, PROP.SLISTS, colour = COVID)) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Maharashtra",
       x = "Month", y = "Group birding proportion")) |
(ggplot(filter(nl_po_sw, STATE == "Assam"), 
       aes(MONTH, PROP.SLISTS, colour = COVID)) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Assam",
       x = "Month", y = "Group birding proportion"))) +
plot_layout(guides = "collect") +
plot_annotation(title = "Change in proportion of group birding per observer with the pandemic",
                subtitle = "Y-axis shows proportion of observers' checklists that are group birding events") -> no_lists_po

no_lists_po # plot

ggsave("03_wrap_figs/01_no_lists_po.png", plot = no_lists_po,
       width = 9, height = 9, units = "in")

rm(list = c("nl_po_sw","nl_po_nw","no_lists_po"))

##### Final analysis: statewise GLMMs ####

# dataframe with empty column to populate with looped values
# total rows: product of distinct values of predictors
data0 <- data0_MY_slice_S %>% 
  filter(STATE %in% anal_states[1,]) %>% 
  tidyr::expand(nesting(STATE, MONTH, M.YEAR)) %>% 
  mutate(PRED = NA,
         SE = NA,
         STATE = factor(STATE, levels = anal_states[1,])) %>% 
  arrange(STATE)

count <- 0
for (i in 1:4) { # for each state
  
  data_a <- data0_MY_slice_S %>% 
    filter(STATE == anal_states[, i]) %>% 
    group_by(M.YEAR, MONTH, OBSERVER.ID, SAMPLING.EVENT.IDENTIFIER) %>% 
    dplyr::summarise(GROUP.BIRDING = if_else(NUMBER.OBSERVERS > 2, 1, 0)) %>%
    ungroup() 
  
  tictoc::tic(glue::glue("GLMM for {anal_states[, i]}"))
  model_a <- glmer(GROUP.BIRDING ~ M.YEAR + M.YEAR:MONTH + (1|OBSERVER.ID),
                   data = data_a, family = binomial(link = "cloglog"),
                   nAGQ = 0, control = glmerControl(optimizer = "bobyqa")) 
  tictoc::toc()

  for (j in 1:n_distinct(data0$MONTH)) {
    for (k in 1:n_distinct(data0$M.YEAR)) {
      
      count <- count + 1
      
      tictoc::tic(glue::glue("Bootstrapped predictions for {anal_states[, i]}"))
      prediction <- boot_conf_GLMM(model_a,
                                   new_data = data.frame(MONTH = data0$MONTH[count],
                                                         M.YEAR = data0$M.YEAR[count]),
                                   nsim = 100)
      tictoc::toc()
      
      data0$PRED[count] <- median(na.omit(prediction$t[count]))
      data0$SE[count] <- sd(na.omit(prediction$t[count]))

    }
  }
} 

      f2$freqt = clogloglink(f2$freq,inverse = T)
      f2$cl = clogloglink((f2$freq-f2$se),inverse = T)
      f2$set = f2$freqt-f2$cl


data0 <- data0 %>% 
    left_join(timeline)


(ggplot(filter(data0, STATE == anal_states$a), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$a,
       x = "Month", y = "Predicted group birding proportion") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period")) /
(ggplot(filter(data0, STATE == anal_states$b), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$b,
       x = "Month", y = "Predicted group birding proportion") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period")) /
(ggplot(filter(data0, STATE == anal_states$c), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$c,
       x = "Month", y = "Predicted group birding proportion") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period")) /
(ggplot(filter(data0, STATE == anal_states$d), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$d,
       x = "Month", y = "Predicted group birding proportion") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period"))

```

### Site fidelity per observer

```{r fidelity, cache=TRUE, message=FALSE}

##### Change in site fidelity per observer ####

# here, doing 24km x 24km (SoIB and also more lenient and appropriate)

set.seed(213) # for bootstrap
fidel_sw <- data0_MY_slice_S %>% 
  group_by(COVID, M.YEAR, MONTH, STATE, OBSERVER.ID) %>% 
  dplyr::summarise(NO.SITES = n_distinct(CELL.ID)) %>% 
  mutate(NO.SITES = replace_na(NO.SITES, 0)) %>% 
  group_by(COVID, M.YEAR, MONTH, STATE) %>% 
  dplyr::summarise(NO.SITES = boot_conf(x = NO.SITES)) %>% 
  group_by(COVID, M.YEAR, MONTH, STATE) %>% 
  dplyr::summarise(SE = sd(NO.SITES),
                   NO.SITES = mean(NO.SITES),
                   CI.L = NO.SITES - 1.96*SE,
                   CI.U = NO.SITES + 1.96*SE)

fidel_nw <- fidel_sw %>% 
  ungroup() %>% 
  group_by(COVID, M.YEAR, MONTH) %>% 
  dplyr::summarise(NO.SITES = mean(NO.SITES),
            SE = sqrt(sum((SE)^2))/n(),
            CI.L = NO.SITES - 1.96*SE,
            CI.U = NO.SITES + 1.96*SE)

  
(ggplot(fidel_nw, aes(MONTH, NO.SITES, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "National level",
         x = "Month", y = "No. of sites")) /
((ggplot(filter(fidel_sw, STATE == "Kerala"), 
         aes(MONTH, NO.SITES, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Kerala",
         x = "Month", y = "No. of sites")) |
(ggplot(filter(fidel_sw, STATE == "Karnataka"), 
         aes(MONTH, NO.SITES, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Karnataka",
         x = "Month", y = "No. of sites"))) /
((ggplot(filter(fidel_sw, STATE == "Maharashtra"), 
         aes(MONTH, NO.SITES, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Maharashtra",
         x = "Month", y = "No. of sites")) |
(ggplot(filter(fidel_sw, STATE == "Assam"), 
         aes(MONTH, NO.SITES, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Assam",
         x = "Month", y = "No. of sites"))) +
plot_layout(guides = "collect") +
plot_annotation(title = "Change in observer site fidelity with the pandemic",
subtitle = "Y-axis shows average number of unique sites (25kmx25km cells) visited by observers") -> fidelity

fidelity # plot

ggsave("03_wrap_figs/02_fidelity.png", plot = fidelity,
       width = 9, height = 9, units = "in")


##### Final analysis: statewise GLMMs ####

# dataframe with empty column to populate with looped values
# total rows: product of distinct values of predictors
data0 <- data0_MY_slice_S %>% 
  filter(STATE %in% anal_states[1,]) %>% 
  tidyr::expand(nesting(STATE, MONTH, M.YEAR)) %>% 
  mutate(PRED = NA,
         STATE = factor(STATE, levels = anal_states[1,])) %>% 
  arrange(STATE)

count <- 0
for (i in 1:4) { # for each state
  
  data_a <- data0_MY_slice_S %>% 
    left_join(obs_mainstates, by = "OBSERVER.ID") %>% 
    filter(MAIN.STATE == anal_states[, i]) %>% 
    group_by(M.YEAR, MONTH, OBSERVER.ID) %>% 
    dplyr::summarise(NO.SITES = n_distinct(CELL.ID)) %>% 
    mutate(NO.SITES = replace_na(NO.SITES, 0)) %>% 
    ungroup()
    
  tictoc::tic(glue::glue("GLMM for {anal_states[, i]}"))
  model_a <- glmer(NO.SITES ~ M.YEAR + M.YEAR:MONTH + (1|OBSERVER.ID),
                   data = data_a, family = poisson,
                   nAGQ = 0, control = glmerControl(optimizer = "bobyqa")) 
  tictoc::toc()

  for (j in 1:n_distinct(data0$MONTH)) {
    for (k in 1:n_distinct(data0$M.YEAR)) {
      count <- count + 1
      
      data0$PRED[count] = predict(model_a,
                                  data.frame(MONTH = data0$MONTH[count],
                                             M.YEAR = data0$M.YEAR[count]),
                                  re.form = NA, type = "response")
    }
  }
} 

data0 <- data0 %>% 
    left_join(timeline)


(ggplot(filter(data0, STATE == anal_states$a), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$a,
       x = "Month", y = "Predicted number of sites an observer visits") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period")) /
(ggplot(filter(data0, STATE == anal_states$b), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$b,
       x = "Month", y = "Predicted number of sites an observer visits") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period")) /
(ggplot(filter(data0, STATE == anal_states$c), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$c,
       x = "Month", y = "Predicted number of sites an observer visits") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period")) /
(ggplot(filter(data0, STATE == anal_states$d), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$d,
       x = "Month", y = "Predicted number of sites an observer visits") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period"))

```

-   At the national level, site fidelity was generally higher during the pandemic than before but mostly only around lockdown/second wave. In other months patterns closely followed pre-pandemic ones.

### Birding time per observer

```{r time, cache=TRUE, message=FALSE}

##### Change in birding time per observer ####

set.seed(423) 
time_sw <- data0_MY_slice_S %>% 
  group_by(COVID, M.YEAR, MONTH, STATE, OBSERVER.ID) %>% 
  dplyr::summarise(B.TIME = sum(DURATION.MINUTES)) %>% 
  mutate(B.TIME = replace_na(B.TIME, 0)) %>% 
  group_by(COVID, M.YEAR, MONTH, STATE) %>% 
  dplyr::summarise(B.TIME = boot_conf(x = B.TIME)) %>% 
  group_by(COVID, M.YEAR, MONTH, STATE) %>% 
  dplyr::summarise(SE = sd(B.TIME),
                   B.TIME = mean(B.TIME),
                   CI.L = B.TIME - 1.96*SE,
                   CI.U = B.TIME + 1.96*SE)

time_nw <- time_sw %>% 
  group_by(COVID, M.YEAR, MONTH) %>% 
  dplyr::summarise(B.TIME = mean(B.TIME),
            SE = sqrt(sum((SE)^2))/n(),
            CI.L = B.TIME - 1.96*SE,
            CI.U = B.TIME + 1.96*SE)


(ggplot(time_nw, aes(MONTH, B.TIME, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "National level",
       x = "Month", y = "Time (mins)")) /
((ggplot(filter(time_sw, STATE == "Kerala"), 
         aes(MONTH, B.TIME, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Kerala",
       x = "Month", y = "Time (mins)")) |
(ggplot(filter(time_sw, STATE == "Karnataka"), 
         aes(MONTH, B.TIME, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Karnataka",
       x = "Month", y = "Time (mins)"))) /
((ggplot(filter(time_sw, STATE == "Maharashtra"), 
         aes(MONTH, B.TIME, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Maharashtra",
       x = "Month", y = "Time (mins)")) |
(ggplot(filter(time_sw, STATE == "Assam"), 
         aes(MONTH, B.TIME, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Assam",
       x = "Month", y = "Time (mins)"))) +
plot_layout(guides = "collect") + 
plot_annotation(title = "Change in birding time per observer with the pandemic",
                subtitle = "Y-axis shows total birding times of observers") -> time

time # plot

ggsave("03_wrap_figs/03_time.png", plot = time,
       width = 9, height = 9, units = "in")

rm(time_nw, time_sw, time)


##### Final analysis: statewise GLMMs ####

# dataframe with empty column to populate with looped values
# total rows: product of distinct values of predictors
data0 <- data0_MY_slice_S %>% 
  filter(STATE %in% anal_states[1,]) %>% 
  tidyr::expand(nesting(STATE, MONTH, M.YEAR)) %>% 
  mutate(PRED = NA,
         STATE = factor(STATE, levels = anal_states[1,])) %>% 
  arrange(STATE)

count <- 0
for (i in 1:4) { # for each state
  
  data_a <- data0_MY_slice_S %>% 
    left_join(obs_mainstates, by = "OBSERVER.ID") %>% 
    filter(MAIN.STATE == anal_states[, i]) %>% 
    group_by(MONTH, M.YEAR, OBSERVER.ID) %>% 
    dplyr::summarise(B.TIME = sum(DURATION.MINUTES)) %>% 
    mutate(B.TIME = replace_na(B.TIME, 0)) %>% 
    ungroup()
  
  tictoc::tic(glue::glue("GLMM for {anal_states[, i]}"))
  model_a <- glmer(B.TIME ~ M.YEAR + M.YEAR:MONTH + (1|OBSERVER.ID),
                   data = data_a, family = poisson,
                   nAGQ = 0, control = glmerControl(optimizer = "bobyqa")) 
  tictoc::toc()

  for (j in 1:12) {
    for (k in 1:4) {
      count <- count + 1
      
      data0$PRED[count] = predict(model_a,
                                  data.frame(MONTH = data0$MONTH[count],
                                             M.YEAR = data0$M.YEAR[count]),
                                  re.form = NA, type = "response")
    }
  }
} 

data0 <- data0 %>% 
    left_join(timeline)


(ggplot(filter(data0, STATE == anal_states$a), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$a,
       x = "Month", y = "Predicted birding time of an observer") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period")) /
(ggplot(filter(data0, STATE == anal_states$b), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$b,
       x = "Month", y = "Predicted birding time of an observer") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period")) /
(ggplot(filter(data0, STATE == anal_states$c), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$c,
       x = "Month", y = "Predicted birding time of an observer") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period")) /
(ggplot(filter(data0, STATE == anal_states$d), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$d,
       x = "Month", y = "Predicted birding time of an observer") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period"))

```

-   At the national level, birding times were and are lowest during monsoon. There was no noticeable negative impact of the pandemic on monthly birding times, except in March 2020 when the pandemic was a novelty. In fact, in the monsoon months of July and August, birding times even increased slightly during the pandemic.\
-   The only other major pattern visible at the state level is in Karnataka, where birding times have increased in the months of August--October.
-   Karnataka and Assam showed a lowered birding duration around the peak months, while Kerala and Maharashtra showed no difference.

### Hotspot birding

Not "per observer", as this metric from data POV.

```{r hotspot, cache=TRUE, message=FALSE}

##### Change in hotspot birding ####

temp1 <- data0_MY_slice_G %>% 
  group_by(COVID, M.YEAR, MONTH, STATE) %>% 
  dplyr::summarise(NO.LISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>%
  ungroup() 

temp2 <- data0_MY_slice_G %>% 
  filter(LOCALITY.TYPE == "H") %>% 
  group_by(COVID, M.YEAR, MONTH, STATE) %>% 
  dplyr::summarise(NO.HLISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>%
  ungroup()

hot_sw <- temp1 %>% 
  left_join(temp2) %>% 
  mutate(NO.HLISTS = replace_na(NO.HLISTS, 0)) %>% 
  group_by(COVID, M.YEAR, MONTH, STATE) %>% 
  mutate(PROP.HLISTS = NO.HLISTS/NO.LISTS,
         SE = sqrt((PROP.HLISTS)*(1 - PROP.HLISTS)/NO.LISTS),
         CI.L = PROP.HLISTS - 1.96*SE,
         CI.U = PROP.HLISTS + 1.96*SE) 

rm(list = c("temp1","temp2"))

hot_nw <- hot_sw %>%
  filter(NO.LISTS != 0) %>% 
  group_by(COVID, M.YEAR, MONTH) %>% 
  dplyr::summarise(PROP.HLISTS = mean(PROP.HLISTS),
            SE = sqrt(sum((SE)^2))/n(),
            CI.L = PROP.HLISTS - 1.96*SE,
            CI.U = PROP.HLISTS + 1.96*SE) 


(ggplot(hot_nw, aes(MONTH, PROP.HLISTS, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "National level",
       x = "Month", y = "Hotspot birding proportion")) /
((ggplot(filter(hot_sw, STATE == "Kerala"),
         aes(MONTH, PROP.HLISTS, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Kerala",
       x = "Month", y = "Hotspot birding proportion")) |
(ggplot(filter(hot_sw, STATE == "Karnataka"),
         aes(MONTH, PROP.HLISTS, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Karnataka",
       x = "Month", y = "Hotspot birding proportion"))) /
((ggplot(filter(hot_sw, STATE == "Maharashtra"),
         aes(MONTH, PROP.HLISTS, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Maharashtra",
       x = "Month", y = "Hotspot birding proportion")) |
(ggplot(filter(hot_sw, STATE == "Assam"),
         aes(MONTH, PROP.HLISTS, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.25, width = 0.4, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  labs(title = "Assam",
       x = "Month", y = "Hotspot birding proportion"))) +
plot_layout(guides = "collect") +
plot_annotation(title = "Change in proportion of hotspot birding with the pandemic",
                subtitle = "Y-axis shows proportion of checklists that are from hotspots") -> hotspot

hotspot # plot

ggsave("03_wrap_figs/04_hotspot.png", plot = hotspot,
       width = 9, height = 9, units = "in")

rm(list = c("hot_nw","hot_sw", "hotspot"))


##### Final analysis: statewise GLMMs ####

# dataframe with empty column to populate with looped values
# total rows: product of distinct values of predictors
data0 <- data0_MY_slice_G %>% 
  filter(STATE %in% anal_states[1,]) %>% 
  tidyr::expand(nesting(STATE, MONTH, M.YEAR)) %>% 
  mutate(PRED = NA,
         STATE = factor(STATE, levels = anal_states[1,])) %>% 
  arrange(STATE)

count <- 0
for (i in 1:4) { # for each state
  
  data_a <- data0_MY_slice_G %>% 
    filter(STATE == anal_states[, i]) %>% 
    group_by(M.YEAR, MONTH, CELL.ID, SAMPLING.EVENT.IDENTIFIER) %>% 
    dplyr::summarise(HOTSPOT = if_else(LOCALITY.TYPE == "H", 1, 0)) %>%
    ungroup() 
  
  tictoc::tic(glue::glue("GLMM for {anal_states[, i]}"))
  model_a <- glmer(HOTSPOT ~ M.YEAR + M.YEAR:MONTH + (1|CELL.ID),
                   data = data_a, family = binomial(link = "cloglog"),
                   nAGQ = 0, control = glmerControl(optimizer = "bobyqa")) 
  tictoc::toc()

  for (j in 1:12) {
    for (k in 1:4) {
      count <- count + 1
      
      data0$PRED[count] = predict(model_a,
                                  data.frame(MONTH = data0$MONTH[count],
                                             M.YEAR = data0$M.YEAR[count]),
                                  re.form = NA, type = "response")
    }
  }
} 

data0 <- data0 %>% 
    left_join(timeline)


(ggplot(filter(data0, STATE == anal_states$a), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$a,
       x = "Month", y = "Predicted hotspot birding proportion") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period")) /
(ggplot(filter(data0, STATE == anal_states$b), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$b,
       x = "Month", y = "Predicted hotspot birding proportion") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period")) /
(ggplot(filter(data0, STATE == anal_states$c), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$c,
       x = "Month", y = "Predicted hotspot birding proportion") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period")) /
(ggplot(filter(data0, STATE == anal_states$d), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$d,
       x = "Month", y = "Predicted hotspot birding proportion") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period"))

```

-   The national level saw lower hotspot birding during the lockdown/second wave months, with some spillover into the months of June and July.\
-   In Karnataka, this impact surprisingly lasted through all of 2020 but returned almost exactly to pre-COVID levels in 2021. The same to an extent in Kerala, though the difference was much less.
-   Maharashtra recovered after the peak period, while there was no effect on Assam.

### Birding protocol

Not "per observer", as this metric from data POV.

```{r protocol, cache=TRUE, message=FALSE}

##### Change in travelling birding ####

temp1 <- data0_MY_slice_G %>% 
  group_by(COVID, M.YEAR, MONTH, STATE) %>% 
  dplyr::summarise(NO.LISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>%
  ungroup() 

temp2 <- data0_MY_slice_G %>% 
  filter(PROTOCOL.TYPE == "Traveling" & EFFORT.DISTANCE.KM >= 0.1) %>% 
  group_by(COVID, M.YEAR, MONTH, STATE) %>% 
  dplyr::summarise(NO.TLISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>%
  ungroup()

prot_sw <- temp1 %>% 
  left_join(temp2) %>% 
  mutate(NO.TLISTS = replace_na(NO.TLISTS, 0)) %>% 
  group_by(COVID, M.YEAR, MONTH, STATE) %>% 
  mutate(PROP.TLISTS = NO.TLISTS/NO.LISTS,
         SE = sqrt((PROP.TLISTS)*(1 - PROP.TLISTS)/NO.LISTS),
         CI.L = PROP.TLISTS - 1.96*SE,
         CI.U = PROP.TLISTS + 1.96*SE) 

rm(list = c("temp1","temp2"))

prot_nw <- prot_sw %>%
  filter(NO.LISTS != 0) %>% 
  group_by(COVID, M.YEAR, MONTH) %>% 
  dplyr::summarise(PROP.TLISTS = mean(PROP.TLISTS),
            SE = sqrt(sum((SE)^2))/n(),
            CI.L = PROP.TLISTS - 1.96*SE,
            CI.U = PROP.TLISTS + 1.96*SE) 


(ggplot(prot_nw, aes(MONTH, PROP.TLISTS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "National level",
         x = "Month", y = "Proportion of travelling lists")) /
((ggplot(filter(prot_sw, STATE == "Kerala"), 
         aes(MONTH, PROP.TLISTS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Kerala",
         x = "Month", y = "Proportion of travelling lists")) |
(ggplot(filter(prot_sw, STATE == "Karnataka"), 
         aes(MONTH, PROP.TLISTS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
   geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Karnataka",
         x = "Month", y = "Proportion of travelling lists"))) /
((ggplot(filter(prot_sw, STATE == "Maharashtra"), 
         aes(MONTH, PROP.TLISTS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Maharashtra",
         x = "Month", y = "Proportion of travelling lists")) |
(ggplot(filter(prot_sw, STATE == "Assam"), 
        aes(MONTH, PROP.TLISTS, colour = COVID)) + 
   scale_x_continuous(breaks = 1:12) + 
   geom_point(size = 3, position = position_dodge(0.5)) +
   geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                 size = 1.25, width = 0.4, position = position_dodge(0.5)) +
   scale_colour_manual(values = covid_palette) +
   labs(title = "Assam",
        x = "Month", y = "Proportion of travelling lists"))) +
plot_layout(guides = "collect") +
plot_annotation(title = "Change in birding mobility with the pandemic",
                subtitle = "Y-axis shows proportion of checklists using the travelling protocol (>= 100m)") -> protocol

protocol 

ggsave("03_wrap_figs/05_protocol.png", plot = protocol,
       width = 9, height = 9, units = "in")

rm(list = c("prot_nw","prot_sw","protocol"))

##### Final analysis: statewise GLMMs ####

# dataframe with empty column to populate with looped values
# total rows: product of distinct values of predictors
data0 <- data0_MY_slice_G %>% 
  filter(STATE %in% anal_states[1,]) %>% 
  tidyr::expand(nesting(STATE, MONTH, M.YEAR)) %>% 
  mutate(PRED = NA,
         STATE = factor(STATE, levels = anal_states[1,])) %>% 
  arrange(STATE)

count <- 0
for (i in 1:4) { # for each state
  
  # using 300m as threshold to define travelling birding
  data_a <- data0_MY_slice_G %>% 
    filter(STATE == anal_states[, i]) %>% 
    group_by(M.YEAR, MONTH, CELL.ID, SAMPLING.EVENT.IDENTIFIER) %>% 
    dplyr::summarise(TRAVELLING = if_else((PROTOCOL.TYPE == "Traveling" & 
                                             EFFORT.DISTANCE.KM > 0.3), 1, 0)) %>%
    ungroup() 
  
  tictoc::tic(glue::glue("GLMM for {anal_states[, i]}"))
  model_a <- glmer(TRAVELLING ~ M.YEAR + M.YEAR:MONTH + (1|CELL.ID),
                   data = data_a, family = binomial(link = "cloglog"),
                   nAGQ = 0, control = glmerControl(optimizer = "bobyqa")) 
  tictoc::toc()

  for (j in 1:n_distinct(data0$MONTH)) {
    for (k in 1:n_distinct(data0$M.YEAR)) {
      count <- count + 1
      
      data0$PRED[count] = predict(model_a,
                                  data.frame(MONTH = data0$MONTH[count],
                                             M.YEAR = data0$M.YEAR[count]),
                                  re.form = NA, type = "response")
    }
  }
} 

data0 <- data0 %>% 
    left_join(timeline)


(ggplot(filter(data0, STATE == anal_states$a), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$a,
       x = "Month", y = "Predicted travelling birding proportion") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period")) /
(ggplot(filter(data0, STATE == anal_states$b), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$b,
       x = "Month", y = "Predicted travelling birding proportion") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period")) /
(ggplot(filter(data0, STATE == anal_states$c), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$c,
       x = "Month", y = "Predicted travelling birding proportion") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period")) /
(ggplot(filter(data0, STATE == anal_states$d), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$d,
       x = "Month", y = "Predicted travelling birding proportion") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period"))

```

-   At the national level, mobility of birders, in terms of proportion of travelling lists submitted, decreased sharply during April--May as expected, and had some spillover into June and July, but matched pre-COVID levels well in the other months of the year.

### Birding distance

Not "per observer", as this metric from data POV. Only considers travelling lists.

```{r distance, cache=TRUE, message=FALSE}

##### Changing in average birding distance ####

set.seed(192) 
dist_sw <- data0_MY_slice_G %>% 
  filter(PROTOCOL.TYPE == "Traveling", !is.na(EFFORT.DISTANCE.KM)) %>% 
  group_by(COVID, M.YEAR, MONTH, STATE) %>% 
  dplyr::summarise(DIST = boot_conf(x = EFFORT.DISTANCE.KM)) %>% 
  group_by(COVID, M.YEAR, MONTH, STATE) %>% 
  dplyr::summarise(SE = sd(DIST),
                   DIST = mean(DIST),
                   CI.L = DIST - 1.96*SE,
                   CI.U = DIST + 1.96*SE)

dist_nw <- dist_sw %>% ungroup() %>% 
  group_by(COVID, M.YEAR, MONTH) %>% 
  dplyr::summarise(DIST = mean(DIST),
            SE = sqrt(sum((SE)^2))/n(),
            CI.L = DIST - 1.96*SE,
            CI.U = DIST + 1.96*SE)


(ggplot(dist_nw, aes(MONTH, DIST, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "National level",
         x = "Month", y = "Distance travelled (km)")) /
((ggplot(filter(dist_sw, STATE == "Kerala"), 
         aes(MONTH, DIST, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Kerala",
         x = "Month", y = "Distance travelled (km)")) |
(ggplot(filter(dist_sw, STATE == "Karnataka"), 
         aes(MONTH, DIST, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Karnataka",
         x = "Month", y = "Distance travelled (km)"))) /
((ggplot(filter(dist_sw, STATE == "Maharashtra"), 
         aes(MONTH, DIST, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Maharashtra",
         x = "Month", y = "Distance travelled (km)")) |
(ggplot(filter(dist_sw, STATE == "Assam"), 
         aes(MONTH, DIST, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Assam",
         x = "Month", y = "Distance travelled (km)"))) +
plot_layout(guides = "collect") +
plot_annotation(title = "Change in birding mobility with the pandemic",
                subtitle = "Y-axis shows average distance travelled in checklists using the travelling protocol") -> distance

distance

ggsave("03_wrap_figs/07_distance.png", plot = distance,
       width = 9, height = 9, units = "in")

rm(dist_sw, dist_nw, distance)


##### Final analysis: statewise GLMMs ####

# dataframe with empty column to populate with looped values
# total rows: product of distinct values of predictors
data0 <- data0_MY_slice_G %>% 
  filter(STATE %in% anal_states[1,]) %>% 
  tidyr::expand(nesting(STATE, MONTH, M.YEAR)) %>% 
  mutate(PRED = NA,
         STATE = factor(STATE, levels = anal_states[1,])) %>% 
  arrange(STATE)

count <- 0
for (i in 1:4) { # for each state
  
  data_a <- data0_MY_slice_G %>% 
    filter(PROTOCOL.TYPE == "Traveling", 
           !is.na(EFFORT.DISTANCE.KM),
           EFFORT.DISTANCE.KM > 0.3,
           STATE == anal_states[, i]) %>% 
    # because no other family fits for distance
    mutate(DISTANCE = log(EFFORT.DISTANCE.KM))
  
  tictoc::tic(glue::glue("LMM for {anal_states[, i]}"))
  model_a <- lmer(DISTANCE ~ M.YEAR + M.YEAR:MONTH + (1|CELL.ID),
                  data = data_a, 
                  control = lmerControl(optimizer = "bobyqa"))  
  tictoc::toc()
  

  for (j in 1:n_distinct(data0$MONTH)) {
    for (k in 1:n_distinct(data0$M.YEAR)) {
      count <- count + 1
      
      data0$PRED[count] = exp(predict(model_a,
                                      data.frame(MONTH = data0$MONTH[count],
                                                 M.YEAR = data0$M.YEAR[count]),
                                      re.form = NA, type = "response"))
    }
  }
} 

data0 <- data0 %>% 
  left_join(timeline)


(ggplot(filter(data0, STATE == anal_states$a), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$a,
       x = "Month", y = "Predicted average birding distance") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period")) /
(ggplot(filter(data0, STATE == anal_states$b), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$b,
       x = "Month", y = "Predicted average birding distance") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period")) /
(ggplot(filter(data0, STATE == anal_states$c), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$c,
       x = "Month", y = "Predicted average birding distance") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period")) /
(ggplot(filter(data0, STATE == anal_states$d), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$d,
       x = "Month", y = "Predicted average birding distance") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period"))

```

-   At the national level, checklist distances were highest during summer and winter before the pandemic. During the pandemic, distances remained fairly similar for the winter months of October--December, but were drastically lower during the summer months of April and May, which coincided with the lockdown/second wave.\
-   Overall, this metric was not affected as much as others. Karnataka and Kerala showed some brief declines, but overall patterns are rather irregular, so nothing major. Maharashtra has shown a consistent decline since the start of the pandemic till now.

### List duration

Not "per observer", as this metric from data POV.

```{r duration, cache=TRUE, message=FALSE}

##### Change in average list duration ####

set.seed(834)
dur_sw <- data0_MY_slice_G %>% 
  filter(!is.na(DURATION.MINUTES)) %>% 
  group_by(COVID, M.YEAR, MONTH, STATE) %>% 
  dplyr::summarise(DURATION = boot_conf(x = DURATION.MINUTES)) %>% 
  group_by(COVID, M.YEAR, MONTH, STATE) %>% 
  dplyr::summarise(SE = sd(DURATION),
                   DURATION = mean(DURATION),
                   CI.L = DURATION - 1.96*SE,
                   CI.U = DURATION + 1.96*SE)

dur_nw <- dur_sw %>% 
  ungroup() %>% 
  group_by(COVID, M.YEAR, MONTH) %>% 
  dplyr::summarise(DURATION = mean(DURATION),
            SE = sqrt(sum((SE)^2))/n(),
            CI.L = DURATION - 1.96*SE,
            CI.U = DURATION + 1.96*SE)


(ggplot(dur_nw, aes(MONTH, DURATION, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "National level",
         x = "Month", y = "Duration (mins)")) /
((ggplot(filter(dur_sw, STATE == "Kerala"), 
         aes(MONTH, DURATION, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Kerala",
         x = "Month", y = "Duration (mins)")) |
(ggplot(filter(dur_sw, STATE == "Karnataka"), 
         aes(MONTH, DURATION, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Karnataka",
         x = "Month", y = "Duration (mins)"))) /
((ggplot(filter(dur_sw, STATE == "Maharashtra"), 
         aes(MONTH, DURATION, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Maharashtra",
         x = "Month", y = "Duration (mins)")) |
(ggplot(filter(dur_sw, STATE == "Assam"), 
         aes(MONTH, DURATION, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Assam",
         x = "Month", y = "Duration (mins)"))) +
plot_layout(guides = "collect") +
plot_annotation(title = "Change in checklist duration with the pandemic",
                subtitle = "Y-axis shows average duration of checklists") -> duration

duration

ggsave("03_wrap_figs/08_duration.png", plot = duration,
       width = 9, height = 9, units = "in")

rm(dur_sw, dur_nw, duration)


##### Final analysis: statewise GLMMs ####

# dataframe with empty column to populate with looped values
# total rows: product of distinct values of predictors
data0 <- data0_MY_slice_G %>% 
  filter(STATE %in% anal_states[1,]) %>% 
  tidyr::expand(nesting(STATE, MONTH, M.YEAR)) %>% 
  mutate(PRED = NA,
         STATE = factor(STATE, levels = anal_states[1,])) %>% 
  arrange(STATE)

count <- 0
for (i in 1:4) { # for each state
  
  data_a <- data0_MY_slice_G %>% 
    filter(STATE == anal_states[, i]) %>% 
    # because no other family fits
    mutate(DURATION = log(DURATION.MINUTES))
  
  tictoc::tic(glue::glue("LMM for {anal_states[, i]}"))
  model_a <- lmer(DURATION ~ M.YEAR + M.YEAR:MONTH + (1|CELL.ID),
                  data = data_a, 
                  control = lmerControl(optimizer = "bobyqa"))  
  tictoc::toc()
  

  for (j in 1:n_distinct(data0$MONTH)) {
    for (k in 1:n_distinct(data0$M.YEAR)) {
      count <- count + 1
      
      data0$PRED[count] = exp(predict(model_a,
                                      data.frame(MONTH = data0$MONTH[count],
                                                 M.YEAR = data0$M.YEAR[count]),
                                      re.form = NA, type = "response"))
    }
  }
} 

data0 <- data0 %>% 
  left_join(timeline)


(ggplot(filter(data0, STATE == anal_states$a), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$a,
       x = "Month", y = "Predicted average checklist duration") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period")) /
(ggplot(filter(data0, STATE == anal_states$b), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$b,
       x = "Month", y = "Predicted average checklist duration") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period")) /
(ggplot(filter(data0, STATE == anal_states$c), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$c,
       x = "Month", y = "Predicted average checklist duration") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period")) /
(ggplot(filter(data0, STATE == anal_states$d), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$d,
       x = "Month", y = "Predicted average checklist duration") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period"))

```

-   In normal circumstances, checklists become longer in the winter, and shorter in the monsoon. The only notable change during the pandemic at the national level was low checklist duration in April--June, due to the national lockdown/second wave.\
-   A point of interest is that the average checklist duration varied greatly between states. Among the 4 compared here, Kerala on average had shortest lists, followed by Karnataka and then Maharashtra.\
-   Patterns somewhat similar to those of birding distance. Generally, briefly reduced during the peak months.

### List length

Not "per observer", as this metric from data POV.

```{r length, cache=TRUE, message=FALSE}

##### Change in list length (number of species) ####

# var(mean(x))/var(median(x)) = 4n/(pi*(2n + 1))

# used mean here because it was proving extremely difficult to work with errors in the case of median

set.seed(98)
length_sw <- data0_MY_slice_G %>% 
  group_by(COVID, M.YEAR, MONTH, STATE) %>% 
  dplyr::summarise(LENGTH = boot_conf(x = NO.SP)) %>% 
  group_by(COVID, M.YEAR, MONTH, STATE) %>% 
  dplyr::summarise(SE = sd(LENGTH),
                   LENGTH = mean(LENGTH),
                   CI.L = LENGTH - 1.96*SE,
                   CI.U = LENGTH + 1.96*SE)

length_nw <- length_sw %>% 
  ungroup() %>% 
  group_by(COVID, M.YEAR, MONTH) %>% 
  dplyr::summarise(LENGTH = mean(LENGTH),
            SE = sqrt(sum((SE)^2))/n(),
            CI.L = LENGTH - 1.96*SE,
            CI.U = LENGTH + 1.96*SE)


(ggplot(length_nw, aes(MONTH, LENGTH, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "National level",
         x = "Month", y = "List length")) /
((ggplot(filter(length_sw, STATE == "Kerala"), 
         aes(MONTH, LENGTH, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Kerala",
         x = "Month", y = "List length")) |
(ggplot(filter(length_sw, STATE == "Karnataka"), 
         aes(MONTH, LENGTH, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Karnataka",
         x = "Month", y = "List length"))) /
((ggplot(filter(length_sw, STATE == "Maharashtra"), 
         aes(MONTH, LENGTH, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Maharashtra",
         x = "Month", y = "List length")) |
(ggplot(filter(length_sw, STATE == "Assam"), 
         aes(MONTH, LENGTH, colour = COVID)) + 
    scale_y_log10() + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Assam",
         x = "Month", y = "List length"))) +
plot_layout(guides = "collect") +
plot_annotation(title = "Change in checklist length with the pandemic",
                subtitle = "Y-axis shows average number of species recorded in checklists (list length)") -> length

length

ggsave("03_wrap_figs/09_length.png", plot = length,
       width = 9, height = 9, units = "in")

rm(length_nw, length_sw, length)


##### Final analysis: statewise GLMMs ####

# dataframe with empty column to populate with looped values
# total rows: product of distinct values of predictors
data0 <- data0_MY_slice_G %>% 
  filter(STATE %in% anal_states[1,]) %>% 
  tidyr::expand(nesting(STATE, MONTH, M.YEAR)) %>% 
  mutate(PRED = NA,
         STATE = factor(STATE, levels = anal_states[1,])) %>% 
  arrange(STATE)

count <- 0
for (i in 1:4) { # for each state
  
  data_a <- data0_MY_slice_G %>% 
    filter(STATE == anal_states[, i])
  
  tictoc::tic(glue::glue("GLMM for {anal_states[, i]}"))
  model_a <- glmer(NO.SP ~ M.YEAR + M.YEAR:MONTH + (1|CELL.ID),
                   data = data_a, family = poisson,
                   nAGQ = 0, control = glmerControl(optimizer = "bobyqa"))  
  tictoc::toc()
  

  for (j in 1:n_distinct(data0$MONTH)) {
    for (k in 1:n_distinct(data0$M.YEAR)) {
      count <- count + 1
      
      data0$PRED[count] = predict(model_a,
                                  data.frame(MONTH = data0$MONTH[count],
                                             M.YEAR = data0$M.YEAR[count]),
                                  re.form = NA, type = "response")
    }
  }
} 

data0 <- data0 %>% 
  left_join(timeline)


(ggplot(filter(data0, STATE == anal_states$a), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$a,
       x = "Month", y = "Predicted average checklist length") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period")) /
(ggplot(filter(data0, STATE == anal_states$b), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$b,
       x = "Month", y = "Predicted average checklist length") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period")) /
(ggplot(filter(data0, STATE == anal_states$c), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$c,
       x = "Month", y = "Predicted average checklist length") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period")) /
(ggplot(filter(data0, STATE == anal_states$d), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$d,
       x = "Month", y = "Predicted average checklist length") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period"))

```

-   Overall, slight decrease during the peak period.\
-   Kerala didn't show any change at all.\
-   The peak months brought average list length in other states to around 10-11 species, which in Kerala's case already used to be the case (shorter lists).

### Spatial spread and coverage

1.  Urban:non-urban ratio in birding effort
    -   Proportion of lists that are urban per grid cell\
    -   Proportion of lists that are urban per grid cell GLMM\
2.  Spatial coverage (min. 10 lists per grid cell)
    -   Proportion of total cells in country/state covered across months
    -   Proportion of total cells in country/state covered across months GLMM\
3.  Spatial spread/evenness (25kmx25km cells; across 9 months) Large spatial and temporal scale---no STATE or MONTH/M.YEAR in question.
    -   Raw change in no. of lists per district
    -   Proportional change in no. of lists per district
    -   Proportional change in grid coverage per district
    -   Proportional change in grid coverage per district with threshold for "coverage"

```{r s_UNU, cache=TRUE, message=FALSE}

##### Proportion of urban lists per grid cell ####

temp <- data0_MY_slice_G %>% 
  group_by(COVID, M.YEAR, MONTH, CELL.ID) %>% # state is meaningless cos CELL.ID
  filter(URBAN == 1) %>% 
  dplyr::summarise(U.LISTS = n_distinct(GROUP.ID)) %>% 
  ungroup()

urban_lists <- data0_MY_slice_G %>% 
  group_by(COVID, M.YEAR, MONTH, CELL.ID) %>%
  dplyr::summarise(TOT.LISTS = n_distinct(GROUP.ID)) %>% 
  left_join(temp) %>% 
  group_by(COVID, M.YEAR, MONTH, CELL.ID) %>%
  dplyr::summarise(U.LISTS = replace_na(U.LISTS, 0),
                   PROP.U = U.LISTS/TOT.LISTS,
                   SE = sqrt((PROP.U)*(1 - PROP.U)/TOT.LISTS)) %>% 
  group_by(COVID, M.YEAR, MONTH) %>% 
  dplyr::summarise(PROP.U = mean(PROP.U),
                   SE = sqrt(sum((SE)^2))/n(),
                   CI.L = PROP.U - 1.96*SE,
                   CI.U = PROP.U + 1.96*SE)

rm(temp)


ggplot(urban_lists, aes(MONTH, PROP.U, colour = COVID)) + 
  scale_x_continuous(breaks = 1:12,
                     labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", 
                                "Sep", "Oct", "Nov", "Dec")) + 
  annotate("rect", xmin = 2.5, xmax = 9.5, ymin = -Inf, ymax = Inf, 
            fill = "#E0DDE3", col = NA, alpha = 0.5) +
  geom_point(size = 4, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.75, width = 0.7, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette,
                      name = "Period", 
                      labels = c("Before", "During (2020)", "During (2021)", "After")) +
  labs(title = "2020 lockdown restricted birders to urban areas",
       x = "Month", y = "Proportion of urban lists per grid cell") -> s_UNU

s_UNU


ggsave("03_wrap_figs/10a_s_UNU.png", 
       plot = s_UNU,
       width = 13, height = 7, units = "in")


##### Final analysis: GLMM of urban birding using weighted proportions ####

# https://github.com/lme4/lme4/issues/281
# Issue with high number of zeroes (or hence low proportion values) and back-transformation of linear variable without accounting for random variation
# So, need to include random effects in prediction.

data_urban <- data0_MY_slice_G %>% 
  group_by(M.YEAR, MONTH, COUNTY, CELL.ID) %>% 
  filter(URBAN == 1) %>% 
  dplyr::summarise(U.LISTS = n_distinct(GROUP.ID)) %>% 
  right_join(data0_MY_slice_G) %>% 
  group_by(M.YEAR, MONTH, COUNTY, CELL.ID) %>% 
  dplyr::summarise(TOT.LISTS = n_distinct(GROUP.ID),
                   U.LISTS = replace_na(min(U.LISTS), 0),
                   PROP.U = round(U.LISTS/TOT.LISTS, 4)) %>% 
  ungroup()
  
counties <- data_urban %>% distinct(COUNTY, CELL.ID)

# no MONTH main effect unlike in birds model
tictoc::tic("UNU GLMM using weighted proportions")
model_urban <- glmer(PROP.U ~ M.YEAR + M.YEAR:MONTH + (1|COUNTY/CELL.ID), 
                      weights = TOT.LISTS,
                      data = data_urban, family = binomial(link = "cloglog"),
                      nAGQ = 0, control = glmerControl(optimizer = "bobyqa")) # 119 sec
tictoc::toc()

# dataframe with empty column to populate with looped values
# total rows: product of distinct values of predictors
urban1 <- data_urban %>% 
  tidyr::expand(nesting(MONTH, M.YEAR, COUNTY, CELL.ID)) %>% 
  filter(!is.na(COUNTY)) %>% 
  mutate(PROP.U.PRED = predict(model_urban, data.frame(MONTH, M.YEAR, COUNTY, CELL.ID),
                               type = "response")) %>% 
  group_by(M.YEAR, MONTH) %>% 
  summarise(SE = sd(PROP.U.PRED)/sqrt(n()),
            PROP.U.PRED = mean(PROP.U.PRED)) %>% 
  summarise(PROP.U.PRED = mean(PROP.U.PRED),
            SE = sqrt(sum((SE)^2))/n(),
            CI.L = PROP.U.PRED - 1.96*SE,
            CI.U = PROP.U.PRED + 1.96*SE)

urban2 <- data_urban %>% 
  tidyr::expand(nesting(MONTH, M.YEAR, COUNTY, CELL.ID)) %>% 
  filter(!is.na(COUNTY)) %>% 
  mutate(PROP.U.PRED = predict(model_urban, data.frame(MONTH, M.YEAR, COUNTY, CELL.ID),
                               type = "response")) %>% 
  filter(MONTH %in% 4:5) %>% 
  group_by(M.YEAR, MONTH) %>% 
  summarise(SE = sd(PROP.U.PRED)/sqrt(n()),
            PROP.U.PRED = mean(PROP.U.PRED)) %>% 
  summarise(PROP.U.PRED = mean(PROP.U.PRED),
            SE = sqrt(sum((SE)^2))/n(),
            CI.L = PROP.U.PRED - 1.96*SE,
            CI.U = PROP.U.PRED + 1.96*SE)


ggplot(urban2, aes(M.YEAR, PROP.U.PRED)) +
    labs(title = "For the months of April and May",
         x = "Migratory year", y = "Predicted urban birding proportion") +
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                  size = 1.5, width = 0.2, position = position_dodge(0.5))

ggplot(urban1, aes(M.YEAR, PROP.U.PRED)) +
    labs(title = "For all months",
         x = "Migratory year", y = "Predicted urban birding proportion") +
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                  size = 1.5, width = 0.2, position = position_dodge(0.5))


```

```{r s_cover, cache=TRUE, message=FALSE}

##### Proportion of total grid cells covered ####

cover_sw <- data0_MY_slice_G %>% 
  group_by(STATE) %>% 
  mutate(TOT.CELLS = n_distinct(CELL.ID)) %>% 
  group_by(COVID, M.YEAR, MONTH, STATE, CELL.ID) %>% 
  dplyr::summarise(TOT.CELLS = min(TOT.CELLS),
                   N.LISTS = n_distinct(GROUP.ID)) %>% 
  filter(N.LISTS >= 10) %>% 
  group_by(COVID, M.YEAR, MONTH, STATE) %>% 
  dplyr::summarise(TOT.CELLS = min(TOT.CELLS),
                   NO.CELLS = n_distinct(CELL.ID),
                   PROP.CELLS = NO.CELLS/TOT.CELLS,
                   SE = sqrt((PROP.CELLS)*(1 - PROP.CELLS)/TOT.CELLS),
                   CI.L = PROP.CELLS - 1.96*SE,
                   CI.U = PROP.CELLS + 1.96*SE)

cover_nw <- data0_MY_slice_G %>% 
  ungroup() %>% 
  mutate(TOT.CELLS = n_distinct(CELL.ID)) %>% 
  group_by(COVID, M.YEAR, MONTH, CELL.ID) %>% 
  dplyr::summarise(TOT.CELLS = min(TOT.CELLS),
                   N.LISTS = n_distinct(GROUP.ID)) %>% 
  filter(N.LISTS >= 10) %>% 
  group_by(COVID, M.YEAR, MONTH) %>% 
  dplyr::summarise(TOT.CELLS = min(TOT.CELLS),
                   NO.CELLS = n_distinct(CELL.ID),
                   PROP.CELLS = NO.CELLS/TOT.CELLS,
                   SE = sqrt((PROP.CELLS)*(1 - PROP.CELLS)/TOT.CELLS),
                   CI.L = PROP.CELLS - 1.96*SE,
                   CI.U = PROP.CELLS + 1.96*SE)


((ggplot(cover_nw, aes(MONTH, PROP.CELLS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "National level",
         x = "Month", y = "Birded proportion")) /
((ggplot(filter(cover_sw, STATE == "Kerala"), 
        aes(MONTH, PROP.CELLS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Kerala",
         x = "Month", y = "Birded proportion")) |
(ggplot(filter(cover_sw, STATE == "Karnataka"), 
        aes(MONTH, PROP.CELLS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Karnataka",
         x = "Month", y = "Birded proportion"))) /
((ggplot(filter(cover_sw, STATE == "Maharashtra"), 
        aes(MONTH, PROP.CELLS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Maharashtra",
         x = "Month", y = "Birded proportion")) |
(ggplot(filter(cover_sw, STATE == "Assam"), 
        aes(MONTH, PROP.CELLS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Assam",
         x = "Month", y = "Birded proportion")))) +
plot_layout(guides = "collect") +
plot_annotation(title = "Change in spatial coverage of monthly birding with the pandemic",
         subtitle = "Proportion of total grid cells (24kmx24km) having birding activity") -> s_cover

s_cover

ggsave("03_wrap_figs/11a_s_cover.png", 
       plot = s_cover,
       width = 9, height = 9, units = "in")


rm(cover_sw, cover_nw, s_cover)


##### Final analysis: statewise GLMMs ####

# dataframe with empty column to populate with looped values
# total rows: product of distinct values of predictors
data0 <- data0_MY_slice_G %>% 
  filter(STATE %in% anal_states[1,]) %>% 
  tidyr::expand(nesting(STATE, MONTH, M.YEAR)) %>% 
  mutate(PRED = NA,
         STATE = factor(STATE, levels = anal_states[1,])) %>% 
  arrange(STATE)

count <- 0
for (i in 1:4) { # for each state
  
  data_a <- data0_MY_slice_G %>% 
    filter(STATE == anal_states[, i]) %>% 
    select(M.YEAR, MONTH, CELL.ID, GROUP.ID) %>% 
    tidyr::complete(M.YEAR, MONTH, CELL.ID) %>% 
    group_by(M.YEAR, MONTH, CELL.ID) %>% 
    dplyr::summarise(COVERED = if_else(n_distinct(GROUP.ID) >= 10, 1, 0))
  
  tictoc::tic(glue::glue("GLMM for {anal_states[, i]}"))
  model_a <- glmer(COVERED ~ M.YEAR + M.YEAR:MONTH + (1|CELL.ID),
                   data = data_a, family = binomial(link = "cloglog"),
                   nAGQ = 0, control = glmerControl(optimizer = "bobyqa"))  
  tictoc::toc()
  
  for (j in 1:n_distinct(data0$MONTH)) {
    for (k in 1:n_distinct(data0$M.YEAR)) {
      count <- count + 1
      
      data0$PRED[count] = predict(model_a,
                                  data.frame(MONTH = data0$MONTH[count],
                                             M.YEAR = data0$M.YEAR[count]),
                                  re.form = NA, type = "response")
    }
  }
} 

data0 <- data0 %>% 
  left_join(timeline)


(ggplot(filter(data0, STATE == anal_states$a), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$a,
       x = "Month", y = "Predicted grid cell coverage") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period")) /
(ggplot(filter(data0, STATE == anal_states$b), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$b,
       x = "Month", y = "Predicted grid cell coverage") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period")) /
(ggplot(filter(data0, STATE == anal_states$c), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$c,
       x = "Month", y = "Predicted grid cell coverage") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period")) /
(ggplot(filter(data0, STATE == anal_states$d), 
       aes(MONTH, PRED, colour = COVID)) +
  labs(title = anal_states$d,
       x = "Month", y = "Predicted grid cell coverage") +
  geom_point(size = 3, position = position_dodge(0.5)) +
  # geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
  #               size = 1.5, width = 0.2, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Period"))

```

For spread, we will not consider change in number or proportion of lists per grid cells because there will be too much noise at that fine a scale. Here, districts provide a nice compromise: large enough to track change, small enough to inform us about local patterns.

For latter 3, only those districts that have a minimum of 5 grid cells will be considered.

```{r s_spread, cache=TRUE, message=FALSE}

load("data/maps_gridmapg1_IN.RData") # all cells within India

gridmapg1_IN <- gridmapg1_IN %>% 
  # setting CRS for join in next step
  st_set_crs(value = st_crs(districtmap))

# creating object linking grid cells to districts
grid_dist <- districtmap %>% 
  st_as_sf() %>% 
  dplyr::select(dtname) %>% 
  rename(DISTRICT.NAME = dtname) %>% 
  st_join(gridmapg1_IN)
# losing some districts and grid cells, but doesn't matter. 
# Will explore patterns with existing.

# Ensuring that a grid cell is only linked with one district
grid_dist_links <- grid_dist %>% 
  st_drop_geometry() %>% 
  group_by(CELL.ID) %>% 
  slice_sample() %>% 
  group_by(DISTRICT.NAME) %>% 
  mutate(NO.CELLS.TOT = n_distinct(CELL.ID)) %>% 
  ungroup()

# # distribution of number of cells covered, across districts
# x <- grid_dist_links %>% 
#   group_by(DISTRICT.NAME) %>% 
#   summarise(NO.CELLS = n_distinct(CELL.ID)) %>% 
#   group_by(NO.CELLS) %>% 
#   summarise(NO.DISTRICTS = n())

grid_dist <- grid_dist %>% 
  right_join(grid_dist_links)


temp0 <- data0_MY_slice_G %>% 
  # selecting only nine months (AFT has only nine)
  filter(MONTH %in% month_compar$MONTH) %>% 
  # combining the two DUR periods
  mutate(COVID = as.character(COVID)) %>% 
  mutate(COVID = case_when(COVID %in% c("DUR_20","DUR_21") ~ "DUR",
                           TRUE ~ COVID)) %>% 
  mutate(COVID = factor(COVID, levels = c("BEF", "DUR", "AFT"))) %>% 
  # although data has fewer cells than grid_dist, using inner so only those cells with 
  # established district links will be considered (removes some cells and districts)
  dplyr::select(-DISTRICT) %>% # removing to avoid confusion
  inner_join(grid_dist)

# considering only districts with at least 5 grid cells for grid coverage
temp1 <- temp0 %>% 
  filter(NO.CELLS.TOT >= 5) %>% 
  # summarising metrics of interest
  group_by(COVID, DISTRICT.NAME, NO.CELLS.TOT) %>% 
  summarise(NO.CELLS = n_distinct(CELL.ID)) %>% 
  ungroup() %>% 
  complete(COVID, nesting(DISTRICT.NAME, NO.CELLS.TOT),
           fill = list(NO.CELLS = 0)) %>% 
  mutate(GRID.COV = 100*NO.CELLS/NO.CELLS.TOT) %>% 
  # moving COVID to columns
  dplyr::select(-NO.CELLS)

# for grid coverage per district with threshold
temp2 <- temp0 %>% 
  filter(NO.CELLS.TOT >= 5) %>% 
  # summarising metrics of interest
  group_by(COVID, DISTRICT.NAME, NO.CELLS.TOT, CELL.ID) %>% 
  summarise(CELL.LISTS = n_distinct(GROUP.ID)) %>% 
  # threshold of 5 lists in a grid cell for the cell to be considered "covered"
  filter(CELL.LISTS >= 5) %>% 
  summarise(NO.CELLS = n_distinct(CELL.ID)) %>% 
  ungroup() %>% 
  complete(COVID, nesting(DISTRICT.NAME, NO.CELLS.TOT),
           fill = list(NO.CELLS = 0)) %>% 
  mutate(GRID.COV.T = 100*NO.CELLS/NO.CELLS.TOT) %>% 
  # moving COVID to columns
  dplyr::select(-NO.CELLS)

data_spread0 <- temp0 %>% 
  # summarising metrics of interest
  group_by(COVID, DISTRICT.NAME, NO.CELLS.TOT) %>% 
  summarise(NO.LISTS = n_distinct(GROUP.ID)) %>% 
  ungroup() %>% 
  complete(COVID, nesting(DISTRICT.NAME, NO.CELLS.TOT),
           fill = list(NO.LISTS = 0)) %>% 
  left_join(temp1) %>% 
  left_join(temp2) %>% 
  pivot_longer(cols = c("NO.LISTS", "GRID.COV", "GRID.COV.T"),
               names_to = "METRIC", values_to = "VALUE") %>% 
  pivot_wider(names_from = "COVID", values_from = "VALUE") %>% 
  # calculating change (produces NA for x == 0 & y == 0 which need to be removed)
  mutate(RAW.CHANGE1 = if_else(METRIC == "NO.LISTS", s_spread_change(BEF, DUR), NA_real_),
         RAW.CHANGE2 = if_else(METRIC == "NO.LISTS", s_spread_change(DUR, AFT), NA_real_),
         PROP.CHANGE1 = s_spread_propchange(BEF, DUR),
         PROP.CHANGE2 = s_spread_propchange(DUR, AFT)) %>% 
  left_join(grid_dist)

transition_names <- data.frame(T.CODE = c("T1", "T2"),
                               T.LABEL = c("Before to During", "During to After"))


##### Change in birding effort per district ####

# 1. map of raw change
# 2. with SEs of raw change
# 3. map of proportional change

# 1. map (to give an idea of overall clustering)

data_spread1 <- data_spread0 %>% 
  filter(METRIC == "NO.LISTS") %>% 
  distinct(DISTRICT.NAME, geometry, RAW.CHANGE1, RAW.CHANGE2) %>% 
  # pivoting longer for facetting
  pivot_longer(cols = c("RAW.CHANGE1", "RAW.CHANGE2"),
               names_to = "METRIC", values_to = "VALUE") %>% 
  mutate(T.CODE = case_when(str_detect(METRIC, "1") ~ "T1",
                            str_detect(METRIC, "2") ~ "T2")) %>% 
  left_join(transition_names)

s_spread_map <- ggplot(data_spread1) +
  geom_polygon(data = indiamap,
               aes(long, lat, group = group),
               fill = "#467C27", colour = "black", size = 0.4) +
  geom_sf(aes(fill = VALUE, geometry = geometry), col ="#ACACAC", size = 0.1) +
  facet_wrap(~ T.LABEL) +
  # scale_fill_continuous_divergingx(palette = "RdBu", mid = 0,
  #                                  n_interp = 7, na.value = "grey") +
  scale_fill_steps2(low = RColorBrewer::brewer.pal(n = 11, name = "RdBu")[1],
                    mid = RColorBrewer::brewer.pal(n = 11, name = "RdBu")[6],
                    high = RColorBrewer::brewer.pal(n = 11, name = "RdBu")[11],
                    midpoint = 0,
                    breaks = c(-2500, -1000, -100, 0, 1, 100, 2500, 5000),
                    limits = c(-4134, 7848),
                    na.value = "#D4B449") +
  labs(x = "Longitude", y = "Latitude",
       fill = "Difference in\nno. of lists",
       title = "Change in no. of lists per district in transition periods",
       subtitle = "Gold: districts in which both periods of the transition had zero lists.\nGreen: districts not considered in analysis.")

ggsave("03_wrap_figs/10c_s_spread_map.png", 
       plot = s_spread_map,
       dpi = 300, width = 14, height = 8, units = "in")


# 2. with SEs

set.seed(23) # for bootstrap

# not directly using data_spread0 object because backtransforming after bootstrapping mean 
# and SE is not wise
data_spread2 <- data_spread1 %>% 
  # removing NA cells (when both periods of the transition are zero)
  filter(!is.na(VALUE)) %>% 
  group_by(T.CODE, T.LABEL) %>% 
  # this gives B number of means of bootstrapped samples
  dplyr::summarise(VALUE = boot_conf(x = VALUE)) %>% 
  group_by(T.CODE, T.LABEL) %>% 
  dplyr::summarise(CI.L = stats::quantile(VALUE, 0.025), # Obtain the CIs
                   CI.U = stats::quantile(VALUE, 0.975),
                   VALUE = median(VALUE)) 

s_spread_change <- ggplot(data_spread2, aes(x = T.LABEL, y = VALUE)) + 
  geom_point(size = 4, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.75, width = 0.08, position = position_dodge(0.5)) +
  # scale_y_continuous(limits = c(1.1, 2.3), breaks = seq(1, 3, 0.2)) +
  labs(title = "Net change in birding effort (no. of lists) across districts\nin transition periods",
       x = "Transition", y = "Difference in no. of lists") 

ggsave("03_wrap_figs/10c_s_spread_rawchange.png", 
       plot = s_spread_change,
       dpi = 300, width = 6, height = 9, units = "in")

# is DUR-AFT really lower than BEF-DUR?
ggplot(data_spread1) +
  geom_histogram(aes(VALUE)) +
  facet_wrap(~ T.LABEL)


# 3. map





##### Proportional change in grid coverage per district ####

# 1. distribution of grid coverage
# 2. with SEs of grid coverage
# 3. distribution of grid coverage with threshold
# 4. with SEs of grid coverage with threshold

data_spread <- data_spread0 %>% 
  filter(METRIC == "DISTRICT.CELLS.COV") %>% 
  distinct(DISTRICT.NAME, geometry.district, CHANGE1, CHANGE2)


ggplot(data_spread) +
  geom_polygon(data = indiamap,
               aes(long, lat, group = group),
               colour = "black", fill = "#E6E9F1", size = 0.2) +
  geom_sf(aes(fill = CHANGE1, geometry = geometry.district), col = NA, na.rm = T) +
  scale_fill_viridis_c(option = "cividis",
                       name = "Proportional\nchange") +
  labs(x = "Longitude", y = "Latitude",
       title = "Proportional change in grid coverage per district in transition periods") +
  # making the colourbar bigger
  theme(legend.key.height = unit(1, "inches")) # height between two breaks of legend

hist(data_spread$CHANGE1, breaks = 100)
hist(data_spread$CHANGE2, breaks = 100)

```

-   The maps and the graphs all show that there was an overall growth in eBirding.\
-   Spatial coverage was entirely unaffected by the pandemic!

### Temporal spread

Not "per observer", as this metric from data POV.

Since month is not considered here (and instead data from multiple months are averaged), it doesn't make sense to keep two years for BEF and AFT---they should instead also be pooled into the averaging.

Here, running models is not necessary, as we just want to visualise the patterns. Robust results for this do not tell us much. Besides, the models that can be run for this metric would be very different from those for the other metrics (no CELL.ID or OBSERVER.ID random effect, and MONTH used as random effect).

```{r t_spread_1, cache=TRUE, message=FALSE}

##### Change in DoW spread ####

t_dow_sw <- data0_MY_slice_G %>% 
  filter(MONTH %in% month_compar$MONTH) %>% 
  mutate(DAY.W = wday(OBSERVATION.DATE, 
                      # start week from Monday
                      week_start = getOption("lubridate.week.start", 1))) %>% 
  group_by(COVID, STATE, M.YEAR, MONTH) %>% 
  mutate(NO.LISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>%
  group_by(COVID, DAY.W, STATE, M.YEAR, MONTH) %>% 
  dplyr::summarise(NO.LISTS = min(NO.LISTS),
                   DAY.LISTS = replace_na(n_distinct(SAMPLING.EVENT.IDENTIFIER)),
                   PROP.LISTS = DAY.LISTS/NO.LISTS,
                   SE = sqrt((PROP.LISTS)*(1 - PROP.LISTS)/NO.LISTS)) %>% 
  group_by(COVID, DAY.W, STATE) %>% 
  dplyr::summarise(PROP.LISTS = mean(PROP.LISTS),
                   SE = sqrt(sum((SE)^2))/n(),
                   CI.L = PROP.LISTS - 1.96*SE,
                   CI.U = PROP.LISTS + 1.96*SE)

t_dow_nw <- t_dow_sw %>% 
  group_by(COVID, DAY.W) %>% 
  dplyr::summarise(PROP.LISTS = mean(PROP.LISTS),
            SE = sqrt(sum((SE)^2))/n(),
            CI.L = PROP.LISTS - 1.96*SE,
            CI.U = PROP.LISTS + 1.96*SE)

(ggplot(t_dow_nw, 
       aes(as.factor(DAY.W), PROP.LISTS, colour = COVID)) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "National level",
         x = "Day", y = "Proportion of lists")) /
((ggplot(filter(t_dow_sw, STATE == "Kerala"), 
         aes(as.factor(DAY.W), PROP.LISTS, colour = COVID)) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Kerala",
         x = "Day", y = "Proportion of lists")) |
ggplot(filter(t_dow_sw, STATE == "Karnataka"), 
         aes(as.factor(DAY.W), PROP.LISTS, colour = COVID)) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Karnataka",
         x = "Day", y = "Proportion of lists")) /
((ggplot(filter(t_dow_sw, STATE == "Maharashtra"), 
         aes(as.factor(DAY.W), PROP.LISTS, colour = COVID)) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Maharashtra",
         x = "Day", y = "Proportion of lists")) |
ggplot(filter(t_dow_sw, STATE == "Assam"), 
         aes(as.factor(DAY.W), PROP.LISTS, colour = COVID)) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Assam",
         x = "Day", y = "Proportion of lists")) +
plot_layout(guides = "collect") +
plot_annotation(title = "Change in temporal spread of checklists with the pandemic (September to March)",
                subtitle = "Y-axis shows average proportion of checklists from respective days of the week on X-axis") -> t_spread_dow

t_spread_dow

ggsave("03_wrap_figs/13_t_spread_dow.png", 
       plot = t_spread_dow,
       width = 9, height = 9, units = "in")


rm(t_dow_sw, t_dow_nw, t_spread_dow)


```

-   At the national level, birding on Saturdays has decreased, and been balanced by more birding on Wednesdays, while Sunday birding seems to have also increased.\
-   In Kerala, birding seems to have increased during Thursday--Saturday and decreased on Sunday.\
-   In Karnataka there hasn't been much of a change with the pandemic, and there is still a heavy weekend bias in birding activity (Saturday and Sunday levels being similar).\
-   In Gujarat and Maharashtra, Sundays see more birding than Saturdays, but there has been no major impact of the pandemic.\
-   This might mean that the national level patterns might be driven by changes in states with generally low birding activity.

```{r t_spread_2, cache=TRUE, message=FALSE}

##### Change in ToD spread ####

t_tod_sw <- data0_MY_slice_G %>% 
  filter(MONTH %in% month_compar$MONTH) %>% 
  group_by(COVID, STATE, M.YEAR, MONTH) %>% 
  mutate(NO.LISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>%
  group_by(COVID, STATE, HOUR, M.YEAR, MONTH) %>% 
  dplyr::summarise(NO.LISTS = min(NO.LISTS),
                   TIME.LISTS = replace_na(n_distinct(SAMPLING.EVENT.IDENTIFIER)),
                   PROP.LISTS = TIME.LISTS/NO.LISTS,
                   SE = sqrt((PROP.LISTS)*(1 - PROP.LISTS)/NO.LISTS)) %>% 
  group_by(COVID, STATE, HOUR) %>% 
  dplyr::summarise(PROP.LISTS = mean(PROP.LISTS),
                   SE = sqrt(sum((SE)^2))/n(),
                   CI.L = PROP.LISTS - 1.96*SE,
                   CI.U = PROP.LISTS + 1.96*SE)

t_tod_nw <- t_tod_sw %>%
  group_by(COVID, HOUR) %>% 
  dplyr::summarise(PROP.LISTS = mean(PROP.LISTS),
            SE = sqrt(sum((SE)^2))/n(),
            CI.L = PROP.LISTS - 1.96*SE,
            CI.U = PROP.LISTS + 1.96*SE)


(ggplot(t_tod_nw, 
       aes(as.factor(HOUR), PROP.LISTS, colour = COVID)) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "National level",
         x = "Time of day (hours)", y = "Proportion of lists")) /
((ggplot(filter(t_tod_sw, STATE == "Kerala"), 
         aes(as.factor(HOUR), PROP.LISTS, colour = COVID)) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Kerala",
         x = "Time of day (hours)", y = "Proportion of lists")) |
ggplot(filter(t_tod_sw, STATE == "Karnataka"), 
         aes(as.factor(HOUR), PROP.LISTS, colour = COVID)) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Karnataka",
         x = "Time of day (hours)", y = "Proportion of lists")) /
((ggplot(filter(t_tod_sw, STATE == "Maharashtra"), 
         aes(as.factor(HOUR), PROP.LISTS, colour = COVID)) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Maharashtra",
         x = "Time of day (hours)", y = "Proportion of lists")) |
ggplot(filter(t_tod_sw, STATE == "Assam"), 
         aes(as.factor(HOUR), PROP.LISTS, colour = COVID)) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Assam",
         x = "Time of day (hours)", y = "Proportion of lists")) +
plot_layout(guides = "collect") +
plot_annotation(title = "Change in temporal spread of checklists with the pandemic (September to March)",
                subtitle = "Y-axis shows average proportion of checklists from respective times of the day on X-axis") -> t_spread_tod

t_spread_tod

ggsave("03_wrap_figs/14_t_spread_tod.png", 
       plot = t_spread_tod,
       width = 13, height = 9, units = "in")


rm(t_tod_nw, t_tod_sw, t_spread_tod)


```

-   At the national level, there are two peaks in the day that see highest birding activity: early morning and evening.\
-   In 2020, there was a sharp rise in birding activity at 17:00, and dip at 16:00.\
-   In 2021, birding increased at 06:00 and 15:00, and decreased at 16:00.\
-   The increase in evening birding was prominent in Karnataka, Gujarat and Maharashtra.

### Media

Not "per observer", as this metric from data POV.

```{r media, cache=TRUE, message=FALSE}

temp1 <- data0_MY_slice_G %>% 
  group_by(COVID, M.YEAR, MONTH, STATE) %>% 
  dplyr::summarise(NO.LISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>%
  ungroup() 

temp2 <- data0_MY_slice_G %>% 
  filter(HAS.MEDIA == 1) %>% 
  group_by(COVID, M.YEAR, MONTH, STATE) %>% 
  dplyr::summarise(NO.MLISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>%
  ungroup()

media_sw <- temp1 %>% 
  left_join(temp2) %>% 
  mutate(NO.MLISTS = replace_na(NO.MLISTS, 0),
         PROP.MLISTS = NO.MLISTS/NO.LISTS,
         SE = sqrt((PROP.MLISTS)*(1 - PROP.MLISTS)/NO.LISTS),
         CI.L = PROP.MLISTS - 1.96*SE,
         CI.U = PROP.MLISTS + 1.96*SE) 

rm(list = c("temp1","temp2"))

media_nw <- media_sw %>% 
  group_by(COVID, M.YEAR, MONTH) %>% 
  dplyr::summarise(PROP.MLISTS = mean(PROP.MLISTS),
            SE = sqrt(sum((SE)^2))/n(),
            CI.L = PROP.MLISTS - 1.96*SE,
            CI.U = PROP.MLISTS + 1.96*SE) 


(ggplot(media_nw, aes(MONTH, PROP.MLISTS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "National level",
         x = "Month", y = "Proportion of media lists")) /
((ggplot(filter(media_sw, STATE == "Kerala"), 
         aes(MONTH, PROP.MLISTS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Kerala",
         x = "Month", y = "Proportion of media lists")) |
(ggplot(filter(media_sw, STATE == "Karnataka"), 
         aes(MONTH, PROP.MLISTS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Karnataka",
         x = "Month", y = "Proportion of media lists"))) /
((ggplot(filter(media_sw, STATE == "Maharashtra"), 
         aes(MONTH, PROP.MLISTS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Maharashtra",
         x = "Month", y = "Proportion of media lists")) |
(ggplot(filter(media_sw, STATE == "Assam"), 
         aes(MONTH, PROP.MLISTS, colour = COVID)) + 
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Assam",
         x = "Month", y = "Proportion of media lists"))) +
plot_layout(guides= "collect") +
plot_annotation(title = "Change in proportion of media lists with the pandemic",
                subtitle = "Y-axis shows proportion of checklists that contain uploaded media") -> media

media

ggsave("03_wrap_figs/15_media.png", plot = media,
       width = 9, height = 9, units = "in")

rm(list = c("media_nw","media_sw","media"))

```

-   There is no major difference in proportion of media lists before, during and after the pandemic.

### New birders

```{r new, cache=TRUE, message=FALSE}

load("data/new_obsr_data.RData")

# not new to the state, but new overall, just grouped by state of first eBirding
new_obsr_sw <- new_obsr_data %>% 
  group_by(COVID, LE.YEAR, LE.MONTH, STATE) %>% 
  dplyr::summarise(NEW = n_distinct(OBSERVER.ID)) %>% 
  ungroup()

set.seed(765)
new_obsr_nw <- new_obsr_sw %>% 
  group_by(COVID, LE.YEAR, LE.MONTH) %>% 
  dplyr::summarise(NEW = boot_conf(x = NEW)) %>% 
  group_by(COVID, LE.YEAR, LE.MONTH) %>% 
  dplyr::summarise(SE = sd(NEW),
                   NEW = mean(NEW),
                   CI.L = NEW - 1.96*SE,
                   CI.U = NEW + 1.96*SE)


(ggplot(new_obsr_nw, aes(LE.MONTH, NEW, colour = COVID)) + 
    scale_y_log10() +
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.25, width = 0.4, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "National level",
         x = "Month", y = "New eBirders")) /
((ggplot(filter(new_obsr_sw, STATE == "Kerala"), 
         aes(LE.MONTH, NEW, colour = COVID)) + 
    scale_y_log10() +
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Kerala",
         x = "Month", y = "New eBirders")) |
(ggplot(filter(new_obsr_sw, STATE == "Karnataka"), 
         aes(LE.MONTH, NEW, colour = COVID)) + 
    scale_y_log10() +
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Karnataka",
         x = "Month", y = "New eBirders"))) /
((ggplot(filter(new_obsr_sw, STATE == "Maharashtra"), 
         aes(LE.MONTH, NEW, colour = COVID)) + 
    scale_y_log10() +
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Maharashtra",
         x = "Month", y = "New eBirders")) |
(ggplot(filter(new_obsr_sw, STATE == "Assam"), 
         aes(LE.MONTH, NEW, colour = COVID)) + 
    scale_y_log10() +
    scale_x_continuous(breaks = 1:12) + 
    geom_point(size = 3, position = position_dodge(0.5)) +
    scale_colour_manual(values = covid_palette) +
    labs(title = "Assam",
         x = "Month", y = "New eBirders"))) +
plot_layout(guides = "collect") +
plot_annotation(title = "Change in eBirder growth with the pandemic",
                subtitle = "Y-axis shows average number of new eBirders") -> new

new # plot

ggsave("03_wrap_figs/16_new.png", plot = new,
       width = 9, height = 9, units = "in")


rm(list = c("new_obsr_nw","new_obsr_sw","new"))

```

-   At the national level, the pandemic has seen a greater number of monthly new eBirders than before, presumably due to greater frequency and impact of outreach initiatives.\
-   While Kerala and Karnataka both have seen increases in monthly new eBirders with the pandemic, the increase is more striking in Karnataka than in Kerala.
