---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Bird behaviour

Two sets of bird species will be selected, one of birds common in human-dominated areas, and the other of common birds that are not very common in human-dominated areas. Reporting frequencies of all species will be calculated for every month and year. If there is an increase during the Mar--May period of birds in the first set and a decrease of birds in the second set, this means that birdwatcher bias has affected bird abundance as seen from the data.

```{r b_setup, cache=TRUE, message=FALSE}

load("data/data0_MY_b.RData")
load("data/data0_MY_b_slice.RData")

### bird species to analyse/compare ####

# Guwahati lacks PBFl, WCBa (but BTBa), InRo, GrFr, PiBu, BWKi, GHSw (in good numbers)
# Ernakulam lacks GrFr, SmMi, BWKi, LTSh, PiBu (in good numbers)
# different from list for BLR and PUN because 5 non-urban species from that list do not occur in sufficient numbers in EKM

species_bang <- data.frame(
  COMMON.NAME = c("House Crow", "Common Myna", "Jungle Myna", "Rock Pigeon", "Black Kite", "Common Tailorbird", "Rose-ringed Parakeet", "Asian Koel", "Purple Sunbird", "Pale-billed Flowerpecker", "White-cheeked Barbet", 
                  "Little Cormorant", "Gray-headed Swamphen", "Indian Pond-Heron",
                  "Green Bee-eater", "Black Drongo", "Pied Bushchat", "Plain Prinia", "Indian Robin", "House Sparrow", "Spotted Dove", "Coppersmith Barbet", "Black-winged Kite", "Small Minivet", "Long-tailed Shrike", "Gray Francolin"),
  SP.CATEGORY = c(rep("U", 11), rep("W", 3), rep("R", 12)))

# removing Black Drongo for models to run
species_pune <- data.frame(
  COMMON.NAME = c("House Crow", "Common Myna", "Jungle Myna", "Rock Pigeon", "Black Kite", "Common Tailorbird", "Rose-ringed Parakeet", "Asian Koel", "Purple Sunbird", "Pale-billed Flowerpecker", "White-cheeked Barbet", 
                  "Little Cormorant", "Gray-headed Swamphen", "Indian Pond-Heron",
                  "Green Bee-eater", "Pied Bushchat", "Plain Prinia", "Indian Robin", "House Sparrow", "Spotted Dove", "Coppersmith Barbet", "Black-winged Kite", "Small Minivet", "Long-tailed Shrike", "Gray Francolin"),
  SP.CATEGORY = c(rep("U", 11), rep("W", 3), rep("R", 11)))

species_koch <- data.frame(
  COMMON.NAME = c("House Crow", "Common Myna", "Jungle Myna", "Rock Pigeon", "Black Kite", "Common Tailorbird", "Rose-ringed Parakeet", "Asian Koel", "Purple Sunbird", "Pale-billed Flowerpecker", "White-cheeked Barbet", 
                  "Little Cormorant", "Gray-headed Swamphen", "Indian Pond-Heron",
                  "Green Bee-eater", "Black Drongo", "Plain Prinia", "Indian Robin", "House Sparrow", "Spotted Dove", "Coppersmith Barbet", "White-throated Kingfisher", "Black-rumped Flameback", "Rufous Treepie"),
  SP.CATEGORY = c(rep("U", 11), rep("W", 3), rep("R", 10)))


species_guwa <- data.frame(
  COMMON.NAME = c("House Crow", "Common Myna", "Jungle Myna", "Rock Pigeon", "Black Kite", "Common Tailorbird", "Rose-ringed Parakeet", "Asian Koel", "Purple Sunbird", "Blue-throated Barbet",
                  "Little Cormorant", "Indian Pond-Heron",
                  "Green Bee-eater", "Black Drongo", "Plain Prinia", "House Sparrow", "Spotted Dove", "Coppersmith Barbet", "White-throated Kingfisher", "Rufous Treepie", "Long-tailed Shrike"),
  SP.CATEGORY = c(rep("U", 10), rep("W", 2), rep("R", 9)))


species_list <- bind_rows("Karnataka" = species_bang,
                          "Maharashtra" = species_pune,
                          "Kerala" = species_koch,
                          "Assam" = species_guwa,
                          .id = "STATE") %>% 
  filter(!(SP.CATEGORY %in% "W")) # removing wetland spp. cos we want to compare UNU

rm(species_bang, species_pune, species_koch, species_guwa)


### choosing grid cells of 4 cities ####

# library(leaflet)
# 
# leaflet() %>%
#   addTiles() %>%
#   addPolygons(data = districtmap, color = "#000000",
#               label = sp::getSpPPolygonsIDSlots(districtmap)) %>%
#   addPolygons(data = gridmapg1, color = "#000000",
#               label = sp::getSpPPolygonsIDSlots(gridmapg1))

# selecting metropolitan subset of cells covered by district
city_bang <- c(13821, 13822, 13823, 13951, 13952, 13953, 14082, 14083)
city_pune <- c(10685, 10686)
city_koch <- c(15636, 15637, 15766, 15767)
city_guwa <- c(6345)

city_list <- data.frame(CITY = c(rep("Bengaluru", 8), rep("Pune", 2), 
                                 rep("Kochi", 4), "Guwahati"),
                        CELLS = c(city_bang, city_pune, city_koch, city_guwa),
                        COUNTY = c(rep("Bengaluru Urban", 8), rep("Pune", 2), 
                                   rep("Ernakulam", 4), "Kamrup Metropolitan"),
                        STATE = c(rep("Karnataka", 8), rep("Maharashtra", 2), 
                                  rep("Kerala", 4), "Assam")) %>% 
  mutate(STATE = factor(STATE, levels = anal_states[1,])) %>% 
  arrange(STATE)

rm(city_bang, city_pune, city_koch, city_guwa)

### highly birded personal locations ####

# eBird user info
eBird_users <- read.delim("data/ebd_users_relMay-2022.txt", 
                          sep = "\t", header = T, quote = "", 
                          stringsAsFactors = F, na.strings = c(""," ",NA)) %>% 
  transmute(OBSERVER.ID = observer_id,
            FULL.NAME = paste(first_name, last_name, sep = " "))

# # list of prolific birders judging from eBird India leaderboard
# prolific_a <- data.frame(FULL.NAME = c("Renju TR", "Jayadev  Menon", "Steffin Babu",
#                          "Panchapakesan Jeganathan", "Premchand Reghuvaran", "Sahana M")) %>% 
#   left_join(eBird_users)

# getting overall checklist leaders from data
prolific_a <- data0_MY_b_slice_S %>% 
  group_by(OBSERVER.ID, LOCALITY.ID, LOCALITY) %>% 
  summarise(TOT.LISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>% 
  ungroup() %>% 
  filter(TOT.LISTS >= 800) %>% # 200 per year threshold (arbitrary)
  arrange(desc(TOT.LISTS)) %>% 
  left_join(eBird_users) %>% 
  select(-TOT.LISTS)
  

# getting checklist leaders with equal effort across years from data
prolific_b <- data0_MY_b_slice_S %>% 
  group_by(OBSERVER.ID, LOCALITY.ID, LOCALITY, M.YEAR, MONTH) %>% 
  summarise(MONTH.LISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>% 
  filter(MONTH.LISTS >= 15) %>% # 200 per year roughly translates to 16 per month
  summarise(N = n_distinct(MONTH),
            TOT.LISTS = sum(MONTH.LISTS)) %>% 
  filter(N >= 10) %>% # at least 10 months in a year
  summarise(N = n_distinct(M.YEAR),
            TOT.LISTS = sum(TOT.LISTS)) %>% 
  filter(N == 4) %>% # all four years
  ungroup() %>% 
  arrange(desc(TOT.LISTS)) %>% 
  left_join(eBird_users) %>% 
  select(-N, -TOT.LISTS)

prolific <- inner_join(prolific_a, prolific_b)

rm(prolific_a, prolific_b)

```

## Overall species reporting

```{r b01_overall_s1, cache=TRUE, message=FALSE}

anal_name <- "b01_overall_s1a_m"

b01_overall_monthly("Karnataka")


anal_name <- "b01_overall_s1b_y"

b01_overall_annual("Karnataka")

```

Nothing conclusively due to the pandemic or in one particular direction, at the scale of Bengaluru city. At the scale of the entire state though, some non-urban species showed lower reporting only during April of 2020 (first lockdown).

At the annual level:

Urban species show a wide range of patterns, often contradictory. But most species show a decrease during the pandemic year, which is contrary to the prediction of a pandemic effect. Most non-urban species saw a decline in reporting in the pandemic year in Bengaluru city. Both these patterns soften out when averaged across space, i.e., at the level of the entire state of Karnataka.

```{r b01_overall_s1c_model, cache=TRUE, message=FALSE}

anal_name <- "b01_overall_s1c_model"


### State 1

state_name <- "Karnataka"

source("scripts/b01_overall_model.R")





b1_KA4_pred <- birds_pred
b1_KA4_graph <- birds_graph

  
  (ggplot(filter(birds_pred, MONTHS.TYPE == "LD"), 
          aes(M.YEAR, REP.FREQ.PRED, col = SP.CATEGORY)) +
      scale_color_manual(values = c("#8F85C1", "#A3383C"),
                         name = "Species category",
                         labels = c("Rural", "Urban")) +
      scale_y_continuous(limits = c(0.1, 1.0)) +
      labs(title = "For the months of April and May",
           x = "Migratory year", y = "Predicted reporting frequency") +
      geom_point(size = 3, position = position_dodge(0.5)) +
      geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                    size = 1.5, width = 0.2, position = position_dodge(0.5)) |
      ggplot(filter(birds_pred, MONTHS.TYPE == "NL"), 
             aes(M.YEAR, REP.FREQ.PRED, col = SP.CATEGORY)) +
      scale_color_manual(values = c("#8F85C1", "#A3383C"),
                         name = "Species category",
                         labels = c("Rural", "Urban")) +
      scale_y_continuous(limits = c(0.1, 1.0)) +
      labs(title = "For other ten months",
           x = "Migratory year", y = "Predicted reporting frequency") +
      geom_point(size = 3, position = position_dodge(0.5)) +
      geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                    size = 1.5, width = 0.2, position = position_dodge(0.5)) |
      ggplot(filter(birds_pred, MONTHS.TYPE == "ALL"), 
             aes(M.YEAR, REP.FREQ.PRED, col = SP.CATEGORY)) +
      scale_color_manual(values = c("#8F85C1", "#A3383C"),
                         name = "Species category",
                         labels = c("Rural", "Urban")) +
      scale_y_continuous(limits = c(0.1, 1.0)) +
      labs(title = "For all twelve months",
           x = "Migratory year", y = "Predicted reporting frequency") +
      geom_point(size = 3, position = position_dodge(0.5)) +
      geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                    size = 1.5, width = 0.2, position = position_dodge(0.5))) +
    plot_layout(guides = "collect") +
    plot_annotation(title = paste(state, "state"),
                    subtitle = paste0(
                      "Predicted reporting frequencies of ",
                      n_distinct(data_occ$COMMON.NAME), " species (",
                      n_distinct(filter(data_occ, SP.CATEGORY == "U")$COMMON.NAME), " urban, ",
                      n_distinct(filter(data_occ, SP.CATEGORY == "R")$COMMON.NAME), " rural)",
                      " in three separate models")) -> birds_graph
  
  assign("birds_pred", birds_pred, envir = .GlobalEnv)
  assign("birds_graph", birds_graph, envir = .GlobalEnv)
  
save(b1_KA4_pred, file = "outputs/b1_KA4_pred.RData")

ggsave("03_wrap_figs/b1_KA4_model.png", b1_KA4_graph,
       width = 21, height = 6, units = "in", dpi = 300)

```

-   Non-urban birds had lower reporting frequency in MY 2020, as seen from the MY:month interaction.
-   However, at the annual level, urban species did not show a higher reporting frequency, which could be because birders in Karnataka compensated during the winter months with exceptional non-urban birding.
-   The crucial months like April and May show higher urban species reporting frequency overall.
-   Changes in birder behaviour during the peak period did not translate to notable changes in reporting frequency of bird species. The slight suggestion (still insignificant) of a change in MY2019 during the peak period completely disappears at the larger temporal scale.

## Kochi

```{r b_koch_01_monthly, cache=TRUE, message=FALSE}

birds_monthly <- data0_MY_b_slice_G %>% 
  filter(COUNTY == "Ernakulam", CELL.ID %in% city_koch) %>% 
  group_by(COVID, M.YEAR, MONTH) %>% 
  summarise(GROUP.ID = GROUP.ID, # not mutating so as to remove unwanted join columns
            TOT.LISTS = n_distinct(GROUP.ID)) %>% 
  left_join(data0) %>% 
  filter(COMMON.NAME %in% species_koch$COMMON.NAME) %>%
  group_by(COMMON.NAME, COVID, M.YEAR, MONTH) %>% 
  summarise(TOT.LISTS = min(TOT.LISTS),
            NO.LISTS = n_distinct(GROUP.ID),
            REP.FREQ = round(NO.LISTS/TOT.LISTS, 4),
            SE = sqrt((REP.FREQ)*(1 - REP.FREQ)/TOT.LISTS),
            CI.L = REP.FREQ - 1.96*SE,
            CI.U = REP.FREQ + 1.96*SE) %>% 
  left_join(species_koch)


((ggplot(filter(birds_monthly, SP.CATEGORY == "U"), 
       aes(MONTH, REP.FREQ, colour = COVID)) +
  geom_point(size = 1.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.6, position = position_dodge(0.5)) +
  scale_x_continuous(breaks = 1:12) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Urban species",
       x = "Month", y = "Reporting frequency")) /
(ggplot(filter(birds_monthly, SP.CATEGORY == "R"), 
       aes(MONTH, REP.FREQ, colour = COVID)) +
  geom_point(size = 1.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.6, position = position_dodge(0.5)) +
  scale_x_continuous(breaks = 1:12) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Non-urban species",
       x = "Month", y = "Reporting frequency")) /
(ggplot(filter(birds_monthly, SP.CATEGORY == "W"), 
       aes(MONTH, REP.FREQ, colour = COVID)) +
  geom_point(size = 1.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.6, position = position_dodge(0.5)) +
  scale_x_continuous(breaks = 1:12) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Wetland species",
       x = "Month", y = "Reporting frequency"))) +
plot_layout(guides = "collect", heights = c(4, 4, 1)) +
plot_annotation(title = "Kochi city") -> b_koch_01_monthly


ggsave("03_wrap_figs/b_koch_01_monthly.png", b_koch_01_monthly,
       width = 11, height = 15, units = "in", dpi = 300)

```

No prominent effect of the pandemic. Biggest change is in wetland birds.

```{r b_koch_02_annual, cache=TRUE, message=FALSE}

birds <- data0_MY_b_slice_G %>% 
  filter(COUNTY == "Ernakulam", CELL.ID %in% city_koch) %>% 
  group_by(M.YEAR, MONTH) %>% 
  summarise(GROUP.ID = GROUP.ID,
            TOT.LISTS = n_distinct(GROUP.ID)) %>% 
  left_join(data0) %>% 
  filter(COMMON.NAME %in% species_koch$COMMON.NAME) %>% 
  group_by(COMMON.NAME, M.YEAR, MONTH) %>% 
  summarise(TOT.LISTS = min(TOT.LISTS),
            NO.LISTS = n_distinct(GROUP.ID),
            REP.FREQ = round(NO.LISTS/TOT.LISTS, 4),
            # not bootstrapping here cos only averaging across 12 months
            # using binomial confidence
            SE = sqrt((REP.FREQ)*(1 - REP.FREQ)/TOT.LISTS)) %>% 
  summarise(REP.FREQ = mean(REP.FREQ),
            SE = sqrt(sum((SE)^2))/n(),
            CI.L = REP.FREQ - 1.96*SE,
            CI.U = REP.FREQ + 1.96*SE) %>% 
  left_join(species_koch)


((ggplot(filter(birds, SP.CATEGORY == "U"), 
       aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.1, width = 0.1, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Urban species",
       x = "Migratory year", y = "Reporting frequency")) /
(ggplot(filter(birds, SP.CATEGORY == "R"), 
       aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.1, width = 0.1, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Non-urban species",
       x = "Migratory year", y = "Reporting frequency")) /
(ggplot(filter(birds, SP.CATEGORY == "W"), 
       aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.1, width = 0.1, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Wetland species",
       x = "Migratory year", y = "Reporting frequency"))) +
plot_layout(guides = "collect", heights = c(4, 4, 1)) +
plot_annotation(title = "Kochi city") -> b_koch_02_annual


ggsave("03_wrap_figs/b_koch_02_annual.png", b_koch_02_annual,
       width = 10, height = 14, units = "in", dpi = 300)


```

No prominent effects directly attributable to the pandemic. Most patterns even out.

```{r b_koch_03_spatial, cache=TRUE, message=FALSE}

birds_monthly <- data0_MY_b_slice_G %>% 
  filter(STATE == "Kerala") %>% 
  group_by(COVID, M.YEAR, MONTH, CELL.ID) %>% 
  summarise(GROUP.ID = GROUP.ID, # not mutating so as to remove unwanted join columns
            TOT.LISTS = n_distinct(GROUP.ID)) %>% 
  left_join(data0) %>% 
  filter(COMMON.NAME %in% species_koch$COMMON.NAME) %>% 
  group_by(COMMON.NAME, COVID, M.YEAR, MONTH, CELL.ID) %>% 
  summarise(TOT.LISTS = min(TOT.LISTS),
            NO.LISTS = n_distinct(GROUP.ID),
            REP.FREQ = round(NO.LISTS/TOT.LISTS, 4),
            # not bootstrapping here cos only averaging across 12 months
            # using binomial confidence
            SE = sqrt((REP.FREQ)*(1 - REP.FREQ)/TOT.LISTS)) %>% 
  summarise(REP.FREQ = mean(REP.FREQ),
            SE = sqrt(sum((SE)^2))/n(),
            CI.L = REP.FREQ - 1.96*SE,
            CI.U = REP.FREQ + 1.96*SE) %>% 
  left_join(species_koch)


((ggplot(filter(birds_monthly, SP.CATEGORY == "U"), 
       aes(MONTH, REP.FREQ, colour = COVID)) +
  geom_point(size = 1.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.6, position = position_dodge(0.5)) +
  scale_x_continuous(breaks = 1:12) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Urban species",
       x = "Month", y = "Reporting frequency")) /
(ggplot(filter(birds_monthly, SP.CATEGORY == "R"), 
       aes(MONTH, REP.FREQ, colour = COVID)) +
  geom_point(size = 1.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.6, position = position_dodge(0.5)) +
  scale_x_continuous(breaks = 1:12) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Non-urban species",
       x = "Month", y = "Reporting frequency")) /
(ggplot(filter(birds_monthly, SP.CATEGORY == "W"), 
       aes(MONTH, REP.FREQ, colour = COVID)) +
  geom_point(size = 1.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.6, position = position_dodge(0.5)) +
  scale_x_continuous(breaks = 1:12) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Wetland species",
       x = "Month", y = "Reporting frequency"))) +
plot_layout(guides = "collect", heights = c(4, 4, 1)) +
plot_annotation(title = "Kerala state") -> b_koch_03_spatial_m

ggsave("03_wrap_figs/b_koch_03_spatial_m.png", b_koch_03_spatial_m,
       width = 11, height = 15, units = "in", dpi = 300)



birds <- data0_MY_b_slice_G %>% 
  filter(STATE == "Kerala") %>% 
  group_by(M.YEAR, MONTH, CELL.ID) %>% 
  summarise(GROUP.ID = GROUP.ID,
            TOT.LISTS = n_distinct(GROUP.ID)) %>% 
  left_join(data0) %>% 
  filter(COMMON.NAME %in% species_koch$COMMON.NAME) %>% 
  group_by(COMMON.NAME, M.YEAR, MONTH, CELL.ID) %>% 
  summarise(TOT.LISTS = min(TOT.LISTS),
            NO.LISTS = n_distinct(GROUP.ID),
            REP.FREQ = round(NO.LISTS/TOT.LISTS, 4),
            # not bootstrapping here cos only averaging across 12 months
            # using binomial confidence
            SE = sqrt((REP.FREQ)*(1 - REP.FREQ)/TOT.LISTS)) %>% 
  summarise(REP.FREQ = mean(REP.FREQ),
            SE = sqrt(sum((SE)^2))/n()) %>% 
  summarise(REP.FREQ = mean(REP.FREQ),
            SE = sqrt(sum((SE)^2))/n(),
            CI.L = REP.FREQ - 1.96*SE,
            CI.U = REP.FREQ + 1.96*SE) %>% 
  left_join(species_koch)


((ggplot(filter(birds, SP.CATEGORY == "U"), 
       aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.1, width = 0.1, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Urban species",
       x = "Migratory year", y = "Reporting frequency")) /
(ggplot(filter(birds, SP.CATEGORY == "R"), 
       aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.1, width = 0.1, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Non-urban species",
       x = "Migratory year", y = "Reporting frequency")) /
(ggplot(filter(birds, SP.CATEGORY == "W"), 
       aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.1, width = 0.1, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Wetland species",
       x = "Migratory year", y = "Reporting frequency"))) +
plot_layout(guides = "collect", heights = c(4, 4, 1)) +
plot_annotation(title = "Kerala state") -> b_koch_03_spatial_a


ggsave("03_wrap_figs/b_koch_03_spatial_a.png", b_koch_03_spatial_a,
       width = 10, height = 14, units = "in", dpi = 300)


```

3-4 non-urban species decreased during the 2-3 months of "peak" and wetland species decreased during 5-6 months of the "peak" and following it. At the annual scale, a few urban species show slightly higher frequencies during migratory year 2019, but others show different patterns. Effect sizes will need to be judged. Non-urban and wetland species were not affected in any particular direction consistently (across species) by the pandemic.

```{r b2_KL4_model, cache=TRUE, message=FALSE}

bird_model_state(state = "Kerala")

b2_KL4_pred <- birds_pred
b2_KL4_graph <- birds_graph
  
save(b2_KL4_pred, file = "outputs/b2_KL4_pred.RData")

ggsave("03_wrap_figs/b2_KL4_model.png", b2_KL4_graph,
       width = 21, height = 6, units = "in", dpi = 300)

```

## Pune

```{r b_pune_01_monthly, cache=TRUE, message=FALSE}

birds_monthly <- data0_MY_b_slice_G %>% 
  filter(COUNTY == "Pune", CELL.ID %in% city_pune) %>% 
  group_by(COVID, M.YEAR, MONTH) %>% 
  summarise(GROUP.ID = GROUP.ID, # not mutating so as to remove unwanted join columns
            TOT.LISTS = n_distinct(GROUP.ID)) %>% 
  left_join(data0) %>% 
  filter(COMMON.NAME %in% species$COMMON.NAME, 
         COMMON.NAME != "White-cheeked Barbet") %>% 
  group_by(COMMON.NAME, COVID, M.YEAR, MONTH) %>% 
  summarise(TOT.LISTS = min(TOT.LISTS),
            NO.LISTS = n_distinct(GROUP.ID),
            REP.FREQ = round(NO.LISTS/TOT.LISTS, 4),
            SE = sqrt((REP.FREQ)*(1 - REP.FREQ)/TOT.LISTS),
            CI.L = REP.FREQ - 1.96*SE,
            CI.U = REP.FREQ + 1.96*SE) %>% 
  left_join(species)


((ggplot(filter(birds_monthly, SP.CATEGORY == "U"), 
       aes(MONTH, REP.FREQ, colour = COVID)) +
  geom_point(size = 1.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.6, position = position_dodge(0.5)) +
  scale_x_continuous(breaks = 1:12) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Urban species",
       x = "Month", y = "Reporting frequency")) /
(ggplot(filter(birds_monthly, SP.CATEGORY == "R"), 
       aes(MONTH, REP.FREQ, colour = COVID)) +
  geom_point(size = 1.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.6, position = position_dodge(0.5)) +
  scale_x_continuous(breaks = 1:12) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Non-urban species",
       x = "Month", y = "Reporting frequency")) /
(ggplot(filter(birds_monthly, SP.CATEGORY == "W"), 
       aes(MONTH, REP.FREQ, colour = COVID)) +
  geom_point(size = 1.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.6, position = position_dodge(0.5)) +
  scale_x_continuous(breaks = 1:12) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Wetland species",
       x = "Month", y = "Reporting frequency"))) +
plot_layout(guides = "collect", heights = c(4, 4, 1)) +
plot_annotation(title = "Pune city") -> b_pune_01_monthly


ggsave("03_wrap_figs/b_pune_01_monthly.png", b_pune_01_monthly,
       width = 11, height = 15, units = "in", dpi = 300)

```

Many non-urban and wetland bird species showed a slight decrease in reporting frequency during the "peak" period, but none of the urban species did. Also, WCBa is rare in Pune city so can be ignored at this level.

```{r b_pune_02_annual, cache=TRUE, message=FALSE}

birds <- data0_MY_b_slice_G %>% 
  filter(COUNTY == "Pune", CELL.ID %in% city_pune) %>% 
  group_by(M.YEAR, MONTH) %>% 
  summarise(GROUP.ID = GROUP.ID,
            TOT.LISTS = n_distinct(GROUP.ID)) %>% 
  left_join(data0) %>% 
  filter(COMMON.NAME %in% species$COMMON.NAME, 
         COMMON.NAME != "White-cheeked Barbet") %>% 
  group_by(COMMON.NAME, M.YEAR, MONTH) %>% 
  summarise(TOT.LISTS = min(TOT.LISTS),
            NO.LISTS = n_distinct(GROUP.ID),
            REP.FREQ = round(NO.LISTS/TOT.LISTS, 4),
            # not bootstrapping here cos only averaging across 12 months
            # using binomial confidence
            SE = sqrt((REP.FREQ)*(1 - REP.FREQ)/TOT.LISTS)) %>% 
  summarise(REP.FREQ = mean(REP.FREQ),
            SE = sqrt(sum((SE)^2))/n(),
            CI.L = REP.FREQ - 1.96*SE,
            CI.U = REP.FREQ + 1.96*SE) %>% 
  left_join(species)


((ggplot(filter(birds, SP.CATEGORY == "U"), 
       aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.1, width = 0.1, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Urban species",
       x = "Migratory year", y = "Reporting frequency")) /
(ggplot(filter(birds, SP.CATEGORY == "R"), 
       aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.1, width = 0.1, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Non-urban species",
       x = "Migratory year", y = "Reporting frequency")) /
(ggplot(filter(birds, SP.CATEGORY == "W"), 
       aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.1, width = 0.1, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Wetland species",
       x = "Migratory year", y = "Reporting frequency"))) +
plot_layout(guides = "collect", heights = c(4, 4, 1)) +
plot_annotation(title = "Pune city") -> b_pune_02_annual


ggsave("03_wrap_figs/b_pune_02_annual.png", b_pune_02_annual,
       width = 10, height = 14, units = "in", dpi = 300)


```

There is no clear direction of change with the pandemic in Pune city. What happens when considering a larger spatial scale?

```{r b_pune_03_spatial, cache=TRUE, message=FALSE}

birds_monthly <- data0_MY_b_slice_G %>% 
  filter(STATE == "Maharashtra") %>% 
  group_by(COVID, M.YEAR, MONTH, CELL.ID) %>% 
  summarise(GROUP.ID = GROUP.ID, # not mutating so as to remove unwanted join columns
            TOT.LISTS = n_distinct(GROUP.ID)) %>% 
  left_join(data0) %>% 
  filter(COMMON.NAME %in% species$COMMON.NAME, 
         COMMON.NAME != "White-cheeked Barbet") %>% 
  group_by(COMMON.NAME, COVID, M.YEAR, MONTH, CELL.ID) %>% 
  summarise(TOT.LISTS = min(TOT.LISTS),
            NO.LISTS = n_distinct(GROUP.ID),
            REP.FREQ = round(NO.LISTS/TOT.LISTS, 4),
            # not bootstrapping here cos only averaging across 12 months
            # using binomial confidence
            SE = sqrt((REP.FREQ)*(1 - REP.FREQ)/TOT.LISTS)) %>% 
  summarise(REP.FREQ = mean(REP.FREQ),
            SE = sqrt(sum((SE)^2))/n(),
            CI.L = REP.FREQ - 1.96*SE,
            CI.U = REP.FREQ + 1.96*SE) %>% 
  left_join(species)


((ggplot(filter(birds_monthly, SP.CATEGORY == "U"), 
       aes(MONTH, REP.FREQ, colour = COVID)) +
  geom_point(size = 1.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.6, position = position_dodge(0.5)) +
  scale_x_continuous(breaks = 1:12) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Urban species",
       x = "Month", y = "Reporting frequency")) /
(ggplot(filter(birds_monthly, SP.CATEGORY == "R"), 
       aes(MONTH, REP.FREQ, colour = COVID)) +
  geom_point(size = 1.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.6, position = position_dodge(0.5)) +
  scale_x_continuous(breaks = 1:12) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Non-urban species",
       x = "Month", y = "Reporting frequency")) /
(ggplot(filter(birds_monthly, SP.CATEGORY == "W"), 
       aes(MONTH, REP.FREQ, colour = COVID)) +
  geom_point(size = 1.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.6, position = position_dodge(0.5)) +
  scale_x_continuous(breaks = 1:12) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Wetland species",
       x = "Month", y = "Reporting frequency"))) +
plot_layout(guides = "collect", heights = c(4, 4, 1)) +
plot_annotation(title = "Maharashtra state") -> b_pune_03_spatial_m

ggsave("03_wrap_figs/b_pune_03_spatial_m.png", b_pune_03_spatial_m,
       width = 11, height = 15, units = "in", dpi = 300)



birds <- data0_MY_b_slice_G %>% 
  filter(STATE == "Maharashtra") %>% 
  group_by(M.YEAR, MONTH, CELL.ID) %>% 
  summarise(GROUP.ID = GROUP.ID,
            TOT.LISTS = n_distinct(GROUP.ID)) %>% 
  left_join(data0) %>% 
  filter(COMMON.NAME %in% species$COMMON.NAME, 
         COMMON.NAME != "White-cheeked Barbet") %>% 
  group_by(COMMON.NAME, M.YEAR, MONTH, CELL.ID) %>% 
  summarise(TOT.LISTS = min(TOT.LISTS),
            NO.LISTS = n_distinct(GROUP.ID),
            REP.FREQ = round(NO.LISTS/TOT.LISTS, 4),
            # not bootstrapping here cos only averaging across 12 months
            # using binomial confidence
            SE = sqrt((REP.FREQ)*(1 - REP.FREQ)/TOT.LISTS)) %>% 
  summarise(REP.FREQ = mean(REP.FREQ),
            SE = sqrt(sum((SE)^2))/n()) %>% 
  summarise(REP.FREQ = mean(REP.FREQ),
            SE = sqrt(sum((SE)^2))/n(),
            CI.L = REP.FREQ - 1.96*SE,
            CI.U = REP.FREQ + 1.96*SE) %>% 
  left_join(species)


((ggplot(filter(birds, SP.CATEGORY == "U"), 
       aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.1, width = 0.1, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Urban species",
       x = "Migratory year", y = "Reporting frequency")) /
(ggplot(filter(birds, SP.CATEGORY == "R"), 
       aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.1, width = 0.1, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Non-urban species",
       x = "Migratory year", y = "Reporting frequency")) /
(ggplot(filter(birds, SP.CATEGORY == "W"), 
       aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.1, width = 0.1, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Wetland species",
       x = "Migratory year", y = "Reporting frequency"))) +
plot_layout(guides = "collect", heights = c(4, 4, 1)) +
plot_annotation(title = "Maharashtra state") -> b_pune_03_spatial_a


ggsave("03_wrap_figs/b_pune_03_spatial_a.png", b_pune_03_spatial_a,
       width = 10, height = 14, units = "in", dpi = 300)


```

At the larger spatial scale, pandemic effects are restricted to the "peak" period, mostly in 2020. Most species decreased (surprisingly, including urban ones) in frequency. But translated to the annual scale, non-urban species also spread out, in that there is no longer one specific direction in which frequencies of all the species changed. Only urban species seem to show a decrease in frequency, probably due to increased birding in non-urban areas over the years.

```{r b3_MH4_model, cache=TRUE, message=FALSE}

bird_model_state(state = "Maharashtra")

b3_MH4_pred <- birds_pred
b3_MH4_graph <- birds_graph
  
save(b3_MH4_pred, file = "outputs/b3_MH4_pred.RData")

ggsave("03_wrap_figs/b3_MH4_model.png", b3_MH4_graph,
       width = 21, height = 6, units = "in", dpi = 300)

```

## Guwahati

Not enough lists in each month to analyse at this scale.

```{r b_guwa_02_annual, cache=TRUE, message=FALSE}

birds <- data0_MY_b_slice_G %>% 
  filter(COUNTY == "Kamrup Metropolitan", CELL.ID %in% city_guwa) %>% 
  group_by(M.YEAR, MONTH) %>% 
  summarise(GROUP.ID = GROUP.ID,
            TOT.LISTS = n_distinct(GROUP.ID)) %>% 
  left_join(data0) %>% 
  filter(COMMON.NAME %in% species_guwa$COMMON.NAME) %>% 
  group_by(COMMON.NAME, M.YEAR, MONTH) %>% 
  summarise(TOT.LISTS = min(TOT.LISTS),
            NO.LISTS = n_distinct(GROUP.ID),
            REP.FREQ = round(NO.LISTS/TOT.LISTS, 4),
            # not bootstrapping here cos only averaging across 12 months
            # using binomial confidence
            SE = sqrt((REP.FREQ)*(1 - REP.FREQ)/TOT.LISTS)) %>% 
  summarise(REP.FREQ = mean(REP.FREQ),
            SE = sqrt(sum((SE)^2))/n(),
            CI.L = REP.FREQ - 1.96*SE,
            CI.U = REP.FREQ + 1.96*SE) %>% 
  left_join(species_guwa)


((ggplot(filter(birds, SP.CATEGORY == "U"), 
       aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.1, width = 0.1, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Urban species",
       x = "Migratory year", y = "Reporting frequency")) /
(ggplot(filter(birds, SP.CATEGORY == "R"), 
       aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.1, width = 0.1, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Non-urban species",
       x = "Migratory year", y = "Reporting frequency")) /
(ggplot(filter(birds, SP.CATEGORY == "W"), 
       aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.1, width = 0.1, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Wetland species",
       x = "Migratory year", y = "Reporting frequency"))) +
plot_layout(guides = "collect", heights = c(4, 4, 1)) +
plot_annotation(title = "Guwahati city") -> b_guwa_02_annual


ggsave("03_wrap_figs/b_guwa_02_annual.png", b_guwa_02_annual,
       width = 10, height = 14, units = "in", dpi = 300)


```

Non-urban and wetland species do not show an impact, while some urban species did have slightly higher frequencies in migratory year 2019. However, confidence intervals are mostly overlapping and effect size is likely small.

```{r b_guwa_03_spatial, cache=TRUE, message=FALSE}

birds_monthly <- data0_MY_b_slice_G %>% 
  filter(STATE == "Assam") %>% 
  group_by(COVID, M.YEAR, MONTH, CELL.ID) %>% 
  summarise(GROUP.ID = GROUP.ID, # not mutating so as to remove unwanted join columns
            TOT.LISTS = n_distinct(GROUP.ID)) %>% 
  left_join(data0) %>% 
  filter(COMMON.NAME %in% species_guwa$COMMON.NAME) %>% 
  group_by(COMMON.NAME, COVID, M.YEAR, MONTH, CELL.ID) %>% 
  summarise(TOT.LISTS = min(TOT.LISTS),
            NO.LISTS = n_distinct(GROUP.ID),
            REP.FREQ = round(NO.LISTS/TOT.LISTS, 4),
            # not bootstrapping here cos only averaging across 12 months
            # using binomial confidence
            SE = sqrt((REP.FREQ)*(1 - REP.FREQ)/TOT.LISTS)) %>% 
  summarise(REP.FREQ = mean(REP.FREQ),
            SE = sqrt(sum((SE)^2))/n(),
            CI.L = REP.FREQ - 1.96*SE,
            CI.U = REP.FREQ + 1.96*SE) %>% 
  left_join(species_guwa)


((ggplot(filter(birds_monthly, SP.CATEGORY == "U"), 
       aes(MONTH, REP.FREQ, colour = COVID)) +
  geom_point(size = 1.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.6, position = position_dodge(0.5)) +
  scale_x_continuous(breaks = 1:12) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Urban species",
       x = "Month", y = "Reporting frequency")) /
(ggplot(filter(birds_monthly, SP.CATEGORY == "R"), 
       aes(MONTH, REP.FREQ, colour = COVID)) +
  geom_point(size = 1.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.6, position = position_dodge(0.5)) +
  scale_x_continuous(breaks = 1:12) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Non-urban species",
       x = "Month", y = "Reporting frequency")) /
(ggplot(filter(birds_monthly, SP.CATEGORY == "W"), 
       aes(MONTH, REP.FREQ, colour = COVID)) +
  geom_point(size = 1.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.6, position = position_dodge(0.5)) +
  scale_x_continuous(breaks = 1:12) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Wetland species",
       x = "Month", y = "Reporting frequency"))) +
plot_layout(guides = "collect", heights = c(4, 4, 1)) +
plot_annotation(title = "Assam state") -> b_guwa_03_spatial_m

ggsave("03_wrap_figs/b_guwa_03_spatial_m.png", b_guwa_03_spatial_m,
       width = 11, height = 15, units = "in", dpi = 300)



birds <- data0_MY_b_slice_G %>% 
  filter(STATE == "Assam") %>% 
  group_by(M.YEAR, MONTH, CELL.ID) %>% 
  summarise(GROUP.ID = GROUP.ID,
            TOT.LISTS = n_distinct(GROUP.ID)) %>% 
  left_join(data0) %>% 
  filter(COMMON.NAME %in% species_guwa$COMMON.NAME) %>% 
  group_by(COMMON.NAME, M.YEAR, MONTH, CELL.ID) %>% 
  summarise(TOT.LISTS = min(TOT.LISTS),
            NO.LISTS = n_distinct(GROUP.ID),
            REP.FREQ = round(NO.LISTS/TOT.LISTS, 4),
            # not bootstrapping here cos only averaging across 12 months
            # using binomial confidence
            SE = sqrt((REP.FREQ)*(1 - REP.FREQ)/TOT.LISTS)) %>% 
  summarise(REP.FREQ = mean(REP.FREQ),
            SE = sqrt(sum((SE)^2))/n()) %>% 
  summarise(REP.FREQ = mean(REP.FREQ),
            SE = sqrt(sum((SE)^2))/n(),
            CI.L = REP.FREQ - 1.96*SE,
            CI.U = REP.FREQ + 1.96*SE) %>% 
  left_join(species_guwa)


((ggplot(filter(birds, SP.CATEGORY == "U"), 
       aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.1, width = 0.1, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Urban species",
       x = "Migratory year", y = "Reporting frequency")) /
(ggplot(filter(birds, SP.CATEGORY == "R"), 
       aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.1, width = 0.1, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Non-urban species",
       x = "Migratory year", y = "Reporting frequency")) /
(ggplot(filter(birds, SP.CATEGORY == "W"), 
       aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2.5, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.1, width = 0.1, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette) +
  facet_wrap(~COMMON.NAME, dir = "h", ncol = 3, strip.position = "left", scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(title = "Wetland species",
       x = "Migratory year", y = "Reporting frequency"))) +
plot_layout(guides = "collect", heights = c(4, 4, 1)) +
plot_annotation(title = "Assam state") -> b_guwa_03_spatial_a


ggsave("03_wrap_figs/b_guwa_03_spatial_a.png", b_guwa_03_spatial_a,
       width = 10, height = 14, units = "in", dpi = 300)


```

Very minor suggestions of changes during the 2 month period in a few species, but overall no pattern. At the annual scale, wetland species show some differences, but others show no overall change in reporting frequency due to the pandemic.

```{r b4_AS4_model, cache=TRUE, message=FALSE}

bird_model_state(state = "Assam")

b4_AS4_pred <- birds_pred
b4_AS4_graph <- birds_graph
  
save(b4_AS4_pred, file = "outputs/b4_AS4_pred.RData")

ggsave("03_wrap_figs/b4_AS4_model.png", b4_AS4_graph,
       width = 21, height = 6, units = "in", dpi = 300)

```

-   Unlike other states, Assam shows a bias in reporting towards urban species even at the annual and state-wide scale.\
-   The increase in urban birding proportion is only relative to MY2018, and not to MY2020, although the trend seems to be negative post-MY2019.\
-   How is this related to Bihu Bird Count (did it start in MY2019?), and to general rise in popularity of birding as well as skill of birders in Assam?

## Bird reporting with standard effort

Out of the most commonly observed species in each of the "prolific" locations, will retain those that are included in species list made for other analysis.

Not modelling for each species separately because small data size.

This is not done for individual states, because focus is individual locations which are anyway spread across the country.

```{r b5_prolific, cache=TRUE, message=FALSE}

# species list
species_prolific <- species_list %>% distinct(COMMON.NAME, SP.CATEGORY)

# to join for presence-absence of various species
temp1 <- data0_MY_b %>% 
  right_join(prolific) %>% 
  group_by(SAMPLING.EVENT.IDENTIFIER, COMMON.NAME) %>% 
  summarise(OBSERVATION.COUNT = max(OBSERVATION.COUNT)) %>% 
  ungroup()

# to later join checklist metadata
temp2 <- data0_MY_b_slice_S %>% 
  right_join(prolific) %>% 
  select(OBSERVER.ID, LOCALITY.ID, SAMPLING.EVENT.IDENTIFIER)

# join patches with patch-specific species
temp3 <- data0_MY_b %>% 
  right_join(prolific) %>% 
  group_by(LOCALITY.ID) %>% 
  filter(COMMON.NAME %in% species_prolific$COMMON.NAME) %>% 
  distinct(LOCALITY.ID, COMMON.NAME)


data_prolific <- temp2 %>% 
  group_by(OBSERVER.ID, LOCALITY.ID, SAMPLING.EVENT.IDENTIFIER) %>% 
  # for every list from a person's patch, adding all possible species for the patch
  left_join(temp3) %>% 
  # joining observation count to species, and will replace NAs with 0
  left_join(temp1) %>% 
  mutate(REPORT = replace_na(OBSERVATION.COUNT, "0"),
         OBSERVATION.COUNT = NULL) %>% 
  # converting counts to presences
  mutate(REPORT = as.numeric(case_when(REPORT != "0" ~ "1", TRUE ~ REPORT))) %>% 
  ungroup()


# FILTERS:
# - calculating overall (12 months, 4 years) repfreq for each patch-species combo to 
#   filter out very uncommon species based on threshold.
# - also removing species only "present" (same threshold) in one locality because 
#   cannot use random effect with only one level in model.
#
# - finally, removing locations where after above filters, both sp. categ. do not exist

temp2a <- temp2 %>% 
  group_by(OBSERVER.ID, LOCALITY.ID) %>% 
  summarise(TOT.LISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>% 
  ungroup()

filt <- data_prolific %>% 
  group_by(OBSERVER.ID, LOCALITY.ID, COMMON.NAME) %>% 
  # data_prolific is list-level, so some patch-species combos will be 0 in certain lists
  # interested in species repfreq for each patch, so here removing 0 to count presences
  filter(REPORT != 0) %>% 
  summarise(NO.LISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>% 
  ungroup() %>% 
  right_join(temp2a) %>% 
  mutate(PROP.LISTS = 100*NO.LISTS/TOT.LISTS) %>% 
  # filtering for species present in threshold lists 
  # (RoPi with 11% not converging in max iterations so setting threshold at 15, not 10)
  filter(PROP.LISTS >= 15) %>%
  # removing species present only in one locality
  group_by(COMMON.NAME) %>% 
  mutate(N.LOC = n_distinct(LOCALITY.ID)) %>%
  ungroup() %>% 
  filter(N.LOC > 1) %>% 
  distinct(COMMON.NAME, LOCALITY.ID) %>%
  arrange(COMMON.NAME, LOCALITY.ID) 

data_prolific <- data_prolific %>% 
  # filtering for patch-species combos present in threshold lists at >1 locality
  right_join(filt) %>% 
  arrange(OBSERVER.ID, LOCALITY.ID, SAMPLING.EVENT.IDENTIFIER, COMMON.NAME) %>% 
  # joining list length, year and month metadata
  left_join(data0_MY_b_slice_S %>% distinct(SAMPLING.EVENT.IDENTIFIER, NO.SP, M.YEAR, MONTH)) %>%
  # joining species info
  left_join(species_prolific) %>% 
  # filtering out locations where both species categories do not exist (after species filters)
  group_by(LOCALITY.ID) %>% 
  mutate(N.SPC = n_distinct(SP.CATEGORY)) %>% 
  ungroup() %>% 
  filter(N.SPC == 2) %>% 
  mutate(N.SPC = NULL)
  

### Model run separately for each location, and species variables included in model 
# Lockdown and non-lockdown months modelled separately, since the month variable in
# model otherwise pulled to a very different average (as seen in third panel in graph here)

data_prolific0 <- bind_rows("LD" = data_prolific %>% filter(MONTH %in% 4:5), 
                            "NL" = data_prolific %>% filter(!(MONTH %in% 4:5)), 
                            "ALL" = data_prolific, 
                            .id = "MONTHS.TYPE") %>% 
  # removing Janardhan Uppada's patch in non-lockdown since it is not converging
  filter(!(MONTHS.TYPE == "NL" & LOCALITY.ID == "L6949429"))

# getting median list length (patch-wise) for prediction later
median_length <- data_prolific0 %>% 
  distinct(MONTHS.TYPE, LOCALITY.ID, M.YEAR, MONTH, SAMPLING.EVENT.IDENTIFIER, NO.SP) %>% 
  group_by(MONTHS.TYPE, LOCALITY.ID, MONTH) %>% 
  summarise(NO.SP.MED = floor(median(NO.SP)))

# dataframe with empty column to populate with looped values
prolific <- data_prolific0 %>% 
  group_by(MONTHS.TYPE, LOCALITY.ID) %>% 
  tidyr::expand(nesting(MONTH), M.YEAR, SP.CATEGORY) %>% 
  ungroup() %>% 
  mutate(REP.FREQ.PRED = NA)


count <- 0
for (m in 1:n_distinct(prolific$MONTHS.TYPE)) {
  
  data_mtype <- data_prolific0 %>% 
    filter(MONTHS.TYPE == unique(prolific$MONTHS.TYPE)[m])
  
  prolific0 <- prolific %>% 
    filter(MONTHS.TYPE == unique(prolific$MONTHS.TYPE)[m]) %>% 
    rename(REP.FREQ.PRED2 = REP.FREQ.PRED)
  
  for (l in 1:n_distinct(prolific0$LOCALITY.ID)) {
    
    data_loc <- data_mtype %>% 
      filter(LOCALITY.ID == unique(prolific0$LOCALITY.ID)[l])
    
    model_spec <- glmer(REPORT ~ M.YEAR + MONTH:NO.SP + MONTH:M.YEAR + SP.CATEGORY +
                          (1|COMMON.NAME),
                        data = data_loc, family = binomial(link = "cloglog"),
                        nAGQ = 0, control = glmerControl(optimizer = "bobyqa"))
    
    for (i in 1:n_distinct(prolific0$MONTH)) {
      
      for (j in 1:n_distinct(prolific0$M.YEAR)) {
        
        for (k in 1:n_distinct(prolific0$SP.CATEGORY)) {
          
          count <- count + 1
          
          prolific0$REP.FREQ.PRED2[count] = predict(
            model_spec,
            data.frame(LOCALITY.ID = prolific0$LOCALITY.ID[count],
                       MONTH = prolific0$MONTH[count],
                       M.YEAR = prolific0$M.YEAR[count],
                       NO.SP = median_length$NO.SP.MED[prolific0$MONTH[count]],
                       SP.CATEGORY = prolific0$SP.CATEGORY[count]),
            re.form = NA, 
            type = "response")
          
          print(glue::glue("Months type {m}: Completed predictions for loc. {l}, month {prolific0$MONTH[count]}, migratory year {prolific0$M.YEAR[count]}"))
          
        }
      }
    }
  }
  
  prolific <- prolific %>% 
    left_join(prolific0) %>% 
    mutate(REP.FREQ.PRED = coalesce(REP.FREQ.PRED, REP.FREQ.PRED2)) %>% 
    select(-REP.FREQ.PRED2)

  # resetting counter for next MONTHS.TYPE
  count <- 0
  
}


# for data points
set.seed(345)
prolific_points <- prolific %>% 
  group_by(MONTHS.TYPE, M.YEAR, SP.CATEGORY, LOCALITY.ID) %>% 
  summarise(REP.FREQ.PRED = boot_conf(REP.FREQ.PRED)) %>% 
  group_by(MONTHS.TYPE, M.YEAR, SP.CATEGORY, LOCALITY.ID) %>% 
  summarise(SE = sd(REP.FREQ.PRED),
            REP.FREQ.PRED = mean(REP.FREQ.PRED)) %>% 
  # to make lines connect per location
  group_by(MONTHS.TYPE, LOCALITY.ID, SP.CATEGORY) %>% 
  arrange(MONTHS.TYPE, LOCALITY.ID, SP.CATEGORY, M.YEAR) %>% 
  mutate(PATH.GROUP = glue("{LOCALITY.ID}{SP.CATEGORY}"))

prolific <- prolific_points %>% 
  group_by(MONTHS.TYPE, M.YEAR, SP.CATEGORY) %>% 
  summarise(REP.FREQ.PRED = mean(REP.FREQ.PRED),
            SE = sqrt(sum((SE)^2))/n(),
            CI.L = REP.FREQ.PRED - 1.96*SE,
            CI.U = REP.FREQ.PRED + 1.96*SE)


(ggplot(filter(prolific, MONTHS.TYPE == "LD"), 
        aes(M.YEAR, REP.FREQ.PRED, col = SP.CATEGORY)) +
    scale_color_manual(values = c("#8F85C1", "#A3383C"),
                       name = "Species category",
                       labels = c("Rural", "Urban")) +
    scale_y_continuous(limits = c(0.1, 1.0)) +
    labs(title = "For the months of April and May",
         x = "Migratory year", y = "Predicted reporting frequency") +
    # data points
    geom_point(data = filter(prolific_points, MONTHS.TYPE == "LD"), 
               aes(group = PATH.GROUP),
               size = 3, alpha = 0.25,
               position = position_dodge(0.25)) + 
    geom_path(data = filter(prolific_points, MONTHS.TYPE == "LD"), 
              aes(x = as.numeric(M.YEAR), y = REP.FREQ.PRED, group = PATH.GROUP),
              size = 1, alpha = 0.15, position = position_dodge(0.25)) + 
    # main points
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.5, width = 0.2, position = position_dodge(0.5)) |
  ggplot(filter(prolific, MONTHS.TYPE == "NL"), 
         aes(M.YEAR, REP.FREQ.PRED, col = SP.CATEGORY)) +
    scale_color_manual(values = c("#8F85C1", "#A3383C"),
                       name = "Species category",
                       labels = c("Rural", "Urban")) +
    scale_y_continuous(limits = c(0.1, 1.0)) +
    labs(title = "For other ten months",
         x = "Migratory year", y = "Predicted reporting frequency") +
    # data points
    geom_point(data = filter(prolific_points, MONTHS.TYPE == "NL"), 
               aes(group = PATH.GROUP),
               size = 3, alpha = 0.25,
               position = position_dodge(0.25)) + 
    geom_path(data = filter(prolific_points, MONTHS.TYPE == "NL"), 
              aes(x = as.numeric(M.YEAR), y = REP.FREQ.PRED, group = PATH.GROUP),
              size = 1, alpha = 0.15, position = position_dodge(0.25)) + 
    # main points
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                   size = 1.5, width = 0.2, position = position_dodge(0.5)) |
  ggplot(filter(prolific, MONTHS.TYPE == "ALL"), 
         aes(M.YEAR, REP.FREQ.PRED, col = SP.CATEGORY)) +
    scale_color_manual(values = c("#8F85C1", "#A3383C"),
                       name = "Species category",
                       labels = c("Rural", "Urban")) +
    scale_y_continuous(limits = c(0.1, 1.0)) +
    labs(title = "For all twelve months",
         x = "Migratory year", y = "Predicted reporting frequency") +
    # data points
    geom_point(data = filter(prolific_points, MONTHS.TYPE == "ALL"), 
               aes(group = PATH.GROUP),
               size = 3, alpha = 0.25,
               position = position_dodge(0.25)) + 
    geom_path(data = filter(prolific_points, MONTHS.TYPE == "ALL"), 
              aes(x = as.numeric(M.YEAR), y = REP.FREQ.PRED, group = PATH.GROUP),
              size = 1, alpha = 0.15, position = position_dodge(0.25)) + 
    # main points
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                  size = 1.5, width = 0.2, position = position_dodge(0.5))) +
  plot_layout(guides = "collect") +
  plot_annotation(title = "Change in bird species reporting from three separate monthwise models",
                  subtitle = "Predicted reporting frequencies of common urban and rural species, from locations with consistent effort.") -> b5_prolific

ggsave("03_wrap_figs/b5_prolific.png", b5_prolific,
       width = 21, height = 6, units = "in", dpi = 300)

```

The prolific birder patterns will be able to suggest something about change in detectability vs. new birds in the area.

When models are run for each location and species included in the model, a bit difficult to interpret because of our migratory year classification. Only two months in 2019 (Apr, May) were affected by COVID, so if pattern in MY 2019 is different from that in MY 2020, difficult to determine which to consider as the COVID effect. However, on including a third graph of all ten months other than the COVID months, this becomes easier to ascertain.

Need to run the model separately for the three time periods because the results from both are greatly different. Because we know that the time periods are different in such a way, we should not ignore this and let the single global model average out and skew the existing patterns.

The months of April and May seem to generally (even in MYs 2018 and 2021) be different from other months, with higher frequencies for both rural and urban species. The different patches under consideration all show considerably more variation (inferred from CIs) during these months than other times of the year. There is slightly more variability in rural species than in urban ones. There is also considerable variation between the patches in how reporting changed from year to year. While some increased in MY 2019, others decreased.

In the other ten months of the year, nothing changed with the pandemic at all. In conclusion, it seems that the pandemic-induced changes were restricted to the lockdown months, but also there is no clear directionality in the changes over different locations and regions across the country.

In some patches where reporting frequencies increased in MY 2019 then went back to normal, it could be due to lower noise levels inducing better detection.
