---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Bird behaviour

Two sets of bird species will be selected, one of birds common in human-dominated areas, and the other of common birds that are not very common in human-dominated areas. Reporting frequencies of all species will be calculated for every month and year. If there is an increase during the Mar--May period of birds in the first set and a decrease of birds in the second set, this means that birdwatcher bias has affected bird abundance as seen from the data.

```{r b_setup, cache=TRUE, message=FALSE}





### preparing data for prolific models ####

# species list (all species included in list for other analyses, across all states)
species_prolific <- species_list %>% distinct(COMMON.NAME, SP.CATEGORY)


# to join for presence-absence of various species
temp1 <- data0_MY_b %>% 
  right_join(prolific) %>% 
  group_by(SAMPLING.EVENT.IDENTIFIER, COMMON.NAME) %>% 
  dplyr::summarise(OBSERVATION.COUNT = max(OBSERVATION.COUNT)) %>% 
  ungroup()

# to later join checklist metadata
temp2 <- data0_MY_b_slice_S %>% 
  right_join(prolific) %>% 
  dplyr::select(OBSERVER.ID, LOCALITY.ID, SAMPLING.EVENT.IDENTIFIER)

# join patches with patch-specific species
temp3 <- data0_MY_b %>% 
  right_join(prolific) %>% 
  group_by(LOCALITY.ID) %>% 
  filter(COMMON.NAME %in% species_prolific$COMMON.NAME) %>% 
  distinct(LOCALITY.ID, COMMON.NAME)


data_prolific <- temp2 %>% 
  group_by(OBSERVER.ID, LOCALITY.ID, SAMPLING.EVENT.IDENTIFIER) %>% 
  # for every list from a person's patch, adding all possible species for the patch
  # many-to-many because we are expanding by species AND checklists
  left_join(temp3, relationship = "many-to-many") %>% 
  # joining observation count to species, and will replace NAs with 0
  left_join(temp1) %>% 
  mutate(REPORT = replace_na(OBSERVATION.COUNT, "0"),
         OBSERVATION.COUNT = NULL) %>% 
  # converting counts to presences
  mutate(REPORT = as.numeric(case_when(REPORT != "0" ~ "1", TRUE ~ REPORT))) %>% 
  ungroup()


# FILTERS:
# - calculating overall (12 months, 4 years) repfreq for each patch-species combo to 
#   filter out very uncommon species based on threshold.
# - also removing species only "present" (same threshold) in one locality because 
#   cannot use random effect with only one level in model.
#
# - finally, removing locations where after above filters, both sp. categ. do not exist

temp2a <- temp2 %>% 
  group_by(OBSERVER.ID, LOCALITY.ID) %>% 
  dplyr::summarise(TOT.LISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>% 
  ungroup()

filt <- data_prolific %>% 
  group_by(OBSERVER.ID, LOCALITY.ID, COMMON.NAME) %>% 
  # data_prolific is list-level, so some patch-species combos will be 0 in certain lists
  # interested in species repfreq for each patch, so here removing 0 to count presences
  filter(REPORT != 0) %>% 
  dplyr::summarise(NO.LISTS = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>% 
  ungroup() %>% 
  right_join(temp2a) %>% 
  mutate(PROP.LISTS = 100*NO.LISTS/TOT.LISTS) %>% 
  # filtering for species present in threshold lists 
  # (RoPi with 11% not converging in max iterations so setting threshold at 15, not 10)
  filter(PROP.LISTS >= 15) %>%
  # removing species present only in one locality
  group_by(COMMON.NAME) %>% 
  mutate(N.LOC = n_distinct(LOCALITY.ID)) %>%
  ungroup() %>% 
  filter(N.LOC > 1) %>% 
  distinct(COMMON.NAME, LOCALITY.ID) %>%
  arrange(COMMON.NAME, LOCALITY.ID) 

data_prolific <- data_prolific %>% 
  # filtering for patch-species combos present in threshold lists at >1 locality
  right_join(filt) %>% 
  arrange(OBSERVER.ID, LOCALITY.ID, SAMPLING.EVENT.IDENTIFIER, COMMON.NAME) %>% 
  # joining list length, year and month metadata
  left_join(data0_MY_b_slice_S %>% distinct(SAMPLING.EVENT.IDENTIFIER, NO.SP, M.YEAR, MONTH)) %>%
  # joining species info
  left_join(species_prolific) %>% 
  # filtering out locations where both species categories do not exist (after species filters)
  group_by(LOCALITY.ID) %>% 
  mutate(N.SPC = n_distinct(SP.CATEGORY)) %>% 
  ungroup() %>% 
  filter(N.SPC == 2) %>% 
  mutate(N.SPC = NULL)
  

```

## Bird reporting with standard effort

Out of the most commonly observed species in each of the "prolific" locations, will retain those that are included in species list made for other analysis.

Not modelling for each species separately because small data size.

This is not done for individual states, because focus is individual locations which are anyway spread across the country.

```{r b_prolific_model, cache=TRUE, message=FALSE}

anal_name <- "b_prolific_model"

rm(data0_MY_b, data0_MY_b_slice_G, data0_MY_b_slice_S)
source("00_scripts/b_model_functions.R")

# models ------------------------------------------------------------

# Model run separately for each location, and species variables included in model 

# Lockdown and all months modelled separately, since the month variable in
# model otherwise pulled to a very different average 
data_prolific0 <- bind_rows("LD" = data_prolific %>% filter(MONTH %in% 4:5), 
                            "ALL" = data_prolific, 
                            .id = "MONTHS.TYPE")

# getting median list length (patch-wise) for prediction later
median_length <- data_prolific0 %>% 
  distinct(MONTHS.TYPE, LOCALITY.ID, M.YEAR, MONTH, SAMPLING.EVENT.IDENTIFIER, NO.SP) %>% 
  group_by(MONTHS.TYPE, LOCALITY.ID, MONTH) %>% 
  dplyr::summarise(NO.SP.MED = floor(median(NO.SP)))

# dataframe with empty column to populate with looped values
prolific_pred <- data_prolific0 %>% 
  group_by(MONTHS.TYPE, LOCALITY.ID) %>% 
  tidyr::expand(nesting(MONTH), M.YEAR, SP.CATEGORY) %>% 
  ungroup() %>% 
  # joining median list length
  left_join(median_length) %>% 
  rename(NO.SP = NO.SP.MED) %>% 
  mutate(PRED.LINK = NA,
         SE.LINK = NA,
         ID = row_number()) # for correct join of iters

to_iter <- prolific_pred %>% distinct(MONTHS.TYPE, LOCALITY.ID)
  # # to test with subset of iters
  # %>% 
  # group_by(MONTHS.TYPE) %>% 
  # slice(1:2)


prolific_pred <- map2(to_iter$MONTHS.TYPE, to_iter$LOCALITY.ID, 
                      ~ fit_pred_prolific(.x, .y, 
                                          data_prolific0, prolific_pred)) %>%
  list_rbind() %>%
  # merge back into main prediction table
  right_join(prolific_pred, 
             by = c("MONTHS.TYPE","LOCALITY.ID","MONTH","M.YEAR","SP.CATEGORY", "NO.SP", "ID")) %>%
  mutate(PRED.LINK = coalesce(PRED.LINK, PRED.LINK2),
         SE.LINK   = coalesce(SE.LINK, SE.LINK2)) %>%
  select(-PRED.LINK2, -SE.LINK2)


# for data points

prolific_points <- prolific_pred %>% 
  left_join(timeline) %>% 
  # averaging across months of year
  group_by(MONTHS.TYPE, M.YEAR, SP.CATEGORY, LOCALITY.ID) %>% 
  summarise_mean_and_se(PRED.LINK, SE.LINK, n_is_sim = FALSE) %>% 
  mutate(PRED = clogloglink(PRED.LINK, inverse = T),
         CI.L = clogloglink((PRED.LINK - 1.96*SE.LINK), inverse = T),
         CI.U = clogloglink((PRED.LINK + 1.96*SE.LINK), inverse = T)) %>% 
  mutate(SE = (PRED - CI.L)/1.96)
  
prolific_points <- prolific_points %>% 
  arrange(MONTHS.TYPE, SP.CATEGORY, LOCALITY.ID, M.YEAR) %>%
  # getting trends values of first year (2018--19)
  group_by(MONTHS.TYPE, SP.CATEGORY, LOCALITY.ID) %>% 
  mutate(PRED.LINK.Y1 = first(PRED.LINK),
         PRED.Y1 = first(PRED),
         SE.LINK.Y1 = first(SE.LINK)) %>% 
  ungroup() %>% 
  # for calculating as % change
  #   back-transformed so value is % of year1 value
  mutate(PRED.PERC = 100*PRED/PRED.Y1) 
   
# calculting simulated CIs for the % change
set.seed(10) 
prolific_points <- prolific_points %>% 
  group_by(MONTHS.TYPE, SP.CATEGORY, LOCALITY.ID, M.YEAR) %>% 
  # 1000 simulations of transformed ratio of present:original values
  # quantiles*100 from these gives us our CI limits for PRED.PERC
  reframe(SIM.RATIOS = simerrordiv(PRED.LINK, PRED.LINK.Y1, SE.LINK, SE.LINK.Y1,
                                     state = NULL, species = NULL) %>% 
            pull(rat)) %>% 
  group_by(MONTHS.TYPE, SP.CATEGORY, LOCALITY.ID, M.YEAR) %>% 
  reframe(CI.L.PERC = 100*as.numeric(quantile(SIM.RATIOS, 0.025)),
          CI.U.PERC = 100*as.numeric(quantile(SIM.RATIOS, 0.975))) %>% 
  right_join(prolific_points, by = c("MONTHS.TYPE", "SP.CATEGORY", "LOCALITY.ID", "M.YEAR")) %>%
  group_by(MONTHS.TYPE, SP.CATEGORY, LOCALITY.ID) %>% 
  # making CI band zero for first year
  mutate(CI.L.PERC = case_when(M.YEAR == first(M.YEAR) ~ PRED.PERC,
                               TRUE ~ CI.L.PERC),
         CI.U.PERC = case_when(M.YEAR == first(M.YEAR) ~ PRED.PERC,
                               TRUE ~ CI.U.PERC)) %>%
  ungroup() %>%
  dplyr::select(MONTHS.TYPE, SP.CATEGORY, LOCALITY.ID, M.YEAR, CI.L, PRED, CI.U,
                CI.L.PERC, PRED.PERC, CI.U.PERC) %>% 
    # calculating SE to propagate
    mutate(SE.PERC = (CI.U.PERC - PRED.PERC)/1.96)
  
# to make lines connect per location                      
prolific_points <- prolific_points %>% 
  arrange(MONTHS.TYPE, LOCALITY.ID, SP.CATEGORY, M.YEAR) %>% 
  mutate(PATH.GROUP = glue("{LOCALITY.ID}{SP.CATEGORY}"))


# summarising and propagating SE for species categories (across locations)
prolific_pred <- prolific_points %>% 
  group_by(MONTHS.TYPE, M.YEAR, SP.CATEGORY) %>% 
  summarise_mean_and_se(PRED.PERC, SE.PERC, n_is_sim = FALSE) |> 
  mutate(CI.L.PERC = PRED.PERC - 1.96*SE.PERC,
         CI.U.PERC = PRED.PERC + 1.96*SE.PERC)



# graphs ------------------------------------------------------------

save(prolific_pred, prolific_points, 
     file = glue("00_outputs/bird_models/{anal_name}.RData"))

prolific_plot <- gg_b_model(prolific_pred, type = "prolific", prolific_points)

ggsave(filename = glue("03_wrap_figs/{anal_name}.png"), plot = prolific_plot,
       dpi = 300, width = 21, height = 6, units = "in")

```

The prolific birder patterns will be able to suggest something about change in detectability vs. new birds in the area. When models are run for each location and species included in the model, a bit difficult to interpret because of our migratory year classification. Only two months in 2019 (Apr, May) were affected by COVID, so if pattern in MY 2019 is different from that in MY 2020, difficult to determine which to consider as the COVID effect. However, on including a third graph of all ten months other than the COVID months, this becomes easier to ascertain. Need to run the model separately for the three time periods because the results from both are greatly different. Because we know that the time periods are different in such a way, we should not ignore this and let the single global model average out and skew the existing patterns.

There is no overall pattern or directionality in changes during the pandemic months of years. There is considerable variation between the patches in how reporting changed from year to year, especially during the peak months: while some increased in MY 2019, others decreased. In the other ten months of the year, nothing changed with the pandemic at all. In conclusion, it seems that the pandemic-induced changes were restricted to the lockdown months, but also there is no clear directionality in the changes over different locations and regions across the country.

Speculation: in some patches where reporting frequencies increased in MY 2019 then went back to normal, it could be due to lower noise levels inducing better detection.
