# Abstract {.unnumbered}

Conventional systematic projects to monitor bird populations, like breeding bird surveys, require large and coordinated volunteer networks that are lacking in many parts of the world. Semi-structured citizen science (CS) data offers an alternative to systematic initiatives in these regions, but the semi-structured nature of such data also presents several challenges. The utility of semi-structured data to monitor species abundance is contingent on how, where, and how comparably birdwatchers watch birds, year-on-year. Trends inferred directly from the data can be confounded during years when birdwatchers may behave differently, such as during the COVID pandemic. We wanted to ascertain how the data uploaded to one such CS platform in India, eBird, was impacted by the deadliest global pandemic of the 21st century. To understand whether eBird data from the pandemic years in India is useful and comparable to data from adjacent years, we explored several metrics of data quantity and quality (such as birdwatcher behaviour). We also compared patterns of modelled reporting frequencies for select species between the different years, to examine whether they were predictably different from the previous and next years, due to changes in either birdwatcher behaviour or bird abundance. We found no negative impact of the pandemic on data quantity. Data quality changed largely only during the peak months involving lockdowns, probably due to decreased human mobility and social interaction, but remained similar to the adjacent years during the rest of this restrictive period. We did not find a direct impact on large-scale patterns of bird species reporting, despite high variability at finer scales. We also found no clear evidence of shifts in bird abundance due to the pandemic. Our results show that birdwatchers as data collectors are resilient to disturbance, and that the effects of the pandemic on birdwatching and birds are highly context- and scale-dependent. In summary, eBird data in India from the pandemic years remains useful and interpretable for most large-scale applications, but will benefit from preliminary data quality checks when applied to fine-scale questions.

# Keywords {.unnumbered}

Bird monitoring, birder behaviour, COVID-19 pandemic, data quality, eBird

\newpage

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}

require(tidyverse)
require(patchwork)
require(glue)
require(lubridate)
require(sf)
require(magick)

theme_set(theme_classic(base_size = 16))

covid_palette <- c("#1B9E77", "#EF4050", "#E89005", "#9678B6")
# covid_palette2 <- c("#1B9E77", "#D95F02", "#7570B3", "#555555")
covid_palette3 <- c("#1B9E77", "#E89005", "#9678B6")


# month labels
month_lab <- data.frame(M.YEAR = 2018:2021) %>% 
  group_by(M.YEAR) %>% 
  summarise(MONTH = 1:12 %>% month()) %>% 
  mutate(MONTH.LAB = month(MONTH, label = T)) %>% 
  mutate(MONTH = factor(MONTH, levels = month(c(6:12, 1:5))),
         M.YEAR = factor(M.YEAR),
         MONTH.LAB = factor(MONTH.LAB, levels = month(c(6:12, 1:5), label = T))) %>% 
  mutate(COL1 = ifelse(MONTH.LAB %in% c("Apr", "May"), "red", "black"), # for normal plots
         COL2 = ifelse((M.YEAR %in% 2019:2020) & 
                         (MONTH.LAB %in% c("Apr", "May")), "red", "black"), # for timeline 
         COL3 = ifelse(M.YEAR %in% 2019:2020, "red", "black")) %>% # for bird model plots
  arrange(M.YEAR, MONTH.LAB)

# loading data required for various summaries, tables and figures
load("outputs/data_for_ms.RData")


knitr::opts_chunk$set(
  fig.retina=3,
  echo = FALSE,
  cache = TRUE, 
  include = TRUE,
  message = FALSE, 
  warning = FALSE
)

```

# Introduction

## Birds and citizen science

Citizen science (CS) today is a research technique that engages volunteers from the general public---most of whom are not trained as scientists---in collecting and/or analysing and interpreting data as part of a scientific enquiry [@silvertown2009]. CS programmes have been around for several decades [see @miller-rushing2012] and facilitate attempts at answering large-scale questions that are otherwise not feasible [@macphail2020]. CS projects are particularly effective when they revolve around popular hobbies or activities that people enjoy, such as birdwatching. Birdwatching (hereafter birding) is a nature-based activity that is enjoyed by a large number of people around the world, making CS projects involving birds especially appealing to the public. Birders, as a matter of practice, make "checklists" of the bird species they observe during birding sessions. By volunteering this data, numerous birders---from trained researchers to experienced birders to laypersons---are able to contribute to the scientific study of birds. Many large-scale bird monitoring programmes designed to better understand bird ecology and inform conservation have relied on volunteer birders for data collection [e.g., @greenwood1995; @barrett2003; @sauer2013a; @harris2021; @praveen2022].

Conventional CS bird monitoring programmes that are used to inform conservation are "systematic", in that volunteers follow strict protocols that ensure homogeneity of data in space and time [@newson2005; @norman2012; @sauer2013a; @birdlifeaustralia2015; @harris2021; @swissornithologicalinstitute2023]. Such programmes are resource- and volunteer-intensive, and not feasible at large scales in many parts of the world. CS programmes therefore vary widely in terms of methodological approach and complexity of data collected [@pocock2017]. In fact, a majority of ecological CS projects are simply a by-product of regular hobbyist nature-watching (e.g., birding) with few specified protocols that make the data "semi-structured" [@pocock2017; @dicecco2021]. eBird [@sullivan2014] is one such CS project that gathers and stores semi-structured data from birders all over the world, enabling large-scale robust scientific analysis and output [@kelling2019].

eBird data has proved very useful in better understanding bird ecology, species ranges [e.g., @sreekumar2021; @sinnott2021], migration [e.g., @aoki2021; @youngflesh2021], behaviour [e.g., @winger2019; @alaniz2020], morphology [e.g., @vrettos2021; @menon2022] and conservation action [@sullivan2017; @soib2020; @ruiz-gutierrez2021]. A drawback of typical eBird data, however, is that the checklists vary in "effort", quality, and quantity (no single fixed protocol), creating apparent biases in space and time [@johnston2021]. Largely associated with birder behaviour, such biases make the data difficult to use to inform yearly abundance trends, but some recent analyses show that the biases can be reduced statistically to mimic inference from more systematic data [@johnston2021, Viswanathan *et al.* in press, see @soib2020]. Biases in semi-structured data become especially enhanced when there are large-scale changes in human (birder) behaviour, such as during the COVID-19 pandemic. Given the widespread utility of semi-structured CS data today for conservation, an important question presents itself: has COVID-19 rendered eBird data from the pandemic years unusable in combination with the rest of the dataset?

## Birding in the time of COVID-19

COVID-19 is a deadly infectious disease that resulted in the deadliest pandemic of the 21st century, with a death toll of more than 6.8 million [around 0.08% of the global population, @wikipedia2022; @worldometer2023], alongside others like the bubonic plague, influenza, smallpox and even cholera. Governments across the world had therefore taken various precautionary and preventive measures, and had implemented various restrictions such as nation- or state-wide lockdowns and curfews, quarantines and self-isolation, and shutting down of non-essential public services and activities [see @hale2020]. Given that the eBird enterprise hinges entirely on public participation, this widespread public upheaval might have changed the very nature of eBird data during this period.

The COVID-19 pandemic has generally posed a big challenge for CS programmes [e.g., @kishimoto2021; @hochachka2021]. Social and travel restrictions, in combination with physical and mental distress [@li2020; @vanbavel2020], may have led to a decline in the amount of hobbyist birding, and therefore in the quantity of data generated. Indeed, some CS projects did show evidence of reduced participation and data generation [e.g., @kishimoto2021]. However, there is also a possibility that the pandemic led to increased data generation [e.g., @basile2021]. Outdoor recreational activities such as birding are known to restore attention and play a vital role in improving physical and mental health [@weng2014]. The pandemic brought an increased desire in people to connect with the natural world [@marsh2021], as well as active promotion and encouragement of nature-related activities such as birding [e.g., @birdcountindia2020]. Thus, many turned to birding as their activity of choice [@randler2020].

Regardless of how much more or less data was generated during the pandemic, the characteristics of this data could also have changed. For example, @randler2020 found differences in aspects of birder behaviour, such as who was birding and with whom, as well as where and when birding (and thus data generation) was happening. Restricted mobility has led to the data being less spread out in space, with observations being concentrated in urbanised locations [@vardi2021] and covering a lower overall diversity of habitats [@hochachka2021], thus reducing its informational value [@zhang2020]. On the other hand, @kishimoto2021 found that the restrictions resulted in lower clustering of data, potentially by forcing birders to upload observations from their erstwhile under-surveyed neighbourhoods and not from birding hotspots (urban or rural).

@hochachka2021 studied changes in birder behaviour during the month-long lockdown across different countries, but found that they were not as substantial as expected from the magnitude of change to daily life. They also found high variability in patterns across the four political units they studied, thus emphasising the importance of regional differences and nuances. Most studies exploring the impacts of the pandemic [e.g., @hochachka2021], however, have not assessed the impacts at broader temporal scales of relevance for large-scale analyses of abundance trends, such as for SoIB [-@soib2020]. As such assessments have serious implications in informing conservation policy, there is an urgent need to ensure robustness of the data driving them.

## Our aim

In this study, we sought to understand how eBird data in India was affected by the pandemic, at multiple scales. We asked whether or not data quantity and quality were similar before, during and after the pandemic. We also explored whether patterns of bird species reporting had changed due to differences in either data characteristics or bird abundance and behaviour. Thus, we wished to understand the implications of the pandemic for the utility of eBird data for long-term bird monitoring.

# Methods

## Data preparation

We used the "relMay-2022" version of the publicly available [eBird Basic Dataset (EBD)](https://ebird.org/data/download/ebd) for India, containing all data uploaded until 31 May 2022 [downloaded on 15 June 2022, follows taxonomy in @clements2019]. We combined [sensitive species data](https://ebird.org/india/news/ebird-sensitive-species/), which is not included in the public download, after requesting separately from eBird. Before analysis, we applied a set of filters to standardise the dataset (Section \@ref(dataprep)), so as to prevent any extreme and outlying cases from skewing patterns. Our final eBird dataset contained `r scales::label_comma()(tot_uniq_lists)` unique "qualifying checklists", i.e., lists that passed the filters.

All analyses presented in this study use "migratory year" (MY) as a unit time period (see Section \@ref(dataprep) for rationale) and we analysed data over a period of four migratory years, from 1 June 2018 to 31 May 2022 (Fig. \@ref(fig:timeline)). We consider a relatively fine temporal scale of months in order to account for within-year variations in bird and birder behaviour. Moreover, in India severe COVID-19 restrictions and lockdowns were limited to a few months in 2020 and 2021 [@theeconomictimes2020; @theindianexpress2021], and this scale allows us to consider these patterns separately from patterns in other months of the "pandemic years" (the migratory years 2019 and 2020, coded as "DUR"; 2018 and 2021 coded as "BEF" and "AFT" respectively).

```{r timeline, fig.cap="Timeline of the study period. The spans of the four migratory years are shown in different colours, and pre-pandemic, pandemic and post-pandemic months are also signified. April and May of both pandemic years (shown in red text) are considered as the peak pandemic months.", fig.width=8, out.width="100%", dev="jpeg", dpi=300}

x <- month_lab %>% 
  mutate(YEAR = ifelse(MONTH %in% 6:12, 
                       as.numeric(as.character(M.YEAR)), 
                       as.numeric(as.character(M.YEAR)) + 1)) %>% 
  mutate(DATE = as.Date(paste(YEAR, MONTH, "01", sep = "-")))

y <- x %>%
  summarise(YEAR.RANGE = as.Date(seq(min(DATE), # buffers for placement on graph
                                     max(DATE) + months(1),
                                     by = "year"),
                                   origin = "1970-01-01")) %>% 
  # we don't need label for M.YEAR 2022
  rows_delete(tibble(YEAR.RANGE = as.Date("2022-06-01")))

# for rect
z <- x %>%
  group_by(M.YEAR) %>%
  summarise(DATE = DATE,
            START = min(DATE),
            END = if_else(M.YEAR == 2021, max(DATE), max(DATE) + months(1)))

# for COVID period labels
covid_lab <- read_csv("00_data/covid_classification.csv") %>% 
  mutate(COVID = ifelse(str_detect(COVID, "DUR"), "DUR", COVID),
         MONTH = factor(MONTH)) %>% 
  right_join(x) %>% 
  # selecting only start and end
  group_by(COVID) %>% 
  slice(c(1, n())) %>% 
  ungroup() %>% 
  arrange(DATE) %>% 
  distinct(COVID, DATE) %>% 
  group_by(COVID) %>% 
  mutate(TEMP = c("START", "END")) %>% 
  pivot_wider(names_from = "TEMP", values_from = "DATE") %>% 
  # label
  mutate(LABEL = case_when(COVID == "BEF" ~ "Pre-pandemic",
                           COVID == "DUR" ~ "Pandemic",
                           COVID == "AFT" ~ "Post-pandemic"))

ggplot(x, aes(x = DATE)) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.line.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line.y = element_blank(),
        legend.position = "top",
        legend.margin = margin(t = 0, unit = "cm")) +
  ylim(-0.15, 0.08) +
  # timeline
  geom_hline(yintercept = 0, colour = "black", size = 1.5) +
  # month labels
  geom_text(aes(x = DATE, y = -0.03, label = month(DATE, label = T)),
            angle = 90, colour = x$COL2, size = 3) +
  # M.YEAR categories
  geom_rect(data = z, 
            aes(xmin = START, xmax = END, ymin = 0.025, ymax = 0.075, fill = M.YEAR)) +
  scale_fill_manual(values = covid_palette, name = "Migratory year") +
  geom_segment(data = z, 
               aes(x = START+days(1), xend = START+days(1), y = -0.02, yend = 0.025),
               linetype = "dashed", size = 0.75) +
  # COVID period labels
  geom_bracket(data = covid_lab, 
               aes(xmin = START, xmax = END, label = LABEL),
               y.position = -0.06, vjust = 3, label.size = 4, tip.length = -0.05)


```

We gridded the spatial extent of the country into 25 km × 25 km cells and used the latitude-longitude information associated with each of these qualifying checklists to place them in corresponding grid cells. We also classified the checklists as "urban" or "rural" using remote sensing data on land cover [MODIS Land Cover Type Product MCD12Q1, @friedl2019] with the help of the *luna* [@ghosh2020] package (see Section \@ref(UNUprep) for details).

## Analysis of data quantity

To test whether the pandemic led to an increase or decrease in eBird data quantity, we explored two different aspects of quantity, data generation and participation. We compared the number of qualifying lists and the number of active eBirders for every month across our years of interest, using simple summaries and visualisations.

## Analyses of data quality

To analyse pandemic-induced changes in the quality of eBird data, we devised a suite of metrics (see below) that help gauge various aspects of "data quality". These metrics were calculated per month, across the four years.

### Metrics of birder behaviour

1.  *Group birding* (observer-level): Were there fewer instances of group birding during the pandemic? For this, we calculated the proportion of an observer's total submitted qualifying lists that were instances of group birding, defined as birding events with three or more observers.

2.  *Site fidelity* (observer-level): Given the various restrictions on movement and travel, did birders selectively visit certain areas more often than they previously did? In other words, did birders show higher "site fidelity" during the pandemic? Here, we did not use the conventional definition of a site as a single location, but instead considered each of `r totalcells[1]` 25 km × 25 km grid cells across the country as being a unique "site". We calculated the number of such unique sites visited by each birder.

3.  *Birding time* (observer-level): Did birders spend less time birding during the pandemic? We calculated the total birding time by summing observation durations across all qualifying checklists uploaded by an observer.

4.  *Hotspot birding*: eBird "hotspots" are public birding locations demarcated by eBird users to aggregate and display data from multiple birders [@ebird2022]. Given the overall decline in public/social interactions during the pandemic, did people also tend to avoid birding in eBird hotspots? We calculated, out of total uploaded qualifying lists, the proportion that was from eBird hotspots.

5.  *Birding protocol*: "Traveling" and "Stationary" are the two most commonly followed protocols in eBird observations. Considering the restrictions on movement during the pandemic, we wondered whether birders also avoided making travelling lists, instead preferring stationary lists from their residences. We calculated, out of total uploaded qualifying lists, the proportion that used the "Traveling" protocol.

6.  *Birding distance*: Among the observations that followed the "Traveling" protocol, we asked whether travelling distances were overall shorter during the pandemic. Here, we simply used the distance travelled, measured as part of the checklist-level effort metadata.

7.  *List duration*: Did eBird checklists become shorter in duration during the pandemic? Rather than calculating the overall time spent birding (as in 3), we considered effort at the checklist level and used checklist duration from the effort metadata.

8.  *List length*: List length is defined as the total number of species observed in a checklist. This metric is known to be a better predictor of effort than distance or duration [@szabo2010; @isaac2014, Viswanathan in press], and we used it to answer, from a different angle, the question of whether checklists became shorter during the pandemic.

9.  *Temporal spread*: Did the pandemic bring a change in when birders chose to observe birds? We looked at how checklists were distributed proportionally over two temporal scales: across the seven days of a week, and across the twenty-four hours of a day (although in the latter case our dataset was truncated due to our nocturnal filter).

All the metrics except temporal spread were analysed at two spatial scales: for the whole country and for each of four states. The four states we considered in our study in order to highlight nuances were Karnataka, Kerala, Maharashtra and Assam. These states have relatively high eBirding activity and cover different regions of the country, but also differ in stage and culture of the birding community as well as in response to the pandemic owing to differences in local governance, policy and general public perspective.

### Urban bias

A majority of eBird users may be urban-dwelling, both in India and across the world. With the pandemic limiting movement, did urban areas see an increase in birding while non-urban areas saw the opposite trend? We calculated the proportion of lists in each 25 km × 25 km cell that were from "urban" sites as defined previously. This analysis was done at both spatial scales (see above).

### Spatial coverage

A decrease in overall coverage of the data from the pandemic was a major concern, due to the restrictions on travel. We calculated the proportion of 25 km × 25 km grid cells across the country that were covered in each time period. The information threshold for a grid cell to be considered "covered" was set at 5 qualifying checklists (per month per year). This analysis was done at both spatial scales (see above).

### Spatial spread

Did the pandemic create spatial clusters of positive or negative change in birding activity across grid cells in the country, or were changes in birding (if any) relatively uniformly spread out? For each grid cell, we calculated proportional change in birding activity (number of qualifying lists uploaded) for transition periods ("BEF" to "DUR" and "DUR" to "AFT", where "DUR" was an aggregate of migratory years 2019 and 2020). Here, the COVID time periods (and not months) were our temporal unit of focus, and the analysis was only done for the entire country. See Section \@ref(sspreadmeth) for details on how this metric was calculated and interpreted.

### Statistics {#model-boot}

The analyses, for all metrics except spatial spread, were first explored through visualisations, followed by simple mean and confidence calculations via bootstrapping (see Section \@ref(explboot)), and finally using the formal Generalised Linear Mixed Model framework [GLMMs, see @bolker2009] where feasible. We used the "glmer()" function in the *lme4* package [@bates2015] for the R environment [@rcoreteam2021] to model GLMMs. Observer identity was used as the random effect for the observer-level metrics (group birding, site fidelity, and birding time). Spatial identity defined by grid cell ID was used as the random effect for all other models (Section \@ref(modelspecs-d)). All GLMMs were fit with binomial error distribution and cloglog link function, using BOBYQA optimiser.

Some metrics like observer-level site fidelity, observer-level birding time, birding distance, list duration and list length were either whole number or continuous numerical variables that did not include zero in the range of possible values. For these, since GLMMs with binomial or Poisson error distributions are not appropriate, we instead fitted LMMs (Linear Mixed Models, "lmer()" function in *lme4*) with the same fixed and random effects specifications and using BOBYQA optimiser, after first log-transforming (natural) the response variables to ensure normality. The specifications of the various models of data metrics are listed in Section \@ref(modelspecs-d). There was no model run for temporal spread due to it being a visual metric and its analytical specifications being very different from those of other metrics (Section \@ref(tspreadmeth)).

For all models, model diagnostics were inspected using the *DHARMa* package [@hartig2022], to ensure that there were no major violations of model assumptions. We used the "bootMer()" function in *lme4* in order to test the significance of the fitted model results, by obtaining mean and CI estimates using semi-parametric bootstrapping. The bootstrapping was done with 1000 simulations and with parallel core processing [with the help of packages *parallel*, @rcoreteam2021; and *VGAM*, @yee2010].

## Analysis of bird species reporting

To ascertain whether any of the changes in data quantity and quality during the pandemic impacted bird species reporting, or whether reporting changed at all due to any other factor, we conducted two major analyses involving two sets of regularly and easily observed bird species. One set comprised "urban" species in common human-dominated areas and the other comprised common non-urban or "rural" species that are often found even just adjacent to urban areas. We expected the pandemic-induced urban bias, if present, to also manifest as consistent declines in reporting frequencies across rural species. On the other hand, if urban species became more abundant or detectable during the pandemic, this should correlate with consistent increases in reporting frequencies across urban species.

Since not all species are widespread and equally abundant in different parts of the country, there were a few replacements in the species sets for different regions. Overall, we considered `r n_distinct(species_list$COMMON.NAME)` species in this study, of which `r n_distinct((species_list %>% filter(SP.CATEGORY == "U"))$COMMON.NAME)` were classified as urban and `r n_distinct((species_list %>% filter(SP.CATEGORY == "R"))$COMMON.NAME)` as rural. Details of the region-wise species sets are provided in Table \@ref(tab:reg-spec).

### Species reporting across all locations

In each of the four states analysed in our study, we also selected one of their large metropolitan (urban) cities. For every species in the corresponding regional species set, we visualised individual species reporting patterns at two temporal (months, migratory years) and spatial (city, state) scales. This was done by calculating the reporting frequency (proportion of qualifying checklists in which a species was reported) for each species at each scale.

We then pooled data from all urban and all rural species, and analysed whether overall species reporting patterns at the large scale (migratory year & state) were affected by the pandemic. This gave us an idea of how urban and rural species idiosyncratically differed in terms of general reporting, how species reporting varied across multiple urban areas and how this averaged out over a large spatial scale, and how species reporting varied over the course of a year and how this variation was expressed at the large temporal scale. Kamrup Metropolitan, the metropolitan capital of the state of Assam, did not have sufficient data for visualisation of month-level or year-level patterns, so this combination was dropped.

```{r reg-spec}
# regionwise species sets used in analyses

species_list <- species_list %>% 
  mutate(STATE = factor(STATE, levels = c("Karnataka", "Kerala", "Maharashtra", "Assam")),
         SP.CATEGORY = factor(SP.CATEGORY, levels = c("U", "R"))) %>% 
  distinct(SP.CATEGORY, STATE, COMMON.NAME) %>% 
  arrange(SP.CATEGORY, STATE) %>% 
  mutate(PRESENT = "+") %>% 
  pivot_wider(names_from = "STATE", values_from = "PRESENT") %>% 
  mutate(SP.CATEGORY = ifelse(SP.CATEGORY == "U", "Urban", "Rural")) %>% 
  rename(Species = COMMON.NAME, `Species category` = SP.CATEGORY)

options(knitr.kable.NA = "")

knitr::kable(species_list, 
             align = "lcccc",
             caption = "Region-wise species sets for which overall reporting patterns were analysed. “+” indicates species that were included in the analysis for the different states.")

```

### Species reporting at locations with consistent effort

The quality of CS data is contingent on observer effort; it is possible for changes in observer effort to produce artefactual changes in metrics such as reporting frequency. Given the existence of biases in effort driven by birder behaviour, it then becomes difficult to make inferences from such data. Hence, in order to ensure that the patterns visible in overall bird species reporting at the large scale are not driven by changes in data quantity and quality, we conducted a similar analysis using data only from "prolific" birders having both sufficient and consistent effort (see Section \@ref(prolific-def) for definitions). Since effort and detection rates are relatively unchanged, this analysis allows us to make a reasonable conclusion regarding true changes in bird abundance, which is a combination of the bird's true abundance and its behavioural characteristics like detectability and boldness.

From the set of species analysed in the previous section, we first filtered out species uncommon at each "prolific" birding location, retaining only those with an overall reporting frequency (over the entire study period) of at least 15%. Next, from the set of species that passed the first filter, we removed those that appeared in only one location (i.e., was observed by only one prolific birder), since a factor with only one level cannot be used as a random effect in a GLMM. Finally, if any prolific birder (their location) was left with species falling in only a single species category (i.e., only urban or only rural) after the filters, they were excluded from the analysis.

### Statistics

We followed the same hierarchy of analyses described in Section \@ref(model-boot). Due to strikingly obvious differences in birding conditions between the lockdown and non-lockdown periods (and due to large differences in results from both approaches), we ran three separate models: one for the months of April and May ("lockdown"), one for the other ten months ("non-lockdown"), and one for all twelve months together ("all"). This was to avoid any disproportionate influence of all other months on the parameter estimates for the "lockdown" period.

For each of these three time periods, overall bird reporting models were run for each species with grid cell ID as the random effect, while models for reporting patterns with consistent effort were run for each highly birded location with species identity as the random effect. The full model specifications are provided in Section \@ref(modelspecs-b). All GLMMs were fit with binomial error distribution and cloglog link function, using the BOBYQA optimiser, except in `r (length(fail_spec_KL) + length(fail_spec_MH))` (`r length(fail_spec_KL)` of Kerala, `r length(fail_spec_MH)` of Maharashtra) of the overall bird reporting models in which the state-species combination could not be modelled successfully due to either step-halving or convergence issues. In such cases, the canonical link function (logit) was used instead.

Diagnostics were not inspected separately for the bird reporting models due to the large number of models fitted. Mean and CI estimates were obtained in the same way as outlined in Section \@ref(model-boot), but with 500 simulations instead of 1000, due to computational limitations.

## Software

All analyses were performed in the R environment [@rcoreteam2021] in RStudio [@positteam2022]. Figures were made using the packages *ggplot2* [@wickham2016] and *patchwork* [@pedersen2020]. In addition, packages *sf* [@pebesma2018], *spdep* [@bivand2022], *raster* [@hijmans2022], *terra* [@hijmans2022a] and *geodata* [@hijmans2022b] for spatial analyses, as well as *lubridate* [@grolemund2011], *glue* [@hester2022] and *tictoc* [@izrailev2021] were used. The source code for the entire analysis, from data preparation to statistical validation of metrics and to producing the final manuscript, is available on the [GitHub repository](https://github.com/rikudoukarthik/covid-ebirding) for the study.

# Results

## Data quantity

Although the number of participating birders and the number of checklists uploaded varied considerably between months, the numbers for each individual month consistently increased across the four-year period (Fig. \@ref(fig:lists-ebirders)), with very few exceptions. One such exception is a lower participation in the first wave of COVID-19 in April of MY 2019 compared to April of MY 2018, but this was still associated with more checklists uploaded. We therefore found no clear evidence for a decrease in data quantity due to the pandemic.

## Data quality

### Birder behaviour

1.  *Group birding* (observer-level): Group birding occurred much less frequently during the peak pandemic months of MY 2019 than before (Fig. \@ref(fig:long-changes)A, Table \@ref(tab:model-summ1)). In the months following the lockdown, this difference in proportions gradually reduced, until group birding decreased again (but to a lesser extent) in the peak months of MY 2020. However, this time around the difference immediately disappeared (Fig. \@ref(fig:long-changes)A). These patterns were largely consistent across different states with minor variations (Fig. \@ref(fig:long-changes-state1)A--\@ref(fig:long-changes-state4)A).

2.  *Site fidelity* (observer-level): Birders during the peak periods of the pandemic visited much fewer sites than usual, instead spending more time in the same places (i.e., higher site fidelity; Fig. \@ref(fig:long-changes)B, Table \@ref(tab:model-summ2)). This change also persisted in the months following the first peak period in MY 2019, but reverted to regular pre-pandemic levels immediately after the peak months of the second year. These patterns were also largely common across different states with minor variations (Fig. \@ref(fig:long-changes-state1)B--\@ref(fig:long-changes-state4)B).

3.  *Birding time* (observer-level): While the average birder spent less overall time birding during the peak months in both pandemic years (Table \@ref(tab:model-summ9)), they soon reverted to their usual levels in the following months (Fig. \@ref(fig:time-cov)A). The time spent birding during some of the winter months was lower during and even after the pandemic, compared to pre-pandemic levels, but this change started even before the pandemic and has continued after (Table \@ref(tab:metric-summ)). Moreover, birders across different states also varied considerably in their response to the pandemic (Fig. \@ref(fig:time-cov-state1)A--\@ref(fig:time-cov-state4)A).

4.  *Hotspot birding*: Birders visited eBird hotspots much less frequently during the peak pandemic months of MY 2019 than before (Fig. \@ref(fig:long-changes)C, Table \@ref(tab:model-summ3)). In the months following the lockdown, this difference in proportions gradually reduced until hotspot birding even returned to "normal", but then decreased again (but to a lesser extent) in the peak months of MY 2020. However, this time around birders recovered soon after the peak months (Fig. \@ref(fig:long-changes)C). These patterns varied slightly across different states (Fig. \@ref(fig:long-changes-state1)C--\@ref(fig:long-changes-state4)C).

5.  *Birding protocol*: Birders made and uploaded considerably fewer travelling lists during the peak pandemic months of MY 2019 than before (Fig. \@ref(fig:long-changes)D, Table \@ref(tab:model-summ4)). In the months following the lockdown, this difference gradually reduced until the proportion of lists with the travelling protocol returned to "normal", but then decreased again (but to a lesser extent) in the peak months of MY 2020. However, this time around birders recovered soon after the peak months (Fig. \@ref(fig:long-changes)D). These patterns were largely consistent across states with minor variations (Fig. \@ref(fig:long-changes-state1)D--\@ref(fig:long-changes-state4)D).

6.  *Birding distance*: Birders covered less distance than previously during the peak periods of MY 2019 and MY 2020 (Fig. \@ref(fig:dist-dur-length)A, Table \@ref(tab:model-summ5)). This change evidently disappeared soon, although birders still seem to be covering less distance overall even after the pandemic has passed (Table \@ref(tab:metric-summ)). These patterns were inconsistent across states (Fig. \@ref(fig:dist-dur-length-state1)A--\@ref(fig:dist-dur-length-state4)A).

7.  *List duration*: The average checklist became shorter in terms of time during the peak periods of MY 2019 and MY 2020 (Fig. \@ref(fig:dist-dur-length)B, Table \@ref(tab:model-summ6)). This change did not last beyond the peak months, although birders still seem to be uploading lower duration lists even after the pandemic has passed (Table \@ref(tab:metric-summ)). These patterns were inconsistent across states (Fig. \@ref(fig:dist-dur-length-state1)B--\@ref(fig:dist-dur-length-state4)B.

8.  *List length*: The average checklist reported fewer species during the peak periods of MY 2019 (Fig. \@ref(fig:dist-dur-length)C, Table \@ref(tab:model-summ7)), but this decrease in list length relative to MY 2018 was evidently also present in certain months before as well as after the pandemic (Table \@ref(tab:metric-summ)). Moreover, unlike birding distance and list duration, this metric stabilised to pre-pandemic levels even in MY 2020. Patterns of list length varied considerably across states (Fig. \@ref(fig:dist-dur-length-state1)C--\@ref(fig:dist-dur-length-state4)C).

```{r long-changes, fig.cap="Graphs of observer-level group birding (A) and site fidelity (B), hotspot birding (C) and travelling protocol birding (D) at the national level. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours, and the peak pandemic months are shown in red text.", fig.width=8, fig.height=12.5, out.width="100%", out.height="100%", dpi=300, dev="jpeg"}

load("outputs/d01_group_po.RData")
d1 <- data0_a %>% left_join(month_lab)
load("outputs/d02_fidelity_po.RData")
d2 <- data0_a %>% left_join(month_lab)
load("outputs/d04_hotspot.RData")
d3 <- data0_a %>% left_join(month_lab)
load("outputs/d05_protocol.RData")
d4 <- data0_a %>% left_join(month_lab)



p1 <- ggplot(d1, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Proportion of\ngroup birding") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p2 <- ggplot(d2, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "No. of sites\nvisited") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p3 <- ggplot(d3, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Proportion of\nhotspot birding") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p4 <- ggplot(d4, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Proportion of\ntravelling birding") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

(p1 / p2 / guide_area() / p3 / p4) +
  plot_layout(guides = "collect", heights = c(3, 3, 0.5, 3, 3)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))


```

### Urban bias

Birding in urban areas was slightly higher during the peak periods of both pandemic years (Table \@ref(tab:model-summ8)), though to a lesser degree in the second (Fig. \@ref(fig:urban)). However, this change was short-lived and returned to regular levels in the immediate months. Birding in urban areas after the pandemic (MY 2021) was generally lower than previously. Urban bias was inconsistent, and was even absent in some states (Fig. \@ref(fig:urban-states)).

```{r urban, fig.cap="Urban bias at the national level: Y-axis shows the proportion of checklists that are from urban areas. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours, and the peak pandemic months are shown in red text.", fig.width=8, out.width="100%", fig.height=5, out.height="100%", dpi=300, dev="jpeg"}

load("outputs/d09a_s_UNU.RData")

ggplot(data0_a %>% left_join(month_lab), 
       aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Proportion of\nurban birding") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))

```

### Spatial coverage

Birders collectively covered slightly fewer grid cells during April of MY 2019 (Fig. \@ref(fig:time-cov)B), but this has not translated to the larger temporal scale of years (Table \@ref(tab:model-summ10)). In fact, coverage has only increased in every successive month since and over the years (Fig. \@ref(fig:time-cov)B). Additionally, the slight decrease at the national level was mirrored only by some states like Kerala, while coverage either stayed the same (e.g., Karnataka) or even kept increasing (e.g., Maharashtra) in others (Fig. \@ref(fig:time-cov-state1)B--\@ref(fig:time-cov-state4)B).

### Spatial spread

There was an overall positive raw change in birding effort across districts in both transition periods ("BEF" to "DUR" and "DUR" to "AFT"; Fig. \@ref(fig:s-spread-figs-raw)A, Fig. \@ref(fig:s-spread-rest1)B). This was true also for grid coverage from "BEF" to "DUR", but there was a slight decline from "DUR" to "AFT" (Fig. \@ref(fig:s-spread-figs-raw)B). Thus, the net change from "DUR" to "AFT" was lower than from "BEF" to "DUR", but was still not negative (Fig. \@ref(fig:s-spread-rest1)D). There were very few major changes, and these were not clustered in space.

Major proportional declines (bright yellow) in both birding effort and grid coverage from "BEF" to "DUR" and from "AFT" to "DUR" were restricted to a few districts in central and northern India, and there was little overlap between the two transitions (Fig. \@ref(fig:s-spread-figs-prop), Fig. \@ref(fig:s-spread-rest3)). There were also similar numbers of districts showing prominent declines in the opposite direction, i.e., from "DUR" to "BEF" and from "DUR" to "AFT". More districts showed major declines when going backwards in time ("DUR" to "BEF" and "AFT" to "DUR") than forwards. Several districts were covered during the pandemic that were not covered before or after.

```{r time-cov, fig.cap="Graphs of observer-level birding time (A), and spatial coverage (B) at the national level. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours, and the peak pandemic months are shown in red text.", fig.width=8, fig.height=9, out.width="100%", dpi=300, dev="jpeg"}

load("outputs/d03_time_po.RData")
d1 <- data0_a %>% left_join(month_lab)
load("outputs/d09b_s_cover.RData")
d2 <- data0_a %>% left_join(month_lab)

p1 <- ggplot(d1, 
             aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Time spent birding\n(mins)") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p2 <- ggplot(d2, 
             aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Proportion of\ncells covered") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(5, 1, 5)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))

```

```{r s-spread-figs-raw, fig.cap="Spatial spread: raw change in number of lists (A) and thresholded percentage grid coverage (B) per district from before to during and from during to after the pandemic. Districts are shaded based on the simple difference in values between the two periods from highly negative (red) to highly positive (blue). Grey districts are those that did not meet the criteria for inclusion in the analysis.", out.width="100%", fig.width=8, fig.height=13, dpi=300, dev="jpeg"}

load("outputs/d09c_s_spread.RData")

d1 <- temp1 %>% left_join(grid_dist_base0)
d2 <- temp5 %>% left_join(grid_dist_base0) %>% filter(METRIC == "GRID.COV.T")

# raw change maps (except gridcov w/o threshold)

p1 <- ggplot(d1) +
  geom_polygon(data = indiamap,
               aes(long, lat, group = group),
               fill = "#ACACAC", colour = "black", size = 0.4) +
  geom_sf(aes(fill = VALUE, geometry = geometry), col ="#ACACAC", size = 0.1) +
  facet_wrap(~ T.LABEL) +
  scale_fill_stepsn(colours = RColorBrewer::brewer.pal(n = 9, name = "RdBu"),
                    breaks = c(-1000, -500, -100, -50, 50, 100, 500, 1000),
                    values = scales::rescale(c(-11000, -1000, -500, -100, -50, 
                                               50, 100, 500, 1000, 11000)), 
                    limits = c(-11000, 11000),
                    na.value = "#ACACAC") +
  labs(x = "Longitude", y = "Latitude",
       fill = "Difference in\nno. of lists")


p2 <- ggplot(d2) +
  geom_polygon(data = indiamap,
               aes(long, lat, group = group),
               fill = "#ACACAC", colour = "black", size = 0.4) +
  geom_sf(aes(fill = VALUE, geometry = geometry), col ="#ACACAC", size = 0.1) +
  facet_wrap(~ T.LABEL) +
  scale_fill_stepsn(colours = RColorBrewer::brewer.pal(n = 9, name = "RdBu"),
                    breaks = c(-40, -25, -10, -1, 1, 10, 25, 40),
                    values = scales::rescale(c(-100, -40, -25, -10, -1, 1, 10, 25, 40, 100)), 
                    limits = c(-100, 100),
                    na.value = "#ACACAC") +
  labs(x = "Longitude", y = "Latitude",
       fill = "Difference in\ngrid coverage")


(p1 / p2) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        legend.key.width = unit(3, "cm"))

```

```{r s-spread-figs-prop, fig.cap="Spatial spread: proportional declines in number of lists per district in forward (top row) and backward (bottom row) transitions. Districts with positive values are shown in grey; declines are shaded based on the magnitude of proportional change from low (dark blue) to high (yellow). White districts are those that did not meet the criteria for inclusion in the analysis.", out.width="100%", fig.width=8, fig.height=13, dpi=300, dev="jpeg"}

load("outputs/d09c_s_spread.RData")

d1 <- temp3 %>% left_join(grid_dist_base0)
d2 <- temp4 %>% left_join(grid_dist_base0) 


p1 <- ggplot(d1) +
  geom_polygon(data = indiamap,
               aes(long, lat, group = group),
               fill = "white", colour = "black", size = 0.4) +
  geom_sf(aes(fill = VALUE, geometry = geometry), col ="#E0E0E0", size = 0.1) +
  facet_wrap(~ T.LABEL) +
  scale_fill_steps2(low = viridisLite::cividis(5, direction = -1)[1],
                    mid = viridisLite::cividis(5, direction = -1)[3],
                    high = viridisLite::cividis(5, direction = -1)[5],
                    midpoint = -0.45,
                    breaks = c(-1, -0.9, -0.65, -0.45, -0.25, -0.1, -0),
                    limits = c(-1, 0),
                    na.value = "#E0E0E0") +
  labs(x = "Longitude", y = "Latitude",
       fill = "Proportional change")

p2 <- ggplot(d2) +
  geom_polygon(data = indiamap,
               aes(long, lat, group = group),
               fill = "white", colour = "black", size = 0.4) +
  geom_sf(aes(fill = VALUE, geometry = geometry), col ="#E0E0E0", size = 0.1) +
  facet_wrap(~ T.LABEL) +
  scale_fill_steps2(low = viridisLite::cividis(5, direction = -1)[1],
                    mid = viridisLite::cividis(5, direction = -1)[3],
                    high = viridisLite::cividis(5, direction = -1)[5],
                    midpoint = -0.45,
                    breaks = c(-1, -0.9, -0.65, -0.45, -0.25, -0.1, -0),
                    limits = c(-1, 0),
                    na.value = "#E0E0E0") +
  labs(x = "Longitude", y = "Latitude",
       fill = "Proportional change")


(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(6, 0.5, 6)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        legend.key.width = unit(2, "cm"),
        legend.margin = margin(t = 0, unit = "cm"))

```

## Bird species reporting

We explored patterns of overall bird species reporting (across all locations) for the states of Karnataka, Kerala, Maharashtra and Assam, for 23 species (11 urban, 12 rural), 21 species (11, 10), 22 species (11, 11) and 19 species (10, 9) respectively. The analysis of reporting patterns with consistent effort considered 20 species (11, 9) common to all `r tot_qual_prolific$LOCATIONS` qualifying locations across the country. In general, urban species are reported more frequently than rural species.

### Species reporting across all locations

During the peak periods of the pandemic, urban species were reported more frequently than usual in all states (Fig. \@ref(fig:overall-212m)A1--A4), but the states varied in how rural species reporting changed. In Karnataka, Kerala and Assam, the increase in urban species reporting was accompanied by an increase in rural species reporting, though to a slightly lesser extent in Karnataka and Assam. In Karnataka, unlike in Kerala and Assam, rural species reporting was higher in MY 2020 than in MY 2019. Maharashtra was the only state where rural species were reported less frequently than usual during the peak pandemic months.

Year-round frequencies in Karnataka and Kerala of both urban and rural species were also marginally higher during the pandemic year MY 2019 (Fig. \@ref(fig:overall-212m)B1,B2). However, from MY 2020 in Karnataka, rural species have been reported more frequently than previously. There was an increase in year-round frequencies of both urban and rural species in Maharashtra in the years since MY 2018, though the increase in rural species is slightly lower than in urban species (Fig. \@ref(fig:overall-212m)B3). The pandemic years saw an increase in urban species reporting in Assam, while rural species were reported as frequently as they previously were (Fig. \@ref(fig:overall-212m)B4).

Different species within the same category sometimes showed contradictory patterns (e.g., Fig. \@ref(fig:overall-non8)). Often, species exhibited patterns contradictory to the expectations of pandemic effects: urban species reporting decreased (e.g., Fig. \@ref(fig:overall-non4)A) or rural species reporting increased (e.g., Fig. \@ref(fig:overall-non8)B). Not only did these patterns of species reporting vary considerably from months to years, between states, and across species, they also became weaker (effect sizes decreased).

```{r metric-summ}
# table of metrics and their overall patterns: + or - (as in BMS2022 poster)

data.frame(Metrics = c("Group birding", "Site fidelity", "Birding time",
                       "Hotspot birding", "Birding protocol", 
                       "Birding distance", "List duration", "List length", 
                       "Urban bias", "Spatial coverage",
                       "Overall bird reporting", "Bird reporting with consistent effort"),
           Peak = c("+", "+", "+", 
                    "+", "+",
                    "+", "+", "+",
                    "+", "",
                    "+", ""),
           Pandemic = c("+", "+", "+",
                        "+", "+",
                        "+", "+", "+",
                        "", "",
                        "", ""),
           Post = c("", "", "+",
                    "", "",
                    "+", "+", "+",
                    "", "",
                    "", "")) %>% 
  arrange(desc(Peak), desc(Pandemic), desc(Post)) %>% 
  knitr::kable(caption = 'Summary of changes to metrics of data quality and bird species reporting at the national level, at various temporal scales: column "Peak" refers to the peak pandemic months; "Pandemic" refers to the pandemic years; "Post-pandemic" denotes changes that lasted even after the pandemic years. "+" denotes the presence of a prominent change (positive or negative) relative to pre-pandemic levels.',
               align = "lccc",
               col.names = c("Metrics", "Peak", "Pandemic", "Post-pandemic"))

```

### Species reporting at locations with consistent effort

With consistent effort, species reporting patterns show a different picture at the small scale: neither urban nor rural species were reported more or less frequently during the peak periods than before, as the highly variable local patterns averaged out across space (Fig. \@ref(fig:prolific), Table \@ref(tab:metric-summ)). This remains true for year-round frequencies as well.

```{r overall-212m, fig.cap="Modelled species reporting frequencies for the months of April and May (top row) and separately for all twelve months of a year (bottom row), for the states of Karnataka (A1, B1), Kerala (A2, B2), Maharashtra (A3, B3) and Assam (A4, B4). Points represent means while error bars show 95% confidence intervals. Values for urban and rural species are shown in maroon and purple respectively, and the pandemic years are shown in red text.", out.width="100%", fig.width=10, fig.height=13, dpi=300, dev="jpeg"}

load("outputs/b01_overall_s1c_model.RData")
d1 <- birds_pred %>% filter(MONTHS.TYPE == "LD")
d5 <- birds_pred %>% filter(MONTHS.TYPE == "ALL")
load("outputs/b01_overall_s2c_model.RData")
d2 <- birds_pred %>% filter(MONTHS.TYPE == "LD")
d6 <- birds_pred %>% filter(MONTHS.TYPE == "ALL")
load("outputs/b01_overall_s3c_model.RData")
d3 <- birds_pred %>% filter(MONTHS.TYPE == "LD")
d7 <- birds_pred %>% filter(MONTHS.TYPE == "ALL")
load("outputs/b01_overall_s4c_model.RData")
d4 <- birds_pred %>% filter(MONTHS.TYPE == "LD") 
d8 <- birds_pred %>% filter(MONTHS.TYPE == "ALL")

p1 <- ggplot(d1, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(values = c("#8F85C1", "#A3383C"),
                     name = "Species category",
                     labels = c("Rural", "Urban")) +
  scale_y_continuous(limits = c(0.13, 0.35)) +
  labs(x = "", y = "Reporting frequency") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) 


p2 <- ggplot(d2, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(values = c("#8F85C1", "#A3383C"),
                     name = "Species category",
                     labels = c("Rural", "Urban")) +
  scale_y_continuous(limits = c(0.05, 0.22)) +
  labs(x = "Migratory year", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) 

p3 <- ggplot(d3, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(values = c("#8F85C1", "#A3383C"),
                     name = "Species category",
                     labels = c("Rural", "Urban")) +
  scale_y_continuous(limits = c(0.175, 0.33)) +
  labs(x = "", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) 

p4 <- ggplot(d4, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(values = c("#8F85C1", "#A3383C"),
                     name = "Species category",
                     labels = c("Rural", "Urban")) +
  scale_y_continuous(limits = c(0.19, 0.5)) +
  labs(x = "Migratory year", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) 


p5 <- ggplot(d5, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(values = c("#8F85C1", "#A3383C"),
                     name = "Species category",
                     labels = c("Rural", "Urban")) +
  scale_y_continuous(limits = c(0.13, 0.35)) +
  labs(x = "", y = "Reporting frequency") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) 


p6 <- ggplot(d6, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(values = c("#8F85C1", "#A3383C"),
                     name = "Species category",
                     labels = c("Rural", "Urban")) +
  scale_y_continuous(limits = c(0.05, 0.22)) +
  labs(x = "Migratory year", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) 

p7 <- ggplot(d7, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(values = c("#8F85C1", "#A3383C"),
                     name = "Species category",
                     labels = c("Rural", "Urban")) +
  scale_y_continuous(limits = c(0.175, 0.33)) +
  labs(x = "", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) 

p8 <- ggplot(d8, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(values = c("#8F85C1", "#A3383C"),
                     name = "Species category",
                     labels = c("Rural", "Urban")) +
  scale_y_continuous(limits = c(0.19, 0.5)) +
  labs(x = "Migratory year", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) 

pA <- (p1 | p2 | p3 | p4) + plot_layout(tag_level = "new")
pB <- (p5 | p6 | p7 | p8) + plot_layout(tag_level = "new")

(pA / guide_area() / pB) +
  plot_layout(guides = "collect", heights = c(6, 1, 6)) +
  plot_annotation(tag_levels = c("A", "1")) &
  theme(legend.position = "bottom",
        plot.tag.position = c(0, 1),
        axis.text.x = element_text(colour = (month_lab %>% distinct(M.YEAR, COL3))$COL3,
                                   angle = 45, vjust = 0.5))


```

```{r prolific, fig.cap="Species reporting frequencies with consistent effort data, separately modelled for the months of April and May (A), the other ten months (B), and all twelve months of a year (C). Values for urban and rural species are shown in maroon and purple respectively, and the pandemic years are shown in red text. Each translucent line connecting points represents reporting frequencies from a single location/patch across the four years. Points with error bars represent means with 95% confidence intervals.", out.width="100%", fig.width=8, fig.height=6, dpi=300, dev="jpeg"}

load("outputs/b02_prolific_model.RData")

d1a <- prolific_pred %>% filter(MONTHS.TYPE == "LD")
d1b <- prolific_points %>% filter(MONTHS.TYPE == "LD")

d2a <- prolific_pred %>% filter(MONTHS.TYPE == "NL")
d2b <- prolific_points %>% filter(MONTHS.TYPE == "NL")

d3a <- prolific_pred %>% filter(MONTHS.TYPE == "ALL")
d3b <- prolific_points %>% filter(MONTHS.TYPE == "ALL")


p1 <- ggplot(d1a, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
                     scale_color_manual(values = c("#8F85C1", "#A3383C"),
                                        name = "Species category",
                                        labels = c("Rural", "Urban")) +
                     scale_y_continuous(limits = c(0.1, 1.0)) +
                     labs(x = "", y = "Reporting frequency") +
                     # data points
                     geom_point(data = d1b, aes(group = PATH.GROUP),
                                size = 2, alpha = 0.25,
                                position = position_dodge(0.25)) + 
                     geom_path(data = d1b, aes(x = as.numeric(M.YEAR), y = PRED, group = PATH.GROUP),
                               size = 1, alpha = 0.15, position = position_dodge(0.25)) + 
                     # main points
                     geom_point(size = 2) +
                     geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                                   size = 1, width = 0.3)
       
p2 <- ggplot(d2a, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
                     scale_color_manual(values = c("#8F85C1", "#A3383C"),
                                        name = "Species category",
                                        labels = c("Rural", "Urban")) +
                     scale_y_continuous(limits = c(0.1, 1.0)) +
                     labs(x = "Migratory year", y = "") +
                     # data points
                     geom_point(data = d2b, aes(group = PATH.GROUP),
                                size = 2, alpha = 0.25,
                                position = position_dodge(0.25)) + 
                     geom_path(data = d2b, aes(x = as.numeric(M.YEAR), y = PRED, group = PATH.GROUP),
                               size = 1, alpha = 0.15, position = position_dodge(0.25)) + 
                     # main points
                     geom_point(size = 2) +
                     geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                                   size = 1, width = 0.3)

p3 <- ggplot(d3a, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
                     scale_color_manual(values = c("#8F85C1", "#A3383C"),
                                        name = "Species category",
                                        labels = c("Rural", "Urban")) +
                     scale_y_continuous(limits = c(0.1, 1.0)) +
                     labs(x = "", y = "") +
                     # data points
                     geom_point(data = d1b, aes(group = PATH.GROUP),
                                size = 2, alpha = 0.25,
                                position = position_dodge(0.25)) + 
                     geom_path(data = d1b, aes(x = as.numeric(M.YEAR), y = PRED, group = PATH.GROUP),
                               size = 1, alpha = 0.15, position = position_dodge(0.25)) + 
                     # main points
                     geom_point(size = 2) +
                     geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                                   size = 1, width = 0.3)

(p1 | p2 | p3) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        plot.tag.position = c(0, 1),
        axis.text.x = element_text(colour = (month_lab %>% distinct(M.YEAR, COL3))$COL3,
                                   angle = 45, vjust = 0.5))


```

# Discussion

## Did data quantity increase?

Despite the various restrictions on movement and social interaction during the strict lockdowns, we found a consistent increase in overall data quantity over the years, in terms of both participation and data generation (Fig. \@ref(fig:lists-ebirders)). This is largely consistent with results from elsewhere in CS programmes such as eBird [@hochachka2021] and iNaturalist [@crimmins2021], but we found no evidence for this large-scale increase in data quantity being directly linked with the pandemic. Instead, our results suggest a consistent and long-term growth of eBird data quantity in India over the years.

## Did birders avoid travel, groups, and public spaces?

During the peak lockdown periods in India, birders behaved differently and in accordance with the guidelines to prevent and minimise the spread of the disease. Birders became more isolated (Fig. \@ref(fig:long-changes)A), and more sedentary by limiting travel (Fig. \@ref(fig:long-changes)B) and moving less (Fig. \@ref(fig:long-changes)C,D), during the first peak lockdown period. In the following months, as restrictions were gradually lifted, birder behaviour also changed slowly back to normal, but only until the peak months of the second pandemic year. In this second lockdown period, birders became isolated to a slightly lesser extent, but recovered soon despite this "wave" being more devastating in terms of COVID-19 cases [@worldometer2023]. The staggered recovery in the first year and the immediate recovery in the second suggest that the novelty of the disease played an important role in changing birder behaviour.

## Did birding effort decrease?

Checklist-level "effort" decreased during the lockdown months, with lists becoming shorter overall (Fig. \@ref(fig:dist-dur-length)), but recovered quickly. The specifics of the changes varied considerably depending on the specific area and scale under consideration (Fig. \@ref(fig:dist-dur-length-state1)--\@ref(fig:dist-dur-length-state4)). While there is no clear evidence for long-lasting impacts of the pandemic in this regard, checklist-level "effort" has been consistently lower since MY 2018 (starting even before the pandemic and continuing after; Table \@ref(tab:metric-summ)). This suggests a gradual evolution in birding culture towards shorter birding sessions. Of the three metrics, list length seems to be more resilient to such perturbations and general changes over time, compared to birding distance and list duration.

This shift towards shorter birding sessions has seemingly also coincided with people spending less overall time birding (lower overall "effort"; Table \@ref(tab:metric-summ)), especially during the winter months after the pandemic (Fig. \@ref(fig:time-cov)A). However, this could be due to the general growth in participation (Fig. \@ref(fig:lists-ebirders)), with many new birders pulling down the average birding time.

Although the lower mobility of birders caused a slight decline in spatial coverage in April of MY 2019 (Fig. \@ref(fig:time-cov)B), this quickly reverted and did not translate to larger or longer-term declines (Fig. \@ref(fig:s-spread-rest3)). In fact, overall spatial coverage has only been increasing since, owing to the birding community in India being in the exponential phase of growth. Importantly, none of the fine-scale changes in effort were concentrated in any notable way in space, i.e., the very few worrisome changes were not located all in certain parts of the country (Fig. \@ref(fig:s-spread-figs-prop), Fig. \@ref(fig:s-spread-rest3)).

## Did birders report more urban and fewer rural species?

Since a large majority of the eBirding community is urban, birders becoming more sedentary also meant a clear bias towards urban areas in the data generated during the lockdown periods (Fig. \@ref(fig:urban)). However, the bias was short-lived (Table \@ref(tab:metric-summ)) and was inconsistent across space (Fig. \@ref(fig:urban-states)), and it therefore did not result in larger or longer-term changes to data quality relative to pre-COVID standards. In the post-pandemic period, birding in urban areas was lower than in previous years. While this could be due to birders overcompensating for their restricted travel during the pandemic, the consistent year-on-year decline in urban birding in the non-pandemic months suggests that this is rather yet another aspect of birding in India that has changed over the years with its growth.

Nevertheless, if this fine-scale urban bias also resulted in complementary changes in species reporting patterns, this would impact the utility of data from these years. Indeed, in some states, birders reported urban species slightly more frequently during the pandemic years than before, but this change remained even after the pandemic, suggesting some other cause for the change (Fig. \@ref(fig:overall-212m)B3). In other states, there was no evidence of higher urban species reporting during the pandemic. Thus, the evidence of directional change was not consistent and cannot be clearly linked to the pandemic.

Importantly, we found no clear, consistent and directional evidence for birders reporting rural species less frequently (Fig. \@ref(fig:overall-212m)). Moreover, the changes in urban and rural species reporting across all locations are absent when birding effort is accounted for (Fig. \@ref(fig:prolific)), suggesting that these changes are driven simply by the differential birding in urban and rural areas, and not by inherent changes in bird abundance or behaviour. The highly variable species-level changes in reporting, caused by differential sampling and spatial bias, can be minimised using spatial subsampling methods.

## Did bird abundance change?

Many studies now suggest that survival and even behavioural traits (e.g., song amplitude and frequency, time of vocal activity, etc.) of birds are affected by human activity, potentially leading to an increase in bird abundance or detectability when human activity is reduced. Indeed, during the period of most stringent COVID-19 restrictions in USA, although a few species showed a decline, most increased in abundance in areas that saw a decrease in human activity [@schrimpf2021]. In contrast, species tightly linked to human presence and activity like the Feral Pigeon showed lower abundances [@soh2021]. Behavioural traits of birds can also change rapidly in response to environmental changes [@derryberry2020; @sanderfoot2022; @hentati-sundberg2021; @gordo2021; but see @seress2021].

However, we found no clear evidence for bird abundance changing during the pandemic (Fig. \@ref(fig:prolific)). This might be because a few months of reduced human activity is insufficient to cause changes in large-scale patterns of bird behaviour or occurrence [see @seress2021]. A few individual locations did separately show signs of increased or decreased reporting of urban or rural species during the peak pandemic months, but any pattern disappeared when averaged across space (Table \@ref(tab:metric-summ)). Moreover, any location-specific pattern became muted when considering entire years, or when excluding the peak pandemic months. In conclusion, there is no consistent evidence of birds being affected by the pandemic, positively or negatively, artefactually or otherwise.

## What happened during the peak pandemic months?

Most changes to data quality (mainly birder behaviour) were concentrated in the months of April and May of MY 2019 and MY 2020. These months involved some of the most stringent restrictions and government regulations (e.g., nationwide lockdown) in order to curb the spread of the pandemic [@bbcnews2020; @gettleman2020; @theeconomictimes2020]. The changes in birder behaviour were more drastic and long-lasting in MY 2019 than in MY 2020, possibly due to the novelty and shock of the experience in the first pandemic year, even though the second was even more devastating in terms of casualties [@worldometer2023].

Soon after the restrictions were lifted, birders returned to their usual behaviours (Table \@ref(tab:metric-summ)), pointing to their resilience. As a result, regardless of whether or not there was a change during the peak pandemic months, all metrics of birder behaviour, spatial coverage, spatial spread and urban bias remained unaffected at the yearly level (see coefficients for years in Tables \@ref(tab:model-summ1)--\@ref(tab:model-summ10)). Additionally, the changes that the pandemic brought to human life did not lead to a prominent change at the large scale in how different bird species were reported.

## How useful is eBird data from the pandemic?

In the unprecedented and disruptive times of the COVID-19 pandemic, conventional monitoring programmes---which are especially contingent on human circumstances---were brought to a halt, and their pitfalls became obvious. However, many of these shortcomings do not apply as strongly to (and hence can be overcome by using) large-scale, crowdsourced, open access data, such as from CS programmes [@bates2020], many of which have continued to operate. Moreover, data from the pandemic, arriving amidst this latest period of intense and extremely rapid expansion of human activity (and thereby modification of the environment), also offers an invaluable opportunity to study questions relating to the influence of anthropogenic factors on organisms [@rutz2020; @bates2020].

There is a general paucity of studies looking at the effects of the pandemic on CS data characteristics, as most focus on changes in the organisms instead, even if using data from CS programmes [e.g., @warrington2022]. We expected changes to the quantity and quality of eBird data during the pandemic due to behavioural changes in the birding community caused directly by the various restrictions and indirectly by lifestyle changes. What we found is a steady increase in data quantity, and certain expected changes to data quality. Since the quality of eBird data from India during the pandemic does not seem different enough, and there is no evidence of it having affected bird species reporting, it remains immensely valuable. In fact, our results highlight certain key points about using this data and making inferences from it.

### Importance of scale

@hochachka2021 emphasised spatial scale considerations by showing the presence of high degrees of regional variability in the effects of the pandemic. However, they only considered a single month across years, and so failed to capture variability across temporal scales. Our findings highlight the importance of both spatial and temporal scales in considering the effects of a major event like the COVID-19 pandemic.

At the small scale of months and states (or even smaller like districts), there were clear and expected changes in birder behaviour and, consequently, to data quality. There was also high variability across states. Most of these changes reverted to "normal" levels after the lockdown months, or averaged out across different states, hence resulting in very few notable nationwide changes at the scale of years. Most analyses of long-term abundance trends, distributions or migratory patterns using eBird data are carried out at such large scales. Moreover, the individual fine-scale variations in metrics of data quality were spread throughout the country and not concentrated in one or a few specific regions.

### Importance of effort

A detailed understanding of how birding effort in India has changed over the years requires a longer-term analysis of metrics of effort, starting from the inception of eBird in India. While this is beyond the scope of our study, we have found that such changes do indeed exist. These were neither caused by nor restricted to the COVID-19 pandemic, but instead represent a consistent directional change as Indian birding evolved, that is likely to stay.

Therefore, it is important to account for this variability in effort in any analysis spanning multiple years. Viswanathan (in press) showed that within individual geographies where species detections and accumulation can be assumed to be relatively constant, list length has a tighter relationship with the data generation process than distance and duration. Our results suggest that list length is also more resilient to sudden perturbations as well as year-to-year variability.

### Guidelines for using pandemic eBird data from India

Our findings suggest that there is still tremendous potential for CS data like eBird, even during the pandemic. When formulating a question to be answered using this data, the researcher must first consider the scale (spatial and temporal) of the question. Questions of larger scale can be tackled directly, whereas small-scale questions will benefit from a deep-dive into artefacts in the raw data that can produce misleading patterns. In the latter cases, the researcher is recommended to conduct similar analyses to the ones in this study in order to first assess the suitability of the data before attempting to answer the question. Variability in sampling effort is likely to be an issue at smaller scales, but its effect can be minimised by including effort variables such as list length as well as by using methods of spatial subsampling in the analyses.

# Conclusion

We found that birders did behave differently during the pandemic, but its consequences for eBird data quality were lower than expected from the impact the pandemic has had on human life. Interestingly, our results suggest that official restrictions introduced during the peak periods of the pandemic were more instrumental in changing birder behaviour, while changes in lifestyle were not drastic or long-lasting enough to alter behaviour in the long run. This resilience of birders helped maintain similar data quality levels at large scales, and also led to a steady increase in data quantity. Importantly, there was no clear evidence of bird reporting having changed due to the pandemic, as a consequence of changes in either birder behaviour or bird abundance. Therefore, the eBird dataset remains valuable, with only certain statistical and scale considerations to be kept in mind.

# References

::: {#refs}
:::

\newpage

Supplementary Online Material (SOM) is available.
