# Alternative titles {.unnumbered}

Birding in the time of COVID: Context-dependence of pandemic effects on birdwatching and birds, and importance of scale considerations

Pandemic-induced changes in birdwatcher behaviour outweighed by lack of corresponding decline in data quantity and of directional change in bird responses

# Abstract {.unnumbered}

Conventional systematic projects to monitor bird populations, like breeding bird surveys, require large and coordinated volunteer networks that are lacking in many parts of the world. Semi-structured citizen science (CS) data offers an alternative to systematic initiatives in these regions, but the semi-structured nature of such data also presents several challenges. The utility of semi-structured data to monitor species abundance is contingent on how, where, and how comparably birdwatchers watch birds, year-on-year. Trends inferred directly from the data can be confounded during years when birdwatchers may behave differently, such as during the COVID pandemic. We wanted to ascertain how the data uploaded to one such CS platform in India, eBird, was impacted by the deadliest global pandemic of the 21st century. To understand whether eBird data from the pandemic years in India is useful and comparable to data from adjacent years, we explored several metrics of data quantity and quality (such as birdwatcher behaviour). We also compared patterns of modelled reporting frequencies for select species between the different years, to examine whether they were predictably different from the previous and next years, due to changes in either birdwatcher behaviour or bird abundance. We found no negative impact of the pandemic on data quantity. Data quality changed largely only during the peak months involving lockdowns, probably due to decreased human mobility and social interaction, but remained similar to the adjacent years during the rest of this restrictive period. We did not find a direct impact on large-scale patterns of bird species reporting, despite high variability at finer scales. We also found no clear evidence of shifts in bird abundance due to the pandemic. Our results show that birdwatchers as data collectors are resilient to disturbance, and that the effects of the pandemic on birdwatching and birds are highly context- and scale- dependent. In summary, eBird data in India from the pandemic years remains useful and interpretable for most large-scale applications, but will benefit from preliminary data quality checks when applied to fine-scale questions.

# Keywords {.unnumbered}

Abundance trends, bird monitoring, birder behaviour, COVID-19 pandemic, data quality, eBird (need to select max. 5)

\newpage

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}

require(tidyverse)
require(patchwork)
require(glue)
require(lubridate)
require(sf)
require(magick)

theme_set(theme_classic(base_size = 16))

covid_palette <- c("#1B9E77", "#EF4050", "#E89005", "#9678B6")
# covid_palette2 <- c("#1B9E77", "#D95F02", "#7570B3", "#555555")
covid_palette3 <- c("#1B9E77", "#E89005", "#9678B6")


# month labels
month_lab <- data.frame(M.YEAR = 2018:2021) %>% 
  group_by(M.YEAR) %>% 
  summarise(MONTH = 1:12 %>% month()) %>% 
  mutate(MONTH.LAB = month(MONTH, label = T)) %>% 
  mutate(MONTH = factor(MONTH, levels = month(c(6:12, 1:5))),
         M.YEAR = factor(M.YEAR),
         MONTH.LAB = factor(MONTH.LAB, levels = month(c(6:12, 1:5), label = T))) %>% 
  mutate(COL1 = ifelse(MONTH.LAB %in% c("Apr", "May"), "red", "black"), # for normal plots
         COL2 = ifelse((M.YEAR %in% 2019:2020) & 
                         (MONTH.LAB %in% c("Apr", "May")), "red", "black")) %>%  # for timeline 
  arrange(M.YEAR, MONTH.LAB)

# loading data required for various summaries, tables and figures
load("outputs/data_for_ms.RData")


knitr::opts_chunk$set(
  fig.retina=3,
  echo = FALSE,
  cache = TRUE, 
  include = TRUE,
  message = FALSE, 
  warning = FALSE
)

```

# Introduction

## Birds and citizen science

Citizen science (CS) today is a research technique that engages volunteers from the general public---most of whom are not trained as scientists---in collecting and/or analysing and interpreting data as part of a scientific enquiry [@silvertown2009]. CS programmes have been around for several decades [see @miller-rushing2012] and facilitate attempts at answering large-scale questions that are otherwise not feasible [@macphail2020]. CS projects are particularly effective when they revolve around popular hobbies or activities that people enjoy, such as birdwatching. Birdwatching (hereafter birding) is a nature-based activity that is enjoyed by a large number of people around the world, making CS projects involving birds especially appealing to the public. Birders, as a matter of practice, make "checklists" of the bird species they observe during birding sessions. By volunteering this data, numerous birders---from trained researchers to experienced birders to laypersons---are able to contribute to the scientific study of birds. Many large-scale bird monitoring programmes designed to better understand bird ecology and inform conservation have relied on volunteer birders for data collection [e.g., @greenwood1995; @praveen2022].

Conventional CS bird monitoring programmes that are used to inform conservation are 'systematic', in that volunteers follow strict protocols that ensure homogeneity of data in space and time [@pocock2017]. Such programmes are resource- and volunteer-intensive, and not feasible at large scales in many parts of the world. CS programmes therefore vary widely in terms of methodological approach and complexity of data collected [@pocock2017]. In fact, a majority of ecological CS projects are simply a by-product of regular hobbyist nature-watching (e.g., birding) with few specified protocols that make the data "semi-structured" [@pocock2017; @dicecco2021]. eBird [@sullivan2014] is one such CS project that gathers and stores semi-structured data from birders all over the world, enabling large-scale robust scientific analysis and output [@kelling2019].

eBird data has proved very useful in better understanding bird ecology, species ranges [e.g., @sreekumar2021; @sinnott2021], migration [e.g., @aoki2021; @youngflesh2021], behaviour [e.g., @winger2019; @alaniz2020], morphology [e.g., @vrettos2021; @menon2022] and conservation action [@sullivan2017; @soib2020; @ruiz-gutierrez2021]. A drawback of typical eBird data, however, is that the checklists vary in "effort", quality, and quantity (no single fixed protocol), creating apparent biases in space and time [@johnston2021]. Such biases make the data difficult to use to inform yearly abundance trends, but some recent analyses show that the biases can be reduced statistically to mimic inference from more systematic data [@johnston2021; Viswanathan *et al.* in press; see @soib2020]. Biases in semi-structured data become especially enhanced when there are large-scale changes in human (birder) behaviour, such as during the COVID-19 pandemic. Given the widespread utility of semi-structured CS data today for conservation, an important question presents itself: has COVID-19 rendered eBird data from the pandemic years unusable in combination with the rest of the dataset?

## eBirding in the time of COVID-19

The COVID-19 pandemic has generally posed a big challenge for CS programmes [e.g., @kishimoto2021; @hochachka2021]. COVID-19 is a deadly infectious disease that resulted in a widespread pandemic. It was the deadliest pandemic of the 21st century and one of the deadliest in human history, with a death toll of more than 6.8 million [around 0.08% of the global population, @wikipedia2022; @worldometer2023], alongside others like the bubonic plague, influenza, smallpox and even cholera. Governments across the world had therefore taken various precautionary and preventive measures, and had implemented various restrictions such as nation- or state-wide lockdowns and curfews, quarantines and self-isolation, and shutting down of non-essential public services and activities [see @hale2020]. Given that the eBird enterprise hinges entirely on public participation, this widespread public upheaval might have changed the very nature of eBird data during this period.

The social and travel restrictions, in combination with physical and mental distress [@li2020; @vanbavel2020], may have led to a considerable decline in data quantity. Indeed, some CS projects did show evidence of reduced participation and data generation [e.g., @kishimoto2021]. However, there is also a possibility that the pandemic led to increased data generation [e.g., @basile2021]. Outdoor recreational activities such as birding are known to restore attention and play a vital role in improving physical and mental health [@weng2014]. The pandemic brought an increased desire in people to connect with the natural world [@marsh2021], as well as active promotion and encouragement of nature-related activities such as birding [e.g., @birdcountindia2020]. Thus, many turned to birding as their activity of choice [@randler2020].

Regardless of how much more or less data was generated during the pandemic, the characteristics of this data could also have changed. For example, @randler2020 found differences in aspects of birder behaviour, such as who was birding and with whom, as well as where and when birding (and thus data generation) was happening. Restricted mobility has led to the data being less spread out in space, with observations being concentrated in urbanised locations [@vardi2021] and covering a lower overall diversity of habitats [@hochachka2021], thus reducing its informational value [@zhang2020]. On the other hand, @kishimoto2021 found that the restrictions resulted in lower clustering of data, potentially by forcing birders to upload observations from their erstwhile under-surveyed neighbourhoods and not from birding hotspots (urban or rural).

@hochachka2021 studied changes in birder behaviour during the month-long lockdown across different countries, but found that they were not as substantial as expected from the magnitude of change to daily life. They also found high variability in patterns across the four political units they studied, thus emphasising the importance of regional differences and nuances. Additionally, most studies exploring the impacts of the pandemic [e.g., @hochachka2021] have not assessed the impacts at broader temporal scales of relevance for large-scale analyses of abundance trends.

For instance, Black Eagle, a forest-dwelling species of Least Concern on the IUCN Global Red List, was shown to have a moderate decline in short-term trend [@soib2020a]. However, the presence of pandemic-induced changes (such as increased urban dominance) at the large scale could artificially result in a steeper and more alarming decline in the trend for these years. As such assessments of abundance trends and distributions have serious implications in informing conservation policy, there is an urgent need to ensure robustness of the data driving them.

## Our aim

In this study, we sought to understand how eBird data in India was affected by the pandemic, at multiple scales. We asked whether or not data quantity and quality were similar before, during and after the pandemic. We also explored whether patterns of bird species reporting had changed due to differences in either data characteristics or bird abundance and behaviour. Thus, we wished to understand the implications of the pandemic for the utility of eBird data.

# Methods

## Data preparation

We used the "relMay-2022" version of the publicly available [eBird Basic Dataset (EBD)](https://ebird.org/data/download/ebd) for India, containing all data uploaded until 31 May 2022 [downloaded on 15 June 2022, follows taxonomy in @clements2019]. We combined [sensitive species data](https://ebird.org/india/news/ebird-sensitive-species/), which is not included in the public download, after requesting separately from eBird. Before analysis, we applied a set of filters to standardise the dataset, so as to prevent any extreme and outlying cases from skewing patterns.

First, we filtered the dataset for "complete checklists", in which observers report all the birds they see or hear. Next, we filtered the data for our timeline of interest. The migratory cycle in India starts in June of one calendar year and ends in May of the next, and this time frame is thus deemed more appropriate than a calendar year. Therefore, all analyses presented in this study use "migratory year" (MY) as a unit time period and we analysed data over a period of four migratory years, from 1 June 2018 to 31 May 2022 Fig. \@ref(fig:timeline). We consider a relatively fine temporal scale of months in order to account for within-year variations in bird and birder behaviour. Moreover, in India severe COVID-19 restrictions and lockdowns were limited to a few months in 2020 and 2021 [@theeconomictimes2020; @theindianexpress2021], and this scale allows us to consider these patterns separately from patterns in other months of the "pandemic years" (the migratory years 2019 and 2020, coded as "DUR"; 2018 and 2021 coded as "BEF" and "AFT" respectively).

```{r timeline, fig.cap="Timeline of our study period, showing migratory years colorued by COVID category.", fig.width=8, out.width="100%", dev="jpeg", dpi=300}

x <- month_lab %>% 
  mutate(YEAR = ifelse(MONTH %in% 6:12, 
                       as.numeric(as.character(M.YEAR)), 
                       as.numeric(as.character(M.YEAR)) + 1)) %>% 
  mutate(DATE = as.Date(paste(YEAR, MONTH, "01", sep = "-")))

y <- x %>%
  summarise(YEAR.RANGE = as.Date(seq(min(DATE), # buffers for placement on graph
                                     max(DATE) + months(1),
                                     by = "year"),
                                   origin = "1970-01-01")) %>% 
  # we don't need label for M.YEAR 2022
  rows_delete(tibble(YEAR.RANGE = as.Date("2022-06-01")))

# for rect
z <- x %>%
  group_by(M.YEAR) %>%
  summarise(DATE = DATE,
            START = min(DATE),
            END = if_else(M.YEAR == 2021, max(DATE), max(DATE) + months(1)))

# for COVID period labels
covid_lab <- read_csv("data/covid_classification.csv") %>% 
  mutate(COVID = ifelse(str_detect(COVID, "DUR"), "DUR", COVID),
         MONTH = factor(MONTH)) %>% 
  right_join(x) %>% 
  # selecting only start and end
  group_by(COVID) %>% 
  slice(c(1, n())) %>% 
  ungroup() %>% 
  arrange(DATE) %>% 
  distinct(COVID, DATE) %>% 
  group_by(COVID) %>% 
  mutate(TEMP = c("START", "END")) %>% 
  pivot_wider(names_from = "TEMP", values_from = "DATE") %>% 
  # label
  mutate(LABEL = case_when(COVID == "BEF" ~ "Pre-pandemic",
                           COVID == "DUR" ~ "Pandemic",
                           COVID == "AFT" ~ "Post-pandemic"))

ggplot(x, aes(x = DATE)) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.line.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line.y = element_blank(),
        legend.position = "top",
        legend.margin = margin(t = 0, unit = "cm")) +
  ylim(-0.15, 0.08) +
  # timeline
  geom_hline(yintercept = 0, colour = "black", size = 1.5) +
  # month labels
  geom_text(aes(x = DATE, y = -0.03, label = month(DATE, label = T)),
            angle = 90, colour = x$COL2, size = 3) +
  # M.YEAR categories
  geom_rect(data = z, 
            aes(xmin = START, xmax = END, ymin = 0.025, ymax = 0.075, fill = M.YEAR)) +
  scale_fill_manual(values = covid_palette, name = "Migratory year") +
  geom_segment(data = z, 
               aes(x = START+days(1), xend = START+days(1), y = -0.02, yend = 0.025),
               linetype = "dashed", size = 0.75) +
  # COVID period labels
  geom_bracket(data = covid_lab, 
               aes(xmin = START, xmax = END, label = LABEL),
               y.position = -0.06, vjust = 3, label.size = 4, tip.length = -0.05)


```

We then removed data from group accounts, which are eBird accounts with which multiple existing eBirders share their lists to track aggregate summaries. Such accounts are commonly used during specific events or surveys, and need to be removed from analyses of observer-related metrics. Next, we applied a nocturnal filter to remove observations made between 8pm and 4am (nocturnal birding is very different from diurnal birding), a pelagic filter to remove observations made from beyond the shoreline of the country, and finally a set of filters to remove observations made with extreme values of birding "effort": checklists that are too short (less than three minutes long, or reporting three or fewer species without information on duration, or reporting three or fewer species at a rate lower than two species per hour), and travelling checklists that are too fast (following "Traveling" protocol and travelling faster than 20 kmph or covering a distance greater than 50 km). These lists with extreme effort are not considered to be "truly complete" checklists. The dataset at this stage was used in analyses of observer-level metrics (see details below), while for non-observer-level data metrics and metrics of bird reporting, a final filter was added: in instances of group birding where each participant eBirder owns a copy of the checklist, duplicate records/checklists were removed by filtering out non-unique combinations of group birding event (column "GROUP.IDENTIFIER") and bird species (column "COMMON.NAME"). The checklists that passed these criteria are hereafter referred to as "qualifying checklists". Our final eBird dataset contained `r scales::label_comma()(tot_uniq_lists)` unique qualifying checklists.

We gridded the spatial extent of the country into 25 km × 25 km cells and used the latitude-longitude information associated with each of these qualifying checklists to place them in corresponding grid cells. In addition, we used remote sensing data to divide the country into urban and non-urban (rural) areas. For this, we downloaded the MODIS Land Cover Type Product [MCD12Q1, @friedl2019] that includes several science datasets (SDSs) mapping global land cover at a 500 m resolution. Although this comes at an annual time step, we used data from the single time point, 31 December 2021. The tiled MODIS data was downloaded using the "getModis()" function in the *luna* [@ghosh2020] package and saved as individual .tif files, which were then merged into a single raster. After cropping and masking to the country's administrative boundary shapefile, the MODIS land cover categories in the raster were reclassified: category 13, "urban and built-up lands" was considered "urban", and all other categories were considered "rural" (technically "non-urban"). We didn't use the MODIS data at 500 m × 500 m resolution because the location information in eBird data, despite being precise, is not very accurate at small grain sizes (due to the very nature of eBird data where sampling often occurs over a distance). So, finally we aggregated the raster to a resolution of 2 km × 2 km which was deemed most appropriate, then classified each qualifying checklist in the dataset as urban or rural by extracting the land cover class from the modified MODIS raster, again using the latitude-longitude information.

## Analysis of data quantity

To test whether the pandemic led to an increase or decrease in eBird data quantity, we explored two different aspects of quantity, data generation and participation. We compared number of qualifying lists and number of active eBirders at the monthly level across our years of interest, using simple summaries and visualisations. `states?`

## Analysis of data quality

To analyse pandemic-induced changes in data quality, we devised a suite of metrics (Table \@ref(tab:metrics-d)) that help gauge various aspects of data quality. These metrics were compared at the monthly level across our years of interest, i.e., the metrics were calculated per month, across the four years. Moreover, all the metrics except for spatial spread and temporal spread were analysed at two spatial scales: for the whole country and for each of four states. The four states we considered in our study in order to highlight nuances were Karnataka, Kerala, Maharashtra and Assam. These states have relatively high eBirding activity and cover different regions of the country, but also differ in stage and culture of the birding community as well as in local governance, policy and general public perspective.

### Metrics of birder behaviour

1.  *Group birding* (observer-level): Were there fewer instances of group birding during the pandemic? For this, we calculated the proportion of an observer's total submitted qualifying lists that were instances of group birding, defined as birding events with three or more observers.

2.  *Site fidelity* (observer-level): Given the various restrictions on movement and travel, did birders show higher site fidelity during the pandemic? In other words, were their birding activities restricted more to certain sites than previously? Here, we considered each of `r totalcells[1]` 25 km × 25 km grid cells across the country as being a unique "site", and calculated the number of such unique sites visited by a birder.

3.  *Birding time* (observer-level): Did birders overall spend less time birding during the pandemic? We calculated total birding time by summing observation durations across all qualifying checklists submitted by an observer.

4.  *Hotspot birding*: eBird "hotspots" are public birding locations demarcated by eBird users to aggregate and display data from multiple birders [@ebird2022]. Given the overall decline in public/social interactions during the pandemic, did people also tend to avoid birding in eBird hotspots? We calculated, out of total submitted qualifying lists, the proportion that was from eBird hotspots.

5.  *Birding protocol*: "Traveling" and "Stationary" are the two most commonly followed protocols in eBird observations. Considering the restrictions on movement during the pandemic, we wondered whether birders also avoided making travelling lists, instead preferring stationary lists from their residences. We calculated, out of total submitted qualifying lists, the proportion that used the "Traveling" protocol.

6.  *Birding distance*: Among the observations that followed the "Traveling" protocol, we asked whether travelling distances were overall shorter during the pandemic. Here, we simply used the distance travelled, measured as part of the checklist-level effort metadata.

7.  *List duration*: Did eBird checklists become shorter in duration during the pandemic? Rather than calculating the overall time spent birding (as in 3), we considered effort at the checklist level and used checklist duration from the effort metadata.

8.  *List length*: List length is defined as the total number of species observed in a checklist. This metric is known to be a better predictor of effort than distance or duration (Viswanathan in press), and we used it to answer, from a different angle, the question of whether checklists became shorter during the pandemic.

9.  *Temporal spread*: Did the pandemic bring a change in when birders chose to observe birds? We looked at how checklists were distributed proportionally over two temporal scales: across the seven days of a week, and across the twenty-four hours of a day (although in the latter case our dataset was truncated due to our nocturnal filter).

### Urban bias

A majority of eBird users in general are urban-dwelling. With the pandemic limiting people's movements, was birding concentrated much more strongly in urban areas during the pandemic than otherwise? At the 25 km × 25 km grid cell level, we calculated the proportion of lists in each cell that were urban.

### Spatial coverage

A decrease in overall coverage of the data from the pandemic was a major concern, due to the restrictions on travel. We calculated the proportion of 25 km × 25 km grid cells across the country that were covered in each time period. The information threshold for a grid cell to be considered "covered" was set at 5 qualifying checklists (per month per year).

### Spatial spread

Did the pandemic create spatial clusters of positive or negative change in birding activity across grid cells in the country, or were changes in birding (if any) relatively uniformly spread out? For each grid cell, we calculated proportional change in birding activity (number of qualifying lists submitted) for transition periods ("BEF" to "DUR" and "DUR" to "AFT", where "DUR" was an aggregate of migratory years 2019 and 2020). Here, the COVID time periods were our temporal unit of focus, and not months. See Section \@ref(sspreadmeth) for details on how this metric was calculated and interpreted.

### Statistical validation {#model-boot}

For all metrics except spatial spread, unless otherwise mentioned, the above analyses were first done using exploratory visualisations, following which emerging patterns were validated using simple confidence calculations or bootstrap validation in some cases (see Section \@ref(explboot)). Finally, after the metrics and analysis approaches were deemed meaningful, they were tested statistically using Generalised Linear Mixed Models (GLMMs). For this, we used the "glmer()" function in the *lme4* package [@bates2015] For the observer-level metrics (group birding, site fidelity, and birding time), observer identity was used as the random effect, whereas spatial identity defined by grid cell ID was used as the random effect for all other models (Table \@ref(tab:metrics-d)). All GLMMs were fit with binomial error distribution and cloglog link function, using BOBYQA optimiser.

Some metrics like observer-level site fidelity, observer-level birding time, birding distance, list duration and list length were either whole number or continuous numerical variables that did not include zero in the range of possible values. For these, since GLMMs with binomial or Poisson error distributions are not appropriate, we instead fitted LMMs (Linear Mixed Models) with the same fixed and random effects specifications and using BOBYQA optimiser, after first log-transforming (natural) the response variables to ensure normality. This was done using the "lmer()" function in *lme4*. The various models of data metrics and their specifications are listed in Table \@ref(tab:metrics-d). There was no model run for temporal spread due to it being a visual metric and its analytical specifications being very different from those of other metrics (Section \@ref(tspreadmeth)).

For all models, model diagnostics were inspected using the *DHARMa* package [@hartig2022], to ensure that there were no major violations of model assumptions. We used the "bootMer()" function in *lme4* in order to test the significance of the fitted model results, by obtaining mean and CI estimates using semi-parametric bootstrapping. The bootstrapping was done with 1000 simulations and with parallel core processing [with the help of packages *parallel*, @rcoreteam2021; and *VGAM*, @yee2010].

```{r metrics-d}
# a large sideways full-page table of metrics of data quantity and quality, and model specifications
knitr::kable(mtcars[1:5, 1:5], caption = "The various metrics of data quantity and quality and their model specifications.")

```

## Analysis of bird species reporting

To ascertain whether any of the changes in data quantity and quality during the pandemic impacted bird species reporting, or whether reporting changed at all due to any other factor, we conducted two major analyses involving two sets of regularly and easily observed bird species, one set common in human-dominated areas ("urban" species) and the other uncommon in urban areas ("rural" species). We expected the pandemic-induced urban bias, if present, to also manifest as consistent declines in reporting frequencies across rural species. On the other hand, if urban species became more abundant or detectable during the pandemic, this should correlate with consistent increases in reporting frequencies across urban species.

Since not all species are widespread and equally abundant in different parts of the country, there were a few replacements in the species sets for different regions. Overall, we considered `r n_distinct(species_list$COMMON.NAME)` species in this study, of which `r n_distinct((species_list %>% filter(SP.CATEGORY == "U"))$COMMON.NAME)` were classified as urban and `r n_distinct((species_list %>% filter(SP.CATEGORY == "R"))$COMMON.NAME)` as rural. Details of the region-wise species sets are provided in Table \@ref(tab:reg-spec).

### Overall species reporting

In each of the four states analysed in our study, we also selected one of their large metropolitan (urban) cities. For every species in the corresponding regional species set, we visualised individual species reporting patterns at two temporal (months, migratory years) and spatial (city, state) scales. (Analysing species reporting patterns at the national level is not appropriate, considering the large variation in species ranges and abundances.) This was done by calculating the reporting frequency (proportion of qualifying checklists in which a species was reported) of each species at each scale.

We then pooled data from all urban and all rural species, and analysed whether overall species reporting patterns at the large scale (migratory years & state) were affected by the pandemic. This gave us an idea of how urban and rural species idiosyncratically differed in terms of general reporting, how species reporting varied across multiple urban areas and how this averaged out over a large spatial scale, and how species reporting varied over the course of a year and how this variation was expressed at the large temporal scale. Kamrup Metropolitan, the metropolitan capital of the state of Assam, did not have sufficient data for visualisation of month-level patterns, so this combination was dropped.

```{r reg-spec}
# regionwise species sets used in analyses
knitr::kable(mtcars[1:5, 1:5], caption = "The regionwise species sets for which reporting patterns were analysed.")

```

### Species reporting with consistent effort

The quality of CS data is contingent on observer effort; it is possible for changes in observer effort to produce artefactual changes in metrics such as reporting frequency. Given the existence of biases in effort driven by birder behaviour, it then becomes difficult to make inferences from such data. Hence, in order to ensure that the patterns visible in overall bird species reporting at the large scale are not driven by changes in data quantity and quality, we conducted a similar analysis using data only from "prolific" birders having consistent effort. An eBirder was considered "prolific" if they had submitted, from a single location, at least 800 qualifying checklists over the course of the four years being analysed (sufficient effort), as well as a minimum of 15 qualifying checklists per month in at least ten months of a year in all four years of the study (consistent effort)---thus translating to about 200 qualifying checklists per year. Since the observer remains constant, and the observer is sufficiently experienced (as inferred from their prolificacy), we assume that detection rates of birds in the location also remains similar over the years. Therefore, this analysis allows us to make a reasonable conclusion regarding true changes in bird abundance, which is a combination of the bird's true abundance and its behavioural characteristics like detectability and boldness.

From the set of species analysed in the previous section, we determined those to be analysed for each prolific birder by first filtering out uncommon species, retaining only those with an overall reporting frequency (over the entire study period) of at least 15%. Next, from the set of species that passed the first filter, we removed those that appeared in only one location (i.e., was observed by only one prolific birder), since a factor with only one level cannot be used as a random effect in a GLMM. Finally, if any prolific birder (their location) was left with species falling in less than two species categories (i.e., urban and rural) after the filters, they were excluded from the analysis.

### Statistical validation

Like with the metrics of data quantity and quality, the bird reporting patterns were first explored using visualisations, then emerging patterns were validated using simple bootstrapping for confidence estimates as in Section \@ref(explboot), except that the bootstrapped values were calculated separately for each spatial scale and propagation was done from the fine temporal scale of month to the scale of migratory year. Eventually, once the specifics of the analyses were finalised, the patterns were statistically tested using GLMMs. Due to strikingly obvious differences in birding conditions between the lockdown and non-lockdown periods (and due to large differences in results from both approaches), we ran three separate models: one for the months of April and May ("lockdown"), one for the other ten months ("non-lockdown"), and one for all twelve months together ("all"). This also helps determine whether the effects of the pandemic on species reporting were restricted to the strict lockdown period.

For each of these three time periods, overall bird reporting models were run for each species with grid cell ID as the random effect, while models for reporting patterns with consistent effort were run for each highly birded location with species identity as the random effect. The full model specifications are provided in Table \@ref(tab:models-b). All GLMMs were fit with binomial error distribution and cloglog link function, using BOBYQA optimiser, except in a few `r (length(fail_spec_KL) + length(fail_spec_MH))` (`r length(fail_spec_KL)` of Kerala, `r length(fail_spec_MH)` of Maharashtra) of the overall bird reporting models in which the state-species combination could not be modelled successfully due to either step-halving or convergence issues. In such cases, the canonical link function (logit) was used instead.

Diagnostics were not inspected separately for the bird reporting models due to the large number of models fitted. Mean and CI estimates were obtained in the same way as outlined in Section \@ref(model-boot), but with 500 simulations instead of 1000, due to computational limitations.

```{r models-b}
# a large sideways full-page table of model specifications of bird reporting patterns
knitr::kable(mtcars[1:5, 1:5], caption = "Model specifications for analyses of bird reporting patterns.")

```

## Software

All analyses were performed in the R environment [@rcoreteam2021] in RStudio [@positteam2022]. Figures were made using the packages *ggplot2* [@wickham2016] and *patchwork* [@pedersen2020]. In addition, packages *sf* [@pebesma2018], *spdep* [@bivand2022], *raster* [@hijmans2022], *terra* [@hijmans2022a] and *geodata* [@hijmans2022b, for spatial analyses], as well as *lubridate* [@grolemund2011], *glue* and *tictoc* were used. The source code for the entire analysis, from data preparation to statistical validation of metrics, is available on the [GitHub repository](https://github.com/rikudoukarthik/covid-ebirding) for the study.

# Results

```{r metric-summ}
# table of metrics and their overall patterns: + or - (as in BMS2022 poster)
knitr::kable(mtcars[1:5, 1:5], caption = "Summary of pandemic-induced changes on metrics of data quantity and quality.")

```

## Data quantity

`data quantity results`

## Data quality

### Birder behaviour

1.  *Group birding* (observer-level): Group birding occurred much less frequently during the peak pandemic months of MY 2019 than before (Fig. \@ref(fig:long-changes)). In the months following the lockdown, this difference in proportions gradually reduced, until group birding decreased again (but to a lesser extent) in the peak months of MY 2020. However, this time birders started interacting with each other like usual soon after the peak months (Fig. \@ref(fig:long-changes)). These patterns were largely consistent across different states with minor variations (Fig. \@ref(fig:not-long-states)).

2.  *Site fidelity* (observer-level): Birders during the peak periods of the pandemic travelled much less and visited much fewer sites than usual, instead spending more time in the same locations (i.e., higher site fidelity; Fig. \@ref(fig:long-changes)). This change also persisted in the months following the first peak period in MY 2019, but reverted to regular pre-pandemic levels immediately after the peak months of the second year. These patterns were also largely common across different states with minor variations (Fig. \@ref(fig:not-long-states)).

3.  *Birding time* (observer-level): While the average birder spent less overall time birding during the peak months in both pandemic years, they soon reverted to their usual levels in the following months (Fig. \@ref(fig:time-cov)). The time spent birding during some of the winter months was lower during and even after the pandemic, compared to before. However, birders across different states varied considerably in their response to the pandemic `(fig: state-level birding time)`.

4.  *Hotspot birding*: Birders visited eBird hotspots much less frequently during the peak pandemic months of MY 2019 than before (Fig. \@ref(fig:long-changes)). In the months following the lockdown, this difference in proportions gradually reduced until hotspot birding even returned to normal, but then decreased again (but to a lesser extent) in the peak months of MY 2020. However, this time around birders recovered soon after the peak months (Fig. \@ref(fig:long-changes)). These patterns varied slightly across different states (Fig. \@ref(fig:not-long-states)).

5.  *Birding protocol*: Birders made and uploaded considerably less travelling lists during the peak pandemic months of MY 2019 than before (Fig. \@ref(fig:long-changes)). In the months following the lockdown, this difference in proportions gradually reduced until hotspot birding even returned to normal, but then decreased again (but to a lesser extent) in the peak months of MY 2020. However, this time around birders recovered soon after the peak months (Fig. \@ref(fig:long-changes)). These patterns were largely consistent across states with minor variations (Fig. \@ref(fig:not-long-states)).

6.  *Birding distance*: Birders covered less distance than usual while birding during the peak periods of MY 2019 and MY 2020 (Fig. \@ref(fig:dist-dur-length)). This change evidently disappeared soon, although birders still seem to be covering less distance overall even after the pandemic has passed. These patterns were inconsistent across states (Fig. \@ref(fig:dist-dur-length-states)).

7.  *List duration*: The average checklist became shorter in terms of time during the peak periods of MY 2019 and MY 2020 (Fig. \@ref(fig:dist-dur-length)). This change did not last beyond the peak months, although birders still seem to be uploading lower duration lists even after the pandemic has passed. These patterns were inconsistent across states (Fig. \@ref(fig:dist-dur-length-states)).

8.  *List length*: The average checklist reported fewer species during the peak periods of MY 2019 and MY 2020 (Fig. \@ref(fig:dist-dur-length)). However, this decrease in list length relative to MY 2018 was evidently also present in certain months before as well as after the pandemic. Patterns of list length varied considerably across states (Fig. \@ref(fig:dist-dur-length-states)).

```{r long-changes, fig.cap="Graphs of observer-level group birding and site fidelity, hotspot birding and travelling protocol birding at the national level.", fig.width=8, fig.height=12.5, out.width="100%", out.height="100%", dpi=300, dev="jpeg"}

load(glue("outputs/d01_group_po.RData"))
d1 <- data0_a %>% left_join(month_lab)
load(glue("outputs/d02_fidelity_po.RData"))
d2 <- data0_a %>% left_join(month_lab)
load(glue("outputs/d04_hotspot.RData"))
d3 <- data0_a %>% left_join(month_lab)
load(glue("outputs/d05_protocol.RData"))
d4 <- data0_a %>% left_join(month_lab)



p1 <- ggplot(d1, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted group\nbirding proportion") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p2 <- ggplot(d2, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted no. of\nsites visited") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p3 <- ggplot(d3, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted hotspot\nbirding proportion") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p4 <- ggplot(d4, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted travelling\nbirding proportion") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

(p1 / p2 / guide_area() / p3 / p4) +
  plot_layout(guides = "collect", heights = c(3, 3, 0.5, 3, 3)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))


```

### Urban bias

Birding in urban areas was slightly higher during the peak periods of both pandemic years, though to a lesser degree in the second (Fig. \@ref(fig:urban)). However, this change was short-lived and returned to regular levels in the immediate months. Urban bias was highly inconsistent, and was even absent in some states.

```{r urban, fig.cap="Graph of urban birding proportion at the national level.", fig.width=8, out.width="100%", fig.height=5, out.height="100%", dpi=300, dev="jpeg"}

load("outputs/d09a_s_UNU.RData")

ggplot(data0_a %>% left_join(month_lab), 
       aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted urban\nbirding proportion") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))

```

### Spatial coverage

Birders collectively covered slightly fewer grid cells during April of MY 2019 (Fig. \@ref(fig:time-cov)), but this has not translated to the larger temporal scale of years `(model in S3)`. In fact, coverage has only increased in every successive month since and over the years (Fig. \@ref(fig:time-cov)). Additionally, the slight decrease at the national level was mirrored only by some states like Kerala, while coverage either stayed the same (e.g., Karnataka) or even kept increasing (e.g., Maharashtra) in others (Fig. \@ref(fig:cov-states)).

### Spatial spread

There was an overall positive raw change in birding effort across districts in both transition periods (Fig. \@ref(fig:s-spread-figs); `figS-netchange`). This was true also for grid coverage from before to during the pandemic, but there was a slight negative change from during to after the pandemic `fig`. Thus, the net change in the latter transition was lower than in the former, but was still not negative `figS-netchange`. There were very few major and worrisome changes, and these were not clustered in space.

Major proportional declines (bright yellow) in both birding effort and grid coverage from before to during and from after to during the pandemic were restricted to a few districts in central and northern India, and there was little overlap between the two transitions (Fig. \@ref(fig:s-spread-figs)). There were also similar numbers of districts showing prominent declines in the opposite direction, i.e., from during to before and from during to after the pandemic respectively. More districts showed major declines in the backward direction than in the forward direction of time. Several districts were covered during the pandemic that were not covered before or after.

```{r time-cov, fig.cap="Graphs of observer-level birding time, and spatial coverage at the national level.", fig.width=8, fig.height=9, out.width="100%", dpi=300, dev="jpeg"}

load("outputs/d03_time_po.RData")
d1 <- data0_a %>% left_join(month_lab)
load("outputs/d09b_s_cover.RData")
d2 <- data0_a %>% left_join(month_lab)

p1 <- ggplot(d1, 
             aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted birding time\nper observer") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p2 <- ggplot(d2, 
             aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted cell coverage\nproportion") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(5, 1, 5)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))

```

```{r s-spread-figs-raw, fig.cap="Graphs of spatial spread visualisations.", out.width="100%", fig.width=8, fig.height=13, dpi=300, dev="jpeg"}

load("outputs/d09c_s_spread.RData")

d1 <- temp1 %>% left_join(grid_dist_base0)
d2 <- temp5 %>% left_join(grid_dist_base0) %>% filter(METRIC == "GRID.COV.T")

# raw change maps (except gridcov w/o threshold)

p1 <- ggplot(d1) +
  geom_polygon(data = indiamap,
               aes(long, lat, group = group),
               fill = "#ACACAC", colour = "black", size = 0.4) +
  geom_sf(aes(fill = VALUE, geometry = geometry), col ="#ACACAC", size = 0.1) +
  facet_wrap(~ T.LABEL) +
  scale_fill_stepsn(colours = RColorBrewer::brewer.pal(n = 9, name = "RdBu"),
                    breaks = c(-1000, -500, -100, -50, 50, 100, 500, 1000),
                    values = scales::rescale(c(-11000, -1000, -500, -100, -50, 
                                               50, 100, 500, 1000, 11000)), 
                    limits = c(-11000, 11000),
                    na.value = "#ACACAC") +
  labs(x = "Longitude", y = "Latitude",
       fill = "Difference in\nno. of lists")


p2 <- ggplot(d2) +
  geom_polygon(data = indiamap,
               aes(long, lat, group = group),
               fill = "#ACACAC", colour = "black", size = 0.4) +
  geom_sf(aes(fill = VALUE, geometry = geometry), col ="#ACACAC", size = 0.1) +
  facet_wrap(~ T.LABEL) +
  scale_fill_stepsn(colours = RColorBrewer::brewer.pal(n = 9, name = "RdBu"),
                    breaks = c(-40, -25, -10, -1, 1, 10, 25, 40),
                    values = scales::rescale(c(-100, -40, -25, -10, -1, 1, 10, 25, 40, 100)), 
                    limits = c(-100, 100),
                    na.value = "#ACACAC") +
  labs(x = "Longitude", y = "Latitude",
       fill = "Difference in\ngrid coverage")


(p1 / p2) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        legend.key.width = unit(3, "cm"))

```

```{r s-spread-figs-prop, fig.cap="Graphs of spatial spread visualisations.", out.width="100%", fig.width=8, fig.height=13, dpi=300, dev="jpeg"}

load("outputs/d09c_s_spread.RData")

d1 <- temp3 %>% left_join(grid_dist_base0)
d2 <- temp4 %>% left_join(grid_dist_base0) 


p1 <- ggplot(d1) +
  geom_polygon(data = indiamap,
               aes(long, lat, group = group),
               fill = "white", colour = "black", size = 0.4) +
  geom_sf(aes(fill = VALUE, geometry = geometry), col ="#ACACAC", size = 0.1) +
  facet_wrap(~ T.LABEL) +
  scale_fill_steps2(low = viridisLite::cividis(5, direction = -1)[1],
                    mid = viridisLite::cividis(5, direction = -1)[3],
                    high = viridisLite::cividis(5, direction = -1)[5],
                    midpoint = -0.45,
                    breaks = c(-1, -0.9, -0.65, -0.45, -0.25, -0.1, -0),
                    limits = c(-1, 0),
                    na.value = "#ACACAC") +
  labs(x = "Longitude", y = "Latitude",
       fill = "Proportional change")

p2 <- ggplot(d2) +
  geom_polygon(data = indiamap,
               aes(long, lat, group = group),
               fill = "white", colour = "black", size = 0.4) +
  geom_sf(aes(fill = VALUE, geometry = geometry), col ="#ACACAC", size = 0.1) +
  facet_wrap(~ T.LABEL) +
  scale_fill_steps2(low = viridisLite::cividis(5, direction = -1)[1],
                    mid = viridisLite::cividis(5, direction = -1)[3],
                    high = viridisLite::cividis(5, direction = -1)[5],
                    midpoint = -0.45,
                    breaks = c(-1, -0.9, -0.65, -0.45, -0.25, -0.1, -0),
                    limits = c(-1, 0),
                    na.value = "#ACACAC") +
  labs(x = "Longitude", y = "Latitude",
       fill = "Proportional change")


(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(6, 0.5, 6)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        legend.key.width = unit(2, "cm"),
        legend.margin = margin(t = 0, unit = "cm"))

```

## Bird species reporting

Overall bird species reporting patterns for the states of Karnataka, Kerala, Maharashtra and Assam were explored for 23 species (11 urban, 12 rural), 21 species (11, 10), 22 species (11, 11) and 19 species (10, 9) respectively. The analysis of reporting patterns with consistent effort considered 20 species (11, 9) common to all `r tot_qual_prolific$LOCATIONS` qualifying locations across the country. In general, urban species are reported more frequently than rural species.

### Overall species reporting

During the peak periods of the pandemic, urban species were reported more frequently than usual in all states (Fig. \@ref(fig:overall2m)), but the states varied in how rural species reporting changed. In Karnataka, Kerala and Assam, the increase in urban species reporting was accompanied by an increase in rural species reporting, though the latter was smaller than the former in Karnataka and Assam. The change in rural species reporting in Karnataka was also higher in the peak period of MY 2020 than of MY 2019. Maharashtra was the only state where rural species were reported less frequently than usual during the peak pandemic months.

Year-round frequencies in Karnataka and Kerala of both urban and rural species were also marginally higher during the pandemic year MY 2019 `fig`. However, from MY 2020 in Karnataka, rural species have been reported more frequently than before. There has been an increase in year-round frequencies of both urban and rural species in Maharashtra in the years after MY 2018, though the latter is smaller than the former. The pandemic years saw an increase in urban species reporting in Assam, while rural species were reported as frequently as before.

Different species within the same category sometimes showed contradictory patterns `(refer annual graph of some state)`. Often, species exhibited patterns contradictory to the expectations of pandemic effects: urban species reporting decreased (e.g., `KA annual graph`) or rural species reporting increased (e.g., `KL annual graph`). Not only did these patterns of species reporting vary considerably from months to years, between states, and across species, they also became weaker (effect sizes decreased).

### Species reporting with consistent effort

Species reporting patterns with consistent effort show a different picture at the small scale: neither urban nor rural species were reported more or less frequently during the peak periods than before, as the highly variable local patterns averaged out across space (Fig. \@ref(fig:prolific)). This remains true for year-round frequencies as well.

```{r overall-212m, fig.cap="Two-months panels of overall bird species reporting models", out.width="100%", fig.width=10, fig.height=13, dpi=300, dev="jpeg"}

load("outputs/b01_overall_s1c_model.RData")
d1 <- birds_pred %>% filter(MONTHS.TYPE == "LD")
d5 <- birds_pred %>% filter(MONTHS.TYPE == "ALL")
load("outputs/b01_overall_s2c_model.RData")
d2 <- birds_pred %>% filter(MONTHS.TYPE == "LD")
d6 <- birds_pred %>% filter(MONTHS.TYPE == "ALL")
load("outputs/b01_overall_s3c_model.RData")
d3 <- birds_pred %>% filter(MONTHS.TYPE == "LD")
d7 <- birds_pred %>% filter(MONTHS.TYPE == "ALL")
load("outputs/b01_overall_s4c_model.RData")
d4 <- birds_pred %>% filter(MONTHS.TYPE == "LD") 
d8 <- birds_pred %>% filter(MONTHS.TYPE == "ALL")

p1 <- ggplot(d1, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(values = c("#8F85C1", "#A3383C"),
                     name = "Species category",
                     labels = c("Rural", "Urban")) +
  scale_y_continuous(limits = c(0.13, 0.35)) +
  labs(x = "", y = "Predicted reporting frequency") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) 


p2 <- ggplot(d2, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(values = c("#8F85C1", "#A3383C"),
                     name = "Species category",
                     labels = c("Rural", "Urban")) +
  scale_y_continuous(limits = c(0.05, 0.22)) +
  labs(x = "Migratory year", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) 

p3 <- ggplot(d3, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(values = c("#8F85C1", "#A3383C"),
                     name = "Species category",
                     labels = c("Rural", "Urban")) +
  scale_y_continuous(limits = c(0.175, 0.33)) +
  labs(x = "", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) 

p4 <- ggplot(d4, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(values = c("#8F85C1", "#A3383C"),
                     name = "Species category",
                     labels = c("Rural", "Urban")) +
  scale_y_continuous(limits = c(0.19, 0.5)) +
  labs(x = "Migratory year", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) 


p5 <- ggplot(d5, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(values = c("#8F85C1", "#A3383C"),
                     name = "Species category",
                     labels = c("Rural", "Urban")) +
  scale_y_continuous(limits = c(0.13, 0.35)) +
  labs(x = "", y = "Predicted reporting frequency") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) 


p6 <- ggplot(d6, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(values = c("#8F85C1", "#A3383C"),
                     name = "Species category",
                     labels = c("Rural", "Urban")) +
  scale_y_continuous(limits = c(0.05, 0.22)) +
  labs(x = "Migratory year", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) 

p7 <- ggplot(d7, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(values = c("#8F85C1", "#A3383C"),
                     name = "Species category",
                     labels = c("Rural", "Urban")) +
  scale_y_continuous(limits = c(0.175, 0.33)) +
  labs(x = "", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) 

p8 <- ggplot(d8, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(values = c("#8F85C1", "#A3383C"),
                     name = "Species category",
                     labels = c("Rural", "Urban")) +
  scale_y_continuous(limits = c(0.19, 0.5)) +
  labs(x = "Migratory year", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) 

pA <- (p1 | p2 | p3 | p4) + plot_layout(tag_level = "new")
pB <- (p5 | p6 | p7 | p8) + plot_layout(tag_level = "new")

(pA / guide_area() / pB) +
  plot_layout(guides = "collect", heights = c(6, 1, 6)) +
  plot_annotation(tag_levels = c("A", "1")) &
  theme(legend.position = "bottom",
        plot.tag.position = c(0, 1),
        axis.text.x = element_text(colour = (month_lab %>% distinct(M.YEAR, COL3))$COL3,
                                   angle = 45, vjust = 0.5))


```

```{r prolific, fig.cap="Models of bird species reporting patterns with consistent effort.", out.width="100%", fig.width=8, fig.height=6, dpi=300, dev="jpeg"}

load("outputs/b02_prolific_model.RData")

d1a <- prolific_pred %>% filter(MONTHS.TYPE == "LD")
d1b <- prolific_points %>% filter(MONTHS.TYPE == "LD")

d2a <- prolific_pred %>% filter(MONTHS.TYPE == "NL")
d2b <- prolific_points %>% filter(MONTHS.TYPE == "NL")

d3a <- prolific_pred %>% filter(MONTHS.TYPE == "ALL")
d3b <- prolific_points %>% filter(MONTHS.TYPE == "ALL")


p1 <- ggplot(d1a, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
                     scale_color_manual(values = c("#8F85C1", "#A3383C"),
                                        name = "Species category",
                                        labels = c("Rural", "Urban")) +
                     scale_y_continuous(limits = c(0.1, 1.0)) +
                     labs(x = "", y = "Predicted reporting frequency") +
                     # data points
                     geom_point(data = d1b, aes(group = PATH.GROUP),
                                size = 2, alpha = 0.25,
                                position = position_dodge(0.25)) + 
                     geom_path(data = d1b, aes(x = as.numeric(M.YEAR), y = PRED, group = PATH.GROUP),
                               size = 1, alpha = 0.15, position = position_dodge(0.25)) + 
                     # main points
                     geom_point(size = 2) +
                     geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                                   size = 1, width = 0.5)
       
p2 <- ggplot(d2a, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
                     scale_color_manual(values = c("#8F85C1", "#A3383C"),
                                        name = "Species category",
                                        labels = c("Rural", "Urban")) +
                     scale_y_continuous(limits = c(0.1, 1.0)) +
                     labs(x = "Migratory year", y = "") +
                     # data points
                     geom_point(data = d2b, aes(group = PATH.GROUP),
                                size = 2, alpha = 0.25,
                                position = position_dodge(0.25)) + 
                     geom_path(data = d2b, aes(x = as.numeric(M.YEAR), y = PRED, group = PATH.GROUP),
                               size = 1, alpha = 0.15, position = position_dodge(0.25)) + 
                     # main points
                     geom_point(size = 2) +
                     geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                                   size = 1, width = 0.5)

p3 <- ggplot(d3a, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
                     scale_color_manual(values = c("#8F85C1", "#A3383C"),
                                        name = "Species category",
                                        labels = c("Rural", "Urban")) +
                     scale_y_continuous(limits = c(0.1, 1.0)) +
                     labs(x = "", y = "") +
                     # data points
                     geom_point(data = d1b, aes(group = PATH.GROUP),
                                size = 2, alpha = 0.25,
                                position = position_dodge(0.25)) + 
                     geom_path(data = d1b, aes(x = as.numeric(M.YEAR), y = PRED, group = PATH.GROUP),
                               size = 1, alpha = 0.15, position = position_dodge(0.25)) + 
                     # main points
                     geom_point(size = 2) +
                     geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                                   size = 1, width = 0.5)

(p1 | p2 | p3) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        plot.tag.position = c(0, 1),
        axis.text.x = element_text(colour = (month_lab %>% distinct(M.YEAR, COL3))$COL3,
                                   angle = 45, vjust = 0.5))


```

# Discussion

## Did data quantity increase?

Despite the various restrictions on movement and social interaction during the strict lockdowns, we found a consistent increase in data quantity over the years, in terms of both participation and data generation (Fig. \@ref(fig:lists-ebirders)). This is largely consistent with results from elsewhere in CS programmes such as eBird [@hochachka2021] and iNaturalist [@crimmins2021], `but unlike these, we found no direct link with the pandemic. Instead, our results suggest a consistent and long-term growth of eBird data quantity in India.`

## Did birders follow precautionary measures?

During the peak lockdown periods in India, birders behaved differently and in accordance with the guidelines to prevent and minimise the spread of the disease. Birders became more isolated (Fig. \@ref(fig:long-changes)A), and more sedentary by limiting travel (Fig. \@ref(fig:long-changes)B) and moving less (Fig. \@ref(fig:long-changes)C,D), during the first peak lockdown period. In the following months, as restrictions were gradually lifted, birder behaviour also changed slowly back to normal, but only until the peak months of the second pandemic year. In this second lockdown period, birders again followed guidelines but not to the same extent as in the first year, despite it being more devastating in terms of COVID-19 cases. Moreover, this time birders recovered immediately after the peak months, suggesting that the novelty of the disease was important in the first year.

## Did birding sessions become shorter?

Checklist-level "effort" decreased during the lockdowns, with lists becoming shorter overall (Fig. \@ref(fig:dist-dur-length)A,B,C), although the specifics of the changes varied considerably depending on the specific area and scale under consideration (Fig. \@ref(fig:dist-dur-length-states)). Moreover, there is no clear evidence for attributing these changes to the pandemic; rather, it seems to be a consistent change in birding culture that started even before the pandemic and has continued after.

## Was overall birding effort lower?

This slow change in birding culture where sessions have become shorter has seemingly resulted in slightly lower overall birding effort even during the winter months after the pandemic (Fig. \@ref(fig:time-cov)A). However, this pattern varied among the different states and may not be directly linked to the pandemic, but could be due to the general growth in participation (Fig. \@ref(fig:lists-ebirders)), with many new birders pulling down the average birding time.

Although the lower mobility of birders caused a slight decline in spatial coverage in April of MY 2019 (Fig. \@ref(fig:time-cov)B), this quickly reverted and did not translate to larger or longer-term changes (Fig. \@ref(fig:s-spread-rest)`net`). In fact, overall spatial coverage has only been increasing since, owing to the birding community in India being in the exponential phase of growth. Importantly, none of the fine-scale changes in effort were concentrated in any notable way in space, i.e., the very few worrisome changes were not located all in certain parts of the country (Fig. \@ref(fig:s-spread-figs)).

## Did birders report more urban and less rural species?

Since a large majority of the eBirding community is urban, birders becoming more sedentary also meant a clear bias towards urban areas in the data generated during the lockdown periods. However, the bias was short-lived and was inconsistent across space, and it therefore did not result in larger or longer-term changes to data quality relative to pre-COVID standards.

Nevertheless, if this fine-scale urban bias also resulted in complementary changes in species reporting patterns, this would impact the utility of data from these years. Indeed, in some states, birders reported urban species slightly more frequently during the pandemic years than before, but this change remained even after the pandemic, suggesting some other cause for the change. Moreover, in other states there was no evidence of higher urban species reporting during the pandemic. Thus the evidence of directional change was not consistent and cannot be clearly linked to the pandemic.

Importantly, we found no clear, consistent and directional evidence for birders reporting rural species less frequently.

## Did bird abundance change?

Not only do anthropogenic factors affect birds via habitat availability, suitability, survival and fitness, etc., but many of their behavioural traits are also directly affected by human activity (e.g., song amplitude and frequency, time of vocal activity, etc.). Indeed, during the period of most stringent restrictions in USA, although a few species showed a decline, most increased in abundance in areas that saw a decrease in human activity [@schrimpf2021]. In contrast, species tightly linked to human presence and activity like the Feral Pigeon showed lower abundances [@soh2021]. Moreover, these behavioural traits can change rapidly in response to environmental changes [@derryberry2020; @sanderfoot2022; @hentati-sundberg2021; @gordo2021; but see @seress2021].

To ascertain whether birds became more or less abundant during the pandemic in the Indian context, we checked once again for differences in bird species reporting, but this time considering only data with similar effort across space and time. If this revealed changes in species reporting, it would suggest some response of birds to the reduced human activity during the pandemic, such as changes in bird abundance or detectability.

However, we found no clear evidence for bird abundance changing during the pandemic. This might be because a few months of reduced human activity is insufficient to cause changes in large-scale patterns of bird behaviour or occurrence [see @seress2021]. A few individual locations did separately show signs of increased or decreased reporting of urban or rural species during the peak pandemic months, but any pattern disappeared when averaged across space. Moreover, any location-specific pattern became muted when considering entire years, or when excluding the peak pandemic months. In conclusion, there is no consistent evidence of birds being affected by the pandemic, positively or negatively, artefactually or otherwise.

## What happened during the peak pandemic months?

Most changes to data quality (mainly birder behaviour) were concentrated in the months of April and May of MY 2019 and MY 2020. These months involved some of the most stringent restrictions and government regulations (e.g., nationwide lockdown) in order to curb the spread of the pandemic [@bbcnews2020; @gettleman2020; @theeconomictimes2020]. The changes in birder behaviour were more drastic and long-lasting in MY 2019 than in MY 2020, possibility due to the novelty and shock of the experience in the first pandemic year, even though the second was even more devastating in terms of casualties [@worldometer2023].

Soon after the restrictions were lifted, birders returned to their usual behaviours, pointing to their resilience. As a result, regardless of whether or not there was a change during the peak pandemic months, all metrics of birder behaviour, spatial coverage, spatial spread and urban bias remained unaffected at the yearly level `(refer to YEAR coefficients in a couple of the model summaries in the supplementary, showing that there were no long-lasting effects)`. Additionally, the changes that the pandemic brought to human life did not lead to a prominent change at the large scale in how different bird species were reported.

## How useful is eBird data from the pandemic?

In the unprecedented and disruptive times of the COVID-19 pandemic, conventional monitoring programmes---which are especially contingent on human circumstances---were brought to a halt, and their pitfalls became obvious. However, many of these shortcomings do not apply as strongly to (and hence can be overcome by using) large-scale, crowdsourced, open access data, such as from CS programmes [@bates2020], many of which have continued to operate. Moreover, data from the pandemic, arriving amidst this latest period of intense and extremely rapid expansion of human activity (and thereby modification of the environment), also offers an invaluable opportunity to study questions relating to the influence of anthropogenic factors on organisms [@rutz2020; @bates2020].

There is a general paucity of studies looking at the effects of the pandemic on CS data characteristics, as most focus on changes in the organisms instead, even if using data from CS programmes [e.g., @warrington2022]. We expected changes to the quantity and quality of eBird data during the pandemic due to behavioural changes in the birding community caused directly by the various restrictions and indirectly by lifestyle changes. What we found is a steady increase in data quantity, and certain expected changes to data quality. Since the quality of eBird data from India during the pandemic does not seem different enough, and there is no evidence of it having affected bird species reporting, it remains immensely valuable. In fact, our results highlight certain key points about using this data and making inferences from it.

### Importance of scale

@hochachka2021 emphasised spatial scale considerations by showing the presence of high degrees of regional variability in the effects of the pandemic. However, they only considered a single month across years, and so failed to capture variability across temporal scales. Our findings highlight the importance of both spatial and temporal scales in considering the effects of a major event like the COVID-19 pandemic.

At the small scale of months and states (or even smaller like districts), there were clear and expected changes in birder behaviour and, consequently, to data quality. There was also high variability across states. Most of these changes reverted to "normal" levels after the lockdown months, or averaged out across different states, hence resulting in very few notable nationwide changes at the larger temporal scale of years. Most analyses of long-term abundance trends or distributions or migratory patterns using eBird data are carried out at such large scales. Moreover, the individual fine-scale variations in metrics of data quality were spread throughout the country and not concentrated in one or a few specific regions.

### Guidelines for using pandemic eBird data from India

Our findings suggest that there is still tremendous potential for CS data like eBird, even during the pandemic. When formulating a question to be answered using this data, the researcher must first consider the scale (spatial and temporal) of the question. Questions of larger scale can be tackled directly, whereas small-scale questions will benefit from a deep-dive into artefacts in the raw data that can produce misleading patterns. In the latter cases, the researcher is recommended to conduct similar analyses to the ones in this study in order to first assess the suitability of the data before attempting to answer the question.

# Conclusion

We found that birders did behave differently during the pandemic, but its consequences for eBird data quality were lower than expected from the impact the pandemic has had on human life. Interestingly, our results suggest that official restrictions introduced during the peak periods of the pandemic were more instrumental in changing birder behaviour, while changes in lifestyle were not drastic or long-lasting enough to alter behaviour in the long run. This resilience of birders helped maintain similar data quality levels at large scales, and also led to a steady increase in data quantity. Importantly, there was no clear evidence of bird reporting having changed due to the pandemic, as a consequence of changes in either birder behaviour or bird abundance. Therefore, the eBird dataset remains valuable, with only certain scale considerations to be kept in mind.

# References

::: {#refs}
:::

\newpage
