# Supplementary Material

## Methods

### Exploratory analyses and bootstrapped estimates {#explboot}

We selected the various metrics of data quantity and quality from the available effort metadata associated with every eBird checklist. To check whether these metrics were useful to answer our question, we first explored basic patterns by calculating the metrics for every month and migratory year, for the four individual states of interest (state-level). These calculations produced a mean or median value and associated confidence interval from non-parametric bootstrapping: first the mean of the values in a group were calculated, following which a new sample of the same size as the original one was selected by resampling the original group with replacement, and a new mean was calculated, and so on until this calculation was done 1000 times. The mean/median of the 1000 sampled means is the group mean/median estimate while its standard deviation is the standard error. We then calculated CIs from the SEs assuming a Gaussian distribution ($CIs = mean \pm 1.96 SE$) due to the large sample size in our dataset.

From the level of states, all the metrics were then calculated for the whole country by averaging the means and propagating the SEs, in order to highlight the high variability across scales, regions, governments, etc. Assuming a Gaussian distribution is necessary to propagate SEs. The error propagation was done as follows:

$$SE_{y} = \frac {\sqrt {\sum_{i=1}^n (SE_{x_{i}}) ^2}} {n}$$

where $n$ is the number of observations/rows in the group.

### Spatial spread {#sspreadmeth}

For analysing spatial spread of the data, we decided to focus on districts as the spatial unit instead of 25 km × 25 km grid cells as in other metrics. This is because birding patterns in general are not uniformly and consistently spread across the country, so the fine scale of grid cells would produce too much noise. Here, districts provide a nice compromise: they are large enough to track change with less noise, while still being small enough to inform us about local patterns. We ensured that a single grid cell was linked with only one district, using random sampling from the list of districts it covered. The temporal unit of focus was the categorical COVID time period ("BEF", "DUR" or "AFT") and not individual months. Only those districts that were "covered" at least once in the study period, i.e., had at least one qualifying checklist in migratory years 2018--2021, were considered in the analyses.

Our aim was to understand whether (and if so, how) the way birding patterns were spread across the country changed during the pandemic years. The two patterns of interest were change in birding effort (number of lists) and change in grid coverage (proportion of total grid cells having birding activity) from one year to the next. The former tells us about the intensity of birding while the latter shows us its breadth or completeness. Here, the changes in metrics rather than the metrics themselves inform us about consistency/variability across years, and would better highlight any anomalies brought by the pandemic. In the case of the period "DUR", both metrics were averaged across the two migratory years it comprised, 2019 and 2020.

The grid coverage metric considered only those districts comprising at least five grid cells after the randomised assignment of grid cells to districts. Although this meant that certain districts which actually cover five or more grid cells might have been disqualified due to the random assignment, the converse is also true so some districts which in reality do not cover five or more grid cells might have qualified. We assume the two cases average out over the entire country, so it is not a statistical concern. However, grid coverage was further analysed in two conditions: one with no threshold for grid cell coverage, and one where a cell was considered "covered" only if it had five or more qualifying checklists per migratory year. The latter condition, by removing outlier cases, would produce clearer patterns with less noise.

Due to the difficulty in quantitatively ascertaining change in spatial spread across such a large area with huge variability, we devised a suite of three analyses which, put together, would provide a clear qualitative picture of change in spatial spread/clustering of the metric.

#### Spatial visualisation of raw change

Raw change in a metric $a$ from period $x$ to period $y$ was calculated as the simple difference of their values $(a_{y}-a_{x})$. It was not calculated for cases where both $a_{x}$ and $a_{y}$ were zero. We plotted the raw change per district from one COVID period to another on a map of India, colouring the districts based on magnitude of change. Here we were concerned with two things: firstly, whether the majority of districts showed a positive or a negative change; second, how many districts showed high magnitude of change, and whether or not these were clustered in space. If most districts showed a negative change in birding effort or grid coverage from BEF to DUR, this would translate to an effect of the pandemic on the quantity and quality of data collected during the pandemic. The higher the magnitude of change, the more worrisome the change would be, and if these high magnitude changes were spatially clustered, this would result in a high spatial bias during the pandemic years, as opposed to if they were spread out across different regions of the country.

#### Net raw change across the country

This is a more quantitative method of analysing whether the majority of districts showed a positive or a negative change. We estimated mean raw change and its 95% CI, by bootstrapping across all the districts (see Section \@ref(explboot) above). If the net change from DUR to AFT was significantly (no overlap of CIs) higher than from BEF to DUR, this would suggest that birding effort or grid coverage was subdued during the pandemic.

#### Spatial visualisation of proportional change in two temporal directions

Proportional change in a metric $a$ from period $x$ to period $y$ was calculated as $(a_{y}-a_{x})/a_{x}$. It was not calculated for cases where both $a_{x}$ and $a_{y}$ were zero. If only $a_{x}$ was zero, a constant $k=1$ was added to both $a_{x}$ and $a_{y}$ in order to avoid zero in the denominator. Moreover, for the birding effort metric, proportional change was not calculated if the first period in the transition had less than 10 qualifying checklists ($a_{x} < 10$). Otherwise, since the metric has a fixed lower limit but not an upper limit, this would result in very high values of proportional change simply due to the low overall birding.

We visualised only the negative values of proportional change, since a general increase in birding effort or grid coverage is not worrisome and is in fact even expected. Here, the values were coloured based on magnitude and once again the high magnitude changes are of interest, but we are also interested in transitions in both directions. For example, if the same districts with high magnitudes of negative proportional change from BEF to DUR have similar values also from AFT to DUR, this would suggest a clear and direct impact of the pandemic in lower birding effort or grid coverage.

### Temporal spread {#tspreadmeth}

#### Days of the week

For each week of a migratory year, we calculated the proportion of lists submitted in each of the seven days of the week. These seven proportions were then averaged across all weeks of the year using bootstrapping (see Section \@ref(explboot)) to get means and associated CIs for the entire year. We visualised this distribution of birding effort across the seven days of a week and compared the four migratory years. These calculations were done separately for the whole country and for the four individual states of interest.

#### Times of the day

For each day of a migratory year, we calculated the proportion of lists submitted in each of the 24 hours of the day. These were then averaged across all days of the year using bootstrapping (see Section \@ref(explboot)) to get means and associated CIs for the entire year. We visualised this distribution of birding effort across the hours of a day and compared the four migratory years. These calculations were done separately for the whole country and for the four individual states of interest.

### Model specifications

#### Data quality {#modelspecs-d}

For the observer-level group birding metric, a GLMM was fitted with the following specification:

$$metric \sim year + year:month + (1/observer)$$

where $year$ represents the migratory years. For observer-level site fidelity and birding time, LMMs were fitted with the following specification to the log-transformed metric:

$$log(metric) \sim year + year:month + (1/observer)$$

For hotspot birding, birding protocol, urban bias and spatial coverage, GLMMs were fitted as follows:

$$metric \sim year + year:month + (1/cell)$$

For birding distance, list duration and list length, LMMs were fitted with the following specification to the log-transformed metric:

$$log(metric) \sim year + year:month + (1/cell)$$

#### Bird species reporting {#modelspecs-b}

Overall reporting GLMMs with detection (yes or no) as the response variable were specified as follows:

$$detection \sim year*month + month:length + (1/cell)$$

and species reporting with consistent effort as follows:

$$detection \sim year + year:month + month:length + category + (1/species)$$

where $length$ refers to the median number of species reported in the respective month, $category$ is the species category and $species$ is the name of the species.

## Figures

### Data quantity increased

```{r lists-ebirders, fig.cap="Graphs of change in data quantity over time, in terms of data generation (number of qualifying lists; A) and participation (number of active eBidrers; B). Values for the four migratory years are shown in different colours (MY 2018 in teal-blue, MY 2019 in red, MY 2020 in orange and MY 2021 in purple), and the peak pandemic months are shown in red text. Note the peak in February of each year, which coincides with the global event, Great Backyard Bird Count.", out.width="100%", fig.width=8, fig.height=9, dpi=300, dev="jpeg"}

x <- month_lab %>% 
  mutate(YEAR = ifelse(MONTH %in% 6:12, 
                       as.numeric(as.character(M.YEAR)), 
                       as.numeric(as.character(M.YEAR)) + 1)) %>% 
  mutate(DATE = as.Date(paste(YEAR, MONTH, "01", sep = "-")))

d1 <- dataquant_a %>% left_join(x) %>% arrange(DATE)


p1 <- ggplot(d1, aes(x = DATE, y = NO.LISTS, colour = M.YEAR)) +
  labs(x = "Months", y = "No. of qualifying lists") +
  geom_point(size = 2) +
  geom_line(size = 1) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  scale_y_continuous(label = comma) +
  scale_x_date(labels = d1$MONTH.LAB, breaks = d1$DATE)

p2 <- ggplot(d1, aes(x = DATE, y = NO.EBIRDERS, colour = M.YEAR)) +
  labs(x = "Months", y = "No. of active eBirders") +
  geom_point(size = 2) +
  geom_line(size = 1) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  scale_y_continuous(label = comma) +
  scale_x_date(labels = d1$MONTH.LAB, breaks = d1$DATE)

(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(6, 0.5, 6)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.margin = margin(t = 0, unit = "cm"),
        legend.position = "bottom",
        axis.text.y = element_text(angle = 90, vjust = 0, hjust = 0.5),
        axis.text.x = element_text(colour = d1$COL2, angle = 90, size = 8, vjust = 0.5))

```

### Birding sessions became shorter

```{r dist-dur-length, fig.cap="Graphs of birding distance, list duration and list length at the national level. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours (MY 2018 in teal-blue, MY 2019 in red, MY 2020 in orange and MY 2021 in purple), and the peak pandemic months are shown in red text.", out.width="100%", fig.width=8, fig.height=12, dpi=300, dev="jpeg"}

load("outputs/d06_distance.RData")
d1 <- data0_a %>% left_join(month_lab)
load("outputs/d07_duration.RData")
d2 <- data0_a %>% left_join(month_lab)
load("outputs/d08_length.RData")
d3 <- data0_a %>% left_join(month_lab)

p1 <- ggplot(d1, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted birding\ndistance (km)") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p2 <- ggplot(d2, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted list\nduration (mins)") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p3 <- ggplot(d3, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted list length\n(no. of species)") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

(p1 / p2 / p3) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))

```

```{r dist-dur-length-state1, fig.cap="Graphs of birding distance, list duration and list length for the state of Karnataka. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours (MY 2018 in teal-blue, MY 2019 in red, MY 2020 in orange and MY 2021 in purple), and the peak pandemic months are shown in red text.", out.width="100%", fig.width=8, fig.height=12, dpi=300, dev="jpeg"}

load("outputs/d06_distance.RData")
d1 <- data0_b %>% filter(STATE == "Karnataka") %>% left_join(month_lab)
load("outputs/d07_duration.RData")
d2 <- data0_b %>% filter(STATE == "Karnataka") %>% left_join(month_lab)
load("outputs/d08_length.RData")
d3 <- data0_b %>% filter(STATE == "Karnataka") %>% left_join(month_lab)

p1 <- ggplot(d1, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted birding\ndistance (km)") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p2 <- ggplot(d2, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted list\nduration (mins)") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p3 <- ggplot(d3, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted list length\n(no. of species)") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

(p1 / p2 / p3) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))

```

```{r dist-dur-length-state2, fig.cap="Graphs of birding distance, list duration and list length for the state of Kerala. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours (MY 2018 in teal-blue, MY 2019 in red, MY 2020 in orange and MY 2021 in purple), and the peak pandemic months are shown in red text.", out.width="100%", fig.width=8, fig.height=12, dpi=300, dev="jpeg"}

load("outputs/d06_distance.RData")
d1 <- data0_b %>% filter(STATE == "Kerala") %>% left_join(month_lab)
load("outputs/d07_duration.RData")
d2 <- data0_b %>% filter(STATE == "Kerala") %>% left_join(month_lab)
load("outputs/d08_length.RData")
d3 <- data0_b %>% filter(STATE == "Kerala") %>% left_join(month_lab)

p1 <- ggplot(d1, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted birding\ndistance (km)") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p2 <- ggplot(d2, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted list\nduration (mins)") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p3 <- ggplot(d3, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted list length\n(no. of species)") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

(p1 / p2 / p3) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))

```

```{r dist-dur-length-state3, fig.cap="Graphs of birding distance, list duration and list length for the state of Maharashtra. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours (MY 2018 in teal-blue, MY 2019 in red, MY 2020 in orange and MY 2021 in purple), and the peak pandemic months are shown in red text.", out.width="100%", fig.width=8, fig.height=12, dpi=300, dev="jpeg"}

load("outputs/d06_distance.RData")
d1 <- data0_b %>% filter(STATE == "Maharashtra") %>% left_join(month_lab)
load("outputs/d07_duration.RData")
d2 <- data0_b %>% filter(STATE == "Maharashtra") %>% left_join(month_lab)
load("outputs/d08_length.RData")
d3 <- data0_b %>% filter(STATE == "Maharashtra") %>% left_join(month_lab)

p1 <- ggplot(d1, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted birding\ndistance (km)") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p2 <- ggplot(d2, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted list\nduration (mins)") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p3 <- ggplot(d3, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted list length\n(no. of species)") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

(p1 / p2 / p3) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))

```

```{r dist-dur-length-state4, fig.cap="Graphs of birding distance, list duration and list length for the state of Assam. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours (MY 2018 in teal-blue, MY 2019 in red, MY 2020 in orange and MY 2021 in purple), and the peak pandemic months are shown in red text.", out.width="100%", fig.width=8, fig.height=12, dpi=300, dev="jpeg"}

load("outputs/d06_distance.RData")
d1 <- data0_b %>% filter(STATE == "Assam") %>% left_join(month_lab)
load("outputs/d07_duration.RData")
d2 <- data0_b %>% filter(STATE == "Assam") %>% left_join(month_lab)
load("outputs/d08_length.RData")
d3 <- data0_b %>% filter(STATE == "Assam") %>% left_join(month_lab)

p1 <- ggplot(d1, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted birding\ndistance (km)") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p2 <- ggplot(d2, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted list\nduration (mins)") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p3 <- ggplot(d3, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted list length\n(no. of species)") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

(p1 / p2 / p3) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))

```

### Birders followed precautionary measures

```{r long-changes-state1, fig.cap="Graphs of observer-level group birding (A) and site fidelity (B), hotspot birding (C) and travelling protocol birding (D) for the state of Karnataka. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours (MY 2018 in teal-blue, MY 2019 in red, MY 2020 in orange and MY 2021 in purple), and the peak pandemic months are shown in red text.", fig.width=8, fig.height=12.5, out.width="100%", out.height="100%", dpi=300, dev="jpeg"}

load(glue("outputs/d01_group_po.RData"))
d1 <- data0_b %>% filter(STATE == "Karnataka") %>% left_join(month_lab)
load(glue("outputs/d02_fidelity_po.RData"))
d2 <- data0_b %>% filter(STATE == "Karnataka") %>% left_join(month_lab)
load(glue("outputs/d04_hotspot.RData"))
d3 <- data0_b %>% filter(STATE == "Karnataka") %>% left_join(month_lab)
load(glue("outputs/d05_protocol.RData"))
d4 <- data0_b %>% filter(STATE == "Karnataka") %>% left_join(month_lab)


p1 <- ggplot(d1, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted group\nbirding proportion") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p2 <- ggplot(d2, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted no. of\nsites visited") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p3 <- ggplot(d3, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted hotspot\nbirding proportion") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p4 <- ggplot(d4, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted travelling\nbirding proportion") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

(p1 / p2 / guide_area() / p3 / p4) +
  plot_layout(guides = "collect", heights = c(3, 3, 0.5, 3, 3)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))

```

```{r long-changes-state2, fig.cap="Graphs of observer-level group birding (A) and site fidelity (B), hotspot birding (C) and travelling protocol birding (D) for the state of Kerala. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours (MY 2018 in teal-blue, MY 2019 in red, MY 2020 in orange and MY 2021 in purple), and the peak pandemic months are shown in red text.", fig.width=8, fig.height=12.5, out.width="100%", out.height="100%", dpi=300, dev="jpeg"}

load(glue("outputs/d01_group_po.RData"))
d1 <- data0_b %>% filter(STATE == "Kerala") %>% left_join(month_lab)
load(glue("outputs/d02_fidelity_po.RData"))
d2 <- data0_b %>% filter(STATE == "Kerala") %>% left_join(month_lab)
load(glue("outputs/d04_hotspot.RData"))
d3 <- data0_b %>% filter(STATE == "Kerala") %>% left_join(month_lab)
load(glue("outputs/d05_protocol.RData"))
d4 <- data0_b %>% filter(STATE == "Kerala") %>% left_join(month_lab)


p1 <- ggplot(d1, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted group\nbirding proportion") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p2 <- ggplot(d2, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted no. of\nsites visited") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p3 <- ggplot(d3, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted hotspot\nbirding proportion") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p4 <- ggplot(d4, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted travelling\nbirding proportion") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

(p1 / p2 / guide_area() / p3 / p4) +
  plot_layout(guides = "collect", heights = c(3, 3, 0.5, 3, 3)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))

```

```{r long-changes-state3, fig.cap="Graphs of observer-level group birding (A) and site fidelity (B), hotspot birding (C) and travelling protocol birding (D) for the state of Maharashtra. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours (MY 2018 in teal-blue, MY 2019 in red, MY 2020 in orange and MY 2021 in purple), and the peak pandemic months are shown in red text.", fig.width=8, fig.height=12.5, out.width="100%", out.height="100%", dpi=300, dev="jpeg"}

load(glue("outputs/d01_group_po.RData"))
d1 <- data0_b %>% filter(STATE == "Maharashtra") %>% left_join(month_lab)
load(glue("outputs/d02_fidelity_po.RData"))
d2 <- data0_b %>% filter(STATE == "Maharashtra") %>% left_join(month_lab)
load(glue("outputs/d04_hotspot.RData"))
d3 <- data0_b %>% filter(STATE == "Maharashtra") %>% left_join(month_lab)
load(glue("outputs/d05_protocol.RData"))
d4 <- data0_b %>% filter(STATE == "Maharashtra") %>% left_join(month_lab)


p1 <- ggplot(d1, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted group\nbirding proportion") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p2 <- ggplot(d2, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted no. of\nsites visited") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p3 <- ggplot(d3, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted hotspot\nbirding proportion") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p4 <- ggplot(d4, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted travelling\nbirding proportion") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

(p1 / p2 / guide_area() / p3 / p4) +
  plot_layout(guides = "collect", heights = c(3, 3, 0.5, 3, 3)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))

```

```{r long-changes-state4, fig.cap="Graphs of observer-level group birding (A) and site fidelity (B), hotspot birding (C) and travelling protocol birding (D) for the state of Assam. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours (MY 2018 in teal-blue, MY 2019 in red, MY 2020 in orange and MY 2021 in purple), and the peak pandemic months are shown in red text.", fig.width=8, fig.height=12.5, out.width="100%", out.height="100%", dpi=300, dev="jpeg"}

load(glue("outputs/d01_group_po.RData"))
d1 <- data0_b %>% filter(STATE == "Assam") %>% left_join(month_lab)
load(glue("outputs/d02_fidelity_po.RData"))
d2 <- data0_b %>% filter(STATE == "Assam") %>% left_join(month_lab)
load(glue("outputs/d04_hotspot.RData"))
d3 <- data0_b %>% filter(STATE == "Assam") %>% left_join(month_lab)
load(glue("outputs/d05_protocol.RData"))
d4 <- data0_b %>% filter(STATE == "Assam") %>% left_join(month_lab)


p1 <- ggplot(d1, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted group\nbirding proportion") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p2 <- ggplot(d2, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted no. of\nsites visited") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p3 <- ggplot(d3, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted hotspot\nbirding proportion") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p4 <- ggplot(d4, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted travelling\nbirding proportion") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

(p1 / p2 / guide_area() / p3 / p4) +
  plot_layout(guides = "collect", heights = c(3, 3, 0.5, 3, 3)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))

```

### Urban bias

```{r urban-states, fig.cap="Urban bias in the states of Karnataka (A), Kerala (B), Maharashtra (C) and Assam (D): Y-axis shows the proportion of checklists that are from urban areas. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours (MY 2018 in teal-blue, MY 2019 in red, MY 2020 in orange and MY 2021 in purple), and the peak pandemic months are shown in red text.", fig.width=8, fig.height=12.5, out.width="100%", out.height="100%", dpi=300, dev="jpeg"}

load("outputs/d09a_s_UNU.RData")
d1 <- data0_b %>% filter(STATE == "Karnataka") %>% left_join(month_lab)
d2 <- data0_b %>% filter(STATE == "Kerala") %>% left_join(month_lab)
d3 <- data0_b %>% filter(STATE == "Maharashtra") %>% left_join(month_lab)
d4 <- data0_b %>% filter(STATE == "Assam") %>% left_join(month_lab)

p1 <- ggplot(d1, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted urban\nbirding proportion") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))

p2 <- ggplot(d2, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted urban\nbirding proportion") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))

p3 <- ggplot(d3, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted urban\nbirding proportion") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))

p4 <- ggplot(d4, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted urban\nbirding proportion") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))

(p1 / p2 / guide_area() / p3 / p4) +
  plot_layout(guides = "collect", heights = c(3, 3, 0.5, 3, 3)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))

```

### Was overall birding effort lower?

```{r time-cov-state1, fig.cap="Graphs of observer-level birding time (A), and spatial coverage (B) for the state of Karnataka. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours (MY 2018 in teal-blue, MY 2019 in red, MY 2020 in orange and MY 2021 in purple), and the peak pandemic months are shown in red text.", fig.width=8, fig.height=9, out.width="100%", dpi=300, dev="jpeg"}

load("outputs/d03_time_po.RData")
d1 <- data0_b %>% filter(STATE == "Karnataka") %>% left_join(month_lab)
load("outputs/d09b_s_cover.RData")
d2 <- data0_b %>% filter(STATE == "Karnataka") %>% left_join(month_lab)

p1 <- ggplot(d1, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted birding time\nper observer") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p2 <- ggplot(d2, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted cell coverage\nproportion") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(6, 0.5, 6)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))

```

```{r time-cov-state2, fig.cap="Graphs of observer-level birding time (A), and spatial coverage (B) for the state of Kerala. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours (MY 2018 in teal-blue, MY 2019 in red, MY 2020 in orange and MY 2021 in purple), and the peak pandemic months are shown in red text.", fig.width=8, fig.height=9, out.width="100%", dpi=300, dev="jpeg"}

load("outputs/d03_time_po.RData")
d1 <- data0_b %>% filter(STATE == "Kerala") %>% left_join(month_lab)
load("outputs/d09b_s_cover.RData")
d2 <- data0_b %>% filter(STATE == "Kerala") %>% left_join(month_lab)

p1 <- ggplot(d1, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted birding time\nper observer") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p2 <- ggplot(d2, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted cell coverage\nproportion") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(6, 0.5, 6)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))

```

```{r time-cov-state3, fig.cap="Graphs of observer-level birding time (A), and spatial coverage (B) for the state of Maharashtra. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours (MY 2018 in teal-blue, MY 2019 in red, MY 2020 in orange and MY 2021 in purple), and the peak pandemic months are shown in red text.", fig.width=8, fig.height=9, out.width="100%", dpi=300, dev="jpeg"}

load("outputs/d03_time_po.RData")
d1 <- data0_b %>% filter(STATE == "Maharashtra") %>% left_join(month_lab)
load("outputs/d09b_s_cover.RData")
d2 <- data0_b %>% filter(STATE == "Maharashtra") %>% left_join(month_lab)

p1 <- ggplot(d1, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted birding time\nper observer") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p2 <- ggplot(d2, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted cell coverage\nproportion") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(6, 0.5, 6)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))

```

```{r time-cov-state4, fig.cap="Graphs of observer-level birding time (A), and spatial coverage (B) for the state of Assam. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours (MY 2018 in teal-blue, MY 2019 in red, MY 2020 in orange and MY 2021 in purple), and the peak pandemic months are shown in red text.", fig.width=8, fig.height=9, out.width="100%", dpi=300, dev="jpeg"}

load("outputs/d03_time_po.RData")
d1 <- data0_b %>% filter(STATE == "Assam") %>% left_join(month_lab)
load("outputs/d09b_s_cover.RData")
d2 <- data0_b %>% filter(STATE == "Assam") %>% left_join(month_lab)

p1 <- ggplot(d1, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted birding time\nper observer") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p2 <- ggplot(d2, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Predicted cell coverage\nproportion") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(6, 0.5, 6)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))

```

```{r s-spread-rest1, fig.cap="(A) Spatial spread: raw change in percentage grid coverage per district with no threshold from before to during and from during to after the pandemic. Districts are shaded based on the simple difference in values between the two periods from highly negative (red) to highly positive (blue). Grey districts are those that did not meet the criteria for inclusion in the analysis. (B,C,D) Net change across districts in number of lists (B), percentage grid coverage with no threshold (C) and thresholded percentage grid coverage (D) from before to during and from during to after the pandemic.", out.width="100%", fig.width=8, fig.height=13, dpi=300, dev="jpeg"}

load("outputs/d09c_s_spread.RData")

d1 <- temp5 %>% left_join(grid_dist_base0) %>% filter(METRIC == "GRID.COV")
d2 <- temp2
d3 <- temp6 %>% filter(METRIC == "GRID.COV")
d4 <- temp6 %>% filter(METRIC == "GRID.COV.T")

p1 <- ggplot(d1) +
  geom_polygon(data = indiamap,
               aes(long, lat, group = group),
               fill = "#ACACAC", colour = "black", size = 0.4) +
  geom_sf(aes(fill = VALUE, geometry = geometry), col ="#ACACAC", size = 0.1) +
  facet_wrap(~ T.LABEL) +
  scale_fill_stepsn(colours = RColorBrewer::brewer.pal(n = 9, name = "RdBu"),
                    breaks = c(-40, -25, -10, -1, 1, 10, 25, 40),
                    values = scales::rescale(c(-100, -40, -25, -10, -1, 1, 10, 25, 40, 100)), 
                    limits = c(-100, 100),
                    na.value = "#ACACAC") +
  labs(x = "Longitude", y = "Latitude",
       fill = "Raw change in\ngrid coverage")

p2 <- ggplot(d2, aes(x = T.LABEL, y = VALUE)) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.25, width = 0.25, position = position_dodge(0.5)) +
  scale_x_discrete(labels = c("Before to\nDuring", "During to\nAfter")) +
  labs(x = "", y = "Net change in no. of lists") 

p3 <- ggplot(d3, aes(x = T.LABEL, y = VALUE)) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.25, width = 0.25, position = position_dodge(0.5)) +
  scale_x_discrete(labels = c("Before to\nDuring", "During to\nAfter")) +
  labs(x = "Transition", y = "Net change in grid coverage (%)") 

p4 <- ggplot(d4, aes(x = T.LABEL, y = VALUE)) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.25, width = 0.25, position = position_dodge(0.5)) +
  scale_x_discrete(labels = c("Before to\nDuring", "During to\nAfter")) +
  labs(x = "", y = "") 


(p1 / (p2 | p3 | p4)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        legend.key.width = unit(3, "cm"),
        legend.margin = margin(t = 0, unit = "cm"),
        axis.text.x = element_text(size = 9))

```

```{r s-spread-rest2, fig.cap="Spatial spread: proportional declines in percentage grid coverage per district with no threshold in forward (top row) and backward (bottom row) transitions. Although the analysis calculated proportional change, districts with positive values are shown in grey; declines are shaded based on the magnitude of proportional change from low (dark blue) to high (yellow). White districts are those that did not meet the criteria for inclusion in the analysis.", out.width="100%", fig.width=8, fig.height=11, dpi=300, dev="jpeg"}

d5 <- temp7 %>% filter(METRIC == "GRID.COV") %>% left_join(grid_dist_base0)

ggplot(d5) +
  geom_polygon(data = indiamap,
               aes(long, lat, group = group),
               fill = "white", colour = "black", size = 0.4) +
  geom_sf(aes(fill = VALUE, geometry = geometry), col ="#ACACAC", size = 0.1) +
  facet_wrap(~ T.LABEL, ncol = 2) +
  scale_fill_steps2(low = viridisLite::cividis(5, direction = -1)[1],
                    mid = viridisLite::cividis(5, direction = -1)[3],
                    high = viridisLite::cividis(5, direction = -1)[5],
                    midpoint = -0.45,
                    breaks = c(-1, -0.9, -0.65, -0.45, -0.25, -0.1, -0),
                    limits = c(-1, 0),
                    na.value = "#ACACAC") +
  labs(x = "Longitude", y = "Latitude",
       fill = "Proportional change") +
  theme(legend.position = "bottom",
        legend.key.width = unit(2, "cm"),
        legend.margin = margin(t = 0, unit = "cm"))

```

```{r s-spread-rest3, fig.cap="Spatial spread: proportional declines in thresholded percentage grid coverage per district in forward (top row) and backward (bottom row) transitions. Although the analysis calculated proportional change, districts with positive values are shown in grey; declines are shaded based on the magnitude of proportional change from low (dark blue) to high (yellow). White districts are those that did not meet the criteria for inclusion in the analysis.", out.width="100%", fig.width=8, fig.height=11, dpi=300, dev="jpeg"}

d6 <- temp7 %>% filter(METRIC == "GRID.COV.T") %>% left_join(grid_dist_base0)

ggplot(d6) +
  geom_polygon(data = indiamap,
               aes(long, lat, group = group),
               fill = "white", colour = "black", size = 0.4) +
  geom_sf(aes(fill = VALUE, geometry = geometry), col ="#ACACAC", size = 0.1) +
  facet_wrap(~ T.LABEL, ncol = 2) +
  scale_fill_steps2(low = viridisLite::cividis(5, direction = -1)[1],
                    mid = viridisLite::cividis(5, direction = -1)[3],
                    high = viridisLite::cividis(5, direction = -1)[5],
                    midpoint = -0.45,
                    breaks = c(-1, -0.9, -0.65, -0.45, -0.25, -0.1, -0),
                    limits = c(-1, 0),
                    na.value = "#ACACAC") +
  labs(x = "Longitude", y = "Latitude",
       fill = "Proportional change") +
  theme(legend.position = "bottom",
        legend.key.width = unit(2, "cm"),
        legend.margin = margin(t = 0, unit = "cm"))

```

### When did birders bird?

Most birding before the pandemic happened on weekends, especially on Sundays, but proportional birding on Sunday decreased in the pandemic years, and there were slight suggestions of birding being more spread out over the weekdays (Fig. \@ref(fig:t-spread)A). This could be because many people started working from home and hence had some time freed up. Similarly, much more birding happened in the evening hours (17:00--18:00 hrs) during the pandemic years than before or after, while there was a little less morning birding, around 06:00 hrs (Fig. \@ref(fig:t-spread)B). People might have done birding in the extra time that would have otherwise gone for the evening commute. Some evidence for these speculations is that the changes were much more notable in states having big metropolitan cities like Karnataka (Fig. \@ref(fig:t-spread-state1)B) and Maharashtra (Fig. \@ref(fig:t-spread-state3)A).

```{r t-spread, fig.cap="Proportional spread of birding (lists submitted) at the national level across the seven days of a week (A) and the twenty-four hours of a day (B). Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours (MY 2018 in teal-blue, MY 2019 in red, MY 2020 in orange and MY 2021 in purple).", fig.width=8, fig.height=9, out.width="100%", dpi=300, dev="jpeg"}

load("outputs/d10a_t_dow.RData")
d1 <- data_a
load("outputs/d10b_t_tod.RData")
d2 <- data_a


p1 <- ggplot(d1, aes(DAY.W.LABEL, PROP.LISTS, colour = M.YEAR)) + 
  geom_point(size = 1.8, position = position_dodge(0.7)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.6, position = position_dodge(0.7)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  labs(x = "Day of week", y = "Proportion of lists")

p2 <- ggplot(d2, aes(HOUR, PROP.LISTS, colour = M.YEAR)) + 
  geom_point(size = 1.8, position = position_dodge(0.9)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.6, position = position_dodge(0.9)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  scale_x_continuous(limits = c(-0.2, 23.2), breaks = 0:23) +
  labs(x = "Time of day (hours)", y = "Proportion of lists")

(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(6, 0.5, 6)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        plot.tag.position = c(0, 1),
        axis.text.y = element_text(angle = 90, hjust = 0.5, vjust = 0))

```

```{r t-spread-state1, fig.cap="Proportional spread of birding (lists submitted) in Karnataka across the seven days of a week (A) and the twenty-four hours of a day (B). Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours (MY 2018 in teal-blue, MY 2019 in red, MY 2020 in orange and MY 2021 in purple).", fig.width=8, fig.height=9, out.width="100%", dpi=300, dev="jpeg"}

load("outputs/d10a_t_dow.RData")
d1 <- data_b %>% filter(STATE == "Karnataka")
load("outputs/d10b_t_tod.RData")
d2 <- data_b %>% filter(STATE == "Karnataka")


p1 <- ggplot(d1, aes(DAY.W.LABEL, PROP.LISTS, colour = M.YEAR)) + 
  geom_point(size = 1.8, position = position_dodge(0.7)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.6, position = position_dodge(0.7)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  labs(x = "Day of week", y = "Proportion of lists")

p2 <- ggplot(d2, aes(HOUR, PROP.LISTS, colour = M.YEAR)) + 
  geom_point(size = 1.8, position = position_dodge(0.9)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.6, position = position_dodge(0.9)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  scale_x_continuous(limits = c(-0.2, 23.2), breaks = 0:23) +
  labs(x = "Time of day (hours)", y = "Proportion of lists")

(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(6, 0.5, 6)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        plot.tag.position = c(0, 1),
        axis.text.y = element_text(angle = 90, hjust = 0.5, vjust = 0))

```

```{r t-spread-state2, fig.cap="Proportional spread of birding (lists submitted) in Kerala across the seven days of a week (A) and the twenty-four hours of a day (B). Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours (MY 2018 in teal-blue, MY 2019 in red, MY 2020 in orange and MY 2021 in purple).", fig.width=8, fig.height=9, out.width="100%", dpi=300, dev="jpeg"}

load("outputs/d10a_t_dow.RData")
d1 <- data_b %>% filter(STATE == "Kerala")
load("outputs/d10b_t_tod.RData")
d2 <- data_b %>% filter(STATE == "Kerala")


p1 <- ggplot(d1, aes(DAY.W.LABEL, PROP.LISTS, colour = M.YEAR)) + 
  geom_point(size = 1.8, position = position_dodge(0.7)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.6, position = position_dodge(0.7)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  labs(x = "Day of week", y = "Proportion of lists")

p2 <- ggplot(d2, aes(HOUR, PROP.LISTS, colour = M.YEAR)) + 
  geom_point(size = 1.8, position = position_dodge(0.9)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.6, position = position_dodge(0.9)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  scale_x_continuous(limits = c(-0.2, 23.2), breaks = 0:23) +
  labs(x = "Time of day (hours)", y = "Proportion of lists")

(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(6, 0.5, 6)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        plot.tag.position = c(0, 1),
        axis.text.y = element_text(angle = 90, hjust = 0.5, vjust = 0))

```

```{r t-spread-state3, fig.cap="Proportional spread of birding (lists submitted) in Maharashtra across the seven days of a week (A) and the twenty-four hours of a day (B). Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours (MY 2018 in teal-blue, MY 2019 in red, MY 2020 in orange and MY 2021 in purple).", fig.width=8, fig.height=9, out.width="100%", dpi=300, dev="jpeg"}

load("outputs/d10a_t_dow.RData")
d1 <- data_b %>% filter(STATE == "Maharashtra")
load("outputs/d10b_t_tod.RData")
d2 <- data_b %>% filter(STATE == "Maharashtra")


p1 <- ggplot(d1, aes(DAY.W.LABEL, PROP.LISTS, colour = M.YEAR)) + 
  geom_point(size = 1.8, position = position_dodge(0.7)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.6, position = position_dodge(0.7)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  labs(x = "Day of week", y = "Proportion of lists")

p2 <- ggplot(d2, aes(HOUR, PROP.LISTS, colour = M.YEAR)) + 
  geom_point(size = 1.8, position = position_dodge(0.9)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.6, position = position_dodge(0.9)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  scale_x_continuous(limits = c(-0.2, 23.2), breaks = 0:23) +
  labs(x = "Time of day (hours)", y = "Proportion of lists")

(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(6, 0.5, 6)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        plot.tag.position = c(0, 1),
        axis.text.y = element_text(angle = 90, hjust = 0.5, vjust = 0))

```

```{r t-spread-state4, fig.cap="Proportional spread of birding (lists submitted) in Assam across the seven days of a week (A) and the twenty-four hours of a day (B). Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours (MY 2018 in teal-blue, MY 2019 in red, MY 2020 in orange and MY 2021 in purple).", fig.width=8, fig.height=9, out.width="100%", dpi=300, dev="jpeg"}

load("outputs/d10a_t_dow.RData")
d1 <- data_b %>% filter(STATE == "Assam")
load("outputs/d10b_t_tod.RData")
d2 <- data_b %>% filter(STATE == "Assam")


p1 <- ggplot(d1, aes(DAY.W.LABEL, PROP.LISTS, colour = M.YEAR)) + 
  geom_point(size = 1.8, position = position_dodge(0.7)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.6, position = position_dodge(0.7)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  labs(x = "Day of week", y = "Proportion of lists")

p2 <- ggplot(d2, aes(HOUR, PROP.LISTS, colour = M.YEAR)) + 
  geom_point(size = 1.8, position = position_dodge(0.9)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.6, position = position_dodge(0.9)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  scale_x_continuous(limits = c(-0.2, 23.2), breaks = 0:23) +
  labs(x = "Time of day (hours)", y = "Proportion of lists")

(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(6, 0.5, 6)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        plot.tag.position = c(0, 1),
        axis.text.y = element_text(angle = 90, hjust = 0.5, vjust = 0))

```

### Overall bird reporting

```{r overall-non1, fig.cap="Calculated monthly reporting frequencies for urban (A) and rural (B) species in Bengaluru city. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours (MY 2018 in teal-blue, MY 2019 in red, MY 2020 in orange and MY 2021 in purple), and the peak pandemic months are shown in red text.", out.width="100%", fig.width=8, fig.height=13, dpi=300, dev="jpeg"}

load("outputs/b01_overall_s1a_m.RData")
d1 <- data_a %>% ungroup() %>% filter(SP.CATEGORY == "U") %>% 
  left_join(month_lab) %>% arrange(M.YEAR, MONTH.LAB)
d2 <- data_a %>% ungroup() %>% filter(SP.CATEGORY == "R") %>% 
  left_join(month_lab) %>% arrange(M.YEAR, MONTH.LAB)

p1 <- ggplot(d1, aes(MONTH.LAB, REP.FREQ, colour = M.YEAR)) +
  geom_point(size = 1.5, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 0.75, width = 0.75, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 2, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Month", y = "Reporting frequency") 

p2 <- ggplot(d2, aes(MONTH.LAB, REP.FREQ, colour = M.YEAR)) +
  geom_point(size = 1.5, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 0.75, width = 0.75, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 2, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Month", y = "Reporting frequency")

(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(6, 0.5, 6)) +
  plot_annotation(tag_levels = "A") &
  theme_classic(base_size = 10) &
  theme(strip.text = element_text(size = 5),
        legend.position = "bottom",
        axis.text.x = element_text(colour = (d1 %>% distinct(MONTH.LAB, COL1))$COL1))

```

```{r overall-non2, fig.cap="Calculated monthly reporting frequencies for urban (A) and rural (B) species in Karnataka state. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours (MY 2018 in teal-blue, MY 2019 in red, MY 2020 in orange and MY 2021 in purple), and the peak pandemic months are shown in red text.", out.width="100%", fig.width=8, fig.height=13, dpi=300, dev="jpeg"}

d1 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "U") %>% 
  left_join(month_lab) %>% arrange(M.YEAR, MONTH.LAB)
d2 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "R") %>% 
  left_join(month_lab) %>% arrange(M.YEAR, MONTH.LAB)

p1 <- ggplot(d1, aes(MONTH.LAB, REP.FREQ, colour = M.YEAR)) +
  geom_point(size = 1.5, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 0.75, width = 0.75, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 2, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Month", y = "Reporting frequency") 

p2 <- ggplot(d2, aes(MONTH.LAB, REP.FREQ, colour = M.YEAR)) +
  geom_point(size = 1.5, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 0.75, width = 0.75, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 2, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Month", y = "Reporting frequency")

(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(6, 0.5, 6)) +
  plot_annotation(tag_levels = "A") &
  theme_classic(base_size = 10) &
  theme(strip.text = element_text(size = 5),
        legend.position = "bottom",
        axis.text.x = element_text(colour = (d1 %>% distinct(MONTH.LAB, COL1))$COL1))

```

```{r overall-non3, fig.cap="Calculated annual reporting frequencies for urban (A) and rural (B) species in Bengaluru city. Points represent means while error bars show 95% confidence intervals, and the peak pandemic years are shown in red text.", out.width="100%", fig.width=8, fig.height=11, dpi=300, dev="jpeg"}

load("outputs/b01_overall_s1b_y.RData")
d1 <- data_a %>% ungroup() %>% filter(SP.CATEGORY == "U") %>% 
  left_join(month_lab) %>% arrange(M.YEAR)
d2 <- data_a %>% ungroup() %>% filter(SP.CATEGORY == "R") %>% 
  left_join(month_lab) %>% arrange(M.YEAR)

p1 <- ggplot(d1, aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.5, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 4, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Migratory year", y = "Reporting frequency") 

p2 <- ggplot(d2, aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.5, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 4, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Migratory year", y = "Reporting frequency")

(p1 / p2) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme_classic(base_size = 10) &
  theme(strip.text = element_text(size = 7),
        legend.position = "bottom",
        axis.text.x = element_text(colour = (d1 %>% distinct(M.YEAR, COL3))$COL3))

```

```{r overall-non4, fig.cap="Calculated annual reporting frequencies for urban (A) and rural (B) species in Karnataka state. Points represent means while error bars show 95% confidence intervals, and the peak pandemic years are shown in red text.", out.width="100%", fig.width=8, fig.height=11, dpi=300, dev="jpeg"}

d1 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "U") %>% 
  left_join(month_lab) %>% arrange(M.YEAR)
d2 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "R") %>% 
  left_join(month_lab) %>% arrange(M.YEAR)

p1 <- ggplot(d1, aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.5, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 4, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Migratory year", y = "Reporting frequency") 

p2 <- ggplot(d2, aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.5, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 4, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Migratory year", y = "Reporting frequency")

(p1 / p2) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme_classic(base_size = 10) &
  theme(strip.text = element_text(size = 7),
        legend.position = "bottom",
        axis.text.x = element_text(colour = (d1 %>% distinct(M.YEAR, COL3))$COL3))

```

```{r overall-non5, fig.cap="Calculated monthly reporting frequencies for urban (A) and rural (B) species in Kochi city. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours (MY 2018 in teal-blue, MY 2019 in red, MY 2020 in orange and MY 2021 in purple), and the peak pandemic months are shown in red text.", out.width="100%", fig.width=8, fig.height=13, dpi=300, dev="jpeg"}

load("outputs/b01_overall_s2a_m.RData")
d1 <- data_a %>% ungroup() %>% filter(SP.CATEGORY == "U") %>% 
  left_join(month_lab) %>% arrange(M.YEAR, MONTH.LAB)
d2 <- data_a %>% ungroup() %>% filter(SP.CATEGORY == "R") %>% 
  left_join(month_lab) %>% arrange(M.YEAR, MONTH.LAB)

p1 <- ggplot(d1, aes(MONTH.LAB, REP.FREQ, colour = M.YEAR)) +
  geom_point(size = 1.5, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 0.75, width = 0.75, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 2, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Month", y = "Reporting frequency") 

p2 <- ggplot(d2, aes(MONTH.LAB, REP.FREQ, colour = M.YEAR)) +
  geom_point(size = 1.5, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 0.75, width = 0.75, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 2, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Month", y = "Reporting frequency")

(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(6, 0.5, 6)) +
  plot_annotation(tag_levels = "A") &
  theme_classic(base_size = 10) &
  theme(strip.text = element_text(size = 5),
        legend.position = "bottom",
        axis.text.x = element_text(colour = (d1 %>% distinct(MONTH.LAB, COL1))$COL1))

```

```{r overall-non6, fig.cap="Calculated monthly reporting frequencies for urban (A) and rural (B) species in Kerala state. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours (MY 2018 in teal-blue, MY 2019 in red, MY 2020 in orange and MY 2021 in purple), and the peak pandemic months are shown in red text.", out.width="100%", fig.width=8, fig.height=13, dpi=300, dev="jpeg"}

d1 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "U") %>% 
  left_join(month_lab) %>% arrange(M.YEAR, MONTH.LAB)
d2 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "R") %>% 
  left_join(month_lab) %>% arrange(M.YEAR, MONTH.LAB)

p1 <- ggplot(d1, aes(MONTH.LAB, REP.FREQ, colour = M.YEAR)) +
  geom_point(size = 1.5, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 0.75, width = 0.75, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 2, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Month", y = "Reporting frequency") 

p2 <- ggplot(d2, aes(MONTH.LAB, REP.FREQ, colour = M.YEAR)) +
  geom_point(size = 1.5, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 0.75, width = 0.75, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 2, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Month", y = "Reporting frequency")

(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(6, 0.5, 6)) +
  plot_annotation(tag_levels = "A") &
  theme_classic(base_size = 10) &
  theme(strip.text = element_text(size = 5),
        legend.position = "bottom",
        axis.text.x = element_text(colour = (d1 %>% distinct(MONTH.LAB, COL1))$COL1))

```

```{r overall-non7, fig.cap="Calculated annual reporting frequencies for urban (A) and rural (B) species in Kochi city. Points represent means while error bars show 95% confidence intervals, and the peak pandemic years are shown in red text.", out.width="100%", fig.width=8, fig.height=11, dpi=300, dev="jpeg"}

load("outputs/b01_overall_s2b_y.RData")
d1 <- data_a %>% ungroup() %>% filter(SP.CATEGORY == "U") %>% 
  left_join(month_lab) %>% arrange(M.YEAR)
d2 <- data_a %>% ungroup() %>% filter(SP.CATEGORY == "R") %>% 
  left_join(month_lab) %>% arrange(M.YEAR)

p1 <- ggplot(d1, aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.5, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 4, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Migratory year", y = "Reporting frequency") 

p2 <- ggplot(d2, aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.5, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 4, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Migratory year", y = "Reporting frequency")

(p1 / p2) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme_classic(base_size = 10) &
  theme(strip.text = element_text(size = 7),
        legend.position = "bottom",
        axis.text.x = element_text(colour = (d1 %>% distinct(M.YEAR, COL3))$COL3))

```

```{r overall-non8, fig.cap="Calculated annual reporting frequencies for urban (A) and rural (B) species in Kerala state. Points represent means while error bars show 95% confidence intervals, and the peak pandemic years are shown in red text.", out.width="100%", fig.width=8, fig.height=11, dpi=300, dev="jpeg"}

d1 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "U") %>% 
  left_join(month_lab) %>% arrange(M.YEAR)
d2 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "R") %>% 
  left_join(month_lab) %>% arrange(M.YEAR)

p1 <- ggplot(d1, aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.5, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 4, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Migratory year", y = "Reporting frequency") 

p2 <- ggplot(d2, aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.5, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 4, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Migratory year", y = "Reporting frequency")

(p1 / p2) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme_classic(base_size = 10) &
  theme(strip.text = element_text(size = 7),
        legend.position = "bottom",
        axis.text.x = element_text(colour = (d1 %>% distinct(M.YEAR, COL3))$COL3))

```

```{r overall-non9, fig.cap="Calculated monthly reporting frequencies for urban (A) and rural (B) species in Pune city. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours (MY 2018 in teal-blue, MY 2019 in red, MY 2020 in orange and MY 2021 in purple), and the peak pandemic months are shown in red text.", out.width="100%", fig.width=8, fig.height=13, dpi=300, dev="jpeg"}

load("outputs/b01_overall_s3a_m.RData")
d1 <- data_a %>% ungroup() %>% filter(SP.CATEGORY == "U") %>% 
  left_join(month_lab) %>% arrange(M.YEAR, MONTH.LAB)
d2 <- data_a %>% ungroup() %>% filter(SP.CATEGORY == "R") %>% 
  left_join(month_lab) %>% arrange(M.YEAR, MONTH.LAB)

p1 <- ggplot(d1, aes(MONTH.LAB, REP.FREQ, colour = M.YEAR)) +
  geom_point(size = 1.5, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 0.75, width = 0.75, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 2, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Month", y = "Reporting frequency") 

p2 <- ggplot(d2, aes(MONTH.LAB, REP.FREQ, colour = M.YEAR)) +
  geom_point(size = 1.5, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 0.75, width = 0.75, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 2, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Month", y = "Reporting frequency")

(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(6, 0.5, 6)) +
  plot_annotation(tag_levels = "A") &
  theme_classic(base_size = 10) &
  theme(strip.text = element_text(size = 5),
        legend.position = "bottom",
        axis.text.x = element_text(colour = (d1 %>% distinct(MONTH.LAB, COL1))$COL1))

```

```{r overall-non10, fig.cap="Calculated monthly reporting frequencies for urban (A) and rural (B) species in Maharashtra state. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours (MY 2018 in teal-blue, MY 2019 in red, MY 2020 in orange and MY 2021 in purple), and the peak pandemic months are shown in red text.", out.width="100%", fig.width=8, fig.height=13, dpi=300, dev="jpeg"}

d1 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "U") %>% 
  left_join(month_lab) %>% arrange(M.YEAR, MONTH.LAB)
d2 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "R") %>% 
  left_join(month_lab) %>% arrange(M.YEAR, MONTH.LAB)

p1 <- ggplot(d1, aes(MONTH.LAB, REP.FREQ, colour = M.YEAR)) +
  geom_point(size = 1.5, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 0.75, width = 0.75, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 2, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Month", y = "Reporting frequency") 

p2 <- ggplot(d2, aes(MONTH.LAB, REP.FREQ, colour = M.YEAR)) +
  geom_point(size = 1.5, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 0.75, width = 0.75, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 2, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Month", y = "Reporting frequency")

(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(6, 0.5, 6)) +
  plot_annotation(tag_levels = "A") &
  theme_classic(base_size = 10) &
  theme(strip.text = element_text(size = 5),
        legend.position = "bottom",
        axis.text.x = element_text(colour = (d1 %>% distinct(MONTH.LAB, COL1))$COL1))

```

```{r overall-non11, fig.cap="Calculated annual reporting frequencies for urban (A) and rural (B) species in Pune city. Points represent means while error bars show 95% confidence intervals, and the peak pandemic years are shown in red text.", out.width="100%", fig.width=8, fig.height=11, dpi=300, dev="jpeg"}

load("outputs/b01_overall_s3b_y.RData")
d1 <- data_a %>% ungroup() %>% filter(SP.CATEGORY == "U") %>% 
  left_join(month_lab) %>% arrange(M.YEAR)
d2 <- data_a %>% ungroup() %>% filter(SP.CATEGORY == "R") %>% 
  left_join(month_lab) %>% arrange(M.YEAR)

p1 <- ggplot(d1, aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.5, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 4, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Migratory year", y = "Reporting frequency") 

p2 <- ggplot(d2, aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.5, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 4, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Migratory year", y = "Reporting frequency")

(p1 / p2) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme_classic(base_size = 10) &
  theme(strip.text = element_text(size = 7),
        legend.position = "bottom",
        axis.text.x = element_text(colour = (d1 %>% distinct(M.YEAR, COL3))$COL3))

```

```{r overall-non12, fig.cap="Calculated annual reporting frequencies for urban (A) and rural (B) species in Maharashtra state. Points represent means while error bars show 95% confidence intervals, and the peak pandemic years are shown in red text.", out.width="100%", fig.width=8, fig.height=11, dpi=300, dev="jpeg"}

d1 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "U") %>% 
  left_join(month_lab) %>% arrange(M.YEAR)
d2 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "R") %>% 
  left_join(month_lab) %>% arrange(M.YEAR)

p1 <- ggplot(d1, aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.5, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 4, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Migratory year", y = "Reporting frequency") 

p2 <- ggplot(d2, aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.5, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 4, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Migratory year", y = "Reporting frequency")

(p1 / p2) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme_classic(base_size = 10) &
  theme(strip.text = element_text(size = 7),
        legend.position = "bottom",
        axis.text.x = element_text(colour = (d1 %>% distinct(M.YEAR, COL3))$COL3))

```

```{r overall-non13, fig.cap="Calculated monthly reporting frequencies for urban (A) and rural (B) species in Assam state. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours (MY 2018 in teal-blue, MY 2019 in red, MY 2020 in orange and MY 2021 in purple), and the peak pandemic months are shown in red text.", out.width="100%", fig.width=8, fig.height=13, dpi=300, dev="jpeg"}

load("outputs/b01_overall_s4a_m.RData")
d1 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "U") %>% 
  left_join(month_lab) %>% arrange(M.YEAR, MONTH.LAB)
d2 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "R") %>% 
  left_join(month_lab) %>% arrange(M.YEAR, MONTH.LAB)

p1 <- ggplot(d1, aes(MONTH.LAB, REP.FREQ, colour = M.YEAR)) +
  geom_point(size = 1.5, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 0.75, width = 0.75, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 2, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Month", y = "Reporting frequency") 

p2 <- ggplot(d2, aes(MONTH.LAB, REP.FREQ, colour = M.YEAR)) +
  geom_point(size = 1.5, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 0.75, width = 0.75, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 2, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Month", y = "Reporting frequency")

(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(6, 0.5, 6)) +
  plot_annotation(tag_levels = "A") &
  theme_classic(base_size = 10) &
  theme(strip.text = element_text(size = 5),
        legend.position = "bottom",
        axis.text.x = element_text(colour = (d1 %>% distinct(MONTH.LAB, COL1))$COL1))

```

```{r overall-non14, fig.cap="Calculated annual reporting frequencies for urban (A) and rural (B) species in Assam state. Points represent means while error bars show 95% confidence intervals, and the peak pandemic years are shown in red text.", out.width="100%", fig.width=8, fig.height=11, dpi=300, dev="jpeg"}

load("outputs/b01_overall_s4b_y.RData")
d1 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "U") %>% 
  left_join(month_lab) %>% arrange(M.YEAR)
d2 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "R") %>% 
  left_join(month_lab) %>% arrange(M.YEAR)

p1 <- ggplot(d1, aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.5, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 4, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Migratory year", y = "Reporting frequency") 

p2 <- ggplot(d2, aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.5, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 4, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Migratory year", y = "Reporting frequency")

(p1 / p2) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme_classic(base_size = 10) &
  theme(strip.text = element_text(size = 7),
        legend.position = "bottom",
        axis.text.x = element_text(colour = (d1 %>% distinct(M.YEAR, COL3))$COL3))

```

```{r overall-10m, fig.cap="Modelled species reporting frequencies for all ten months excluding April and May, for the states of Karnataka (A), Kerala (B), Maharashtra (C) and Assam (D). Points represent means while error bars show 95% confidence intervals. Values for urban and rural species are shown in maroon and purple respectively, and the pandemic years are shown in red text.", out.width="100%", fig.width=10, fig.height=6, dpi=300, dev="jpeg"}

load("outputs/b01_overall_s1c_model.RData")
d1 <- birds_pred %>% filter(MONTHS.TYPE == "NL")
load("outputs/b01_overall_s2c_model.RData")
d2 <- birds_pred %>% filter(MONTHS.TYPE == "NL")
load("outputs/b01_overall_s3c_model.RData")
d3 <- birds_pred %>% filter(MONTHS.TYPE == "NL")
load("outputs/b01_overall_s4c_model.RData")
d4 <- birds_pred %>% filter(MONTHS.TYPE == "NL") 

p1 <- ggplot(d1, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(values = c("#8F85C1", "#A3383C"),
                     name = "Species category",
                     labels = c("Rural", "Urban")) +
  labs(x = "", y = "Predicted reporting frequency") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) 


p2 <- ggplot(d2, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(values = c("#8F85C1", "#A3383C"),
                     name = "Species category",
                     labels = c("Rural", "Urban")) +
  labs(x = "Migratory year", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) 

p3 <- ggplot(d3, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(values = c("#8F85C1", "#A3383C"),
                     name = "Species category",
                     labels = c("Rural", "Urban")) +
  labs(x = "", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) 

p4 <- ggplot(d4, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(values = c("#8F85C1", "#A3383C"),
                     name = "Species category",
                     labels = c("Rural", "Urban")) +
  labs(x = "Migratory year", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) 


(p1 | p2 | p3 | p4) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        plot.tag.position = c(0, 1),
        axis.text.x = element_text(colour = (month_lab %>% distinct(M.YEAR, COL3))$COL3,
                                   angle = 45, vjust = 0.5))

```

## Tables

Tables of model summaries for all the data metrics

```{r model-summ1, fig.cap="Model summary for observer-level group birding at the national level.", out.width="100%", fig.width=8, fig.height=9, dpi=300, dev="jpeg"}

load(glue("outputs/d01_group_po.RData"))



```
