# Results

## Bird species reporting

We explored patterns of overall bird species reporting (across all locations) for the states of Karnataka, Kerala, Maharashtra and Assam, for 23 species (11 urban, 12 rural), 21 species (11, 10), 22 species (11, 11) and 19 species (10, 9) respectively. The analysis of reporting patterns with consistent effort considered 20 species (11, 9) common to all `r tot_qual_prolific$LOCATIONS` qualifying locations across the country. In general, urban species are reported more frequently than rural species.

### Species reporting across all locations

During the peak periods of the pandemic, urban species were reported more frequently than usual in all states (Fig. \@ref(fig:overall-212m)A1--A4), but the states varied in how rural species reporting changed. In Karnataka, Kerala and Assam, the increase in urban species reporting was accompanied by an increase in rural species reporting, though to a slightly lesser extent in Karnataka and Assam. In Karnataka, unlike in Kerala and Assam, rural species reporting was higher in MY 2020 than in MY 2019. Maharashtra was the only state where rural species were reported less frequently than usual during the peak pandemic months.

Year-round frequencies in Karnataka and Kerala of both urban and rural species were also marginally higher during the pandemic year MY 2019 (Fig. \@ref(fig:overall-212m)B1,B2). However, from MY 2020 in Karnataka, rural species have been reported more frequently than previously. There was an increase in year-round frequencies of both urban and rural species in Maharashtra in the years since MY 2018, though the increase in rural species is slightly lower than in urban species (Fig. \@ref(fig:overall-212m)B3). The pandemic years saw an increase in urban species reporting in Assam, while rural species were reported as frequently as they previously were (Fig. \@ref(fig:overall-212m)B4).

Different species within the same category sometimes showed contradictory patterns (e.g., Fig. \@ref(fig:overall-non8)). Often, species exhibited patterns contradictory to the expectations of pandemic effects: urban species reporting decreased (e.g., Fig. \@ref(fig:overall-non4)A) or rural species reporting increased (e.g., Fig. \@ref(fig:overall-non8)B). Not only did these patterns of species reporting vary considerably from months to years, between states, and across species, they also became weaker (effect sizes decreased).

```{r metric-summ}

# table of metrics and their overall patterns: + or - (as in BMS2022 poster)

data.frame(
  Metrics = c(
    "Group birding",
    "Site fidelity",
    "Birding time",
    "Hotspot birding",
    "Birding protocol",
    "Birding distance",
    "List duration",
    "List length",
    "Urban bias",
    "Spatial coverage",
    "Overall bird reporting",
    "Bird reporting with consistent effort"
  ),
  Peak = c("+", "+", "+",
           "+", "+",
           "+", "+", "+",
           "+", "",
           "+", ""),
  Pandemic = c("+", "+", "+",
               "+", "+",
               "+", "+", "+",
               "", "",
               "", ""),
  Post = c("", "", "+",
           "", "",
           "+", "+", "+",
           "", "",
           "", "")
) %>%
  arrange(desc(Peak), desc(Pandemic), desc(Post)) %>%
  knitr::kable(
    caption = 'Summary of changes to metrics of data quality and bird species reporting at the national level, at various temporal scales: column "Peak" refers to the peak pandemic months; "Pandemic" refers to the pandemic years; "Post-pandemic" denotes changes that lasted even after the pandemic years. "+" denotes the presence of a prominent change (positive or negative) relative to pre-pandemic levels.',
    align = "lccc",
    col.names = c("Metrics", "Peak", "Pandemic", "Post-pandemic")
  )

```

### Species reporting at locations with consistent effort

With consistent effort, species reporting patterns show a different picture at the small scale: neither urban nor rural species were reported more or less frequently during the peak periods than before, as the highly variable local patterns averaged out across space (Fig. \@ref(fig:prolific), Table \@ref(tab:metric-summ)). This remains true for year-round frequencies as well.

```{r overall-212m, fig.cap="Modelled species reporting frequencies for the months of April and May (top row) and separately for all twelve months of a year (bottom row), for the states of Karnataka (A1, B1), Kerala (A2, B2), Maharashtra (A3, B3) and Assam (A4, B4). Points represent means while error bars show 95% confidence intervals. Values for urban and rural species are shown in maroon and purple respectively, and the pandemic years are shown in red text.", out.width="100%", fig.width=10, fig.height=13, dpi=300, dev="jpeg"}

load("00_outputs/b_overall_s1c_model.RData")
d1 <-
  birds_pred %>% filter(MONTHS.TYPE == "LD")
d5 <-
  birds_pred %>% filter(MONTHS.TYPE == "ALL")
load("00_outputs/b_overall_s2c_model.RData")
d2 <-
  birds_pred %>% filter(MONTHS.TYPE == "LD")
d6 <-
  birds_pred %>% filter(MONTHS.TYPE == "ALL")
load("00_outputs/b_overall_s3c_model.RData")
d3 <-
  birds_pred %>% filter(MONTHS.TYPE == "LD")
d7 <-
  birds_pred %>% filter(MONTHS.TYPE == "ALL")
load("00_outputs/b_overall_s4c_model.RData")
d4 <-
  birds_pred %>% filter(MONTHS.TYPE == "LD")
d8 <-
  birds_pred %>% filter(MONTHS.TYPE == "ALL")

p1 <-
  ggplot(d1, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(
    values = c("#8F85C1", "#A3383C"),
    name = "Species category",
    labels = c("Rural", "Urban")
  ) +
  scale_y_continuous(limits = c(0.13, 0.35)) +
  labs(x = "", y = "Reporting frequency") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(
    aes(ymin = CI.L, ymax = CI.U),
    size = 1,
    width = 0.5,
    position = position_dodge(0.5)
  )


p2 <-
  ggplot(d2, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(
    values = c("#8F85C1", "#A3383C"),
    name = "Species category",
    labels = c("Rural", "Urban")
  ) +
  scale_y_continuous(limits = c(0.05, 0.22)) +
  labs(x = "Migratory year", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(
    aes(ymin = CI.L, ymax = CI.U),
    size = 1,
    width = 0.5,
    position = position_dodge(0.5)
  )

p3 <-
  ggplot(d3, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(
    values = c("#8F85C1", "#A3383C"),
    name = "Species category",
    labels = c("Rural", "Urban")
  ) +
  scale_y_continuous(limits = c(0.175, 0.33)) +
  labs(x = "", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(
    aes(ymin = CI.L, ymax = CI.U),
    size = 1,
    width = 0.5,
    position = position_dodge(0.5)
  )

p4 <-
  ggplot(d4, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(
    values = c("#8F85C1", "#A3383C"),
    name = "Species category",
    labels = c("Rural", "Urban")
  ) +
  scale_y_continuous(limits = c(0.19, 0.5)) +
  labs(x = "Migratory year", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(
    aes(ymin = CI.L, ymax = CI.U),
    size = 1,
    width = 0.5,
    position = position_dodge(0.5)
  )


p5 <-
  ggplot(d5, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(
    values = c("#8F85C1", "#A3383C"),
    name = "Species category",
    labels = c("Rural", "Urban")
  ) +
  scale_y_continuous(limits = c(0.13, 0.35)) +
  labs(x = "", y = "Reporting frequency") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(
    aes(ymin = CI.L, ymax = CI.U),
    size = 1,
    width = 0.5,
    position = position_dodge(0.5)
  )


p6 <-
  ggplot(d6, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(
    values = c("#8F85C1", "#A3383C"),
    name = "Species category",
    labels = c("Rural", "Urban")
  ) +
  scale_y_continuous(limits = c(0.05, 0.22)) +
  labs(x = "Migratory year", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(
    aes(ymin = CI.L, ymax = CI.U),
    size = 1,
    width = 0.5,
    position = position_dodge(0.5)
  )

p7 <-
  ggplot(d7, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(
    values = c("#8F85C1", "#A3383C"),
    name = "Species category",
    labels = c("Rural", "Urban")
  ) +
  scale_y_continuous(limits = c(0.175, 0.33)) +
  labs(x = "", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(
    aes(ymin = CI.L, ymax = CI.U),
    size = 1,
    width = 0.5,
    position = position_dodge(0.5)
  )

p8 <-
  ggplot(d8, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(
    values = c("#8F85C1", "#A3383C"),
    name = "Species category",
    labels = c("Rural", "Urban")
  ) +
  scale_y_continuous(limits = c(0.19, 0.5)) +
  labs(x = "Migratory year", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(
    aes(ymin = CI.L, ymax = CI.U),
    size = 1,
    width = 0.5,
    position = position_dodge(0.5)
  )

pA <-
  (p1 | p2 | p3 | p4) + plot_layout(tag_level = "new")
pB <-
  (p5 | p6 | p7 | p8) + plot_layout(tag_level = "new")

(pA / guide_area() / pB) +
  plot_layout(guides = "collect", heights = c(6, 1, 6)) +
  plot_annotation(tag_levels = c("A", "1")) &
  theme(
    legend.position = "bottom",
    plot.tag.position = c(0, 1),
    axis.text.x = element_text(
      colour = (month_lab %>% distinct(M.YEAR, COL3))$COL3,
      angle = 45,
      vjust = 0.5
    )
  )


```

```{r prolific, fig.cap="Species reporting frequencies with consistent effort data, separately modelled for the months of April and May (A), the other ten months (B), and all twelve months of a year (C). Values for urban and rural species are shown in maroon and purple respectively, and the pandemic years are shown in red text. Each translucent line connecting points represents reporting frequencies from a single location/patch across the four years. Points with error bars represent means with 95% confidence intervals.", out.width="100%", fig.width=8, fig.height=6, dpi=300, dev="jpeg"}

load("00_outputs/b_prolific_model.RData")

d1a <-
  prolific_pred %>% filter(MONTHS.TYPE == "LD")
d1b <-
  prolific_points %>% filter(MONTHS.TYPE == "LD")

d2a <-
  prolific_pred %>% filter(MONTHS.TYPE == "NL")
d2b <-
  prolific_points %>% filter(MONTHS.TYPE == "NL")

d3a <-
  prolific_pred %>% filter(MONTHS.TYPE == "ALL")
d3b <-
  prolific_points %>% filter(MONTHS.TYPE == "ALL")


p1 <-
  ggplot(d1a, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(
    values = c("#8F85C1", "#A3383C"),
    name = "Species category",
    labels = c("Rural", "Urban")
  ) +
  scale_y_continuous(limits = c(0.1, 1.0)) +
  labs(x = "", y = "Reporting frequency") +
  # data points
  geom_point(
    data = d1b,
    aes(group = PATH.GROUP),
    size = 2,
    alpha = 0.25,
    position = position_dodge(0.25)
  ) +
  geom_path(
    data = d1b,
    aes(
      x = as.numeric(M.YEAR),
      y = PRED,
      group = PATH.GROUP
    ),
    size = 1,
    alpha = 0.15,
    position = position_dodge(0.25)
  ) +
  # main points
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.3)

p2 <-
  ggplot(d2a, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(
    values = c("#8F85C1", "#A3383C"),
    name = "Species category",
    labels = c("Rural", "Urban")
  ) +
  scale_y_continuous(limits = c(0.1, 1.0)) +
  labs(x = "Migratory year", y = "") +
  # data points
  geom_point(
    data = d2b,
    aes(group = PATH.GROUP),
    size = 2,
    alpha = 0.25,
    position = position_dodge(0.25)
  ) +
  geom_path(
    data = d2b,
    aes(
      x = as.numeric(M.YEAR),
      y = PRED,
      group = PATH.GROUP
    ),
    size = 1,
    alpha = 0.15,
    position = position_dodge(0.25)
  ) +
  # main points
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.3)

p3 <-
  ggplot(d3a, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(
    values = c("#8F85C1", "#A3383C"),
    name = "Species category",
    labels = c("Rural", "Urban")
  ) +
  scale_y_continuous(limits = c(0.1, 1.0)) +
  labs(x = "", y = "") +
  # data points
  geom_point(
    data = d1b,
    aes(group = PATH.GROUP),
    size = 2,
    alpha = 0.25,
    position = position_dodge(0.25)
  ) +
  geom_path(
    data = d1b,
    aes(
      x = as.numeric(M.YEAR),
      y = PRED,
      group = PATH.GROUP
    ),
    size = 1,
    alpha = 0.15,
    position = position_dodge(0.25)
  ) +
  # main points
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.3)

(p1 |
    p2 | p3) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(
    legend.position = "bottom",
    plot.tag.position = c(0, 1),
    axis.text.x = element_text(
      colour = (month_lab %>% distinct(M.YEAR, COL3))$COL3,
      angle = 45,
      vjust = 0.5
    )
  )


```

# Discussion

## Did birders report more urban and fewer rural species?

Since a large majority of the eBirding community is urban, birders becoming more sedentary also meant a clear bias towards urban areas in the data generated during the lockdown periods (Fig. \@ref(fig:urban-cov)). However, the bias was short-lived (Table \@ref(tab:metric-summ)) and was inconsistent across space (Fig. \@ref(fig:urban-cov-states)), and it therefore did not result in larger or longer-term changes to data quality relative to pre-COVID standards. In the post-pandemic period, birding in urban areas was lower than in previous years. While this could be due to birders overcompensating for their restricted travel during the pandemic, the consistent year-on-year decline in urban birding in the non-pandemic months suggests that this is rather yet another aspect of birding in India that has changed over the years with its growth.

Nevertheless, if this fine-scale urban bias also resulted in complementary changes in species reporting patterns, this would impact the utility of data from these years. Indeed, in some states, birders reported urban species slightly more frequently during the pandemic years than before, but this change remained even after the pandemic, suggesting some other cause for the change (Fig. \@ref(fig:overall-212m)B3). In other states, there was no evidence of higher urban species reporting during the pandemic. Thus, the evidence of directional change was not consistent and cannot be clearly linked to the pandemic.

Importantly, we found no clear, consistent and directional evidence for birders reporting rural species less frequently (Fig. \@ref(fig:overall-212m)). Moreover, the changes in urban and rural species reporting across all locations are absent when birding effort is accounted for (Fig. \@ref(fig:prolific)), suggesting that these changes are driven simply by the differential birding in urban and rural areas, and not by inherent changes in bird abundance or behaviour. The highly variable species-level changes in reporting, caused by differential sampling and spatial bias, can be minimised using spatial subsampling methods.

## Did bird abundance change?

Many studies now suggest that survival and even behavioural traits (e.g., song amplitude and frequency, time of vocal activity, etc.) of birds are affected by human activity, potentially leading to an increase in bird abundance or detectability when human activity is reduced. Indeed, during the period of most stringent COVID-19 restrictions in USA, although a few species showed a decline, most increased in abundance in areas that saw a decrease in human activity [@schrimpf2021]. In contrast, species tightly linked to human presence and activity like the Feral Pigeon showed lower abundances [@soh2021]. Behavioural traits of birds can also change rapidly in response to environmental changes [@derryberry2020; @sanderfoot2022; @hentati-sundberg2021; @gordo2021; but see @seress2021].

However, we found no clear evidence for bird abundance changing during the pandemic (Fig. \@ref(fig:prolific)). This might be because a few months of reduced human activity is insufficient to cause changes in large-scale patterns of bird behaviour or occurrence [see @seress2021]. A few individual locations did separately show signs of increased or decreased reporting of urban or rural species during the peak pandemic months, but any pattern disappeared when averaged across space (Table \@ref(tab:metric-summ)). Moreover, any location-specific pattern became muted when considering entire years, or when excluding the peak pandemic months. In conclusion, there is no consistent evidence of birds being affected by the pandemic, positively or negatively, artefactually or otherwise.

## What happened during the peak pandemic months?

... Additionally, the changes that the pandemic brought to human life did not lead to a prominent change at the large scale in how different bird species were reported.

## Paper 2

If so, this would signal some behavioural (changes in detectability or boldness) or population-level (true abundance, occupancy, etc.) changes associated with the pandemic. Not only would this illuminate the impact that human activity has on birds in India, but it would also show that birds are plastic and can respond to environmental changes at the short time scale.

We know that there was a long-lasting (relative to other pandemic-induced changes) urban bias in eBirding in India during the pandemic years [@thrikkadeeriCOVIDp1]. This difference in effort between urban and rural areas could theoretically, depending on the scale, result in higher apparent reporting frequencies for species largely restricted to urban areas. In this case, any adverse effects of the reduced human activity during the peak pandemic months would be cancelled out by this overinflated reporting of urban species, whereas any increases in urban bird detectability or abundance would become even more prominent. On the other hand, if comparisons between urban and rural species were made while keeping effort constant, any difference in reporting frequencies related to real changes in bird behaviour or abundance should become unambiguously evident.

"Disentangling true patterns from potential confounds"

rationale behind prolific birders

# Conclusion

We found that birders did behave differently during the pandemic, but its consequences for eBird data quality were lower than expected from the impact the pandemic has had on human life. Interestingly, our results suggest that official restrictions introduced during the peak periods of the pandemic were more instrumental in changing birder behaviour, while changes in lifestyle were not drastic or long-lasting enough to alter behaviour in the long run. This resilience of birders helped maintain similar data quality levels at large scales, and also led to a steady increase in data quantity. Importantly, there was no clear evidence of bird reporting having changed due to the pandemic, as a consequence of changes in either birder behaviour or bird abundance. Therefore, the eBird dataset remains valuable, with only certain statistical and scale considerations to be kept in mind.
