# Abstract {.unnumbered}

The recent anthropause has had impacts on not just human life but also wildlife. In many cases across North America and Europe, birds seemed to respond to lowered human activity with rapid changes in behaviour (e.g., song, detectability), abundance or even occurrence. The anthropause and associated changes, however, did not always benefit the birds, suggesting that the relationships between birds and human activity are complex and context-dependent. To investigate how birds responded to lowered human activity during the pandemic in India, we explored how bird species reporting on a citizen science platform, eBird, changed during the pandemic. We examined patterns of modelled reporting frequencies for select species across different years. Our analysis revealed no predictable difference between these reporting frequencies in the pandemic years and those in the previous and following years. We also found no clear evidence of bird communities becoming more or less abundant during the pandemic, or of urban bird species being reported more frequently than birds of rural areas where the magnitude of the anthropause was likely to be lower (after accounting for existing differences). Therefore, our study finds no compelling evidence for the popular sentiment of birds "returning" to human-inhabited areas in India during the COVID-19 anthropause. This underscores the importance of appropriate, data-driven, and context-specific analyses, and cautions against making broad global generalisations.

# Keywords {.unnumbered}

Anthropause, bird monitoring, COVID-19 pandemic, eBird, urban birds

\newpage

# Introduction

## The anthropause in India

COVID-19 was the deadliest pandemic of the 21st century, and resulted in major widespread disruption in human life [see @bates2020]. This anthropause [@rutz2020] has had several cascading impacts, including on wildlife and the environment. Researchers put in effort towards utilising this unique opportunity (of greatly reduced human activity) to better understand the impact of human activities on wildlife, by studying how wildlife responded to the anthropause [@bates2020].

The government in India, like elsewhere, had taken several steps to minimise the spread of the COVID-19 pandemic. The most severe restrictions and regulations such as curfews, quarantines, and shutting down of non-essential public services and activities, happened to fall in the months of April--May in both 2020 and 2021. A nationwide total lockdown was implemented just before April 2020 [@bbcnews2020] and continued until late May [@banerjea2020] when the first wave started to subside. The second wave in India started to peak in May 2021, with a higher fatality rate than the first wave, provoking numerous statewide lockdowns and regulations in the absence of a nationwide lockdown [@theindianexpress2021]. The other months of the pandemic years saw lower and varying degrees of regulations and preventive measures across space and time.

<!-- edit below para -->
In the big picture, this could have led to lower overall impact of, and adherence to, regulations due to India's unique socioeconomic, cultural and political position. For instance, social distancing measures are thought to have much lower epidemiological and welfare value in lower-income countries, such as India, than in higher-income countries, because they usually necessitate economic sacrifices [@barnett-howell2021]. Indeed, there was a decline in adherence to social distancing measures that cut across socioeconomic strata in India [@schaner2022]. This was indeed the case for birdwatcher behaviour in India, which underwent expected changes in accordance with government restrictions during the peak pandemic months, but quickly recovered to "normal" in the other months [@thrikkadeeri2024]. For all these reasons, the magnitude of the anthropause itself could have been more muted in India than in other countries, particularly those in the West.

Nevertheless, these periods of restricted human activity were accompanied by the strong media narrative of wildlife making a comeback, "[reclaiming] their territory" [@karmakar2020] both in India as well as across the globe [@downtoearth2020; see @grandoni2023]. While this surge of anecdotal accounts was followed by empirical tests in certain countries, such formal analyses have been very limited in India, and more broadly, the Global South. Moreover, when the few existing studies in India did find pandemic-induced changes, particularly with respect to novel ranging dynamics, they were focused on mammals which are a group of large-ranging animals, and the patterns were inconsistent across species [@behera2022; see also @tucker2023]. As such, there is currently no clear and consistent evidence in India of direct pandemic impacts on bird life, i.e., of birds "making a comeback" to urban areas in India.

## Birds in the anthropause

Our knowledge of bird behavioural traits and abundances changing rapidly during the anthropause---and evidence that the popular "comeback" narrative was not just a result of people paying more attention---is limited to individual, often contradictory, cases from outside India. Birds adjusted song traits in response to lowered anthropogenic noise in San Francisco, USA [@derryberry2020], and in parts of Spain became more detectable in urban areas despite similar occurrence rates [@gordo2021]. In the Pacific Northwest of USA, many bird species responded behaviourally to day-to-day changes in human activity [@sanderfoot2022]. Across North America, during the period of most stringent restrictions, some species showed a decline but most increased in abundance in areas that saw a decrease in human activity [@schrimpf2021].

In contrast, abundances of some species tightly linked to human presence and activity like the Feral Pigeon in Singapore went down [@soh2021]. Reduced human disturbance in urban areas did not necessarily result in increased reproductive output [@seress2021], and in fact even had adverse fitness consequences on some species through increased predation rates [e.g., @hentati-sundberg2021].

Further, from the definition of anthropause as the substantial decline in human mobility and activity during COVID-19 lockdowns [@rutz2020], it follows that the anthropause was more pronounced in urban areas where human mobility and activity are highest. Therefore, any impacts of the pandemic should be stronger in urban bird species than species of rural areas where the decline in human activity was likely less substantial. For instance, species that prefer urban habitats could have been reported more regularly during the pandemic than in the years before and after.

India is a large country with diverse landscapes, habitats, and even social and policy environments, in addition to harbouring 12% of the world's bird species---all of which can affect how birds responded to the anthropause and the consequent large-scale patterns of pandemic impacts. Importantly, there is substantial evidence that humans have both positive and negative impacts on wildlife and the environment [@bates2021], and that there are several idiosyncrasies and nuances that need to be taken into consideration before generalising local patterns. Given the uniqueness of India's response to the pandemic outlined earlier, we can expect bird responses to the Indian anthropause to have been unique as well.

## Our aim

The question therefore remains about whether or not abundances of Indian birds were impacted by the pandemic, and whether notable changes were consistent across species, space and time. To answer this question, a large-scale analytical approach that considers nuances over this vast area is required. The data uploaded by birdwatchers (hereafter birders) to citizen science (CS) projects like eBird [@sullivan2014] enables robust scientific analysis and output at large scales [@kelling2019], and thus represents a good option. Among its various utilities is the estimation and monitoring of large-scale trends in bird abundance over time, using an index of abundance [see e.g., @viswanathanetalSoIB]. Therefore, we wanted to understand whether or not this index of abundance changed predictably during the COVID-19 pandemic for Indian bird species, and how consistently it did so across species, space and time.

# Methods

We obtained the "relMay-2022" version of the publicly available [eBird Basic Dataset (EBD)](https://ebird.org/data/download/ebd) for India [downloaded on 15 June 2022, follows taxonomy in @clements2019], and combined it with [sensitive species data](https://ebird.org/india/news/ebird-sensitive-species/), which was requested separately from eBird. This dataset thus contained all data uploaded to eBird until 31 May 2022, and we converted it into a form appropriate for our analyses using a series of filtering and processing steps outlined in @thrikkadeerip1inpress. Our final eBird dataset contained `r scales::label_comma()(tot_uniq_lists)` unique "qualifying checklists", i.e., lists that passed the filters, with each list classified as being from an urban or rural area, and being associated with one of `r grid_size_deg`$^\circ$ $\times$ `r grid_size_deg`$^\circ$ grid cells.

Since the magnitude of the anthropause was higher in urban areas than in rural areas, its impacts on bird life would also have been more prominent in urban birds. For instance, if species became more abundant or detectable during the pandemic due to lowered noise levels and other human disturbances, this should correlate with consistent increases in reporting frequencies across urban species. Thus, our analyses (see below) to ascertain whether bird species reporting changed at all during the pandemic, involved two sets of regularly and easily observed bird species. One set comprised "urban" species common in human-dominated areas and the other comprised common non-urban or "rural" species that are often found even just adjacent to urban areas. Since not all species are widespread and equally abundant in different parts of the country, there were a few replacements in the species sets for different regions. Overall, we considered `r n_distinct(species_list$COMMON.NAME)` species in this study, of which `r n_distinct((species_list %>% filter(SP.CATEGORY == "U"))$COMMON.NAME)` were classified as urban and `r n_distinct((species_list %>% filter(SP.CATEGORY == "R"))$COMMON.NAME)` as rural. Details of the region-wise species sets are provided in Table \@ref(tab:reg-spec).

## Species reporting across all locations

To analyse the impact of the Indian anthropause on birds, we selected the four Indian states of Karnataka, Kerala, Maharashtra and Assam. These states have relatively high eBirding activity and are located in different parts of the country, but also differ in local governance, policy and general public perspective, thus helping to bring in important nuance. In each of the four states analysed in our study, we also selected one of their large metropolitan (urban) cities.

For every species in the corresponding regional species set, we visualised individual species reporting patterns at two temporal (months, years) and spatial (city, state) scales. This was done by calculating the reporting frequency (proportion of qualifying checklists in which a species was reported), taken as an index of abundance, for each species at each scale. In this study, "years" refer to "migratory years" (MY) which start in June of one calendar year and end in May of the next [see Supplementary Material in @thrikkadeerip1inpress for rationale].

We then pooled data from all urban and all rural species, and analysed whether overall species reporting patterns at the large scale (migratory year & state) were affected by the pandemic. This gave us an idea of how urban and rural species idiosyncratically differed in terms of general reporting, how species reporting varied across multiple urban areas and how this averaged out over a large spatial scale, and how species reporting varied over the course of a year and how this variation was expressed at the large temporal scale. Kamrup Metropolitan, the metropolitan capital of the state of Assam, did not have sufficient data for visualisation of month-level or year-level patterns, so these combinations were dropped.

## Species reporting at locations with consistent effort

caveat:

On the other hand, spurious patterns due to the sensitivity of eBird data to effort expended for data collection by birders [@johnston2021] could make it difficult to infer about actual responses of birds. For instance, the short-term pandemic-induced urban bias [@thrikkadeerip1inpress] could have manifested as consistent declines in reporting frequencies across rural species.

we wanted to determine whether these patterns were due to birds indeed responding to the anthropause by changes in their behaviour or abundance (true abundance x detectability, as a caveat of method used), or if the few short-lived changes in birder behaviour were in fact responsible for artefactual patterns in bird reporting.

rationale behind prolific birders

An eBirder was considered "prolific" if they had uploaded, from a single location, at least 800 qualifying checklists over the course of the four years being analysed (sufficient effort), as well as a minimum of 15 qualifying checklists per month in at least ten months of a year in all four years of the study (consistent effort)---thus translating to about 200 qualifying checklists per year. Since the observer remains constant, and the observer is sufficiently experienced (as inferred from their prolificacy), we assume that detection rates of birds in the location also remains similar over the years (given the short time period).

...

The quality of CS data is contingent on observer effort; it is possible for changes in observer effort (driven by observer behaviour) to produce artefactual changes in metrics such as reporting frequency [@johnston2021], which makes it difficult to make inferences from such data. Hence, in order to ensure that the patterns visible in overall bird species reporting at the large scale are not driven by changes in data quantity and quality, we conducted a similar analysis using data only from "prolific" birders having both sufficient and consistent effort (see Section \@ref(prolific-def) for definitions). Since effort and detection rates are relatively unchanged, this analysis allows us to make a reasonable conclusion regarding true changes in bird abundance, which is a combination of the bird's true abundance and its behavioural characteristics like detectability and boldness.

From the set of species analysed in the previous section, we first filtered out species uncommon at each "prolific" birding location, retaining only those with an overall reporting frequency (over the entire study period) of at least 15%. Next, from the set of species that passed the first filter, we removed those that appeared in only one location (i.e., was observed by only one prolific birder), since a factor with only one level cannot be used as a random effect in a GLMM. Finally, if any prolific birder (their location) was left with species falling in only a single species category (i.e., only urban or only rural) after the filters, they were excluded from the analysis.

## Statistics {#statistics}

We followed the same hierarchy of analyses described in Section \@ref(model-boot). Due to notable differences in birding conditions and birder behaviour between the lockdown and non-lockdown periods [@thrikkadeeri2024], we ran two separate models: one for the months of April and May ("lockdown"), and one for the other ten months ("non-lockdown"). This was to avoid any disproportionate influence of all other months on the parameter estimates for the "lockdown" period.

For each of these three time periods, overall bird reporting models were run for each species with grid cell ID as the random effect, while models for reporting patterns with consistent effort were run for each highly birded location with species identity as the random effect. The full model specifications are provided below. All `how many?` GLMMs were fit with binomial error distribution and cloglog link function, using the BOBYQA optimiser, except in `r (length(fail_spec_KL) + length(fail_spec_MH))` (`r length(fail_spec_KL)` of Kerala, `r length(fail_spec_MH)` of Maharashtra) of the overall bird reporting models in which the state-species combination could not be modelled successfully due to either step-halving or convergence issues. In such cases, the canonical link function (logit) was used instead.

Diagnostics were not inspected separately for the bird reporting models due to the large number of models fitted. Mean and CI estimates were obtained in the same way as outlined in Section \@ref(model-boot), but with 500 simulations instead of 1000, due to computational limitations.

...

GLMMs for species reporting across all locations were fit for each species and for different periods ("lockdown", "non-lockdown", "all") separately. These models, with detection (yes or no) as the response variable, were specified as follows:

$$detection \sim year*month + month:length + (1/cell)$$

Models for species reporting at locations with consistent effort were fit for each locations and for different periods ("lockdown", "non-lockdown", "all") separately. The models were specified as follows:

$$detection \sim year + year:month + month:length + category + (1/species)$$

where $length$ refers to the median number of species reported in the respective month, $category$ is the species category and $species$ is the name of the species.

## Software

All analyses were performed in the R environment [@rcoreteam2021] in RStudio [@positteam2022]. Figures were made using the packages *ggplot2* [@wickham2016] and *patchwork* [@pedersen2020]. In addition, packages *sf* [@pebesma2018], *spdep* [@bivand2022], *raster* [@hijmans2022], *terra* [@hijmans2022a] and *geodata* [@hijmans2022b] for spatial analyses, as well as *lubridate* [@grolemund2011], *glue* [@hester2022] and *tictoc* [@izrailev2021] were used. The source code for the entire analysis, from data preparation to statistical validation of metrics and to producing the final manuscript, is available on the [GitHub repository](https://github.com/rikudoukarthik/covid-ebirding) for the study.

# Results

We explored patterns of overall bird species reporting (across all locations) for the states of Karnataka, Kerala, Maharashtra and Assam, for 23 species (11 urban, 12 rural), 21 species (11, 10), 22 species (11, 11) and 19 species (10, 9) respectively. The analysis of reporting patterns with consistent effort considered 20 species (11, 9) common to all `r tot_qual_prolific$LOCATIONS` qualifying locations across the country. In general, urban species are reported more frequently than rural species.

## Species reporting across all locations

During the peak periods of the pandemic, urban species were reported more frequently than usual in all states (Fig. \@ref(fig:overall-212m)A1--A4), but the states varied in how rural species reporting changed. In Karnataka, Kerala and Assam, the increase in urban species reporting was accompanied by an increase in rural species reporting, though to a slightly lesser extent in Karnataka and Assam. In Karnataka, unlike in Kerala and Assam, rural species reporting was higher in MY 2020 than in MY 2019. Maharashtra was the only state where rural species were reported less frequently than usual during the peak pandemic months.

Year-round frequencies in Karnataka and Kerala of both urban and rural species were also marginally higher during the pandemic year MY 2019 (Fig. \@ref(fig:overall-212m)B1,B2). However, from MY 2020 in Karnataka, rural species have been reported more frequently than previously. There was an increase in year-round frequencies of both urban and rural species in Maharashtra in the years since MY 2018, though the increase in rural species is slightly lower than in urban species (Fig. \@ref(fig:overall-212m)B3). The pandemic years saw an increase in urban species reporting in Assam, while rural species were reported as frequently as they previously were (Fig. \@ref(fig:overall-212m)B4).

Different species within the same category sometimes showed contradictory patterns (e.g., Fig. \@ref(fig:overall-non8)). Often, species exhibited patterns contradictory to the expectations of pandemic effects: urban species reporting decreased (e.g., Fig. \@ref(fig:overall-non4)A) or rural species reporting increased (e.g., Fig. \@ref(fig:overall-non8)B). Not only did these patterns of species reporting vary considerably from months to years, between states, and across species, they also became weaker (effect sizes decreased).

```{r metric-summ}

# table of metrics and their overall patterns: + or - (as in BMS2022 poster)

data.frame(
  Metrics = c(
    "Group birding",
    "Site fidelity",
    "Birding time",
    "Hotspot birding",
    "Birding protocol",
    "Birding distance",
    "List duration",
    "List length",
    "Urban bias",
    "Spatial coverage",
    "Overall bird reporting",
    "Bird reporting with consistent effort"
  ),
  Peak = c("+", "+", "+",
           "+", "+",
           "+", "+", "+",
           "+", "",
           "+", ""),
  Pandemic = c("+", "+", "+",
               "+", "+",
               "+", "+", "+",
               "", "",
               "", ""),
  Post = c("", "", "+",
           "", "",
           "+", "+", "+",
           "", "",
           "", "")
) %>%
  arrange(desc(Peak), desc(Pandemic), desc(Post)) %>%
  knitr::kable(
    caption = 'Summary of changes to metrics of data quality and bird species reporting at the national level, at various temporal scales: column "Peak" refers to the peak pandemic months; "Pandemic" refers to the pandemic years; "Post-pandemic" denotes changes that lasted even after the pandemic years. "+" denotes the presence of a prominent change (positive or negative) relative to pre-pandemic levels.',
    align = "lccc",
    col.names = c("Metrics", "Peak", "Pandemic", "Post-pandemic")
  )

```

## Species reporting at locations with consistent effort

With consistent effort, species reporting patterns show a different picture at the small scale: neither urban nor rural species were reported more or less frequently during the peak periods than before, as the highly variable local patterns averaged out across space (Fig. \@ref(fig:prolific), Table \@ref(tab:metric-summ)). This remains true for year-round frequencies as well.

```{r overall-212m, fig.cap="Modelled species reporting frequencies for the months of April and May (top row) and separately for all twelve months of a year (bottom row), for the states of Karnataka (A1, B1), Kerala (A2, B2), Maharashtra (A3, B3) and Assam (A4, B4). Points represent means while error bars show 95% confidence intervals. Values for urban and rural species are shown in maroon and purple respectively, and the pandemic years are shown in red text.", out.width="100%", fig.width=10, fig.height=13, dpi=300, dev="jpeg"}

load("00_outputs/b_overall_s1c_model.RData")
d1 <-
  birds_pred %>% filter(MONTHS.TYPE == "LD")
d5 <-
  birds_pred %>% filter(MONTHS.TYPE == "ALL")
load("00_outputs/b_overall_s2c_model.RData")
d2 <-
  birds_pred %>% filter(MONTHS.TYPE == "LD")
d6 <-
  birds_pred %>% filter(MONTHS.TYPE == "ALL")
load("00_outputs/b_overall_s3c_model.RData")
d3 <-
  birds_pred %>% filter(MONTHS.TYPE == "LD")
d7 <-
  birds_pred %>% filter(MONTHS.TYPE == "ALL")
load("00_outputs/b_overall_s4c_model.RData")
d4 <-
  birds_pred %>% filter(MONTHS.TYPE == "LD")
d8 <-
  birds_pred %>% filter(MONTHS.TYPE == "ALL")

p1 <-
  ggplot(d1, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(
    values = c("#8F85C1", "#A3383C"),
    name = "Species category",
    labels = c("Rural", "Urban")
  ) +
  scale_y_continuous(limits = c(0.13, 0.35)) +
  labs(x = "", y = "Reporting frequency") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(
    aes(ymin = CI.L, ymax = CI.U),
    size = 1,
    width = 0.5,
    position = position_dodge(0.5)
  )


p2 <-
  ggplot(d2, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(
    values = c("#8F85C1", "#A3383C"),
    name = "Species category",
    labels = c("Rural", "Urban")
  ) +
  scale_y_continuous(limits = c(0.05, 0.22)) +
  labs(x = "Migratory year", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(
    aes(ymin = CI.L, ymax = CI.U),
    size = 1,
    width = 0.5,
    position = position_dodge(0.5)
  )

p3 <-
  ggplot(d3, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(
    values = c("#8F85C1", "#A3383C"),
    name = "Species category",
    labels = c("Rural", "Urban")
  ) +
  scale_y_continuous(limits = c(0.175, 0.33)) +
  labs(x = "", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(
    aes(ymin = CI.L, ymax = CI.U),
    size = 1,
    width = 0.5,
    position = position_dodge(0.5)
  )

p4 <-
  ggplot(d4, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(
    values = c("#8F85C1", "#A3383C"),
    name = "Species category",
    labels = c("Rural", "Urban")
  ) +
  scale_y_continuous(limits = c(0.19, 0.5)) +
  labs(x = "Migratory year", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(
    aes(ymin = CI.L, ymax = CI.U),
    size = 1,
    width = 0.5,
    position = position_dodge(0.5)
  )


p5 <-
  ggplot(d5, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(
    values = c("#8F85C1", "#A3383C"),
    name = "Species category",
    labels = c("Rural", "Urban")
  ) +
  scale_y_continuous(limits = c(0.13, 0.35)) +
  labs(x = "", y = "Reporting frequency") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(
    aes(ymin = CI.L, ymax = CI.U),
    size = 1,
    width = 0.5,
    position = position_dodge(0.5)
  )


p6 <-
  ggplot(d6, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(
    values = c("#8F85C1", "#A3383C"),
    name = "Species category",
    labels = c("Rural", "Urban")
  ) +
  scale_y_continuous(limits = c(0.05, 0.22)) +
  labs(x = "Migratory year", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(
    aes(ymin = CI.L, ymax = CI.U),
    size = 1,
    width = 0.5,
    position = position_dodge(0.5)
  )

p7 <-
  ggplot(d7, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(
    values = c("#8F85C1", "#A3383C"),
    name = "Species category",
    labels = c("Rural", "Urban")
  ) +
  scale_y_continuous(limits = c(0.175, 0.33)) +
  labs(x = "", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(
    aes(ymin = CI.L, ymax = CI.U),
    size = 1,
    width = 0.5,
    position = position_dodge(0.5)
  )

p8 <-
  ggplot(d8, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(
    values = c("#8F85C1", "#A3383C"),
    name = "Species category",
    labels = c("Rural", "Urban")
  ) +
  scale_y_continuous(limits = c(0.19, 0.5)) +
  labs(x = "Migratory year", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(
    aes(ymin = CI.L, ymax = CI.U),
    size = 1,
    width = 0.5,
    position = position_dodge(0.5)
  )

pA <-
  (p1 | p2 | p3 | p4) + plot_layout(tag_level = "new")
pB <-
  (p5 | p6 | p7 | p8) + plot_layout(tag_level = "new")

(pA / guide_area() / pB) +
  plot_layout(guides = "collect", heights = c(6, 1, 6)) +
  plot_annotation(tag_levels = c("A", "1")) &
  theme(
    legend.position = "bottom",
    plot.tag.position = c(0, 1),
    axis.text.x = element_text(
      colour = (month_lab %>% distinct(M.YEAR, COL3))$COL3,
      angle = 45,
      vjust = 0.5
    )
  )


```

```{r prolific, fig.cap="Species reporting frequencies with consistent effort data, separately modelled for the months of April and May (A), the other ten months (B), and all twelve months of a year (C). Values for urban and rural species are shown in maroon and purple respectively, and the pandemic years are shown in red text. Each translucent line connecting points represents reporting frequencies from a single location/patch across the four years. Points with error bars represent means with 95% confidence intervals.", out.width="100%", fig.width=8, fig.height=6, dpi=300, dev="jpeg"}

load("00_outputs/b_prolific_model.RData")

d1a <-
  prolific_pred %>% filter(MONTHS.TYPE == "LD")
d1b <-
  prolific_points %>% filter(MONTHS.TYPE == "LD")

d2a <-
  prolific_pred %>% filter(MONTHS.TYPE == "ALL")
d2b <-
  prolific_points %>% filter(MONTHS.TYPE == "ALL")

d3a <-
  prolific_pred %>% filter(MONTHS.TYPE == "ALL")
d3b <-
  prolific_points %>% filter(MONTHS.TYPE == "ALL")


p1 <-
  ggplot(d1a, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(
    values = c("#8F85C1", "#A3383C"),
    name = "Species category",
    labels = c("Rural", "Urban")
  ) +
  scale_y_continuous(limits = c(0.1, 1.0)) +
  labs(x = "", y = "Reporting frequency") +
  # data points
  geom_point(
    data = d1b,
    aes(group = PATH.GROUP),
    size = 2,
    alpha = 0.25,
    position = position_dodge(0.25)
  ) +
  geom_path(
    data = d1b,
    aes(
      x = as.numeric(M.YEAR),
      y = PRED,
      group = PATH.GROUP
    ),
    size = 1,
    alpha = 0.15,
    position = position_dodge(0.25)
  ) +
  # main points
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.3)

p2 <-
  ggplot(d2a, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(
    values = c("#8F85C1", "#A3383C"),
    name = "Species category",
    labels = c("Rural", "Urban")
  ) +
  scale_y_continuous(limits = c(0.1, 1.0)) +
  labs(x = "Migratory year", y = "") +
  # data points
  geom_point(
    data = d2b,
    aes(group = PATH.GROUP),
    size = 2,
    alpha = 0.25,
    position = position_dodge(0.25)
  ) +
  geom_path(
    data = d2b,
    aes(
      x = as.numeric(M.YEAR),
      y = PRED,
      group = PATH.GROUP
    ),
    size = 1,
    alpha = 0.15,
    position = position_dodge(0.25)
  ) +
  # main points
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.3)

p3 <-
  ggplot(d3a, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(
    values = c("#8F85C1", "#A3383C"),
    name = "Species category",
    labels = c("Rural", "Urban")
  ) +
  scale_y_continuous(limits = c(0.1, 1.0)) +
  labs(x = "", y = "") +
  # data points
  geom_point(
    data = d1b,
    aes(group = PATH.GROUP),
    size = 2,
    alpha = 0.25,
    position = position_dodge(0.25)
  ) +
  geom_path(
    data = d1b,
    aes(
      x = as.numeric(M.YEAR),
      y = PRED,
      group = PATH.GROUP
    ),
    size = 1,
    alpha = 0.15,
    position = position_dodge(0.25)
  ) +
  # main points
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.3)

(p1 |
    p2 | p3) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(
    legend.position = "bottom",
    plot.tag.position = c(0, 1),
    axis.text.x = element_text(
      colour = (month_lab %>% distinct(M.YEAR, COL3))$COL3,
      angle = 45,
      vjust = 0.5
    )
  )


```

# Discussion

## Did the pandemic change reporting patterns of urban and rural species?

### Did birders report more urban and fewer rural species?

Nevertheless, if this fine-scale urban bias also resulted in complementary changes in species reporting patterns, this would impact the utility of data from these years. Indeed, in some states, birders reported urban species slightly more frequently during the pandemic years than before, but this change remained even after the pandemic, suggesting some other cause for the change (Fig. \@ref(fig:overall-212m)B3). In other states, there was no evidence of higher urban species reporting during the pandemic. Thus, the evidence of directional change was not consistent and cannot be clearly linked to the pandemic.

Importantly, we found no clear, consistent and directional evidence for birders reporting rural species less frequently (Fig. \@ref(fig:overall-212m)). Moreover, the changes in urban and rural species reporting across all locations are absent when birding effort is accounted for (Fig. \@ref(fig:prolific)), suggesting that these changes are driven simply by the differential birding in urban and rural areas, and not by inherent changes in bird abundance or behaviour. The highly variable species-level changes in reporting, caused by differential sampling and spatial bias, can be minimised using spatial subsampling methods.

## Did bird abundance change?

Many studies now suggest that survival and even behavioural traits (e.g., song amplitude and frequency, time of vocal activity, etc.) of birds are affected by human activity, potentially leading to an increase in bird abundance or detectability when human activity is reduced. Indeed, during the period of most stringent COVID-19 restrictions in USA, although a few species showed a decline, most increased in abundance in areas that saw a decrease in human activity [@schrimpf2021]. In contrast, species tightly linked to human presence and activity like the Feral Pigeon showed lower abundances [@soh2021]. Behavioural traits of birds can also change rapidly in response to environmental changes [@derryberry2020; @sanderfoot2022; @hentati-sundberg2021; @gordo2021; but see @seress2021].

However, we found no clear evidence for bird abundance changing during the pandemic (Fig. \@ref(fig:prolific)). This might be because a few months of reduced human activity is insufficient to cause changes in large-scale patterns of bird behaviour or occurrence [see @seress2021]. A few individual locations did separately show signs of increased or decreased reporting of urban or rural species during the peak pandemic months, but any pattern disappeared when averaged across space (Table \@ref(tab:metric-summ)). Moreover, any location-specific pattern became muted when considering entire years, or when excluding the peak pandemic months. In conclusion, there is no consistent evidence of birds being affected by the pandemic, positively or negatively, artefactually or otherwise.

## What happened during the peak pandemic months?

... Additionally, the changes that the pandemic brought to human life did not lead to a prominent change at the large scale in how different bird species were reported.

<!--  -->

If so, this would signal some behavioural (changes in detectability or boldness) or population-level (true abundance, occupancy, etc.) changes associated with the pandemic. Not only would this illuminate the impact that human activity has on birds in India, but it would also show that birds are plastic and can respond to environmental changes at the short time scale.

We know that there was a long-lasting (relative to other pandemic-induced changes) urban bias in eBirding in India during the pandemic years [@thrikkadeeriCOVIDp1]. This difference in effort between urban and rural areas could theoretically, depending on the scale, result in higher apparent reporting frequencies for species largely restricted to urban areas. In this case, any adverse effects of the reduced human activity during the peak pandemic months would be cancelled out by this overinflated reporting of urban species, whereas any increases in urban bird detectability or abundance would become even more prominent. On the other hand, if comparisons between urban and rural species were made while keeping effort constant, any difference in reporting frequencies related to real changes in bird behaviour or abundance should become unambiguously evident.

"Disentangling true patterns from potential confounds"

# Conclusion

Importantly, there was no clear evidence of bird reporting having changed due to the pandemic, as a consequence of changes in either birder behaviour or bird abundance. Therefore, the eBird dataset remains valuable, with only certain statistical and scale considerations to be kept in mind.


# References

::: {#refs}
:::

\newpage

Supplementary Online Material (SOM) is available.
