# Supplementary Material

## Methods

### Prolific birders {#prolific-def}

An eBirder was considered "prolific" if they had uploaded, from a single location, at least 800 qualifying checklists over the course of the four years being analysed (sufficient effort), as well as a minimum of 15 qualifying checklists per month in at least ten months of a year in all four years of the study (consistent effort)---thus translating to about 200 qualifying checklists per year. Since the observer remains constant, and the observer is sufficiently experienced (as inferred from their prolificacy), we assume that detection rates of birds in the location also remains similar over the years (given the short time period).

### Model specifications {#modelspecs}

GLMMs for species reporting across all locations were fit for each species and for different periods ("lockdown", "non-lockdown", "all") separately. These models, with detection (yes or no) as the response variable, were specified as follows:

$$detection \sim year*month + month:length + (1/cell)$$

Models for species reporting at locations with consistent effort were fit for each locations and for different periods ("lockdown", "non-lockdown", "all") separately. The models were specified as follows:

$$detection \sim year + year:month + month:length + category + (1/species)$$

where $length$ refers to the median number of species reported in the respective month, $category$ is the species category and $species$ is the name of the species.

## Figures

### Overall bird reporting

```{r overall-non1, fig.cap="Calculated monthly reporting frequencies for urban (A) and rural (B) species in Bengaluru city. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours, and the peak pandemic months are shown in red text.", out.width="100%", fig.width=8, fig.height=13, dpi=300, dev="jpeg"}

load("00_outputs/b01_overall_s1a_m.RData")
d1 <- data_a %>% ungroup() %>% filter(SP.CATEGORY == "U") %>% 
  left_join(month_lab) %>% arrange(M.YEAR, MONTH.LAB)
d2 <- data_a %>% ungroup() %>% filter(SP.CATEGORY == "R") %>% 
  left_join(month_lab) %>% arrange(M.YEAR, MONTH.LAB)

p1 <- ggplot(d1, aes(MONTH.LAB, REP.FREQ, colour = M.YEAR)) +
  geom_point(size = 1.5, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 0.75, width = 0.75, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 2, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Month", y = "Reporting frequency") 

p2 <- ggplot(d2, aes(MONTH.LAB, REP.FREQ, colour = M.YEAR)) +
  geom_point(size = 1.5, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 0.75, width = 0.75, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 2, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Month", y = "Reporting frequency")

(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(6, 0.5, 6)) +
  plot_annotation(tag_levels = "A") &
  theme_classic(base_size = 10) &
  theme(strip.text = element_text(size = 5),
        legend.position = "bottom",
        axis.text.x = element_text(colour = (d1 %>% distinct(MONTH.LAB, COL1))$COL1))

```

```{r overall-non2, fig.cap="Calculated monthly reporting frequencies for urban (A) and rural (B) species in Karnataka state. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours, and the peak pandemic months are shown in red text.", out.width="100%", fig.width=8, fig.height=13, dpi=300, dev="jpeg"}

d1 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "U") %>% 
  left_join(month_lab) %>% arrange(M.YEAR, MONTH.LAB)
d2 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "R") %>% 
  left_join(month_lab) %>% arrange(M.YEAR, MONTH.LAB)

p1 <- ggplot(d1, aes(MONTH.LAB, REP.FREQ, colour = M.YEAR)) +
  geom_point(size = 1.5, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 0.75, width = 0.75, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 2, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Month", y = "Reporting frequency") 

p2 <- ggplot(d2, aes(MONTH.LAB, REP.FREQ, colour = M.YEAR)) +
  geom_point(size = 1.5, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 0.75, width = 0.75, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 2, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Month", y = "Reporting frequency")

(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(6, 0.5, 6)) +
  plot_annotation(tag_levels = "A") &
  theme_classic(base_size = 10) &
  theme(strip.text = element_text(size = 5),
        legend.position = "bottom",
        axis.text.x = element_text(colour = (d1 %>% distinct(MONTH.LAB, COL1))$COL1))

```

```{r overall-non3, fig.cap="Calculated annual reporting frequencies for urban (A) and rural (B) species in Bengaluru city. Points represent means while error bars show 95% confidence intervals, and the peak pandemic years are shown in red text.", out.width="100%", fig.width=8, fig.height=11, dpi=300, dev="jpeg"}

load("00_outputs/b01_overall_s1b_y.RData")
d1 <- data_a %>% ungroup() %>% filter(SP.CATEGORY == "U") %>% 
  left_join(month_lab) %>% arrange(M.YEAR)
d2 <- data_a %>% ungroup() %>% filter(SP.CATEGORY == "R") %>% 
  left_join(month_lab) %>% arrange(M.YEAR)

p1 <- ggplot(d1, aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.5, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 4, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Migratory year", y = "Reporting frequency") 

p2 <- ggplot(d2, aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.5, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 4, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Migratory year", y = "Reporting frequency")

(p1 / p2) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme_classic(base_size = 10) &
  theme(strip.text = element_text(size = 7),
        legend.position = "bottom",
        axis.text.x = element_text(colour = (d1 %>% distinct(M.YEAR, COL3))$COL3))

```

```{r overall-non4, fig.cap="Calculated annual reporting frequencies for urban (A) and rural (B) species in Karnataka state. Points represent means while error bars show 95% confidence intervals, and the peak pandemic years are shown in red text.", out.width="100%", fig.width=8, fig.height=11, dpi=300, dev="jpeg"}

d1 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "U") %>% 
  left_join(month_lab) %>% arrange(M.YEAR)
d2 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "R") %>% 
  left_join(month_lab) %>% arrange(M.YEAR)

p1 <- ggplot(d1, aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.5, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 4, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Migratory year", y = "Reporting frequency") 

p2 <- ggplot(d2, aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.5, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 4, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Migratory year", y = "Reporting frequency")

(p1 / p2) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme_classic(base_size = 10) &
  theme(strip.text = element_text(size = 7),
        legend.position = "bottom",
        axis.text.x = element_text(colour = (d1 %>% distinct(M.YEAR, COL3))$COL3))

```

```{r overall-non5, fig.cap="Calculated monthly reporting frequencies for urban (A) and rural (B) species in Kochi city. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours, and the peak pandemic months are shown in red text.", out.width="100%", fig.width=8, fig.height=13, dpi=300, dev="jpeg"}

load("00_outputs/b01_overall_s2a_m.RData")
d1 <- data_a %>% ungroup() %>% filter(SP.CATEGORY == "U") %>% 
  left_join(month_lab) %>% arrange(M.YEAR, MONTH.LAB)
d2 <- data_a %>% ungroup() %>% filter(SP.CATEGORY == "R") %>% 
  left_join(month_lab) %>% arrange(M.YEAR, MONTH.LAB)

p1 <- ggplot(d1, aes(MONTH.LAB, REP.FREQ, colour = M.YEAR)) +
  geom_point(size = 1.5, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 0.75, width = 0.75, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 2, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Month", y = "Reporting frequency") 

p2 <- ggplot(d2, aes(MONTH.LAB, REP.FREQ, colour = M.YEAR)) +
  geom_point(size = 1.5, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 0.75, width = 0.75, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 2, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Month", y = "Reporting frequency")

(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(6, 0.5, 6)) +
  plot_annotation(tag_levels = "A") &
  theme_classic(base_size = 10) &
  theme(strip.text = element_text(size = 5),
        legend.position = "bottom",
        axis.text.x = element_text(colour = (d1 %>% distinct(MONTH.LAB, COL1))$COL1))

```

```{r overall-non6, fig.cap="Calculated monthly reporting frequencies for urban (A) and rural (B) species in Kerala state. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours, and the peak pandemic months are shown in red text.", out.width="100%", fig.width=8, fig.height=13, dpi=300, dev="jpeg"}

d1 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "U") %>% 
  left_join(month_lab) %>% arrange(M.YEAR, MONTH.LAB)
d2 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "R") %>% 
  left_join(month_lab) %>% arrange(M.YEAR, MONTH.LAB)

p1 <- ggplot(d1, aes(MONTH.LAB, REP.FREQ, colour = M.YEAR)) +
  geom_point(size = 1.5, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 0.75, width = 0.75, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 2, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Month", y = "Reporting frequency") 

p2 <- ggplot(d2, aes(MONTH.LAB, REP.FREQ, colour = M.YEAR)) +
  geom_point(size = 1.5, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 0.75, width = 0.75, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 2, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Month", y = "Reporting frequency")

(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(6, 0.5, 6)) +
  plot_annotation(tag_levels = "A") &
  theme_classic(base_size = 10) &
  theme(strip.text = element_text(size = 5),
        legend.position = "bottom",
        axis.text.x = element_text(colour = (d1 %>% distinct(MONTH.LAB, COL1))$COL1))

```

```{r overall-non7, fig.cap="Calculated annual reporting frequencies for urban (A) and rural (B) species in Kochi city. Points represent means while error bars show 95% confidence intervals, and the peak pandemic years are shown in red text.", out.width="100%", fig.width=8, fig.height=11, dpi=300, dev="jpeg"}

load("00_outputs/b01_overall_s2b_y.RData")
d1 <- data_a %>% ungroup() %>% filter(SP.CATEGORY == "U") %>% 
  left_join(month_lab) %>% arrange(M.YEAR)
d2 <- data_a %>% ungroup() %>% filter(SP.CATEGORY == "R") %>% 
  left_join(month_lab) %>% arrange(M.YEAR)

p1 <- ggplot(d1, aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.5, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 4, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Migratory year", y = "Reporting frequency") 

p2 <- ggplot(d2, aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.5, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 4, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Migratory year", y = "Reporting frequency")

(p1 / p2) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme_classic(base_size = 10) &
  theme(strip.text = element_text(size = 7),
        legend.position = "bottom",
        axis.text.x = element_text(colour = (d1 %>% distinct(M.YEAR, COL3))$COL3))

```

```{r overall-non8, fig.cap="Calculated annual reporting frequencies for urban (A) and rural (B) species in Kerala state. Points represent means while error bars show 95% confidence intervals, and the peak pandemic years are shown in red text.", out.width="100%", fig.width=8, fig.height=11, dpi=300, dev="jpeg"}

d1 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "U") %>% 
  left_join(month_lab) %>% arrange(M.YEAR)
d2 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "R") %>% 
  left_join(month_lab) %>% arrange(M.YEAR)

p1 <- ggplot(d1, aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.5, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 4, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Migratory year", y = "Reporting frequency") 

p2 <- ggplot(d2, aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.5, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 4, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Migratory year", y = "Reporting frequency")

(p1 / p2) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme_classic(base_size = 10) &
  theme(strip.text = element_text(size = 7),
        legend.position = "bottom",
        axis.text.x = element_text(colour = (d1 %>% distinct(M.YEAR, COL3))$COL3))

```

```{r overall-non9, fig.cap="Calculated monthly reporting frequencies for urban (A) and rural (B) species in Pune city. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours, and the peak pandemic months are shown in red text.", out.width="100%", fig.width=8, fig.height=13, dpi=300, dev="jpeg"}

load("00_outputs/b01_overall_s3a_m.RData")
d1 <- data_a %>% ungroup() %>% filter(SP.CATEGORY == "U") %>% 
  left_join(month_lab) %>% arrange(M.YEAR, MONTH.LAB)
d2 <- data_a %>% ungroup() %>% filter(SP.CATEGORY == "R") %>% 
  left_join(month_lab) %>% arrange(M.YEAR, MONTH.LAB)

p1 <- ggplot(d1, aes(MONTH.LAB, REP.FREQ, colour = M.YEAR)) +
  geom_point(size = 1.5, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 0.75, width = 0.75, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 2, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Month", y = "Reporting frequency") 

p2 <- ggplot(d2, aes(MONTH.LAB, REP.FREQ, colour = M.YEAR)) +
  geom_point(size = 1.5, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 0.75, width = 0.75, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 2, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Month", y = "Reporting frequency")

(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(6, 0.5, 6)) +
  plot_annotation(tag_levels = "A") &
  theme_classic(base_size = 10) &
  theme(strip.text = element_text(size = 5),
        legend.position = "bottom",
        axis.text.x = element_text(colour = (d1 %>% distinct(MONTH.LAB, COL1))$COL1))

```

```{r overall-non10, fig.cap="Calculated monthly reporting frequencies for urban (A) and rural (B) species in Maharashtra state. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours, and the peak pandemic months are shown in red text.", out.width="100%", fig.width=8, fig.height=13, dpi=300, dev="jpeg"}

d1 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "U") %>% 
  left_join(month_lab) %>% arrange(M.YEAR, MONTH.LAB)
d2 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "R") %>% 
  left_join(month_lab) %>% arrange(M.YEAR, MONTH.LAB)

p1 <- ggplot(d1, aes(MONTH.LAB, REP.FREQ, colour = M.YEAR)) +
  geom_point(size = 1.5, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 0.75, width = 0.75, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 2, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Month", y = "Reporting frequency") 

p2 <- ggplot(d2, aes(MONTH.LAB, REP.FREQ, colour = M.YEAR)) +
  geom_point(size = 1.5, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 0.75, width = 0.75, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 2, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Month", y = "Reporting frequency")

(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(6, 0.5, 6)) +
  plot_annotation(tag_levels = "A") &
  theme_classic(base_size = 10) &
  theme(strip.text = element_text(size = 5),
        legend.position = "bottom",
        axis.text.x = element_text(colour = (d1 %>% distinct(MONTH.LAB, COL1))$COL1))

```

```{r overall-non11, fig.cap="Calculated annual reporting frequencies for urban (A) and rural (B) species in Pune city. Points represent means while error bars show 95% confidence intervals, and the peak pandemic years are shown in red text.", out.width="100%", fig.width=8, fig.height=11, dpi=300, dev="jpeg"}

load("00_outputs/b01_overall_s3b_y.RData")
d1 <- data_a %>% ungroup() %>% filter(SP.CATEGORY == "U") %>% 
  left_join(month_lab) %>% arrange(M.YEAR)
d2 <- data_a %>% ungroup() %>% filter(SP.CATEGORY == "R") %>% 
  left_join(month_lab) %>% arrange(M.YEAR)

p1 <- ggplot(d1, aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.5, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 4, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Migratory year", y = "Reporting frequency") 

p2 <- ggplot(d2, aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.5, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 4, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Migratory year", y = "Reporting frequency")

(p1 / p2) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme_classic(base_size = 10) &
  theme(strip.text = element_text(size = 7),
        legend.position = "bottom",
        axis.text.x = element_text(colour = (d1 %>% distinct(M.YEAR, COL3))$COL3))

```

```{r overall-non12, fig.cap="Calculated annual reporting frequencies for urban (A) and rural (B) species in Maharashtra state. Points represent means while error bars show 95% confidence intervals, and the peak pandemic years are shown in red text.", out.width="100%", fig.width=8, fig.height=11, dpi=300, dev="jpeg"}

d1 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "U") %>% 
  left_join(month_lab) %>% arrange(M.YEAR)
d2 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "R") %>% 
  left_join(month_lab) %>% arrange(M.YEAR)

p1 <- ggplot(d1, aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.5, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 4, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Migratory year", y = "Reporting frequency") 

p2 <- ggplot(d2, aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.5, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 4, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Migratory year", y = "Reporting frequency")

(p1 / p2) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme_classic(base_size = 10) &
  theme(strip.text = element_text(size = 7),
        legend.position = "bottom",
        axis.text.x = element_text(colour = (d1 %>% distinct(M.YEAR, COL3))$COL3))

```

```{r overall-non13, fig.cap="Calculated monthly reporting frequencies for urban (A) and rural (B) species in Assam state. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours, and the peak pandemic months are shown in red text.", out.width="100%", fig.width=8, fig.height=13, dpi=300, dev="jpeg"}

load("00_outputs/b01_overall_s4a_m.RData")
d1 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "U") %>% 
  left_join(month_lab) %>% arrange(M.YEAR, MONTH.LAB)
d2 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "R") %>% 
  left_join(month_lab) %>% arrange(M.YEAR, MONTH.LAB)

p1 <- ggplot(d1, aes(MONTH.LAB, REP.FREQ, colour = M.YEAR)) +
  geom_point(size = 1.5, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 0.75, width = 0.75, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 2, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Month", y = "Reporting frequency") 

p2 <- ggplot(d2, aes(MONTH.LAB, REP.FREQ, colour = M.YEAR)) +
  geom_point(size = 1.5, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 0.75, width = 0.75, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 2, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Month", y = "Reporting frequency")

(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(6, 0.5, 6)) +
  plot_annotation(tag_levels = "A") &
  theme_classic(base_size = 10) &
  theme(strip.text = element_text(size = 5),
        legend.position = "bottom",
        axis.text.x = element_text(colour = (d1 %>% distinct(MONTH.LAB, COL1))$COL1))

```

```{r overall-non14, fig.cap="Calculated annual reporting frequencies for urban (A) and rural (B) species in Assam state. Points represent means while error bars show 95% confidence intervals, and the peak pandemic years are shown in red text.", out.width="100%", fig.width=8, fig.height=11, dpi=300, dev="jpeg"}

load("00_outputs/b01_overall_s4b_y.RData")
d1 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "U") %>% 
  left_join(month_lab) %>% arrange(M.YEAR)
d2 <- data_b %>% ungroup() %>% filter(SP.CATEGORY == "R") %>% 
  left_join(month_lab) %>% arrange(M.YEAR)

p1 <- ggplot(d1, aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.5, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 4, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Migratory year", y = "Reporting frequency") 

p2 <- ggplot(d2, aes(M.YEAR, REP.FREQ)) +
  geom_point(size = 2, position = position_dodge(0.8)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1, width = 0.5, position = position_dodge(0.8)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year") +
  facet_wrap(~ COMMON.NAME, dir = "h", ncol = 4, 
             strip.position = "left", scales = "free_y") +
  labs(x = "Migratory year", y = "Reporting frequency")

(p1 / p2) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme_classic(base_size = 10) &
  theme(strip.text = element_text(size = 7),
        legend.position = "bottom",
        axis.text.x = element_text(colour = (d1 %>% distinct(M.YEAR, COL3))$COL3))

```

```{r overall-10m, fig.cap="Modelled species reporting frequencies for all ten months excluding April and May, for the states of Karnataka (A), Kerala (B), Maharashtra (C) and Assam (D). Points represent means while error bars show 95% confidence intervals. Values for urban and rural species are shown in maroon and purple respectively, and the pandemic years are shown in red text.", out.width="100%", fig.width=10, fig.height=6, dpi=300, dev="jpeg"}

load("00_outputs/b01_overall_s1c_model.RData")
d1 <- birds_pred %>% filter(MONTHS.TYPE == "NL")
load("00_outputs/b01_overall_s2c_model.RData")
d2 <- birds_pred %>% filter(MONTHS.TYPE == "NL")
load("00_outputs/b01_overall_s3c_model.RData")
d3 <- birds_pred %>% filter(MONTHS.TYPE == "NL")
load("00_outputs/b01_overall_s4c_model.RData")
d4 <- birds_pred %>% filter(MONTHS.TYPE == "NL") 

p1 <- ggplot(d1, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(values = c("#8F85C1", "#A3383C"),
                     name = "Species category",
                     labels = c("Rural", "Urban")) +
  labs(x = "", y = "Reporting frequency") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) 


p2 <- ggplot(d2, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(values = c("#8F85C1", "#A3383C"),
                     name = "Species category",
                     labels = c("Rural", "Urban")) +
  labs(x = "Migratory year", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) 

p3 <- ggplot(d3, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(values = c("#8F85C1", "#A3383C"),
                     name = "Species category",
                     labels = c("Rural", "Urban")) +
  labs(x = "", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) 

p4 <- ggplot(d4, aes(M.YEAR, PRED, col = SP.CATEGORY)) +
  scale_color_manual(values = c("#8F85C1", "#A3383C"),
                     name = "Species category",
                     labels = c("Rural", "Urban")) +
  labs(x = "Migratory year", y = "") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) 


(p1 | p2 | p3 | p4) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        plot.tag.position = c(0, 1),
        axis.text.x = element_text(colour = (month_lab %>% distinct(M.YEAR, COL3))$COL3,
                                   angle = 45, vjust = 0.5))

```

## Tables
