---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Abstract {.unnumbered}

Many parts of the world lack the large and coordinated volunteer networks required for systematic monitoring of bird populations. In these regions, citizen science (CS) programmes offer an alternative with their semi-structured data, but the utility of these data is contingent on how, where, and how comparably birdwatchers watch birds, year on year. Trends inferred from the data can be confounded during years when birdwatchers may behave differently, such as during the COVID-19 pandemic. We wanted to ascertain how the data uploaded from India to one such CS platform, eBird, were impacted by this deadly global pandemic. To understand whether eBird data from the pandemic years in India are comparable to data from adjacent years, we explored several characteristics of the data, such as how often people watched birds in groups or at public locations, at multiple spatial and temporal scales. We found that the volume of data generated increased during the pandemic years 2020--21 compared to 2019. Data characteristics changed largely only during the peak pandemic months (April--May 2020 and April--May 2021) associated with high fatality rates and strict lockdowns, possibly due to the decreased human mobility and social interaction. The data from the remainder of these restrictive years remained similar to those of the adjacent years, thereby reducing the impact of the aberrant peak months on any annual inference. Our findings show that birdwatchers in India as contributors to CS were resilient to pandemic-induced changes, and that the effects of the pandemic on birdwatching effort and birdwatcher behaviour are scale- and context-dependent. In summary, eBird data from the pandemic years in India remain useful for abundance trend estimation and similar large-scale applications, but will benefit from preliminary data quality checks when utilised at a fine scale.

# Keywords {.unnumbered}

Birdwatcher behaviour, COVID-19 pandemic, data characteristics, eBird, effort

# Lay summary {.unnumbered}

-   Citizen science platforms like eBird comprise vast repositories of data generated by casual birdwatching.
-   Such data are vital to understanding bird population trends, but large changes in how birdwatchers watch birds from year to year can be very challenging to account for when estimating trends.
-   Given the impact of the COVID-19 pandemic on our everyday lives, we investigated whether it also impacted the way people reported birds, thereby potentially reducing the value of the data for trend analyses.
-   We analysed data uploaded to eBird from India, and found that the impacts of the pandemic on how these data were collected were largely restricted to April and May in 2020. During these months that coincided with the greatest health impacts and movement restrictions, birdwatchers made fewer travelling lists, and made fewer lists in groups and at public locations.
-   Birdwatchers were resilient; they bounced back soon after these difficult periods, and started birdwatching like they had done before the pandemic.
-   Because the pandemic impact on characteristics of eBird data from India was limited to short periods, we conclude that the data still remain useful for analyses of bird abundance trends.

\newpage

# Introduction

## Birds and citizen science

Citizen science (CS) today is a practice that engages volunteers from the general public---most of whom are not trained as scientists---in various processes of scientific enquiry [@mckinley2017], from collecting to analysing or interpreting data [@silvertown2009]. CS programmes are particularly effective when they revolve around popular hobbies or activities that people enjoy, such as birdwatching. Birdwatching (hereafter birding) is a nature-based activity that is enjoyed by a large number of people around the world, making CS projects involving birds especially appealing to the public. Birders, as a matter of practice, make "checklists" of the bird species they observe during birding sessions. By pooling this data into a common public database, numerous birders with varying degrees of experience---from trained researchers to seasoned birders to laypersons---are able to contribute to the scientific study of birds.

Studies that assess the health of birds in a region, such as "State of Birds" assessments, are entirely reliant on trends derived from CS programmes. These are mostly systematic monitoring programmes [see @miller-rushing2012] that follow fixed protocols year after year. This makes it straightforward to estimate bird population sizes and derive their trends over years. These systematic CS programmes that engage birders are resource- and volunteer-intensive [e.g., @greenwood1995; @sauer2013a; @birdlifeaustralia2015; @praveen2022; @swissornithologicalinstitute2023] and are not yet feasible at large scales in many parts of the world. However, in some of these places, CS projects with "semi-structured" protocols [@pocock2017; @dicecco2021] are popular, such as eBird [@sullivan2014] which has been widely used by birders in India since 2014 (see <https://ebird.org/india/about> for more details). Like the structured data from systematic monitoring programmes, semi-structured data on eBird also enable large-scale trend assessments [@neate-clegg2020; @kittelberger2023], such as the State of India's Birds 2023 report [@soib2023] which analysed abundance trends, range size, and conservation status of 942 Indian bird species.

A drawback of the semi-structured data on eBird is that checklists vary in aspects such as effort and the data are not homogeneous in space and time (as there is no single fixed protocol), thereby creating noticeable biases in the data [@johnston2021]. Such biases largely arise from variation in birder behaviour, making the data difficult to use to inform yearly abundance trends. Although some recent analyses show that appropriate statistical adjustments can reduce these biases such that results closely match those from systematic data [@johnston2021], biases in semi-structured data become especially pronounced when there are large-scale changes in human (birder) behaviour, such as presumably during the COVID-19 pandemic (see next). Given the widespread utility of semi-structured CS data today for conservation, an urgent question is whether the pandemic has dramatically altered the underlying observation process, thereby making it difficult to use eBird data from these years in combination with the rest of the dataset for estimating yearly abundance trends.

## Birding in the time of COVID-19

COVID-19 is a highly infectious disease that resulted in the deadliest pandemic of the 21st century, with a death toll of more than 6.9 million (around 0.1% of the global population$;$ @worldometer2023). Governments across the world therefore took various precautionary and preventive measures, and implemented restrictions such as nation- or state-wide lockdowns and curfews, quarantines and self-isolation, and shutdown of non-essential public services and activities [see @hale2020]. Given that the eBird enterprise hinges entirely on public participation, this widespread public upheaval might have changed the very nature of eBird data during this period.

### How much birding?

The COVID-19 pandemic has generally posed a big challenge for CS programmes [e.g., @hochachka2021; @kishimoto2021]. Social and travel restrictions, in combination with physical and mental distress [@li2020; @vanbavel2020], may have led to a decline in the amount of hobbyist as well as systematic birding, and therefore in the quantity of data generated. Some CS projects did indeed show evidence of reduced participation and data generation [e.g., @kishimoto2021], but others in fact witnessed an increase in data generated [e.g., @basile2021]. Outdoor recreational activities such as birding are known to restore attention and play a vital role in improving physical and mental health [@weng2014]. The pandemic brought an increased desire in people to connect with the natural world [@marsh2021], as well as active promotion and encouragement of nature-related activities such as birding [e.g., @birdcountindia2020]. Thus, many turned to birding as their activity of choice [@randler2020].

### What kinds of birding?

Regardless of how much more or less data were generated during the pandemic, the characteristics of these data could also have changed. For example, @randler2020 found differences in aspects of birder behaviour, such as who was birding and with whom, as well as where and when birding (and thus data generation) was happening. Restricted mobility led to data being less spread out in space, with observations being concentrated in urbanised locations [@vardi2021] and covering a lower overall diversity of habitats [@hochachka2021], thus reducing their informational value [@zhang2020]. On the other hand, @kishimoto2021 found that the restrictions resulted in lower clustering of data, potentially because birders were limited to their erstwhile under-surveyed neighbourhoods and could not upload as many observations as usual from birding hotspots (either urban or rural).

### Scale and generalisability

@hochachka2021 studied changes in birder behaviour during a month-long lockdown across three countries: USA (states of California and New York), Spain and Portugal. The changes were not as substantial as expected from the magnitude of change to daily life. Notably, they found differential responses to the pandemic in the respective birding communities, and suggested that such differences are likely to exist amongst geographical units across the world, thus cautioning against a global generalisation of how the pandemic influenced birding and birder behaviour. Moreover, such studies of pandemic impacts have been limited to the short temporal scale of individual months, and have not explored the time scales of individual years relevant for large-scale analyses of abundance trends, such as those contained in the State of India's Birds report [@soib2023].

## Our aim

India's rapid recent progress in volunteer-driven bird monitoring is both unique and pioneering, in the context of the Global South where large-scale resource-intensive systematic surveys may not be possible in the near future. The State of India's Birds reports [@soib2020; @soib2023] have demonstrated that semi-structured monitoring can be immensely valuable for assessing the health of a region's birds. Because such assessments are intended to inform conservation policy, there is an urgent need to ensure robustness of the data from birding to disruptive events like the COVID-19 pandemic or even potential future climate crises.

This approach of examining the observation process that generates the data is more appropriate than direct comparisons of trends estimated from these data, where there is no clear benchmark to compare patterns against and thereby reach any meaningful conclusion. The observers (i.e., birders) in India and elsewhere in the Global South might behave differently from those in North America and Europe, and therefore might have responded differently to the COVID-19 pandemic, due to different socioeconomic realities, aspirations, and politics. For instance, adherence to social distancing measures rapidly declined across socioeconomic strata in India after the initial months of lockdowns [@schaner2022], likely because the epidemiological and welfare value of such measures in countries such as India is often lower than in higher-income countries due to the economic sacrifices involved [@barnett-howell2021].

In this study, we sought a comprehensive understanding of how birders in India may have changed their birding behaviour during the pandemic, and what this may indicate about the utility of the generated data for applications such as abundance trend estimation. We specifically focused on the use of the platform eBird in India, and investigated how the nature of eBird data changed during the pandemic: both the amount of data generated as well as how birders collected the data. We then tried to ascertain the extent of these changes in data characteristics, by assessing whether they were limited to the peak lockdown months or applied to entire years. If the changes applied across entire years, the data would not be easily usable in yearly trend estimation. Similarly, we examined the utility of the data for trend estimation at small and large spatial scales (states vs country). Finally, we discuss the implications of our results on the utility of eBird data for abundance trend estimation in India, and also outline guidelines for using these data for the same.

# Methods

## Data preparation {#dataprepmaintext}

We used the *relMay-2022* version of the publicly available [eBird Basic Dataset (EBD)](https://ebird.org/data/download/ebd) for India, containing all data uploaded until 31 May 2022 [downloaded on 15 June 2022, follows taxonomy in @clements2021]. We combined [sensitive species data](https://ebird.org/india/news/ebird-sensitive-species/), which is not included in the public download, after requesting this separately from eBird. Before analysis, we applied a set of filters to standardise the dataset (Section \@ref(dataprep)), so as to prevent any extreme and outlying cases from skewing the main patterns. Our final eBird dataset contained `r scales::label_comma()(tot_uniq_lists)` unique "qualifying checklists", i.e., lists that passed the filters.

All analyses presented in this study use "migratory year" (MY; June of one calendar year to May of the next; see Section \@ref(dataprep) for rationale) as a unit time period and we analysed data over a period of four migratory years, from 1 June 2018 to 31 May 2022 (Fig. \@ref(fig:timeline)). We consider a relatively fine temporal scale of months in order to account for within-year variations in birder behaviour. Moreover, in India severe COVID-19 restrictions and lockdowns were limited to a few months in 2020 and 2021 [@theeconomictimes2020; @theindianexpress2021], and this scale allows us to consider these patterns separately from patterns in other months of the "pandemic years" (the migratory years 2019--20 and 2020--21).

```{r timeline, fig.cap="Timeline of the study period. The spans of the four migratory years are shown in different colours, and pre-pandemic, pandemic and post-pandemic months are also signified. April and May of both pandemic years (shown in red text) are considered as the peak pandemic months.", fig.width=8, out.width="100%", dev="jpeg", dpi=300}

x <- month_lab %>% 
  mutate(YEAR = ifelse(MONTH %in% 6:12, 
                       as.numeric(as.character(M.YEAR)), 
                       as.numeric(as.character(M.YEAR)) + 1)) %>% 
  mutate(DATE = as.Date(paste(YEAR, MONTH, "01", sep = "-")))

y <- x %>%
  summarise(YEAR.RANGE = as.Date(seq(min(DATE), # buffers for placement on graph
                                     max(DATE) + months(1),
                                     by = "year"),
                                 origin = "1970-01-01")) %>% 
  # we don't need label for M.YEAR 2022
  rows_delete(tibble(YEAR.RANGE = as.Date("2022-06-01")))

# for rect
z <- x %>%
  group_by(M.YEAR) %>%
  summarise(DATE = DATE,
            START = min(DATE),
            END = if_else(M.YEAR == 2021, max(DATE), max(DATE) + months(1)))

# for COVID period labels
covid_lab <- read_csv("../../00_data/covid_classification.csv") %>% 
  mutate(COVID = ifelse(str_detect(COVID, "DUR"), "DUR", COVID),
         MONTH = factor(MONTH)) %>% 
  right_join(x) %>% 
  # selecting only start and end
  group_by(COVID) %>% 
  slice(c(1, n())) %>% 
  ungroup() %>% 
  arrange(DATE) %>% 
  distinct(COVID, DATE) %>% 
  group_by(COVID) %>% 
  mutate(TEMP = c("START", "END")) %>% 
  pivot_wider(names_from = "TEMP", values_from = "DATE") %>% 
  # label
  mutate(LABEL = case_when(COVID == "BEF" ~ "Pre-pandemic",
                           COVID == "DUR" ~ "Pandemic",
                           COVID == "AFT" ~ "Post-pandemic"))

ggplot(x, aes(x = DATE)) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.line.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line.y = element_blank(),
        legend.position = "top",
        legend.margin = margin(t = 0, unit = "cm")) +
  ylim(-0.15, 0.08) +
  # timeline
  geom_hline(yintercept = 0, colour = "black", size = 1.5) +
  # month labels
  geom_text(aes(x = DATE, y = -0.03, label = month(DATE, label = T)),
            angle = 90, colour = x$COL2, size = 3) +
  # M.YEAR categories
  geom_rect(data = z, 
            aes(xmin = START, xmax = END, ymin = 0.025, ymax = 0.075, fill = M.YEAR)) +
  scale_fill_manual(values = covid_palette, name = "Migratory year",
                    labels = unique(x$M.YEAR.LAB)) +
  geom_segment(data = z, 
               aes(x = START+days(1), xend = START+days(1), y = -0.02, yend = 0.025),
               linetype = "dashed", size = 0.75) +
  # COVID period labels
  geom_bracket(data = covid_lab, 
               aes(xmin = START, xmax = END, label = LABEL),
               y.position = -0.06, vjust = 3, label.size = 4, tip.length = -0.05)


```

We gridded the spatial extent of the country into `r glue("{grid_size_deg}\u00b0")` $\times$ `r glue("{grid_size_deg}\u00b0")` cells (`r min(grid_dimrange_km$X)`--`r max(grid_dimrange_km$X)` km $\times$ `r unique(grid_dimrange_km$Y)` km, depending on latitude) and used the latitude-longitude information associated with each of the qualifying checklists to link them with corresponding grid cells. We also classified the checklists as "urban" or "non-urban" (i.e., as being from urban or non-urban sites) using remote sensing data on land cover [MODIS Land Cover Type Product MCD12Q1, @friedl2019] with the help of the *luna* [@ghosh2020] package (see Section \@ref(UNUprep) for details). The spatial resolution of our grid is large (relative to practices elsewhere like North America), and we avoided associating eBird checklist locations with remotely sensed habitat variables, because of the significant inherent imprecision in where and how checklist locations are assigned (e.g., eBird hotspots often encompass large areas and numerous habitats).

## Analysis of data characteristics

To analyse changes in the nature of eBird data, we devised a suite of metrics (see below) that help gauge various characteristics of the data. These metrics were all calculated per month across the four years, and these values were compared in order to investigate changes in data characteristics over time. Except for data quantity (Section \@ref(dataquant)), all metrics were analysed at two spatial scales: for the whole country and for each of four states. The four states we considered in our study in order to highlight nuances were Karnataka, Kerala, Maharashtra and Assam. These states have relatively high eBirding activity and cover different regions of the country, but also differ in characteristics of the birding community as well as in response to the pandemic owing to differences in local governance, policy and general public perspective.

The metrics of birder behaviour (Section \@ref(birderbehav)) and spatial coverage (Section \@ref(spat-cov)) all have a common directionality: if the pandemic did indeed change these characteristics of the data, our expectation was declines in the values of each of these metrics from pre-pandemic levels. The metrics are described below, and the process of summarising them and interpreting their combined meaning is detailed in Section \@ref(composite).

### Data quantity {#dataquant}

To check whether the pandemic led to an increase or decrease in eBird data quantity, we explored two separate aspects of quantity: data volume and participation. We compared the number of qualifying lists and the number of active eBirders (those who uploaded at least one list) respectively for every month across our years of interest, using simple summaries and visualisations.

### Birder behaviour {#birderbehav}

1.  *Group birding*: Were there fewer instances of group birding during the pandemic? For this, we calculated the proportion of an observer's total qualifying lists that were instances of group birding, defined as lists with three or more observers as specified in the "party size" field.

2.  *Site fidelity*: Given the various restrictions on movement and travel, did birders selectively visit certain areas more often than they previously did, and therefore visit fewer areas overall? In other words, did birders show higher "site fidelity" during the pandemic? We calculated the number of grid cells visited by each birder, i.e., the inverse of site fidelity.

3.  *Hotspot birding*: eBird "hotspots" are public birding locations demarcated by eBird users to aggregate and display data from multiple birders [@ebird2022], and are different from "personal" birding locations which are usually visited by just one or few birders. Given the overall decline in public/social interactions during the pandemic, did people also tend to avoid birding in eBird hotspots? We calculated, out of an observer's total qualifying lists, the proportion that was from eBird hotspots.

4.  *Birding protocol*: "Traveling" and "Stationary" are the two most commonly followed protocols in eBird observations. Considering the restrictions on movement during the pandemic, we asked whether birders made fewer travelling lists (which we defined as lists using the "Traveling" protocol and covering $\gt$ 0.3 km), instead preferring stationary lists (from, say, their homes). We calculated, out of an observer's total qualifying lists, the proportion that was of travelling lists.

5.  *List distance*: Among the observations that followed the "Traveling" protocol, we asked whether travelling distances were shorter during the pandemic. Here, we simply used the distance travelled, measured as part of the checklist-level effort metadata.

6.  *List duration*: Did birders make shorter-duration checklists during the pandemic? Here, we used checklist duration, from the checklist-level effort metadata.

7.  *List length*: List length is defined as the total number of species observed in a checklist. This has been used as a metric of effort in some studies [e.g., @szabo2010; @isaac2014], and is often used as a substitute for list duration or distance in such analyses [e.g., @soib2023]. We used list length to answer, from a different angle, the question of whether checklists became shorter during the pandemic.

8.  *Urban bias*: A majority of eBird users are likely to be urban-dwelling, both in India and across the world, but often move far and wide for birding. Birders from urban areas may therefore account for a large proportion of the birding in non-urban areas as well. With the pandemic limiting movement, did urban areas see an increase in birding while non-urban areas saw the opposite trend? If so, this would have implications for how well these two areas are represented in the data compared to other years, potentially biasing analyses. We calculated the proportion of lists out of an observer's total qualifying lists that were from "urban" sites (as defined previously in Section \@ref(dataprepmaintext)) to estimate urban bias. For the composite analysis with the common expected directionality (Section \@ref(composite)), we used the additive inverse of this metric, i.e., non-urban bias.

### Spatial coverage {#spat-cov}

Given the restrictions on travel, a decrease in overall spatial coverage of the data from the pandemic was a major concern, because the region (whole country or one of four states) as a whole would be less well represented than in other years, thus biasing comparisons. We calculated the proportion of grid cells across the region of interest (country or states) that were covered in each time period. The information threshold for a grid cell to be considered "covered" was set to five qualifying checklists (per month per year).

### Modelling framework {#model-boot}

All metrics except data quantity were analysed using the formal mixed-effects model framework, mostly using Generalised Linear Mixed Models [GLMMs, see @bolker2009]. We used the "glmer()" function in the *lme4* package [@bates2015] to fit GLMMs. Observer identity defined by unique observer number was used as the random effect for the metrics of birder behaviour; spatial identity defined by grid cell ID was used as the random effect for spatial coverage. All GLMMs were fitted with binomial error distribution and cloglog link function, using BOBYQA optimiser:

$$metric \sim year + year:month + (1/observer)$$

$$metric \sim year + year:month + (1/cell)$$

where $year$ represents the migratory years.

Some metrics like site fidelity, list distance, list duration and list length were either whole-number or continuous numerical variables that did not include zero in the range of possible values. For these, since GLMMs with binomial or Poisson error distributions are not appropriate, we instead fitted LMMs (Linear Mixed Models, "lmer()" function in *lme4*) with the same fixed and random effects specifications and using BOBYQA optimiser, after first log-transforming (natural log) the response variables to approximate normality:

$$log(metric) \sim year + year:month + (1/observer)$$ For all models, model diagnostics were inspected using the *DHARMa* package [@hartig2022], to ensure that there were no major violations of model assumptions. We used the "bootMer()" function in *lme4* in order to test the significance of the fitted model results, by obtaining mean and CI estimates using semi-parametric bootstrapping. The bootstrapping was done with `r scales::label_comma()(1000)` simulations and with parallel core processing [with the help of packages *parallel*, @rcoreteam2021; and *VGAM*, @yee2010].

### Composite of modelled metrics {#composite}

The model results of the nine metrics of data characteristics were summarised as follows to give an understanding of the overall impact of the pandemic (and the scale of impact) on birding and birder behaviour in India. Model predictions for each metric (per month and year) were first transformed into proportional change relative to corresponding months in MY 2018--19, while also transforming the associated uncertainty appropriately. In the case of urban bias, before transforming to proportional change, we initially calculated the additive inverse of predicted values (i.e., non-urban bias), in order to match our expectation of decline during the pandemic. After the transformation, the months of a year were classified into whether they were peak impact months (April--May) or not, and values within these were averaged across individual months and metrics (accompanied by propagation of uncertainty).

Thus, for each year and month category (peak impact vs rest of year), we finally obtained an overall measure and associated confidence intervals of proportional change in data characteristics relative to the standard of MY 2018--19. This allowed us to assess the nature of the pandemic impact on data characteristics as a whole, as well as to compare the pandemic impact during the peak lockdown months with its broader impact across entire years. Due to the common expected directionality in each of the nine metrics, we expected the pandemic impact to manifest as a decline from pre-pandemic levels when the metrics were combined into this single index. The absence of any perceivable difference in a given period would indicate that there is no evidence of fundamental characteristics of the data having changed during that period.

## Software

All analyses were performed in the R environment [@rcoreteam2021] in RStudio [@positteam2022]. Figures were made using the packages *ggplot2* [@wickham2016] and *patchwork* [@pedersen2020]. In addition, packages *sf* [@pebesma2018], *spdep* [@bivand2022], *raster* [@hijmans2022], *terra* [@hijmans2022a] and *geodata* [@hijmans2022b] for spatial analyses, as well as *lubridate* [@grolemund2011], *glue* [@hester2022] and *tictoc* [@izrailev2021] were used. The source code for the entire work, from data preparation to statistical validation of metrics and to producing the final manuscript, is available on the [GitHub repository](https://github.com/rikudoukarthik/covid-ebirding) for the study.

# Results

## Data quantity

The numbers of uploaded checklists and participating eBirders for each month increased year on year across the four-year period (Fig. \@ref(fig:fig-quant)) with very few exceptions. One such exception is a slightly lower participation (number of birders) in the first wave of COVID-19 in April of MY 2019--20 compared with April of MY 2018--19, but this was still associated with more checklists uploaded.

```{r fig-quant, fig.cap="Graphs of change in data quantity over time, in terms of data generation (number of qualifying lists; A) and participation (number of active eBirders; B). Values for the four migratory years are shown in different colours, and the peak pandemic months are shown in red text. Data contributions to eBird increased consistently every year starting from MY 2018--19, even during the pandemic. Note the separate peak in data generation and participation in February of each year, which coincides with the global event, Great Backyard Bird Count.", out.width="70%", out.height="50%", dpi=300, dev="jpeg"}

month_lab_dates <- month_lab %>% 
  mutate(YEAR = ifelse(MONTH %in% 6:12, 
                       as.numeric(as.character(M.YEAR)), 
                       as.numeric(as.character(M.YEAR)) + 1)) %>% 
  mutate(DATE = as.Date(paste(YEAR, MONTH, "01", sep = "-")))

d1 <- dataquant_a %>% 
  left_join(month_lab_dates) %>% 
  arrange(DATE)

p1 <- ggplot(d1, aes(x = DATE, y = NO.LISTS, colour = M.YEAR)) +
  labs(x = "Months", y = "No. of qualifying lists") +
  geom_point(size = 2) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = covid_palette, name = "Migratory year",
                      labels = unique(d1$M.YEAR.LAB)) +
  scale_y_continuous(label = comma) +
  scale_x_date(labels = d1$MONTH.LAB, breaks = d1$DATE)

p2 <- ggplot(d1, aes(x = DATE, y = NO.EBIRDERS, colour = M.YEAR)) +
  labs(x = "Months", y = "No. of active eBirders") +
  geom_point(size = 2) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = covid_palette, name = "Migratory year",
                      labels = unique(d1$M.YEAR.LAB)) +
  scale_y_continuous(label = comma) +
  scale_x_date(labels = d1$MONTH.LAB, breaks = d1$DATE)

fig_quant <- (p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(5, 1, 5), ncol = 1) +
  plot_annotation(tag_levels = "A") &
  theme(legend.margin = margin(t = 0, unit = "cm"),
        legend.position = "bottom",
        axis.text.y = element_text(angle = 90, vjust = 0, hjust = 0.5),
        axis.text.x = element_text(colour = d1$COL2, angle = 90, size = 8, vjust = 0.5))

fig_path <- "fig_quant.png"

ggsave(plot = fig_quant, fig_path,
       width = 10, height = 10, units = "in", dpi = 300)


knitr::include_graphics(fig_path)

```

## Changes in data characteristics

We derived a composite of nine metrics that characterise the data based on birder behaviour and coverage, with each individual metric expected to decline during the pandemic years. We found that this composite remained near-identical across years during all months of the year except during the peak pandemic months (April--May of MY 2019--20; red ribbon in Fig. \@ref(fig:fig-main)), indicating that birder behaviour changed notably only during these two months. Some individual states also showed an indication of a similar pattern, though uncertainty in estimates was high (Fig. \@ref(fig:fig-main-states)). Across both spatial scales, although data characteristics in MY 2021--22 are indistinguishable from those in MY 2018--19, they have much higher variability, but note the magnitude of growth in eBird data generation and participation over this period (Fig. \@ref(fig:fig-quant)A, B).

```{r fig-main, fig.cap="Proportional change per year and month category in the nine metrics of the national-level composite of eBird data characteristics, relative to values in MY 2018--19 (represented by the y = 1 line). The four years of interest are on the X-axis, with the pandemic years labelled in red text. Each coloured ribbon represents the 95% confidence interval, which is the combined uncertainty estimated across individual uncertainties from model predictions for each of the nine metrics (means have not been depicted). Note the decline in the composite during the first wave of pandemic lockdowns (decline in red ribbon in MY 2019--20), but its recovery to original levels soon after (blue ribbon in MY 2020--21) and the minimal changes even during the second wave (red ribbon in MY 2020--21).", out.width="50%", out.height="50%", dpi=300, dev="jpeg"}

fig_path <- "fig_main.png"
fig_main <- create_composite_graph("country")

ggsave(plot = fig_main, fig_path,
       width = 7, height = 11, units = "in", dpi = 300)

knitr::include_graphics(fig_path)

```

```{r long-changes, fig.cap="Graphs of group birding (A) and inverse site fidelity (B), hotspot birding (C) and travelling birding (D) at the national level. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours, and the peak pandemic months are shown in red text. Note that major changes to metrics were concentrated in the peak pandemic months (April and May of MY 2019--20 and 2020--21), and were more substantial in the first wave than the second.", fig.width=8, fig.height=12.5, out.width="100%", out.height="100%", dpi=300, dev="jpeg"}

d1 <- d01_group_po %>% pluck(1)
d2 <- d02_fidelity_po %>% pluck(1)
d3 <- d04_hotspot %>% pluck(1)
d4 <- d05_protocol %>% pluck(1)


p1 <-
  ggplot(d1, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Proportion of\ngroup birding") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(
    aes(ymin = CI.L, ymax = CI.U),
    size = 1,
    width = 0.5,
    position = position_dodge(0.5)
  ) +
  scale_colour_manual(values = covid_palette, name = "Migratory year",
                      labels = unique(d1$M.YEAR.LAB))

p2 <-
  ggplot(d2, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "No. of sites\nvisited") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(
    aes(ymin = CI.L, ymax = CI.U),
    size = 1,
    width = 0.5,
    position = position_dodge(0.5)
  ) +
  scale_colour_manual(values = covid_palette, name = "Migratory year",
                      labels = unique(d2$M.YEAR.LAB))

p3 <-
  ggplot(d3, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Proportion of\nhotspot birding") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(
    aes(ymin = CI.L, ymax = CI.U),
    size = 1,
    width = 0.5,
    position = position_dodge(0.5)
  ) +
  scale_colour_manual(values = covid_palette, name = "Migratory year",
                      labels = unique(d3$M.YEAR.LAB))

p4 <-
  ggplot(d4, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Proportion of\ntravelling birding") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(
    aes(ymin = CI.L, ymax = CI.U),
    size = 1,
    width = 0.5,
    position = position_dodge(0.5)
  ) +
  scale_colour_manual(values = covid_palette, name = "Migratory year",
                      labels = unique(d4$M.YEAR.LAB))

(p1 / p2 / guide_area() / p3 / p4) +
  plot_layout(guides = "collect", heights = c(3, 3, 0.5, 3, 3)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))


```

Among the individual metrics, the pandemic month of April involved prominent declines in the proportion of group birding, which lowered from roughly 30% in 2019 to nearly 5% in 2020 (Fig. \@ref(fig:long-changes)A, Table \@ref(tab:model-summ1)), and in the proportion of travelling birding, which fell from 60% in 2019 to around 20% in 2020 (birders preferred stationary checklists instead; Fig. \@ref(fig:long-changes)D, Table \@ref(tab:model-summ4)). There was also a prominent decline in proportion of birding in eBird hotspots (Fig. \@ref(fig:long-changes)C, Table \@ref(tab:model-summ3)). In this peak period of the pandemic, birders showed comparatively higher site fidelity, i.e., travelled to fewer grid cells than usual (Fig. \@ref(fig:long-changes)B, Table \@ref(tab:model-summ2)), were more confined to urban areas (Fig. \@ref(fig:urban-cov)A, Table \@ref(tab:model-summ8)), and collectively covered slightly fewer grid cells overall (Fig. \@ref(fig:urban-cov)B; Table \@ref(tab:model-summ10)). However, like with the composite, these changes were short-lived and did not translate to the larger temporal scale of years, the metrics returning to their original levels in the period of June 2020--March 2021 (blue ribbon). For instance, coverage in fact only increased in every successive month since and over the years (Fig. \@ref(fig:urban-cov)B).

```{r urban-cov, fig.cap="Graphs of urban bias (A) and spatial coverage (B) at the national level. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours, and the peak pandemic months are shown in red text. Note that major changes to metrics were concentrated in the peak pandemic months (April and May of MY 2019--20 and 2020--21), and were more substantial in the first wave than the second.", fig.width=8, out.width="100%", fig.height=9, out.height="100%", dpi=300, dev="jpeg"}

d1 <- d09a_s_UNU %>% pluck(1)
d2 <- d09b_s_cover %>% pluck(1)

p1 <- ggplot(d1,
             aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Proportion of\nurban birding") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(
    aes(ymin = CI.L, ymax = CI.U),
    size = 1,
    width = 0.5,
    position = position_dodge(0.5)
  ) +
  scale_colour_manual(values = covid_palette, name = "Migratory year",
                      labels = unique(d1$M.YEAR.LAB))

p2 <- ggplot(d2,
             aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Proportion of\ncells covered") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(
    aes(ymin = CI.L, ymax = CI.U),
    size = 1,
    width = 0.5,
    position = position_dodge(0.5)
  ) +
  scale_colour_manual(values = covid_palette, name = "Migratory year",
                      labels = unique(d2$M.YEAR.LAB))


(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(5, 1, 5)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))

```

```{r dist-dur-length, fig.cap="Graphs of birding distance, list duration and list length at the national level. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours, and the peak pandemic months are shown in red text. Note that all three metrics have been consistently declining over the long term, which is independent of the prominent decline during the peak pandemic months (April and May of MY 2019--20 and 2020--21).", out.width="100%", fig.width=8, fig.height=12, dpi=300, dev="jpeg"}

d1 <- d06_distance %>% pluck(1)
d2 <- d07_duration %>% pluck(1)
d3 <- d08_length %>% pluck(1)

p1 <- ggplot(d1, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Birding distance\n(km)") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year",
                      labels = unique(d1$M.YEAR.LAB))

p2 <- ggplot(d2, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "List duration\n(mins)") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year",
                      labels = unique(d2$M.YEAR.LAB))

p3 <- ggplot(d3, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "List length\n(no. of species)") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year",
                      labels = unique(d3$M.YEAR.LAB))

(p1 / p2 / p3) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))

```

Aside from these suggestions of direct impacts of the pandemic on data characteristics (metrics changing during the peak months and reverting to "normal" soon after), there were also other changes that seem to reflect broader and more general changes over time. For instance, although the pandemic-induced shift to a more sedentary birding style did result in shorter distances covered by the minority birders that made travelling checklists (Fig. \@ref(fig:dist-dur-length)A, Table \@ref(tab:model-summ5)), there is also a general trend towards shorter travelling lists that started before the pandemic and has persisted even after. Similarly, birders started making shorter duration checklists (Fig. \@ref(fig:dist-dur-length)B, Table \@ref(tab:model-summ6)) that reported fewer species (Fig. \@ref(fig:dist-dur-length)C, Table \@ref(tab:model-summ7)), both during the peak pandemic months as well as more generally. Temporally, the impact of the pandemic on these aspects seems minimal, as these changes were not restricted to the peak months or even the pandemic years; instead, these broader trends could be related to the general growth of birding and eBirding in India and a potential accompanying change in birding attitudes (see Section \@ref(evol-ebirding)). It is worth noting, however, that a decline in checklist-level effort does not imply a decline in overall effort.

## Scale of changes

Pandemic-induced changes were largely transient: they were limited to the peak months of the pandemic, and these unsurprising fine-scale changes did not translate over to the larger scale of years. Among these short-term changes to birder behaviour, the decline in group birding lasted the longest, extending even to a few months after the first peak pandemic period (Fig. \@ref(fig:long-changes)A). Notably, birders recovered much quicker in the second year than in the first (Fig. \@ref(fig:fig-main)). These patterns of birder behaviour---particularly the transience of changes---were largely consistent across different states (Fig. \@ref(fig:fig-main-states)) with mostly only minor variations (e.g., Fig. \@ref(fig:long-changes-state)). Major differences were rare, but one example is the complete absence of urban bias in Kerala. Interestingly, Kerala was also the only one of the four states to perfectly mirror the national-level pattern of consistent and unambiguous decline in the composite of metrics during the peak months of MY 2019--20 (Fig. \@ref(fig:fig-main-states)B); others showed high variability within the state, with some instances even of increases (e.g., Maharashtra). Overall, across these spatial units, data characteristics therefore remained similar in the years before, during and after the pandemic.

# Discussion

## Birders in the pandemic: Early impact but innate resilience

The months of April and May in 2020 and 2021 were the peak periods of the pandemic in India and involved some of the most stringent restrictions and government regulations (e.g., nationwide lockdown) in order to curb the spread of the disease [@bbcnews2020; @gettleman2020; @theeconomictimes2020]. Accordingly, most changes to eBird data were also concentrated in these peak pandemic months (Fig. \@ref(fig:fig-main)). Unsurprisingly, birders became more isolated during these periods (especially in the first lockdown in April--May 2020), avoiding travel, groups, and public spaces (Fig. \@ref(fig:long-changes)). They switched to a more sedentary birding style, minimising movement and travel, showing higher site fidelity and being more confined to urban areas than usual (Fig. \@ref(fig:urban-cov)A). Naturally, this also resulted in a slight decline in spatial coverage across the country (Fig. \@ref(fig:urban-cov)B). Birders also made shorter checklists that reported fewer species (Fig. \@ref(fig:dist-dur-length)). Thus, birders seem to have adapted their behaviour during the first peak pandemic period in accordance with the various official guidelines to prevent and minimise the spread of the disease.

As restrictions were gradually lifted in the following months, birder behaviour also changed back to normal, until the peak months of the second pandemic year. Despite this second wave of the pandemic and lockdowns being more devastating than the first in terms of COVID-19 cases and fatalities [@worldometer2023], changes in birder behaviour (such as isolation) were relatively muted and also recovered more quickly this time. The staggered recovery in the first year and the almost immediate recovery in the second suggest that the novelty of the disease played an important role in changing birder behaviour, and also point to the innate resilience of birders. As a result of this, all metrics of eBird data characteristics remained unaffected at the yearly level (see coefficients for years in Tables \@ref(tab:model-summ1)--\@ref(tab:model-summ10)).

## Evolution of eBirding in India {#evol-ebirding}

We found that eBird data quantity has been consistently increasing over the years, in terms of both participation and data generation. This was true even during the peak pandemic periods, despite the various restrictions on movement and social interaction. Such increases in data quantity have been observed elsewhere, both in eBird [@hochachka2021] and in other CS programmes like iNaturalist [@crimmins2021], but in our case we found little evidence for this increase being directly linked to the pandemic. Instead, our results suggest a consistent and long-term growth of eBirding in India over the years, which has also resulted in an increase in overall spatial coverage of eBird data.

Furthermore, this steady growth seems to have been accompanied by shifts in India's birding and eBirding culture, although specifics vary considerably depending on the region and scale under consideration. Some of the observed trends in birder behaviour, although exaggerated during the peak pandemic months, have existed since 2018--19 (i.e., even before the pandemic) and have persisted much after the pandemic has passed. In particular, checklist-level effort has been consistently lower in terms of duration, distance and length (number of species reported). This suggests a gradual evolution towards shorter birding sessions (checklists) and finer scale data generation, even if overall birding effort is not necessarily lower. Of these three metrics of effort, it is worth noting that list length appears to be more robust than distance and duration to both abrupt (like the pandemic) and gradual changes in birder behaviour (note the relatively lower month-on-month and year-on-year deviations, as well as lower deviation during the peak pandemic months; Fig. \@ref(fig:dist-dur-length)).

## Utility of eBird data from the pandemic

The appreciation and use of large-scale, crowdsourced, open access data such as from CS programmes have been growing in recent years, and came into the spotlight during the unprecedented and disruptive times of the COVID-19 pandemic which drastically altered human circumstances worldwide. Most studies linking the pandemic and CS data have focused on changes at the organismal level [e.g., @warrington2022], exploring the phenomenon of anthropause [@rutz2020] to better understand the influence of anthropogenic factors on organisms [@bates2020].

In contrast, there is a paucity of studies looking at the effects of the pandemic on the very characteristics of CS data (i.e., observer-/human-level changes). We expected changes to the nature of eBird data during the pandemic due to behavioural changes in the birding community, caused directly by the various restrictions and indirectly by the changes to everyday life. What we found was a steady increase in data quantity, and certain expected but short-lived changes to birder behaviour. Our findings echo the interpretation of @hochachka2021 of their own results: compared with the magnitude of change brought by the pandemic, its impacts on birder behaviour and thereby eBird data are not substantial. While our results pertain to birders and eBird, there is precedent to find similar trends in other CS programmes as well. We posit that this is due to a stronger urge during the pandemic (and perhaps similar challenging circumstances) for people to connect with nature [@marsh2021], as well as an inherent resilience in birders that buffers them from radical changes to their birding lifestyle. In other, more aphoristic words, birders will be birders.

Thus, we find that the pandemic has not caused major changes in the utility of eBird data, particularly for applications such as abundance trend estimation which are vulnerable to changes in data characteristics (more so than, say, range estimation). Our results suggest that eBird data from the pandemic remain immensely valuable, as well as highlight certain key points regarding the use of these data and making inferences from it.

### Importance of scale

@hochachka2021 emphasised the importance of considering spatial scale, by showing the presence of high degrees of regional variability in the effects of the pandemic. However, they only considered a single month across years, and so failed to capture variability across the temporal scale. Our findings highlight that considerations of both spatial and temporal scales are crucial in assessing the impacts of a major event like the COVID-19 pandemic.

At the fine scale of months and states, there were indeed clear and expected changes in birder behaviour and, consequently, to the nature of the data. Notably, most changes reverted to regular levels soon after the lockdown months, highlighting their temporal transience. These changes also showed high variability (and thus averaged out) across different states (Fig. \@ref(fig:fig-main-states)), pointing to spatial nuances caused by differences in myriad factors that are difficult to predict, such as social and political scenarios.

As a result, there were very few notable nationwide changes across all of the four years (i.e., at the large spatial and temporal scale). Most analyses of long-term abundance trends, distributions or migratory patterns using eBird data are carried out at such large scales, and hence can continue to be performed without concerns of fundamental changes in the data.

### Importance of effort

A detailed understanding of how birding effort in India has changed over the years requires a longer-term analysis of metrics of effort, starting from the inception of eBird in India. While this is beyond the scope of our study, we did find suggestions of such a change, neither caused by nor restricted to the COVID-19 pandemic, but instead representing a consistent directional change as Indian birding evolved, and one that is likely to stay. Therefore, it becomes important to appropriately account for this variability in effort in any analysis spanning multiple years. Our results from the Indian context suggest that list length as a metric of effort may be more resilient than list duration or distance to abrupt short-term as well as to year-to-year variability.

### Guidelines for using pandemic eBird data from India

We outline some guidelines for making use of the tremendous potential in CS programmes like eBird even during the pandemic. When formulating a question to be answered using these data, the researcher must first consider what scale (both spatial and temporal) is appropriate for the question. Questions of larger scale can be tackled directly, whereas fine-scale questions will benefit from a deep-dive into pandemic-induced artefacts in the raw data that can produce misleading patterns. In the latter cases, the researcher is recommended to conduct analyses of data characteristics (similar to the ones in this study) in order to first assess the suitability of the data before attempting to answer the question. Variability in sampling effort is likely to be an issue at smaller scales, but its effect can be minimised by including effort variables such as list length as well as by using methods of spatial subsampling in the analyses. In both cases (large and small scales), it is important to ensure that analyses being conducted are locally appropriate: after identifying the relevant scale and effort metrics, these should be accounted for appropriately, considering the various idiosyncrasies and nuances that may manifest across space and time.

# Conclusion

We found that birders did indeed behave differently during the pandemic, but its consequences for eBird data characteristics were lower than expected from the impact the pandemic has had on human life. While the pandemic did bring about several changes in day-to-day life across the country, this interestingly did not translate to drastic or long-lasting changes in birder behaviour. Our results suggest that official restrictions introduced during the peak periods of the pandemic, together with its novelty in 2020, were most instrumental in changing birder behaviour. Outside these periods, an innate resilience in birders helped maintain characteristics of the data at the large scale, and also led to a steady increase in overall data quantity. Therefore, the eBird dataset from the pandemic years can still be used for inferring year-on-year abundance trends, while smaller-scale applications will benefit from preliminary data quality checks.

# References

::: {#refs}
:::

\newpage
