---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Supplementary Material

## Methods

### Data filters {#dataprep}

We standardised the dataset before analysis with the following set of filters. First, we filtered the dataset for "complete checklists", in which observers report all the birds they see or hear. Next, we filtered the data for our timeline of interest. The migratory cycle in India starts in June of one calendar year and ends in May of the next, and this time frame is thus deemed more appropriate than a calendar year. Therefore, all analyses presented in this study use "migratory year" (MY) as a unit time period.

We then removed data from group accounts, which are eBird accounts with which multiple existing eBirders share their lists to track aggregate summaries. Such accounts are commonly used during specific events or surveys, and need to be removed from analyses of observer-related metrics. Next, we applied a nocturnal filter to remove observations made between 20:00 h and 04:00 h (nocturnal birding is very different from diurnal birding), a pelagic filter to remove observations made from beyond the shoreline of the country, and finally a set of filters to remove observations made with extreme values of birding effort: checklists that are too short (less than three minutes long, or reporting three or fewer species without information on duration, or reporting three or fewer species at a rate lower than two species per hour), and travelling checklists that are too fast (following "Traveling" protocol and travelling faster than 20 kmph or covering a distance greater than 50 km). These lists with extreme effort are not considered to be "truly complete" checklists. Finally, in instances of group birding (column "GROUP.IDENTIFIER") where each participant eBirder owns a copy of the checklist, duplicate checklists were removed by randomly sampling one out of the multiple checklists (column "SAMPLING.EVENT.IDENTIFIER").

The checklists that passed these criteria are referred to as "qualifying checklists".

### Urban/non-urban classification {#UNUprep}

We used remote sensing data to divide the country into urban and non-urban areas. For this, we downloaded the MODIS Land Cover Type Product (MCD12Q1) that includes several science datasets (SDSs) mapping global land cover at a 500 m resolution. Although this comes at an annual time step, we used data from the single time point, 31 December 2021. The tiled MODIS data was downloaded using the "getModis()" function in the *luna* package and saved as individual .tif files, which were then merged into a single raster. After cropping and masking to the country's administrative boundary shapefile, the MODIS land cover categories in the raster were reclassified: category 13, "urban and built-up lands" was considered "urban", and all other categories were considered "non-urban". We didn't use the MODIS data at 500 m × 500 m resolution because the location information in eBird data, despite being precise, is not very accurate at small grain sizes (due to the very nature of eBird data where sampling often occurs over a distance). So, we finally aggregated the raster to a resolution of 2 km $\times$ 2 km which was deemed more appropriate, using a 25% threshold for classification as urban, i.e., a 2 km $\times$ 2 km cell was classified as urban if four or more of the 16 500 m $\times$ 500 m cells being aggregated were urban. We then linked each qualifying checklist in the dataset to this modified MODIS raster, using its associated latitude-longitude information, and thus classified each checklist as urban or non-urban.

## Figures

```{r fig-main-states, fig.cap="Proportional change per year and month category in the nine metrics of state-level composites of eBird data characteristics, relative to values in MY 2018--19 (represented by the y = 1 line). This is visualised for the four states of Karnataka (A), Kerala (B), Maharashtra (C), and Assam (D). The four years of interest are on the X-axis, with the pandemic years labelled in red text. Each coloured ribbon represents the 95% confidence interval, which is the combined uncertainty estimated across individual uncertainties from model predictions for each of the nine metrics (means have not been depicted). Note that state-level changes largely mirror national-level patterns such as transience of pandemic impacts, but exhibit higher internal variability (as suggested by the higher uncertainty) across all years.", out.width="100%", dpi=300, dev="jpeg"}

fig_path <- "fig_main_states.png"

fig1 <- create_composite_graph("state", "Karnataka")
fig2 <- create_composite_graph("state", "Kerala", ylab = FALSE)
fig3 <- create_composite_graph("state", "Maharashtra", ylab = FALSE)
fig4 <- create_composite_graph("state", "Assam", ylab = FALSE)

fig_main_states <- (fig1 | fig2 | fig3 | fig4) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        plot.tag.position = c(0, 0.995))

ggsave(plot = fig_main_states, fig_path,
       width = 27, height = 11, units = "in", dpi = 300)

knitr::include_graphics(fig_path)

```

```{r long-changes-state, fig.cap="Graphs of group birding for the four states of Karnataka (A), Kerala (B), Maharashtra (C), and Assam (D). Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours, and the peak pandemic months are shown in red text. Note that state-level changes largely mirror national-level patterns such as transience of pandemic impacts during the first wave of pandemic lockdowns.", fig.width=8, fig.height=12.5, out.width="100%", out.height="100%", dpi=300, dev="jpeg"}

d1 <- d01_group_po %>% 
  pluck(2) %>% 
  filter(STATE == "Karnataka")
d2 <- d01_group_po %>% 
  pluck(2) %>% 
  filter(STATE == "Kerala")
d3 <- d01_group_po %>% 
  pluck(2) %>% 
  filter(STATE == "Maharashtra")
d4 <- d01_group_po %>% 
  pluck(2) %>% 
  filter(STATE == "Assam")


p1 <- ggplot(d1, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Proportion of\ngroup birding") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year",
                      labels = unique(d1$M.YEAR.LAB))

p2 <- ggplot(d2, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Proportion of\ngroup birding") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year",
                      labels = unique(d2$M.YEAR.LAB))

p3 <- ggplot(d3, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Proportion of\ngroup birding") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year",
                      labels = unique(d3$M.YEAR.LAB))

p4 <- ggplot(d4, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Proportion of\ngroup birding") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year",
                      labels = unique(d4$M.YEAR.LAB))

(p1 / p2 / guide_area() / p3 / p4) +
  plot_layout(guides = "collect", heights = c(3, 3, 0.5, 3, 3)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))

```

\newpage

## Tables

```{r re-vars, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}

# getting re_vars for table captions
get_re_vars <- function(data) {
  revar <- data %>% 
    pluck(3) %>% 
    lme4::VarCorr() %>% 
    as.data.frame() %>% 
    filter(grp != "Residual") %>%  # LMMs output additional row for residuals
    pull(vcov) %>% 
    round(digits = 3)
  
  return(revar)
}

re_var01 <- get_re_vars(d01_group_po)
re_var02 <- get_re_vars(d02_fidelity_po)
re_var04 <- get_re_vars(d04_hotspot)
re_var05 <- get_re_vars(d05_protocol)
re_var06 <- get_re_vars(d06_distance)
re_var07 <- get_re_vars(d07_duration)
re_var08 <- get_re_vars(d08_length)
re_var09a <- get_re_vars(d09a_s_UNU)
re_var09b <- get_re_vars(d09b_s_cover)

```

```{r model-summ1, eval.after = "tab.cap", tab.cap=glue('Model summary for group birding at the national level. Numerical years refer to "migratory years". Random effect variance of Observer ID is {re_var01}. See model specifications described in Section 2.2.4 of the main text.'), out.width="100%"}

model_a <- d01_group_po %>% pluck(3)

# random effect variance
re_var <- model_a %>% 
  lme4::VarCorr() %>% 
  as.data.frame() %>% 
  pull(vcov) %>% 
  round(digits = 3)

summary(model_a) %>% 
  coef() %>% 
  as.data.frame() %>% 
  dplyr::select(1:2) %>% # estimate and SE only
  rownames_to_column() %>% # fixef
  mutate(rowname = str_replace(rowname, "M.YEAR", "")) %>% 
  mutate(rowname = str_replace_all(rowname, c("MONTH2" = "February",
                                              "MONTH3" = "March",
                                              "MONTH4" = "April",
                                              "MONTH5" = "May",
                                              "MONTH6" = "June",
                                              "MONTH7" = "July",
                                              "MONTH8" = "August",
                                              "MONTH9" = "September",
                                              "MONTH10" = "October",
                                              "MONTH11" = "November",
                                              "MONTH12" = "December"))) %>% 
  magrittr::set_colnames(c("Fixed effect", "Estimate", "Standard error")) %>% 
  mutate(across(c("Estimate", "Standard error"), ~ round(., 3))) %>% 
  flextable() %>%
  set_table_properties(layout = "autofit") %>%
  align(align = "center", part = "all") %>%
  align(align = "left", j = "Fixed effect", part = "all") %>%
  line_spacing(space = 1, part = "all") %>% 
  padding(padding = 1.5, part = "body") %>%
  fontsize(size = 7, part = "all") %>%
  flextable::font(fontname = "Times New Roman", part = "all") %>% 
  paginate(init = TRUE, hdr_ftr = TRUE)

```

\newpage

```{r model-summ2, eval.after = "tab.cap", tab.cap=glue('Model summary for site fidelity at the national level. Numerical years refer to "migratory years". Random effect variance of Observer ID is {re_var02}. See model specifications described in Section 2.2.4 of the main text.'), out.width="100%"}

model_a <- d02_fidelity_po %>% pluck(3)

# random effect variance
re_var <- model_a %>% 
  lme4::VarCorr() %>% 
  as.data.frame() %>% 
  pull(vcov) %>% 
  round(digits = 3)

summary(model_a) %>% 
  coef() %>% 
  as.data.frame() %>% 
  dplyr::select(1:2) %>% # estimate and SE only
  rownames_to_column() %>% # fixef
  mutate(rowname = str_replace(rowname, "M.YEAR", "")) %>% 
  mutate(rowname = str_replace_all(rowname, c("MONTH2" = "February",
                                              "MONTH3" = "March",
                                              "MONTH4" = "April",
                                              "MONTH5" = "May",
                                              "MONTH6" = "June",
                                              "MONTH7" = "July",
                                              "MONTH8" = "August",
                                              "MONTH9" = "September",
                                              "MONTH10" = "October",
                                              "MONTH11" = "November",
                                              "MONTH12" = "December"))) %>% 
  magrittr::set_colnames(c("Fixed effect", "Estimate", "Standard error")) %>% 
  mutate(across(c("Estimate", "Standard error"), ~ round(., 3))) %>% 
  flextable() %>%
  set_table_properties(layout = "autofit") %>%
  align(align = "center", part = "all") %>%
  align(align = "left", j = "Fixed effect", part = "all") %>%
  line_spacing(space = 1, part = "all") %>% 
  padding(padding = 1.5, part = "body") %>%
  fontsize(size = 7, part = "all") %>%
  flextable::font(fontname = "Times New Roman", part = "all") %>% 
  paginate(init = TRUE, hdr_ftr = TRUE)

```

\newpage

```{r model-summ3, eval.after = "tab.cap", tab.cap=glue('Model summary for hotspot birding at the national level. Numerical years refer to "migratory years". Random effect variance of Observer ID is {re_var04}. See model specifications described in Section 2.2.4 of the main text.'), out.width="100%"}

model_a <- d04_hotspot %>% pluck(3)

# random effect variance
re_var <- model_a %>% 
  lme4::VarCorr() %>% 
  as.data.frame() %>% 
  pull(vcov) %>% 
  round(digits = 3)

summary(model_a) %>% 
  coef() %>% 
  as.data.frame() %>% 
  dplyr::select(1:2) %>% # estimate and SE only
  rownames_to_column() %>% # fixef
  mutate(rowname = str_replace(rowname, "M.YEAR", "")) %>% 
  mutate(rowname = str_replace_all(rowname, c("MONTH2" = "February",
                                              "MONTH3" = "March",
                                              "MONTH4" = "April",
                                              "MONTH5" = "May",
                                              "MONTH6" = "June",
                                              "MONTH7" = "July",
                                              "MONTH8" = "August",
                                              "MONTH9" = "September",
                                              "MONTH10" = "October",
                                              "MONTH11" = "November",
                                              "MONTH12" = "December"))) %>% 
  magrittr::set_colnames(c("Fixed effect", "Estimate", "Standard error")) %>% 
  mutate(across(c("Estimate", "Standard error"), ~ round(., 3))) %>% 
  flextable() %>%
  set_table_properties(layout = "autofit") %>%
  align(align = "center", part = "all") %>%
  align(align = "left", j = "Fixed effect", part = "all") %>%
  line_spacing(space = 1, part = "all") %>% 
  padding(padding = 1.5, part = "body") %>%
  fontsize(size = 7, part = "all") %>%
  flextable::font(fontname = "Times New Roman", part = "all") %>% 
  paginate(init = TRUE, hdr_ftr = TRUE)

```

\newpage

```{r model-summ4, eval.after = "tab.cap", tab.cap=glue('Model summary for travelling protocol birding at the national level. Numerical years refer to "migratory years". Random effect variance of Observer ID is {re_var05}. See model specifications described in Section 2.2.4 of the main text.'), out.width="100%"}

model_a <- d05_protocol %>% pluck(3)

# random effect variance
re_var <- model_a %>% 
  lme4::VarCorr() %>% 
  as.data.frame() %>% 
  pull(vcov) %>% 
  round(digits = 3)

summary(model_a) %>% 
  coef() %>% 
  as.data.frame() %>% 
  dplyr::select(1:2) %>% # estimate and SE only
  rownames_to_column() %>% # fixef
  mutate(rowname = str_replace(rowname, "M.YEAR", "")) %>% 
  mutate(rowname = str_replace_all(rowname, c("MONTH2" = "February",
                                              "MONTH3" = "March",
                                              "MONTH4" = "April",
                                              "MONTH5" = "May",
                                              "MONTH6" = "June",
                                              "MONTH7" = "July",
                                              "MONTH8" = "August",
                                              "MONTH9" = "September",
                                              "MONTH10" = "October",
                                              "MONTH11" = "November",
                                              "MONTH12" = "December"))) %>% 
  magrittr::set_colnames(c("Fixed effect", "Estimate", "Standard error")) %>% 
  mutate(across(c("Estimate", "Standard error"), ~ round(., 3))) %>% 
  flextable() %>%
  set_table_properties(layout = "autofit") %>%
  align(align = "center", part = "all") %>%
  align(align = "left", j = "Fixed effect", part = "all") %>%
  line_spacing(space = 1, part = "all") %>% 
  padding(padding = 1.5, part = "body") %>%
  fontsize(size = 7, part = "all") %>%
  flextable::font(fontname = "Times New Roman", part = "all") %>% 
  paginate(init = TRUE, hdr_ftr = TRUE)

```

\newpage

```{r model-summ5, eval.after = "tab.cap", tab.cap=glue('Model summary for list distance at the national level. Numerical years refer to "migratory years". Random effect variance of Observer ID is {re_var06}. See model specifications described in Section 2.2.4 of the main text.'), out.width="100%"}

model_a <- d06_distance %>% pluck(3)

# random effect variance
re_var <- model_a %>% 
  lme4::VarCorr() %>% 
  as.data.frame() %>% 
  pull(vcov) %>% 
  round(digits = 3)

summary(model_a) %>% 
  coef() %>% 
  as.data.frame() %>% 
  dplyr::select(1:2) %>% # estimate and SE only
  rownames_to_column() %>% # fixef
  mutate(rowname = str_replace(rowname, "M.YEAR", "")) %>% 
  mutate(rowname = str_replace_all(rowname, c("MONTH2" = "February",
                                              "MONTH3" = "March",
                                              "MONTH4" = "April",
                                              "MONTH5" = "May",
                                              "MONTH6" = "June",
                                              "MONTH7" = "July",
                                              "MONTH8" = "August",
                                              "MONTH9" = "September",
                                              "MONTH10" = "October",
                                              "MONTH11" = "November",
                                              "MONTH12" = "December"))) %>% 
  magrittr::set_colnames(c("Fixed effect", "Estimate", "Standard error")) %>% 
  mutate(across(c("Estimate", "Standard error"), ~ round(., 3))) %>% 
  flextable() %>%
  set_table_properties(layout = "autofit") %>%
  align(align = "center", part = "all") %>%
  align(align = "left", j = "Fixed effect", part = "all") %>%
  line_spacing(space = 1, part = "all") %>% 
  padding(padding = 1.5, part = "body") %>%
  fontsize(size = 7, part = "all") %>%
  flextable::font(fontname = "Times New Roman", part = "all") %>% 
  paginate(init = TRUE, hdr_ftr = TRUE)

```

\newpage

```{r model-summ6, eval.after = "tab.cap", tab.cap=glue('Model summary for list duration at the national level. Numerical years refer to "migratory years". Random effect variance of Observer ID is {re_var07}. See model specifications described in Section 2.2.4 of the main text.'), out.width="100%"}

model_a <- d07_duration %>% pluck(3)

# random effect variance
re_var <- model_a %>% 
  lme4::VarCorr() %>% 
  as.data.frame() %>% 
  pull(vcov) %>% 
  round(digits = 3)

summary(model_a) %>% 
  coef() %>% 
  as.data.frame() %>% 
  dplyr::select(1:2) %>% # estimate and SE only
  rownames_to_column() %>% # fixef
  mutate(rowname = str_replace(rowname, "M.YEAR", "")) %>% 
  mutate(rowname = str_replace_all(rowname, c("MONTH2" = "February",
                                              "MONTH3" = "March",
                                              "MONTH4" = "April",
                                              "MONTH5" = "May",
                                              "MONTH6" = "June",
                                              "MONTH7" = "July",
                                              "MONTH8" = "August",
                                              "MONTH9" = "September",
                                              "MONTH10" = "October",
                                              "MONTH11" = "November",
                                              "MONTH12" = "December"))) %>% 
  magrittr::set_colnames(c("Fixed effect", "Estimate", "Standard error")) %>% 
  mutate(across(c("Estimate", "Standard error"), ~ round(., 3))) %>% 
  flextable() %>%
  set_table_properties(layout = "autofit") %>%
  align(align = "center", part = "all") %>%
  align(align = "left", j = "Fixed effect", part = "all") %>%
  line_spacing(space = 1, part = "all") %>% 
  padding(padding = 1.5, part = "body") %>%
  fontsize(size = 7, part = "all") %>%
  flextable::font(fontname = "Times New Roman", part = "all") %>% 
  paginate(init = TRUE, hdr_ftr = TRUE)

```

\newpage

```{r model-summ7, eval.after = "tab.cap", tab.cap=glue('Model summary for list length at the national level. Numerical years refer to "migratory years". Random effect variance of Observer ID is {re_var08}. See model specifications described in Section 2.2.4 of the main text.'), out.width="100%"}

model_a <- d08_length %>% pluck(3)

# random effect variance
re_var <- model_a %>% 
  lme4::VarCorr() %>% 
  as.data.frame() %>% 
  pull(vcov) %>% 
  round(digits = 3)

summary(model_a) %>% 
  coef() %>% 
  as.data.frame() %>% 
  dplyr::select(1:2) %>% # estimate and SE only
  rownames_to_column() %>% # fixef
  mutate(rowname = str_replace(rowname, "M.YEAR", "")) %>% 
  mutate(rowname = str_replace_all(rowname, c("MONTH2" = "February",
                                              "MONTH3" = "March",
                                              "MONTH4" = "April",
                                              "MONTH5" = "May",
                                              "MONTH6" = "June",
                                              "MONTH7" = "July",
                                              "MONTH8" = "August",
                                              "MONTH9" = "September",
                                              "MONTH10" = "October",
                                              "MONTH11" = "November",
                                              "MONTH12" = "December"))) %>% 
  magrittr::set_colnames(c("Fixed effect", "Estimate", "Standard error")) %>% 
  mutate(across(c("Estimate", "Standard error"), ~ round(., 3))) %>% 
  flextable() %>%
  set_table_properties(layout = "autofit") %>%
  align(align = "center", part = "all") %>%
  align(align = "left", j = "Fixed effect", part = "all") %>%
  line_spacing(space = 1, part = "all") %>% 
  padding(padding = 1.5, part = "body") %>%
  fontsize(size = 7, part = "all") %>%
  flextable::font(fontname = "Times New Roman", part = "all") %>% 
  paginate(init = TRUE, hdr_ftr = TRUE)

```

\newpage

```{r model-summ8, eval.after = "tab.cap", tab.cap=glue('Model summary for urban birding at the national level. Numerical years refer to "migratory years". Random effect variance of Observer ID is {re_var09a}. See model specifications described in Section 2.2.4 of the main text.'), out.width="100%"}

model_a <- d09a_s_UNU %>% pluck(3)

# random effect variance
re_var <- model_a %>% 
  lme4::VarCorr() %>% 
  as.data.frame() %>% 
  pull(vcov) %>% 
  round(digits = 3)

summary(model_a) %>% 
  coef() %>% 
  as.data.frame() %>% 
  dplyr::select(1:2) %>% # estimate and SE only
  rownames_to_column() %>% # fixef
  mutate(rowname = str_replace(rowname, "M.YEAR", "")) %>% 
  mutate(rowname = str_replace_all(rowname, c("MONTH2" = "February",
                                              "MONTH3" = "March",
                                              "MONTH4" = "April",
                                              "MONTH5" = "May",
                                              "MONTH6" = "June",
                                              "MONTH7" = "July",
                                              "MONTH8" = "August",
                                              "MONTH9" = "September",
                                              "MONTH10" = "October",
                                              "MONTH11" = "November",
                                              "MONTH12" = "December"))) %>% 
  magrittr::set_colnames(c("Fixed effect", "Estimate", "Standard error")) %>% 
  mutate(across(c("Estimate", "Standard error"), ~ round(., 3))) %>% 
  flextable() %>%
  set_table_properties(layout = "autofit") %>%
  align(align = "center", part = "all") %>%
  align(align = "left", j = "Fixed effect", part = "all") %>%
  line_spacing(space = 1, part = "all") %>% 
  padding(padding = 1.5, part = "body") %>%
  fontsize(size = 7, part = "all") %>%
  flextable::font(fontname = "Times New Roman", part = "all") %>% 
  paginate(init = TRUE, hdr_ftr = TRUE)

```

\newpage

```{r model-summ10, eval.after = "tab.cap", tab.cap=glue('Model summary for spatial coverage at the national level. Numerical years refer to "migratory years". Random effect variance of Cell ID is {re_var09b}. See model specifications described in Section 2.2.4 of the main text.'), out.width="100%"}

model_a <- d09b_s_cover %>% pluck(3)

# random effect variance
re_var <- model_a %>% 
  lme4::VarCorr() %>% 
  as.data.frame() %>% 
  pull(vcov) %>% 
  round(digits = 3)

summary(model_a) %>% 
  coef() %>% 
  as.data.frame() %>% 
  dplyr::select(1:2) %>% # estimate and SE only
  rownames_to_column() %>% # fixef
  mutate(rowname = str_replace(rowname, "M.YEAR", "")) %>% 
  mutate(rowname = str_replace_all(rowname, c("MONTH2" = "February",
                                              "MONTH3" = "March",
                                              "MONTH4" = "April",
                                              "MONTH5" = "May",
                                              "MONTH6" = "June",
                                              "MONTH7" = "July",
                                              "MONTH8" = "August",
                                              "MONTH9" = "September",
                                              "MONTH10" = "October",
                                              "MONTH11" = "November",
                                              "MONTH12" = "December"))) %>% 
  magrittr::set_colnames(c("Fixed effect", "Estimate", "Standard error")) %>% 
  mutate(across(c("Estimate", "Standard error"), ~ round(., 3))) %>% 
  flextable() %>%
  set_table_properties(layout = "autofit") %>%
  align(align = "center", part = "all") %>%
  align(align = "left", j = "Fixed effect", part = "all") %>%
  line_spacing(space = 1, part = "all") %>% 
  padding(padding = 1.5, part = "body") %>%
  fontsize(size = 7, part = "all") %>%
  flextable::font(fontname = "Times New Roman", part = "all") %>% 
  paginate(init = TRUE, hdr_ftr = TRUE)

```
