---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Supplementary Material

## Methods

### Data preparation

#### Data filters {#dataprep}

We standardised the dataset before analysis with the following set of filters. First, we filtered the dataset for "complete checklists", in which observers report all the birds they see or hear. Next, we filtered the data for our timeline of interest. The migratory cycle in India starts in June of one calendar year and ends in May of the next, and this time frame is thus deemed more appropriate than a calendar year. Therefore, all analyses presented in this study use "migratory year" (MY) as a unit time period.

We then removed data from group accounts, which are eBird accounts with which multiple existing eBirders share their lists to track aggregate summaries. Such accounts are commonly used during specific events or surveys, and need to be removed from analyses of observer-related metrics. Next, we applied a nocturnal filter to remove observations made between 8pm and 4am (nocturnal birding is very different from diurnal birding), a pelagic filter to remove observations made from beyond the shoreline of the country, and finally a set of filters to remove observations made with extreme values of birding "effort": checklists that are too short (less than three minutes long, or reporting three or fewer species without information on duration, or reporting three or fewer species at a rate lower than two species per hour), and travelling checklists that are too fast (following "Traveling" protocol and travelling faster than 20 kmph or covering a distance greater than 50 km). These lists with extreme effort are not considered to be "truly complete" checklists. The dataset at this stage was used in analyses of observer-level metrics, while for non-observer-level data metrics and metrics of bird reporting, a final filter was added: in instances of group birding where each participant eBirder owns a copy of the checklist, duplicate records/checklists were removed by filtering out non-unique combinations of group birding event (column "GROUP.IDENTIFIER") and bird species (column "COMMON.NAME").

The checklists that passed these criteria are referred to as "qualifying checklists".

#### Urban or rural {#UNUprep}

We used remote sensing data to divide the country into urban and non-urban (rural) areas. For this, we downloaded the MODIS Land Cover Type Product (MCD12Q1) that includes several science datasets (SDSs) mapping global land cover at a 500 m resolution. Although this comes at an annual time step, we used data from the single time point, 31 December 2021. The tiled MODIS data was downloaded using the "getModis()" function in the *luna* package and saved as individual .tif files, which were then merged into a single raster. After cropping and masking to the country's administrative boundary shapefile, the MODIS land cover categories in the raster were reclassified: category 13, "urban and built-up lands" was considered "urban", and all other categories were considered "rural" (technically "non-urban"). We didn't use the MODIS data at 500 m × 500 m resolution because the location information in eBird data, despite being precise, is not very accurate at small grain sizes (due to the very nature of eBird data where sampling often occurs over a distance). So, finally we aggregated the raster to a resolution of 2 km × 2 km which was deemed more appropriate, then classified each qualifying checklist in the dataset as urban or rural by extracting the land cover class from the modified MODIS raster, again using the latitude-longitude information.

### Exploratory analyses and bootstrapped estimates {#explboot}

We selected the various metrics of data quantity and quality from the available effort metadata associated with every eBird checklist. To check whether these metrics were useful to answer our question, we first explored basic patterns by calculating the metrics for every month and migratory year, for the four individual states of interest (state-level). These calculations produced a mean or median value and associated confidence interval from non-parametric bootstrapping: first the mean of the values in a group were calculated, following which a new sample of the same size as the original one was selected by resampling the original group with replacement, and a new mean was calculated, and so on until this calculation was done 1000 times. The mean/median of the 1000 sampled means is the group mean/median estimate while its standard deviation is the standard error. We then calculated CIs from the SEs assuming a Gaussian distribution ($CIs = mean \pm 1.96 SE$) due to the large sample size in our dataset.

From the level of states, all the metrics were then calculated for the whole country by averaging the means and propagating the SEs, in order to highlight the high variability across scales, regions, governments, etc. Assuming a Gaussian distribution is necessary to propagate SEs. The error propagation was done as follows:

$$SE_{y} = \frac {\sqrt {\sum_{i=1}^n (SE_{x_{i}}) ^2}} {n}$$

where $n$ is the number of observations/rows in the group.

### Spatial spread {#sspreadmeth}

For analysing spatial spread of the data, we decided to focus on districts as the spatial unit instead of 25 km × 25 km grid cells as in other metrics. This is because birding patterns in general are not uniformly and consistently spread across the country, so the fine scale of grid cells would produce too much noise. Here, districts provide a nice compromise: they are large enough to track change with less noise, while still being small enough to inform us about local patterns. We ensured that a single grid cell was linked with only one district, using random sampling from the list of districts it covered. The temporal unit of focus was the categorical COVID time period ("BEF", "DUR" or "AFT") and not individual months. Only those districts that were "covered" at least once in the study period, i.e., had at least one qualifying checklist in migratory years 2018--2021, were considered in the analyses.

Our aim was to understand whether (and if so, how) the way birding patterns were spread across the country changed during the pandemic years. The two patterns of interest were change in birding effort (number of lists) and change in grid coverage (proportion of total grid cells having birding activity) from one year to the next. The former tells us about the intensity of birding while the latter shows us its breadth or completeness. Here, the changes in metrics rather than the metrics themselves inform us about consistency/variability across years, and would better highlight any anomalies brought by the pandemic. In the case of the period "DUR", both metrics were averaged across the two migratory years it comprised, 2019 and 2020.

The grid coverage metric considered only those districts comprising at least five grid cells after the randomised assignment of grid cells to districts. Although this meant that certain districts which actually cover five or more grid cells might have been disqualified due to the random assignment, the converse is also true so some districts which in reality do not cover five or more grid cells might have qualified. We assume the two cases average out over the entire country, so it is not a statistical concern. However, grid coverage was further analysed in two conditions: one with no threshold for grid cell coverage, and one where a cell was considered "covered" only if it had five or more qualifying checklists per migratory year. The latter condition, by removing outlier cases, would produce clearer patterns with less noise.

Due to the difficulty in quantitatively ascertaining change in spatial spread across such a large area with huge variability, we devised a suite of three analyses which, put together, would provide a clear qualitative picture of change in spatial spread/clustering of the metric.

#### Spatial visualisation of raw change

Raw change in a metric $a$ from period $x$ to period $y$ was calculated as the simple difference of their values $(a_{y}-a_{x})$. It was not calculated for cases where both $a_{x}$ and $a_{y}$ were zero. We plotted the raw change per district from one COVID period to another on a map of India, colouring the districts based on magnitude of change. Here we were concerned with two things: firstly, whether the majority of districts showed a positive or a negative change; second, how many districts showed high magnitude of change, and whether or not these were clustered in space. If most districts showed a negative change in birding effort or grid coverage from BEF to DUR, this would translate to an effect of the pandemic on the quantity and quality of data collected during the pandemic. The higher the magnitude of change, the more worrisome the change would be, and if these high magnitude changes were spatially clustered, this would result in a high spatial bias during the pandemic years, as opposed to if they were spread out across different regions of the country.

#### Net raw change across the country

This is a more quantitative method of analysing whether the majority of districts showed a positive or a negative change. We estimated mean raw change and its 95% CI, by bootstrapping across all the districts (see Section \@ref(explboot) above). If the net change from DUR to AFT was significantly (no overlap of CIs) higher than from BEF to DUR, this would suggest that birding effort or grid coverage was subdued during the pandemic.

#### Spatial visualisation of proportional change in two temporal directions

Proportional change in a metric $a$ from period $x$ to period $y$ was calculated as $(a_{y}-a_{x})/a_{x}$. It was not calculated for cases where both $a_{x}$ and $a_{y}$ were zero. If only $a_{x}$ was zero, a constant $k=1$ was added to both $a_{x}$ and $a_{y}$ in order to avoid zero in the denominator. Moreover, for the birding effort metric, proportional change was not calculated if the first period in the transition had less than 10 qualifying checklists ($a_{x} < 10$). Otherwise, since the metric has a fixed lower limit but not an upper limit, this would result in very high values of proportional change simply due to the low overall birding.

We visualised only the negative values of proportional change, since a general increase in birding effort or grid coverage is not worrisome and is in fact even expected. Here, the values were coloured based on magnitude and once again the high magnitude changes are of interest, but we are also interested in transitions in both directions. For example, if the same districts with high magnitudes of negative proportional change from BEF to DUR have similar values also from AFT to DUR, this would suggest a clear and direct impact of the pandemic in lower birding effort or grid coverage.

### Model specifications {#modelspecs}

For the observer-level group birding metric, a GLMM was fitted with the following specification:

$$metric \sim year + year:month + (1/observer)$$

where $year$ represents the migratory years. For observer-level site fidelity and birding time, LMMs were fitted with the following specification to the log-transformed metric:

$$log(metric) \sim year + year:month + (1/observer)$$

For hotspot birding, birding protocol, urban bias and spatial coverage, GLMMs were fitted as follows:

$$metric \sim year + year:month + (1/cell)$$

For birding distance, list duration and list length, LMMs were fitted with the following specification to the log-transformed metric:

$$log(metric) \sim year + year:month + (1/cell)$$

## Figures

```{r fig-main-states, fig.cap="Proportional change in data quality metrics per year and month category, relative to values in MY 2018.", fig.width=6, fig.height=9.5, out.width="100%", out.height="100%", dpi=300, dev="jpeg"}

fig1 <- create_composite_graph("state", "Karnataka")
fig2 <- create_composite_graph("state", "Kerala", ylab = FALSE)
fig3 <- create_composite_graph("state", "Maharashtra", ylab = FALSE)
fig4 <- create_composite_graph("state", "Assam", ylab = FALSE)

fig_main_states <- (fig1 | fig2 | fig3 | fig4) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

```

```{r time-cov, fig.cap="Graphs of observer-level birding time (A), and spatial coverage (B) at the national level. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours, and the peak pandemic months are shown in red text.", fig.width=8, fig.height=9, out.width="100%", dpi=300, dev="jpeg"}

load("../../00_outputs/d03_time_po.RData")
d1 <-
  data0_a %>% left_join(month_lab)
load("../../00_outputs/d09b_s_cover.RData")
d2 <-
  data0_a %>% left_join(month_lab)

p1 <-
  ggplot(d1,
         aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Time spent birding\n(mins)") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(
    aes(ymin = CI.L, ymax = CI.U),
    size = 1,
    width = 0.5,
    position = position_dodge(0.5)
  ) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p2 <-
  ggplot(d2,
         aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Proportion of\ncells covered") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(
    aes(ymin = CI.L, ymax = CI.U),
    size = 1,
    width = 0.5,
    position = position_dodge(0.5)
  ) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(5, 1, 5)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))

```

```{r dist-dur-length, fig.cap="Graphs of birding distance, list duration and list length at the national level. Points represent means while error bars show 95% confidence intervals. Values for the four migratory years are shown in different colours, and the peak pandemic months are shown in red text.", out.width="100%", fig.width=8, fig.height=12, dpi=300, dev="jpeg"}

load("00_outputs/d06_distance.RData")
d1 <- data0_a %>% left_join(month_lab)
load("00_outputs/d07_duration.RData")
d2 <- data0_a %>% left_join(month_lab)
load("00_outputs/d08_length.RData")
d3 <- data0_a %>% left_join(month_lab)

p1 <- ggplot(d1, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "Birding distance\n(km)") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p2 <- ggplot(d2, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "List duration\n(mins)") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

p3 <- ggplot(d3, aes(MONTH.LAB, PRED, colour = M.YEAR)) +
  labs(x = "Month", y = "List length\n(no. of species)") +
  geom_point(size = 2, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U),
                size = 1, width = 0.5, position = position_dodge(0.5)) +
  scale_colour_manual(values = covid_palette, name = "Migratory year")

(p1 / p2 / p3) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        axis.text.x = element_text(colour = month_lab$COL1))

```

```{r s-spread-figs-raw, fig.cap="Spatial spread: raw change in number of lists (A) and thresholded percentage grid coverage (B) per district from before to during and from during to after the pandemic. Districts are shaded based on the simple difference in values between the two periods from highly negative (red) to highly positive (blue). Grey districts are those that did not meet the criteria for inclusion in the analysis.", out.width="100%", fig.width=8, fig.height=13, dpi=300, dev="jpeg"}

load("../../00_outputs/d09c_s_spread.RData")

d1 <-
  temp1 %>% left_join(grid_dist_base0)
d2 <-
  temp5 %>% left_join(grid_dist_base0) %>% filter(METRIC == "GRID.COV.T")

# raw change maps (except gridcov w/o threshold)

p1 <-
  ggplot(d1) +
  geom_polygon(
    data = indiamap,
    aes(long, lat, group = group),
    fill = "#ACACAC",
    colour = "black",
    size = 0.4
  ) +
  geom_sf(aes(fill = VALUE, geometry = geometry),
          col = "#ACACAC",
          size = 0.1) +
  facet_wrap( ~ T.LABEL) +
  scale_fill_stepsn(
    colours = RColorBrewer::brewer.pal(n = 9, name = "RdBu"),
    breaks = c(-1000, -500, -100, -50, 50, 100, 500, 1000),
    values = scales::rescale(c(
      -11000, -1000, -500, -100, -50,
      50, 100, 500, 1000, 11000
    )),
    limits = c(-11000, 11000),
    na.value = "#ACACAC"
  ) +
  labs(x = "Longitude", y = "Latitude",
       fill = "Difference in\nno. of lists")


p2 <-
  ggplot(d2) +
  geom_polygon(
    data = indiamap,
    aes(long, lat, group = group),
    fill = "#ACACAC",
    colour = "black",
    size = 0.4
  ) +
  geom_sf(aes(fill = VALUE, geometry = geometry),
          col = "#ACACAC",
          size = 0.1) +
  facet_wrap( ~ T.LABEL) +
  scale_fill_stepsn(
    colours = RColorBrewer::brewer.pal(n = 9, name = "RdBu"),
    breaks = c(-40, -25, -10, -1, 1, 10, 25, 40),
    values = scales::rescale(c(-100, -40, -25, -10, -1, 1, 10, 25, 40, 100)),
    limits = c(-100, 100),
    na.value = "#ACACAC"
  ) +
  labs(x = "Longitude", y = "Latitude",
       fill = "Difference in\ngrid coverage")


(p1 / p2) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        legend.key.width = unit(3, "cm"))

```

```{r s-spread-figs-prop, fig.cap="Spatial spread: proportional declines in number of lists per district in forward (top row) and backward (bottom row) transitions. Districts with positive values are shown in grey; declines are shaded based on the magnitude of proportional change from low (dark blue) to high (yellow). White districts are those that did not meet the criteria for inclusion in the analysis.", out.width="100%", fig.width=8, fig.height=13, dpi=300, dev="jpeg"}

load("../../00_outputs/d09c_s_spread.RData")

d1 <-
  temp3 %>% left_join(grid_dist_base0)
d2 <-
  temp4 %>% left_join(grid_dist_base0)


p1 <-
  ggplot(d1) +
  geom_polygon(
    data = indiamap,
    aes(long, lat, group = group),
    fill = "white",
    colour = "black",
    size = 0.4
  ) +
  geom_sf(aes(fill = VALUE, geometry = geometry),
          col = "#E0E0E0",
          size = 0.1) +
  facet_wrap( ~ T.LABEL) +
  scale_fill_steps2(
    low = viridisLite::cividis(5, direction = -1)[1],
    mid = viridisLite::cividis(5, direction = -1)[3],
    high = viridisLite::cividis(5, direction = -1)[5],
    midpoint = -0.45,
    breaks = c(-1, -0.9, -0.65, -0.45, -0.25, -0.1, -0),
    limits = c(-1, 0),
    na.value = "#E0E0E0"
  ) +
  labs(x = "Longitude", y = "Latitude",
       fill = "Proportional change")

p2 <-
  ggplot(d2) +
  geom_polygon(
    data = indiamap,
    aes(long, lat, group = group),
    fill = "white",
    colour = "black",
    size = 0.4
  ) +
  geom_sf(aes(fill = VALUE, geometry = geometry),
          col = "#E0E0E0",
          size = 0.1) +
  facet_wrap( ~ T.LABEL) +
  scale_fill_steps2(
    low = viridisLite::cividis(5, direction = -1)[1],
    mid = viridisLite::cividis(5, direction = -1)[3],
    high = viridisLite::cividis(5, direction = -1)[5],
    midpoint = -0.45,
    breaks = c(-1, -0.9, -0.65, -0.45, -0.25, -0.1, -0),
    limits = c(-1, 0),
    na.value = "#E0E0E0"
  ) +
  labs(x = "Longitude", y = "Latitude",
       fill = "Proportional change")


(p1 / guide_area() / p2) +
  plot_layout(guides = "collect", heights = c(6, 0.5, 6)) +
  plot_annotation(tag_levels = "A") &
  theme(
    legend.position = "bottom",
    legend.key.width = unit(2, "cm"),
    legend.margin = margin(t = 0, unit = "cm")
  )

```

```{r s-spread-rest1, fig.cap="(A) Spatial spread: raw change in percentage grid coverage per district with no threshold from before to during and from during to after the pandemic. Districts are shaded based on the simple difference in values between the two periods from highly negative (red) to highly positive (blue). Grey districts are those that did not meet the criteria for inclusion in the analysis. (B,C,D) Net change across districts in number of lists (B), percentage grid coverage with no threshold (C) and thresholded percentage grid coverage (D) from before to during and from during to after the pandemic.", out.width="100%", fig.width=8, fig.height=13, dpi=300, dev="jpeg"}

load("00_outputs/d09c_s_spread.RData")

d1 <- temp5 %>% left_join(grid_dist_base0) %>% filter(METRIC == "GRID.COV")
d2 <- temp2
d3 <- temp6 %>% filter(METRIC == "GRID.COV")
d4 <- temp6 %>% filter(METRIC == "GRID.COV.T")

p1 <- ggplot(d1) +
  geom_polygon(data = indiamap,
               aes(long, lat, group = group),
               fill = "#ACACAC", colour = "black", size = 0.4) +
  geom_sf(aes(fill = VALUE, geometry = geometry), col ="#ACACAC", size = 0.1) +
  facet_wrap(~ T.LABEL) +
  scale_fill_stepsn(colours = RColorBrewer::brewer.pal(n = 9, name = "RdBu"),
                    breaks = c(-40, -25, -10, -1, 1, 10, 25, 40),
                    values = scales::rescale(c(-100, -40, -25, -10, -1, 1, 10, 25, 40, 100)), 
                    limits = c(-100, 100),
                    na.value = "#ACACAC") +
  labs(x = "Longitude", y = "Latitude",
       fill = "Raw change in\ngrid coverage")

p2 <- ggplot(d2, aes(x = T.LABEL, y = VALUE)) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.25, width = 0.25, position = position_dodge(0.5)) +
  scale_x_discrete(labels = c("Before to\nDuring", "During to\nAfter")) +
  labs(x = "", y = "Net change in no. of lists") 

p3 <- ggplot(d3, aes(x = T.LABEL, y = VALUE)) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.25, width = 0.25, position = position_dodge(0.5)) +
  scale_x_discrete(labels = c("Before to\nDuring", "During to\nAfter")) +
  labs(x = "Transition", y = "Net change in grid coverage (%)") 

p4 <- ggplot(d4, aes(x = T.LABEL, y = VALUE)) + 
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = CI.L, ymax = CI.U), 
                size = 1.25, width = 0.25, position = position_dodge(0.5)) +
  scale_x_discrete(labels = c("Before to\nDuring", "During to\nAfter")) +
  labs(x = "", y = "") 


(p1 / (p2 | p3 | p4)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        legend.key.width = unit(3, "cm"),
        legend.margin = margin(t = 0, unit = "cm"),
        axis.text.x = element_text(size = 9))

```

## Tables

```{r model-summ1, out.width="100%"}

load("00_outputs/d01_group_po.RData")

summary(model_a) %>% 
  coef() %>% 
  as.data.frame() %>% 
  dplyr::select(1:2) %>% # estimate and SE only
  rownames_to_column() %>% # fixef
  mutate(rowname = str_replace(rowname, "M.YEAR", "")) %>% 
  mutate(rowname = str_replace_all(rowname, c("MONTH2" = "February",
                                              "MONTH3" = "March",
                                              "MONTH4" = "April",
                                              "MONTH5" = "May",
                                              "MONTH6" = "June",
                                              "MONTH7" = "July",
                                              "MONTH8" = "August",
                                              "MONTH9" = "September",
                                              "MONTH10" = "October",
                                              "MONTH11" = "November",
                                              "MONTH12" = "December"))) %>% 
  magrittr::set_colnames(c("Fixed effect", "Estimate", "Standard error")) %>% 
  knitr::kable(caption = 'Model summary for observer-level group birding at the national level. Numerical years refer to "migratory years". Random effect variance of Observer ID is 2.796. See model specifications described earlier.',
               align = "lccc")

```

```{r model-summ2, out.width="100%"}

load("00_outputs/d02_fidelity_po.RData")

summary(model_a) %>% 
  coef() %>% 
  as.data.frame() %>% 
  dplyr::select(1:2) %>% # estimate and SE only
  rownames_to_column() %>% # fixef
  mutate(rowname = str_replace(rowname, "M.YEAR", "")) %>% 
  mutate(rowname = str_replace_all(rowname, c("MONTH2" = "February",
                                              "MONTH3" = "March",
                                              "MONTH4" = "April",
                                              "MONTH5" = "May",
                                              "MONTH6" = "June",
                                              "MONTH7" = "July",
                                              "MONTH8" = "August",
                                              "MONTH9" = "September",
                                              "MONTH10" = "October",
                                              "MONTH11" = "November",
                                              "MONTH12" = "December"))) %>% 
  magrittr::set_colnames(c("Fixed effect", "Estimate", "Standard error")) %>% 
  knitr::kable(caption = 'Model summary for observer-level site fidelity at the national level. Numerical years refer to "migratory years". Random effect variance of Observer ID is 0.0988. See model specifications described earlier.',
               align = "lccc")

```

```{r model-summ3, out.width="100%"}

load("00_outputs/d04_hotspot.RData")

summary(model_a) %>% 
  coef() %>% 
  as.data.frame() %>% 
  dplyr::select(1:2) %>% # estimate and SE only
  rownames_to_column() %>% # fixef
  mutate(rowname = str_replace(rowname, "M.YEAR", "")) %>% 
  mutate(rowname = str_replace_all(rowname, c("MONTH2" = "February",
                                              "MONTH3" = "March",
                                              "MONTH4" = "April",
                                              "MONTH5" = "May",
                                              "MONTH6" = "June",
                                              "MONTH7" = "July",
                                              "MONTH8" = "August",
                                              "MONTH9" = "September",
                                              "MONTH10" = "October",
                                              "MONTH11" = "November",
                                              "MONTH12" = "December"))) %>% 
  magrittr::set_colnames(c("Fixed effect", "Estimate", "Standard error")) %>% 
  knitr::kable(caption = 'Model summary for hotspot birding at the national level. Numerical years refer to "migratory years". Random effect variance of Cell ID is 2.635. See model specifications described earlier.',
               align = "lccc")

```

```{r model-summ4, out.width="100%"}

load("00_outputs/d05_protocol.RData")

summary(model_a) %>% 
  coef() %>% 
  as.data.frame() %>% 
  dplyr::select(1:2) %>% # estimate and SE only
  rownames_to_column() %>% # fixef
  mutate(rowname = str_replace(rowname, "M.YEAR", "")) %>% 
  mutate(rowname = str_replace_all(rowname, c("MONTH2" = "February",
                                              "MONTH3" = "March",
                                              "MONTH4" = "April",
                                              "MONTH5" = "May",
                                              "MONTH6" = "June",
                                              "MONTH7" = "July",
                                              "MONTH8" = "August",
                                              "MONTH9" = "September",
                                              "MONTH10" = "October",
                                              "MONTH11" = "November",
                                              "MONTH12" = "December"))) %>% 
  magrittr::set_colnames(c("Fixed effect", "Estimate", "Standard error")) %>% 
  knitr::kable(caption = 'Model summary for travelling protocol birding at the national level. Numerical years refer to "migratory years". Random effect variance of Cell ID is 0.3118. See model specifications described earlier.',
               align = "lccc")

```

```{r model-summ5, out.width="100%"}

load("00_outputs/d06_distance.RData")

summary(model_a) %>% 
  coef() %>% 
  as.data.frame() %>% 
  dplyr::select(1:2) %>% # estimate and SE only
  rownames_to_column() %>% # fixef
  mutate(rowname = str_replace(rowname, "M.YEAR", "")) %>% 
  mutate(rowname = str_replace_all(rowname, c("MONTH2" = "February",
                                              "MONTH3" = "March",
                                              "MONTH4" = "April",
                                              "MONTH5" = "May",
                                              "MONTH6" = "June",
                                              "MONTH7" = "July",
                                              "MONTH8" = "August",
                                              "MONTH9" = "September",
                                              "MONTH10" = "October",
                                              "MONTH11" = "November",
                                              "MONTH12" = "December"))) %>% 
  magrittr::set_colnames(c("Fixed effect", "Estimate", "Standard error")) %>% 
  knitr::kable(caption = 'Model summary for birding distance at the national level. Numerical years refer to "migratory years". Random effect variance of Cell ID is 0.3059. See model specifications described earlier.',
               align = "lccc")

```

```{r model-summ6, out.width="100%"}

load("00_outputs/d07_duration.RData")

summary(model_a) %>% 
  coef() %>% 
  as.data.frame() %>% 
  dplyr::select(1:2) %>% # estimate and SE only
  rownames_to_column() %>% # fixef
  mutate(rowname = str_replace(rowname, "M.YEAR", "")) %>% 
  mutate(rowname = str_replace_all(rowname, c("MONTH2" = "February",
                                              "MONTH3" = "March",
                                              "MONTH4" = "April",
                                              "MONTH5" = "May",
                                              "MONTH6" = "June",
                                              "MONTH7" = "July",
                                              "MONTH8" = "August",
                                              "MONTH9" = "September",
                                              "MONTH10" = "October",
                                              "MONTH11" = "November",
                                              "MONTH12" = "December"))) %>% 
  magrittr::set_colnames(c("Fixed effect", "Estimate", "Standard error")) %>% 
  knitr::kable(caption = 'Model summary for list duration at the national level. Numerical years refer to "migratory years". Random effect variance of Cell ID is 0.3324. See model specifications described earlier.',
               align = "lccc")

```

```{r model-summ7, out.width="100%"}

load("00_outputs/d08_length.RData")

summary(model_a) %>% 
  coef() %>% 
  as.data.frame() %>% 
  dplyr::select(1:2) %>% # estimate and SE only
  rownames_to_column() %>% # fixef
  mutate(rowname = str_replace(rowname, "M.YEAR", "")) %>% 
  mutate(rowname = str_replace_all(rowname, c("MONTH2" = "February",
                                              "MONTH3" = "March",
                                              "MONTH4" = "April",
                                              "MONTH5" = "May",
                                              "MONTH6" = "June",
                                              "MONTH7" = "July",
                                              "MONTH8" = "August",
                                              "MONTH9" = "September",
                                              "MONTH10" = "October",
                                              "MONTH11" = "November",
                                              "MONTH12" = "December"))) %>% 
  magrittr::set_colnames(c("Fixed effect", "Estimate", "Standard error")) %>% 
  knitr::kable(caption = 'Model summary for list length at the national level. Numerical years refer to "migratory years". Random effect variance of Cell ID is 0.2146. See model specifications described earlier.',
               align = "lccc")

```

```{r model-summ8, out.width="100%"}

load("00_outputs/d09a_s_UNU.RData")

summary(model_a) %>% 
  coef() %>% 
  as.data.frame() %>% 
  dplyr::select(1:2) %>% # estimate and SE only
  rownames_to_column() %>% # fixef
  mutate(rowname = str_replace(rowname, "M.YEAR", "")) %>% 
  mutate(rowname = str_replace_all(rowname, c("MONTH2" = "February",
                                              "MONTH3" = "March",
                                              "MONTH4" = "April",
                                              "MONTH5" = "May",
                                              "MONTH6" = "June",
                                              "MONTH7" = "July",
                                              "MONTH8" = "August",
                                              "MONTH9" = "September",
                                              "MONTH10" = "October",
                                              "MONTH11" = "November",
                                              "MONTH12" = "December"))) %>% 
  magrittr::set_colnames(c("Fixed effect", "Estimate", "Standard error")) %>% 
  knitr::kable(caption = 'Model summary for urban birding at the national level. Numerical years refer to "migratory years". Random effect variance of Cell ID is 8.563. See model specifications described earlier.',
               align = "lccc")

```

```{r model-summ9, out.width="100%"}

load("00_outputs/d03_time_po.RData")

summary(model_a) %>% 
  coef() %>% 
  as.data.frame() %>% 
  dplyr::select(1:2) %>% # estimate and SE only
  rownames_to_column() %>% # fixef
  mutate(rowname = str_replace(rowname, "M.YEAR", "")) %>% 
  mutate(rowname = str_replace_all(rowname, c("MONTH2" = "February",
                                              "MONTH3" = "March",
                                              "MONTH4" = "April",
                                              "MONTH5" = "May",
                                              "MONTH6" = "June",
                                              "MONTH7" = "July",
                                              "MONTH8" = "August",
                                              "MONTH9" = "September",
                                              "MONTH10" = "October",
                                              "MONTH11" = "November",
                                              "MONTH12" = "December"))) %>% 
  magrittr::set_colnames(c("Fixed effect", "Estimate", "Standard error")) %>% 
  knitr::kable(caption = 'Model summary for observer-level birding time at the national level. Numerical years refer to "migratory years". Random effect variance of Observer ID is 0.9489. See model specifications described earlier.',
               align = "lccc")

```

```{r model-summ10, out.width="100%"}

load("00_outputs/d09b_s_cover.RData")

summary(model_a) %>% 
  coef() %>% 
  as.data.frame() %>% 
  dplyr::select(1:2) %>% # estimate and SE only
  rownames_to_column() %>% # fixef
  mutate(rowname = str_replace(rowname, "M.YEAR", "")) %>% 
  mutate(rowname = str_replace_all(rowname, c("MONTH2" = "February",
                                              "MONTH3" = "March",
                                              "MONTH4" = "April",
                                              "MONTH5" = "May",
                                              "MONTH6" = "June",
                                              "MONTH7" = "July",
                                              "MONTH8" = "August",
                                              "MONTH9" = "September",
                                              "MONTH10" = "October",
                                              "MONTH11" = "November",
                                              "MONTH12" = "December"))) %>% 
  magrittr::set_colnames(c("Fixed effect", "Estimate", "Standard error")) %>% 
  knitr::kable(caption = 'Model summary for spatial coverage at the national level. Numerical years refer to "migratory years". Random effect variance of Cell ID is 10.22. See model specifications described earlier.',
               align = "lccc")

```
